### 2012-03-09 flux.tex
Back to LaTeX

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}

% comments
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf Integrating influenza antigenic dynamics with molecular evolution}

\author[1]{Trevor Bedford}
\author[2,3,4]{Marc A. Suchard}
\author[5]{Philippe Lemey}
\author[1]{Gytis Dudas}
\author[6]{Colin Russell}
\author[6,7]{Derek Smith}
\author[1,8]{Andrew Rambaut}

\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
\affil[4]{Department of Biostatistics, UCLA School of Public Health, University of California, Los Angeles CA, USA}
\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
\affil[6]{Department of Zoology, University of Cambridge, Cambridge, UK.}
\affil[7]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
\affil[8]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}

\date{\today}

\begin{document}

\maketitle

%%% ABSTRACT %%%
\section*{Abstract}

%%% INTRODUCTION %%%
\section*{Introduction}

\tbc{Introduction focuses on vaccine selection to tie things together with the planned conclusion.}

Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to
500,000 deaths annually \cite{flufactsheet}. While individuals develop long-lasting immunity to particular
influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that
are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future
infection. The influenza virus population continually evolves in antigenic phenotype in a process known as
antigenic drift. A large proportion of the disease burden of influenza stems from antigenic drift; it is why
vaccines remain only transiently effective. A thorough understanding of the process of antigenic drift is
essential to our efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.

There are currently three major clades of influenza circulating within the human population: influenza A
subtype H3N2, influenza A subtype H1N1 and influenza B. Subtypes refer to the genes, hemagglutinin (H) and
neuraminidase (N), that are primarily responsible for the antigenic character of a strain. Currently, seasonal
influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain
of influenza B. The World Health Organization (WHO) Global Influenza Surveillance Network issues twice-yearly
recommendations on which strains of influenza to use as vaccine strains in 9--12 month’s time, i.e.\ a
February recommendation for the Northern Hemisphere flu season and an August recommendation for the Southern
Hemisphere flu season. These recommendations are provided by a panel of experts after review of the available
data.

Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic
drift in the influenza virus \cite{Nelson07NatRevGenet}. Experimental characterization of antigenic phenotype
is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the
cross-reactivity of one virus strain to serum raised against another strain. Sera from older strains react
poorly with more recent viruses resulting in new strains having a transmission advantage over established
strains. The results of many HI assays across a multitude of virus strains can be combined to yield a
two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified
measure \cite{Smith04}. The antigenic map of influenza A (H3N2) has shown largely linear movement of the
influenza virus population since its introduction in 1968. However, evolution of antigenic phenotype appears
punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution
appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or
combinations of genetic changes may drive changes in antigenic phenotype. The process of antigenic drift
results in the rapid turnover of the virus population. Although mutation occurs rapidly, standing genetic
diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant
trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.

Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in
isolation. An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is
constructed from sequence data. However, the opportunity for a combined approach exists as both the antigenic
map and the phylogenetic tree often contain many of the same isolates. Here, we implement a flexible Bayesian
approach to jointly analyze the antigenic and genetic dynamics of the influenza virus population. We apply
this approach to investigate the dynamics of influenza A (H3N2), influenza A (H1N1) and influenza B.
\tbc{Brief conclusions.}

%%% RESULTS %%%
\section*{Results}

%%% Figure 1: mds %%%
\begin{figure}[tb]
	\centering		
	\makebox[\textwidth]{	
%		\includegraphics{figures/mds}
	}
	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the
mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  The map has been
oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2)
orthogonal to this axis.  (C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year
of virus isolation.  The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX}
for H3N2 and \tbc{XXX} for H1N1.  (E) and (F) Antigenic location along the secondary axis of variation (AG2)
vs.\ year of virus isolation.  Antigenic units represent two-fold dilutions of the HI assay, and strains have
been colored based on year of isolation.  \tbc{Decide whether to just show mean posterior or show
distributions.}}
	\label{mds}
\end{figure}

%%% DISCUSSION %%%
\section*{Discussion}

%%% METHODS %%%
\section*{Methods}

\subsection*{Genetic and antigenic data}

We compiled an antigenic dataset for hemagglutination inhibition (HI) measurements for influenza A (H3N2) by
combining data used in Hay et al. \cite{Hay01}, Smith et al.\ \cite{Smith04}, Russell et al.\
\cite{Russell08}, Barr et al.\ \cite{Barr10} and Cox et al.\ \tbc{WHO report}. This combined dataset had 1651
influenza isolates (present as either virus or sera in HI comparisons) dating from 1968 to 2011. However, the
majority of isolates date from 2002 to 2007. Because we are interested in longer-term antigenic evolution, we
censored the data to have at most 20 strains per year, preferentially keeping those strains with more
antigenic comparisons. This censoring left 338 strains present as 320 viruses and 438 sera (replicate sera are
often constructed from the same strain). Across these viruses and sera, we observe 7240 HI measurements. We
queried the IRD \cite{IRD} and GISAID \tbc{CITE} sequence databases for HA nucleotide sequences based on
strain names, e.g.\ A/HongKong/1/1968, of these strains. If a strain had multiple sequences in the databases
we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. Sequences were
aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.

Antigenic data for influenza A (H1N1) was collected from sources... \tbc{ask Gytis}.

\subsection*{Bayesian multidimensional scaling}

We follow Smith et al. \cite{Smith04} and represent antigenic locations on a 2D antigenic map. Through the
hemagglutination inhibition (HI) assay, there exist measurements of the cross-reactivity of hemagglutinin (HA)
from one virus strain to serum raised against another strain \cite{Hirst43}. Thus, antigenic phenotype is
measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to sera from strain
$j$. Due to experimental constraints, the distance matrix $\mathbf{H}$ is sparse; most comparisons have not be
made. Our goal is to find an optimal projection of the high-dimensional distance matrix into a lower number of
dimensions. We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a
probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations
to the observed matrix HI measurements.

Let $\mathbf{X}_i \in \Re^{P}$ represent the cartographic location of strain $i$ for $i = 1,\ldots, N$.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  This gives set of distances
between cartographic locations 
\begin{equation}
	\delta_{ij} =  || \mathbf{X}_i - \mathbf{X}_i ||_2.
\end{equation}
We define the antigenic distance between virus from strain $i$ and antisera from strain $j$
\begin{equation}
	d_{ij} =  \max{ H_{ij} } + 
	\log_2 \left(   
		\frac{ H_{ij} }{ \max{ H_j } }
	\right).
\end{equation}
and let the set $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
The goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}_1,\ldots,\mathbf{X}_N$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized. 
A probabilistic interpretation reformulates the optimization as
\begin{equation} 
d_{ij} \sim \mbox{Normal}( \delta_{ij}, \omega ) \times 1( d_{ij} > 0 ) \mbox{ for all } (i,j) \in \cal I
\end{equation}
where $1 ( \cdot )$ is the indicator function for truncation. The likelihood of the antigenic distance
measures is proportional to
\begin{equation}
\omega^{m/2} \mbox{exp} \left[
- \frac{\omega \times SSR}{2} - \sum_{(i,j) \in \cal I} \log \Phi (\omega \times \delta_{ij})
\right],
\end{equation}
where $m$ is the cardinality of $\cal I$, $SSR$ is the sum of the squared residuals and $\Phi(\cdot)$ is the
standard normal CDF. We consider the conjugate prior $\omega \sim \mbox{Gamma}(a, b)$, and assume a uniform
prior over $\mathbf{X}$.

%%% ACKNOWLEDGMENTS %%%
\subsection*{Acknowledgments}

%%% FUNDING %%%
\subsection*{Funding}

%%% REFERENCES %%%
\bibliographystyle{plos}
\bibliography{/Users/bedfordt/Documents/bedford}

\pagebreak

\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thepage}{S\arabic{page}}

%%% SUPPORTING INFORMATION %%%
\section*{Supporting Information}

\pagebreak

\end{document}

### 2012-03-09 flux.tex
Commit.

@@
 
-Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to
-500,000 deaths annually \cite{flufactsheet}. While individuals develop long-lasting immunity to particular
-influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that
-are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future
-infection. The influenza virus population continually evolves in antigenic phenotype in a process known as
-antigenic drift. A large proportion of the disease burden of influenza stems from antigenic drift; it is why
-vaccines remain only transiently effective. A thorough understanding of the process of antigenic drift is
-essential to our efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
+Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
+While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection.
+The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
+A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
+A thorough understanding of the process of antigenic drift is essential to our efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
 
-There are currently three major clades of influenza circulating within the human population: influenza A
-subtype H3N2, influenza A subtype H1N1 and influenza B. Subtypes refer to the genes, hemagglutinin (H) and
-neuraminidase (N), that are primarily responsible for the antigenic character of a strain. Currently, seasonal
-influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain
-of influenza B. The World Health Organization (WHO) Global Influenza Surveillance Network issues twice-yearly
-recommendations on which strains of influenza to use as vaccine strains in 9--12 month’s time, i.e.\ a
-February recommendation for the Northern Hemisphere flu season and an August recommendation for the Southern
-Hemisphere flu season. These recommendations are provided by a panel of experts after review of the available
-data.
+There are currently three major clades of influenza circulating within the human population: influenza A subtype H3N2, influenza A subtype H1N1 and influenza B. 
+Subtypes refer to the genes, hemagglutinin (H) and neuraminidase (N), that are primarily responsible for the antigenic character of a strain. 
+Currently, seasonal influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain of influenza B. 
+The World Health Organization (WHO) Global Influenza Surveillance Network issues twice-yearly recommendations on which strains of influenza to use as vaccine strains in 9--12 month’s time, i.e.\ a February recommendation for the Northern Hemisphere flu season and an August recommendation for the Southern Hemisphere flu season.
+These recommendations are provided by a panel of experts after review of the available data.
 
-Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic
-drift in the influenza virus \cite{Nelson07NatRevGenet}. Experimental characterization of antigenic phenotype
-is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the
-cross-reactivity of one virus strain to serum raised against another strain. Sera from older strains react
-poorly with more recent viruses resulting in new strains having a transmission advantage over established
-strains. The results of many HI assays across a multitude of virus strains can be combined to yield a
-two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified
-measure \cite{Smith04}. The antigenic map of influenza A (H3N2) has shown largely linear movement of the
-influenza virus population since its introduction in 1968. However, evolution of antigenic phenotype appears
-punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution
-appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or
-combinations of genetic changes may drive changes in antigenic phenotype. The process of antigenic drift
-results in the rapid turnover of the virus population. Although mutation occurs rapidly, standing genetic
-diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant
-trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
+Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
+Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain.
+Here is a new sentence.
+Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over established strains. 
+The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified measure \cite{Smith04}.
+The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968.
+However, evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
+The process of antigenic drift results in the rapid turnover of the virus population. 
+Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
 
-Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in
-isolation. An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is
-constructed from sequence data. However, the opportunity for a combined approach exists as both the antigenic
-map and the phylogenetic tree often contain many of the same isolates. Here, we implement a flexible Bayesian
-approach to jointly analyze the antigenic and genetic dynamics of the influenza virus population. We apply
-this approach to investigate the dynamics of influenza A (H3N2), influenza A (H1N1) and influenza B.
+Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
+An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
+However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
+Here, we implement a flexible Bayesian approach to jointly analyze the antigenic and genetic dynamics of the influenza virus population. 
+We apply this approach to investigate the dynamics of influenza A (H3N2), influenza A (H1N1) and influenza B. 
 \tbc{Brief conclusions.}
@@
 	\centering		
-	\makebox[\textwidth]{	
-%		\includegraphics{figures/mds}
+	\includegraphics[width=\textwidth]{figures/mds}
+	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
+	The map has been oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
+	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
+	The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX} for H3N2 and \tbc{XXX} for H1N1.  
+	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
+	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.  
+	\tbc{Decide whether to just show mean posterior or show distributions.}
 	} 
-	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the
-mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  The map has been
-oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2)
-orthogonal to this axis.  (C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year
-of virus isolation.  The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX}
-for H3N2 and \tbc{XXX} for H1N1.  (E) and (F) Antigenic location along the secondary axis of variation (AG2)
-vs.\ year of virus isolation.  Antigenic units represent two-fold dilutions of the HI assay, and strains have
-been colored based on year of isolation.  \tbc{Decide whether to just show mean posterior or show
-distributions.}}
 	\label{mds} 
@@
 
-We compiled an antigenic dataset for hemagglutination inhibition (HI) measurements for influenza A (H3N2) by
-combining data used in Hay et al. \cite{Hay01}, Smith et al.\ \cite{Smith04}, Russell et al.\
-\cite{Russell08}, Barr et al.\ \cite{Barr10} and Cox et al.\ \tbc{WHO report}. This combined dataset had 1651
-influenza isolates (present as either virus or sera in HI comparisons) dating from 1968 to 2011. However, the
-majority of isolates date from 2002 to 2007. Because we are interested in longer-term antigenic evolution, we
-censored the data to have at most 20 strains per year, preferentially keeping those strains with more
-antigenic comparisons. This censoring left 338 strains present as 320 viruses and 438 sera (replicate sera are
-often constructed from the same strain). Across these viruses and sera, we observe 7240 HI measurements. We
-queried the IRD \cite{IRD} and GISAID \tbc{CITE} sequence databases for HA nucleotide sequences based on
-strain names, e.g.\ A/HongKong/1/1968, of these strains. If a strain had multiple sequences in the databases
-we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. Sequences were
-aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
+We compiled an antigenic dataset for hemagglutination inhibition (HI) measurements for influenza A (H3N2) by combining data used in Hay et al.
+\cite{Hay01}, Smith et al.\ \cite{Smith04}, Russell et al.\ \cite{Russell08}, Barr et al.\ \cite{Barr10} and Cox et al.\ \tbc{WHO report}.
+This combined dataset had 1651 influenza isolates (present as either virus or sera in HI comparisons) dating from 1968 to 2011. 
+However, the majority of isolates date from 2002 to 2007. 
+Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 strains per year, preferentially keeping those strains with more antigenic comparisons. 
+This censoring left 338 strains present as 320 viruses and 438 sera (replicate sera are often constructed from the same strain). 
+Across these viruses and sera, we observe 7240 HI measurements. 
+We queried the IRD \cite{IRD} and GISAID \tbc{CITE} sequence databases for HA nucleotide sequences based on strain names, e.g.\ A/HongKong/1/1968, of these strains. 
+If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
+Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
 
@@
 
-We follow Smith et al. \cite{Smith04} and represent antigenic locations on a 2D antigenic map. Through the
-hemagglutination inhibition (HI) assay, there exist measurements of the cross-reactivity of hemagglutinin (HA)
-from one virus strain to serum raised against another strain \cite{Hirst43}. Thus, antigenic phenotype is
-measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to sera from strain
-$j$. Due to experimental constraints, the distance matrix $\mathbf{H}$ is sparse; most comparisons have not be
-made. Our goal is to find an optimal projection of the high-dimensional distance matrix into a lower number of
-dimensions. We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a
-probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations
-to the observed matrix HI measurements.
+We follow Smith et al. \cite{Smith04} and represent antigenic locations on a 2D antigenic map. 
+Through the hemagglutination inhibition (HI) assay, there exist measurements of the cross-reactivity of hemagglutinin (HA) from one virus strain to serum raised against another strain \cite{Hirst43}. 
+Thus, antigenic phenotype is measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to sera from strain $j$. 
+Due to experimental constraints, the distance matrix $\mathbf{H}$ is sparse; most comparisons have not be made. 
+Our goal is to find an optimal projection of the high-dimensional distance matrix into a lower number of dimensions. 
+We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix HI measurements.
 
 Let $\mathbf{X}_i \in \Re^{P}$ represent the cartographic location of strain $i$ for $i = 1,\ldots, N$.
-Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  This gives set of distances
-between cartographic locations 
+Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
+This gives set of distances between cartographic locations 
 \begin{equation}
@@
 \end{equation}
-where $1 ( \cdot )$ is the indicator function for truncation. The likelihood of the antigenic distance
-measures is proportional to
+where $1 ( \cdot )$ is the indicator function for truncation. 
+The likelihood of the antigenic distance measures is proportional to
 \begin{equation}
@@
 \end{equation}
-where $m$ is the cardinality of $\cal I$, $SSR$ is the sum of the squared residuals and $\Phi(\cdot)$ is the
-standard normal CDF. We consider the conjugate prior $\omega \sim \mbox{Gamma}(a, b)$, and assume a uniform
+where $m$ is the cardinality of $\cal I$, $SSR$ is the sum of the squared residuals and $\Phi(\cdot)$ is the standard normal CDF. 
+We consider the conjugate prior $\omega \sim \mbox{Gamma}(a, b)$, and assume a uniform
 prior over $\mathbf{X}$.

### 2012-04-03 flux.tex
Reverted.
Moving BibTeX references into the repo.
Cleaning up Gytis's references.
Including some initial results on MDS error and antigenic drift.

@@
 % cite package, to clean up citations in the main text. Do not remove.
-%\usepackage{cite}
+\usepackage{cite}
 %\renewcommand\citeleft{(}
@@
 \author[2,3,4]{Marc A. Suchard}
+\author[1]{Gytis Dudas}
 \author[5]{Philippe Lemey}
-\author[1]{Gytis Dudas}
 \author[6]{Colin Russell}
@@
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain.
-Here is a new sentence.
 Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over established strains. 
@@
 
+\subsection*{Bayesian multidimensional scaling}
+
+In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS.
+In this model, each virus strain is given a $N$ dimensional antigenic location, and the distance between strains in this antigenic space is proportional to their level of cross-reactivity, as measured by the hemagglutination inhibition (HI) assay.
+The BMDS model provides an expectation for HI titers, such that a map distance of one antigenic unit between strains translates to a 2-fold drop in HI titer.
+A map that produces pairwise distances congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
+We integrate over sources of uncertainty, such as strain location, in flexible Bayesian fashion.
+
+We started with a BMDS implementation of antigenic model used by Smith et al.
+\cite{Smith04}, where each ferret sera and each virus isolate receive a unique location in a 2D antigenic landscape, and map distance is proportional to drop in HI titer relative to the maximum for a particular ferret sera.
+This model performed well, yielding an average absolute predictive error of 0.71 $\mathrm{log}_2$ HI titers on the 6545 training measurements used to build the model, and an average absolute error of 1.23 on 723 test measurements (Table \ref{errortable}).
+We extended this model by forcing sera and virus from the same influenza strain to share an antigenic location, reducing the number of required parameters and decreasing test error to 1.05.
+We further extended this model by sampling the expected titer for the homologous comparison to the sera strain (column effects) and the expected titer for the homologous comparison to the virus strain (row effects).
+We found that adding these column and row effects, decreased training error to 0.67 and test error to 0.87.
+We experimented with increasing the number of dimensions beyond two, but found that further increases in dimensionality did not significantly improve test error (Table \ref{errortable}).
+
+\subsection*{Antigenic drift in across influenza types and subtypes}
+
+Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{fluAGrid}A).  
+Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
+Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+To solve this problem of identifiability, we enforced a weak prior that favors linear movement in the 2D antigen space.
+We find that influenza A (H3N2) moved XXX antigenic units per year (XXX--XXX 95\% credible interval) from 1968 to 2011 (Figure~\ref{fluAGrid}C).
+However, the rate of antigenic drift was not constant, with year-to-year movement of the mean antigenic phenotype averaging XXX units, but showing an interquartile range of XXX--XXX.
+
 %%% Figure 1: mds %%%
@@
 	\centering		
-	\includegraphics[width=\textwidth]{figures/mds}
+	\includegraphics[width=\textwidth]{figures/fluAGrid}
 	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
@@
 	} 
-	\label{mds} 
+	\label{fluAGrid} 
 \end{figure}
+
+\begin{table}[tb]
+	\centering
+	\caption{\textbf{MDS precision and absolute error of $\mathrm{log}_2$ HI titer prediction for training and test data across models.}}
+	\label{errortable}
+	\begin{tabular*}{1.0\textwidth}{ c c c c c c c } 
+	\hline
+	Locations 	& 	Column effects 	&	Row effects	&	MDS		&	Precision	& 	Training error	& 	Test error	\\
+	\hline				
+	Unique 		&	Fixed 			&	None		&	2D		&	0.79		& 	0.71			&	1.23		\\
+	Shared		&	Fixed			&	None		&	2D		&	0.52		&	0.91			&	1.20		\\
+	Shared		&	Sampled			&	None		&	2D		&	0.62		&	0.88			&	1.05		\\
+	Shared		&	Sampled			&	Sampled		&	2D		&	0.87		&	0.72			&	0.88		\\
+	Shared		&	Sampled			&	Sampled		&	3D		&	0.92		&	0.67			&	0.87		\\
+	Shared		&	Sampled			&	Sampled		&	4D		&	XXX			&	XXX				&	XXX			\\
+	Shared		&	Sampled			&	Sampled		&	5D		&	XXX			&	XXX				&	XXX			\\	
+	\hline
+	\end{tabular*}
+\end{table}
 
@@
 
-Antigenic data for influenza A (H1N1) was collected from sources... \tbc{ask Gytis}.
+Antigenic data for influenza A (H1N1) was collected from \cite{Faress05,Chakraverty82,Chakraverty86,Cox83,Daniels85,Daum02,Donatelli93,Kendal78,McDonald07,Nakajima79,Nakajima81,Pereira82,Stevens87,Webster79,Raymond86}.
 
@@
 \bibliographystyle{plos}
-\bibliography{/Users/bedfordt/Documents/bedford}
+\bibliography{flux}
 

### 2012-04-09 flux.tex
Including description of training vs test error.  Some description of antigenic drift.

--- 2012-03-09 flux.tex
@@
 
-%%% Figure 1: mds %%%
+% Antigenic drift with 95% CI
+% H3N2: 1.159 ()
+% H1N1: 0.643 ()
+% Vic:  0.445 ()
+% Yam:  0.203 (0.150-0.260)
+
+\subsection*{Bayesian multidimensional scaling}
+
+In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS.
+In this model, each virus strain is given an $N$ dimensional location, thus specifying an `antigenic map', and the distance between strains in this space is inversely proportional to their level of cross-reactivity.
+In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
+Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
+We integrate over sources of uncertainty, such as strain location, in a flexible Bayesian fashion.
+
+We began with a BMDS analog of the MDS model used by Smith et al.
+\cite{Smith04}, where viruses and sera are represented as 2D locations and HI expectation is relative to the maximum titer of a particular ferret sera.
+This model performed well, yielding an average absolute predictive error of 0.66 log$_2$ HI titers on the 6545 training measurements used to build the model, and an average error of 0.86 on an additional 723 test measurements (Table \ref{errortable}).
+We find that a model with two dimensional antigenic locations better predicts additional HI titers than models with higher (or lower) dimensionality (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
+We extend this model by estimating the expected homologous titer of the sera rather than fixing this at its maximum titer, and additionally, by including the expected homologous titer of the virus in calculating an expectated HI titer.
+We refer to these estimates as column effects and row effects, respectively.
+We found that including these effects decreased training error to 0.50 and test error to 0.75 log$_2$ HI titers.
+
+%%% Table 1: errortable %%%
+\begin{table}[tb]
+	\centering
+	\caption{\textbf{Absolute prediction error of log$_2$ HI titer for training and test data across models.}}
+	\label{errortable}
+	\begin{tabular}{ c c c c c } 
+	\hline
+	MDS 	& 	Column effects 	&	Row effects	& 	Training error	& 	Test error	\\
+	\hline	
+	1D 		&	Fixed 			&	None		& 	0.91			&	1.03		\\	
+	2D 		&	Fixed 			&	None		& 	0.66			&	0.86		\\
+	3D 		&	Fixed 			&	None		& 	0.65			&	0.88		\\
+	4D 		&	Fixed 			&	None		& 	0.71			&	0.96		\\
+	5D 		&	Fixed 			&	None		& 	0.71			&	1.06		\\	
+	2D 		&	Estimated 		&	None		& 	0.55			&	0.77		\\	
+	2D 		&	Estimated 		&	Estimated	& 	0.50			&	0.75		\\		
+	\hline
+	\end{tabular}
+\end{table}
+
+\subsection*{Antigenic drift across influenza types and subtypes}
+
+Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{fluAGrid}A).  
+Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
+Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model.
+
+We find that influenza A (H3N2) evolved XXX antigenic units per year (XXX--XXX 95\% credible interval) from 1968 to 2011 (Figure~\ref{fluAGrid}C).
+However, the rate of antigenic drift was not constant, with year-to-year movement of the mean antigenic phenotype averaging XXX units, but showing an interquartile range of XXX--XXX.  
+Large jumps in antigenic phenotype (Figure~\ref{fluAGrid}C) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
+Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{fluAGrid}E), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
+
+We find that influenza A (H1N1) evolved in antigenic phenotype substantially slower than H3N2, with an average rate of XXX antigenic units per year (XXX--XXX 95\% credible interval).  
+Year-to-year antigenic drift in H1N1 influenza averages XXX units, buts shows an interquartile range of XXX--XXX.
+H1N1 influenza shows a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands 2006 cluster.  
+It is unclear whether H1N1 undergoes fewer antigenic transitions of comparable magnitude to H3N2, or whether antigenic transitions in H1N1 are of smaller scale.
+
+%%% Figure 1: fluAGrid %%%
 \begin{figure}[tb]
 	\centering		
-	\includegraphics[width=\textwidth]{figures/mds}
+	\includegraphics[width=\textwidth]{figures/fluAGrid}
 	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
-	The map has been oriented so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
+	The model has been constructed so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
 	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
@@
 	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
-	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.  
-	\tbc{Decide whether to just show mean posterior or show distributions.}
-	} 
-	\label{mds} 
+	Points represent median strain locations and translucent areas represent 50\% highest density regions.
+	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.} 
+	\label{fluAGrid} 
 \end{figure}

### 2012-04-09 flux.tex
Description of surveillance comparison.

@@
 
-\tbc{Introduction focuses on vaccine selection to tie things together with the planned conclusion.}
-
 Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
@@
 We apply this approach to investigate the dynamics of influenza A (H3N2), influenza A (H1N1) and influenza B.
-\tbc{Brief conclusions.}
 
@@
 \section*{Results}
-
-% Antigenic drift with 95% CI
-% H3N2: 1.159 ()
-% H1N1: 0.643 ()
-% Vic:  0.445 ()
-% Yam:  0.203 (0.150-0.260)
 
@@
 
-%%% Table 1: errortable %%%
+%%% errortable %%%
 \begin{table}[tb]
@@
 
-\subsection*{Antigenic drift across influenza types and subtypes}
+\subsection*{Antigenic drift across influenza lineages}
 
-Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{fluAGrid}A).  
+% Antigenic drift with 95% CI
+% H3N2: 1.159 (1.126-1.203)
+% H1N1: 0.643 (0.585-0.703)
+% Vic:  0.449 (0.397-0.506)
+% Yam:  0.203 (0.150-0.260)
+
+Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{flu_grid}A).  
 Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
 Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model.
+To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
 
-We find that influenza A (H3N2) evolved XXX antigenic units per year (XXX--XXX 95\% credible interval) from 1968 to 2011 (Figure~\ref{fluAGrid}C).
-However, the rate of antigenic drift was not constant, with year-to-year movement of the mean antigenic phenotype averaging XXX units, but showing an interquartile range of XXX--XXX.  
-Large jumps in antigenic phenotype (Figure~\ref{fluAGrid}C) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
-Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{fluAGrid}E), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
+We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{flu_grid}C).
+However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.40--1.79.  
+Large jumps in antigenic phenotype (Figure~\ref{flu_grid}C) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
+Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{flu_grid}E), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
 
-We find that influenza A (H1N1) evolved in antigenic phenotype substantially slower than H3N2, with an average rate of XXX antigenic units per year (XXX--XXX 95\% credible interval).  
-Year-to-year antigenic drift in H1N1 influenza averages XXX units, buts shows an interquartile range of XXX--XXX.
+We find that other types and subtypes of influenza evolved in antigenic phenotype substantially slower than influenza A (H3N2).
+Influenza A (H1N1) evolved at an average rate of 0.64 antigenic units per year (HPD 0.59--0.70) since its reemergence in 1977 to 2009.  
 H1N1 influenza shows a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands 2006 cluster.  
 It is unclear whether H1N1 undergoes fewer antigenic transitions of comparable magnitude to H3N2, or whether antigenic transitions in H1N1 are of smaller scale.  
+Both the Victoria and Yamagata lineages of influenza B also evolved relatively slowely, with average rates of 0.45 (HPD 0.40--0.51) and 0.20 (HPD 0.15--0.26) antigenic units per year, respectively.  
+From 2000 to 2011, both the Victoria and Yamagata lineages show slow antigenic movement, but no major antigenic transitions.
 
-%%% Figure 1: fluAGrid %%%
+%%% flu_grid %%%
 \begin{figure}[tb]
 	\centering		
-	\includegraphics[width=\textwidth]{figures/fluAGrid}
-	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} (A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
+	\includegraphics[width=\textwidth]{figures/flu_grid}
+	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} 
+	(A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
 	The model has been constructed so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
@@
 	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.} 
-	\label{fluAGrid} 
+	\label{flu_grid} 
+\end{figure}
+
+\subsection*{Competition between influenza lineages}
+
+We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata.
+From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Victoria and B/Yamagata split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
+Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
+Influenza A (H1N1) accounted for 23\% of incidence, B/Victoria accounted for 12\% of incidence and B/Yamagata accounted for 10\% of incidence.
+These proportions are significantly similar to one another (randomization test $p = 0.010$).
+
+%%% drift_incidence %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=\textwidth]{figures/drift_incidence}
+	\caption{\textbf{Relative antigenic drift and incidence between lineages of influenza from 1999 to 2009.} 
+	(A) Relative antigenic drift of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
+	Relative change in antigenic dimension 1 between year $i$ and year $i-1$ is shown for each of the 4 lineages.
+	(B) Relative incidence of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
+	Relative isolation counts of each lineage in the USA for each winter influenza season.
+	} 
+	\label{drift_incidence} 
 \end{figure}

### 2012-05-01 flux.tex
Clarifying methods in the BMDS section.

@@
 \renewcommand\Affilfont{\small \normalfont}
+
+% math notation
+\usepackage{amssymb}
+\newcommand{\point}{f_{\scriptscriptstyle \vert}}
+\newcommand{\threshold}{f_{\textstyle \lrcorner}}
+\newcommand{\interval}{f_{\sqcup}}
 
@@
 Through the hemagglutination inhibition (HI) assay, there exist measurements of the cross-reactivity of hemagglutinin (HA) from one virus strain to serum raised against another strain \cite{Hirst43}. 
-Thus, antigenic phenotype is measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to sera from strain $j$. 
+Thus, antigenic phenotype is measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to serum from strain $j$. 
 Due to experimental constraints, the distance matrix $\mathbf{H}$ is sparse; most comparisons have not be made. 
 Our goal is to find an optimal projection of the high-dimensional distance matrix into a lower number of dimensions. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix HI measurements.
+We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of HI measurements.
 
-Let $\mathbf{X}_i \in \Re^{P}$ represent the cartographic location of strain $i$ for $i = 1,\ldots, N$.
+Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots, n$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots, s$.
+Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
 Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
-This gives set of distances between cartographic locations 
+This gives set of distances between virus and serum cartographic locations 
 \begin{equation}
-	\delta_{ij} =  || \mathbf{X}_i - \mathbf{X}_i ||_2.
+	\delta_{ij} =  || X_i - Y_j ||_2.
 \end{equation}
-We define the antigenic distance between virus from strain $i$ and antisera from strain $j$
+Here, $H_{ij}$ represents the $\mathrm{log}_2$ HI titer of virus $i$ against serum $j$, and HI distance is defined as
 \begin{equation}
-	d_{ij} =  \max{ H_{ij} } + 
-	\log_2 \left(   
-		\frac{ H_{ij} }{ \max{ H_j } }
-	\right).
+	d_{ij} =  \max{ H_j } - H_{ij}.
 \end{equation}
-and let the set $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
-The goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}_1,\ldots,\mathbf{X}_N$ such that
+Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
 \begin{equation}
@@
 \end{equation}
-is minimized. 
-A probabilistic interpretation reformulates the optimization as
+is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
+Here, we instead assume a probabilistic interpretation in which a point estimate of HI distance is normally distributed 
 \begin{equation} 
-d_{ij} \sim \mbox{Normal}( \delta_{ij}, \omega ) \times 1( d_{ij} > 0 ) \mbox{ for all } (i,j) \in \cal I
+	d_{ij} \sim \mbox{Normal}( \delta_{ij}, \sigma^2 ),
 \end{equation}
-where $1 ( \cdot )$ is the indicator function for truncation. 
-The likelihood of the antigenic distance measures is proportional to
+and the likelihood of observing some HI distance given the placement of antigenic locations is 
 \begin{equation} 
-\omega^{m/2} \mbox{exp} \left[
-- \frac{\omega \times SSR}{2} - \sum_{(i,j) \in \cal I} \log \Phi (\omega \times \delta_{ij})
-\right],
+	\point(d_{ij}) = \phi \left( \frac{x-\delta_{ij}}{\sigma} \right),
 \end{equation}
-where $m$ is the cardinality of $\cal I$, $SSR$ is the sum of the squared residuals and $\Phi(\cdot)$ is the standard normal CDF. 
-We consider the conjugate prior $\omega \sim \mbox{Gamma}(a, b)$, and assume a uniform
-prior over $\mathbf{X}$.
+where $\sigma^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
+However, HI assays sometimes show no inhibition of agglutination at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
+In this case, the likelihood of observing the threshold measurement follows the cumulative density of the upper tail of the normal distribution
+\begin{equation} 
+	\threshold(d_{ij}) = 1 - \Phi \left( \frac{x-\delta_{ij}}{\sigma} \right),
+\end{equation}
+where $\Phi(\cdot)$ represents the standard normal CDF.
+Although it is simplest to assume that HI measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80.
+Rather taking the HI titer as 1:40, we can instead treat this as an interval measurement, assuming that the exact HI titer for this measurement would occur somewhere between 1:40 and 1:80.
+Assuming an HI distance represents an interval estimate, we calculcate its probability according to
+\begin{equation} 
+	\interval(d_{ij}) = \Phi \left( \frac{x-\delta_{ij}}{\sigma} \right) - \Phi \left( \frac{x-\delta_{ij}-1}{\sigma} \right).
+\end{equation}
+Throughout the analysis, we use interval probabilities $\interval$ rather than point probabilities $\point$ unless otherwise noted.
+
+We calculcate the likelihood of a set of antigenic locations by multiplying probabilities of individual measurements
+\begin{equation} 
+	L(\mathbf{X}) = \prod_{(i,j) \in \cal I} f(d_{ij}),
+\end{equation}
+using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
+We assume that the prior precision $\omega = 1/\sigma^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we assume a uniform prior over $\mathbf{X}$ and $\mathbf{Y}$.
+
+The preceding model represents HI distance as a drop in titer against the most reactive comparison for a particular antiserum.  
+Here, we relax this assumption and treat the expected homologous titer as a random variable.
+In this case, HI distance follows
+\begin{equation} 
+	d_{ij} =  C_j - H_{ij},
+\end{equation}
+with the set of column effects $C_j$ for $j = 1,\ldots, s$ estimated.
+We assume that $C_j$ values are hierarchically distributed according to a normal distribution.
+The mean value of this distribution is taken from the empirical mean across columns in the $\mathrm{log}_2$ HI matrix, and the variance across sera is estimated.
+This formulation assumes that particular sera are more reactive in general than other sera, we follow the same logic and by including an effective of a particular viruses being more reactive in general than other viruses.
+Here, HI distance follows
+\begin{equation} 
+	d_{ij} =  \frac{1}{2} (R_i+C_j) - H_{ij},
+\end{equation}
+and the set of row effects $R_i$ for $i = 1,\ldots, n$ is modeled in an analogous hierarchical fashion, with the variance across viruses estimated.
+We assume separate Gamma(0.001,1000) hyperpriors on the precisions of the $C$ and $R$ distributions.
+
+We integrate over uncertainty using the Markov chain Monto Carlo (MCMC) procedures implemented in the phylogenetic package BEAST.
+Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, moves to MDS precision $\omega$ and moves to $C$ and $R$ precision.
 

### 2012-05-09 flux.tex
Revising.

@@
 
-\usepackage{authblk}
-\renewcommand\Authands{ \& }
-\renewcommand\Authfont{\normalsize \bf}
-\renewcommand\Affilfont{\small \normalfont}
+%\usepackage{authblk}
+%\renewcommand\Authands{ \& }
+%\renewcommand\Authfont{\normalsize \bf}
+%\renewcommand\Affilfont{\small \normalfont}
 
@@
 \title{\vspace{1.0cm} \LARGE \bf Integrating influenza antigenic dynamics with molecular evolution}
-
-\author[1]{Trevor Bedford}
-\author[2,3,4]{Marc A. Suchard}
-\author[1]{Gytis Dudas}
-\author[5]{Philippe Lemey}
-\author[6]{Colin Russell}
-\author[6,7]{Derek Smith}
-\author[1,8]{Andrew Rambaut}
-
-\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[4]{Department of Biostatistics, UCLA School of Public Health, University of California, Los Angeles CA, USA}
-\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
-\affil[6]{Department of Zoology, University of Cambridge, Cambridge, UK.}
-\affil[7]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
-\affil[8]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
-
-\date{\today}
+\author{}
+\date{}
 

### 2012-10-16 flux.tex
Revisions.

@@
 There are currently three major clades of influenza circulating within the human population: influenza A subtype H3N2, influenza A subtype H1N1 and influenza B. 
-Subtypes refer to the genes, hemagglutinin (H) and neuraminidase (N), that are primarily responsible for the antigenic character of a strain. 
+Subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
 Currently, seasonal influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain of influenza B. 
@@
 
-Mutations to the HA1 region of the hemagglutinin (HA) protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
+Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain.
-Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over established strains. 
-The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, representing antigenic similarity and distance as an easily visualized and quantified measure \cite{Smith04}.
+Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains. 
+The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
 The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968. 
-However, evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
+Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
 The process of antigenic drift results in the rapid turnover of the virus population. 
@@
 Here, we implement a flexible Bayesian approach to jointly analyze the antigenic and genetic dynamics of the influenza virus population. 
-We apply this approach to investigate the dynamics of influenza A (H3N2), influenza A (H1N1) and influenza B.
+We apply this approach to investigate the dynamics of influenza lineages A/H3N2,  A/H1N1, B/Victoria and B/Yamagata.
 
@@
 In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS.
-In this model, each virus strain is given an $N$ dimensional location, thus specifying an `antigenic map', and the distance between strains in this space is inversely proportional to their level of cross-reactivity.
+In this model, viruses and sera are given $N$ dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
 In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
 Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
-We integrate over sources of uncertainty, such as strain location, in a flexible Bayesian fashion.
+We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
 
@@
 This model performed well, yielding an average absolute predictive error of 0.66 log$_2$ HI titers on the 6545 training measurements used to build the model, and an average error of 0.86 on an additional 723 test measurements (Table \ref{errortable}).
-We find that a model with two dimensional antigenic locations better predicts additional HI titers than models with higher (or lower) dimensionality (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
-We extend this model by estimating the expected homologous titer of the sera rather than fixing this at its maximum titer, and additionally, by including the expected homologous titer of the virus in calculating an expectated HI titer.
+We find that a model with two dimensional antigenic locations better predicts test data than models with higher (or lower) dimensionality (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
+We extend this model by estimating the strength of overall reactivity of each sera rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus in calculating an expectated HI titer.
 We refer to these estimates as column effects and row effects, respectively.

### 2012-11-29 flux.tex
Revise H3N2 methods paragraph.

@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf Integrating influenza antigenic dynamics with molecular evolution}
+\title{\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography}
 \author{}
@@
 %%% ABSTRACT %%%
+
+% Abstracts explain to the general reader why the research was done and why the results are important.
+They should start with some brief BACKGROUND information: a sentence giving a broad introduction to the field comprehensible to the general reader, and then a sentence of more detailed background specific to your study.
+This should be followed by the RESULTS, or if the paper is more methods/technique oriented an explanation of OBJECTIVES/METHODS and then the RESULTS.
+The final sentence should outline the main CONCLUSIONS of the study, in terms that will be comprehensible to all our readers.
+The abstract should be 125 words or less.
+
 \section*{Abstract}
@@
 
-% Antigenic drift with 95% CI
-% H3N2: 1.159 (1.126-1.203)
-% H1N1: 0.643 (0.585-0.703)
-% Vic:  0.449 (0.397-0.506)
-% Yam:  0.203 (0.150-0.260)
+%%% dynamicstable %%%
+\begin{table}[tb]
+	\centering
+	\caption{\textbf{Antigenic dynamics of influenza lineages.}}
+	\label{errortable}
+	\begin{tabular}{ l c c c c } 
+	\hline
+	 													&	A/H3N2	& 	A/H1N1 	& 	B/Vic	& 	B/Yam	\\
+	\hline	
+	Rate of population drift 							&	1.004	&	0.554	&	0.406	&	0.205	\\
+	Population standard deviation 						&	0.787	&	0.790	&	0.917	&	0.631	\\
+
+	Rate of drift on side branches						&	0.067	&	0.192	&	-0.024	&	0.025	\\	
+	Rate of drift on internal branches					&	0.172	&	0.017	&	0.035	&	0.009	\\
+	Rate of drift on trunk								&	1.011	&	0.465	&	0.015	&	0.417	\\
+
+	Coefficient of diffusion on side branches 			&	0.425	&	0.387	&	0.461	&	0.251	\\
+	Coefficient of diffusion on internal branches		&	0.290	&	0.161	&	0.259	&	0.154	\\
+	Coefficient of diffusion on trunk  					&	0.511	&	0.442 	&	1.098	&	0.260	\\	
+	\hline
+	\end{tabular}
+\end{table}
 
@@
 
-We compiled an antigenic dataset for hemagglutination inhibition (HI) measurements for influenza A (H3N2) by combining data used in Hay et al.
-\cite{Hay01}, Smith et al.\ \cite{Smith04}, Russell et al.\ \cite{Russell08}, Barr et al.\ \cite{Barr10} and Cox et al.\ \tbc{WHO report}.
-This combined dataset had 1651 influenza isolates (present as either virus or sera in HI comparisons) dating from 1968 to 2011. 
-However, the majority of isolates date from 2002 to 2007. 
-Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 strains per year, preferentially keeping those strains with more antigenic comparisons. 
-This censoring left 338 strains present as 320 viruses and 438 sera (replicate sera are often constructed from the same strain). 
-Across these viruses and sera, we observe 7240 HI measurements. 
-We queried the IRD \cite{IRD} and GISAID \tbc{CITE} sequence databases for HA nucleotide sequences based on strain names, e.g.\ A/HongKong/1/1968, of these strains. 
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR2002,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
+We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
 If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
-Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
+This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
+However, the majority of isolates were present from 2002 to 2007. 
+Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
+We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
+This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.
 
 Antigenic data for influenza A (H1N1) was collected from \cite{Faress05,Chakraverty82,Chakraverty86,Cox83,Daniels85,Daum02,Donatelli93,Kendal78,McDonald07,Nakajima79,Nakajima81,Pereira82,Stevens87,Webster79,Raymond86}.
+
+Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
 

### 2012-11-29 flux.tex
Include paragraph on H1N1 data.

@@
 
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR2002,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
 We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
@@
 
-Antigenic data for influenza A (H1N1) was collected from \cite{Faress05,Chakraverty82,Chakraverty86,Cox83,Daniels85,Daum02,Donatelli93,Kendal78,McDonald07,Nakajima79,Nakajima81,Pereira82,Stevens87,Webster79,Raymond86}.
+Antigenic data for influenza A (H1N1) was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
+The same procedure was followed as was followed for H3N2 to match sequence data and to censor antigenic comparisons.
+This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
 

### 2012-11-29 flux.tex
Add paragraphs on Vic and Yam data.

@@
 
+Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Hay01} \tbc{Include in BibTex: Ansaldi04, AusWHO06, Barr06, Daum06, Lin07, Muyanga01, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+Here, the sequence matching and censoring procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
+
+Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Hay01} \tbc{Include in BibTex: Abed03, Ansaldi03, Ansaldi04, AusWHO06, Barr06, Daum06, Kanegae90, Lin07, Matsuzaki04, Muyanga01, Nakagawa02, Nakajima92, Nerome98, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+For B/Yamagata, the matching and censoring procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
+
 Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.

### 2012-11-29 flux.tex
Include authors.

@@
 
-%\usepackage{authblk}
-%\renewcommand\Authands{ \& }
-%\renewcommand\Authfont{\normalsize \bf}
-%\renewcommand\Affilfont{\small \normalfont}
+\usepackage{authblk}
+\renewcommand\Authands{ \& }
+\renewcommand\Authfont{\normalsize \bf}
+\renewcommand\Affilfont{\small \normalfont}
 
@@
 \title{\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography}
-\author{}
+
+\author[1]{Trevor Bedford}
+\author[2,3,4]{Marc A. Suchard}
+\author[5]{Philippe Lemey}
+\author[1]{Gytis Dudas}
+\author[6]{Colin Russell}
+\author[6,7]{Derek Smith}
+\author[1,8]{Andrew Rambaut}
+
+\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
+\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
+\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
+\affil[4]{Department of Biostatistics, UCLA School of Public Health, University of California, Los Angeles CA, USA}
+\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
+\affil[6]{Department of Zoology, University of Cambridge, Cambridge, UK.}
+\affil[7]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
+\affil[8]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
+
 \date{}

### 2012-11-29 flux.tex
Revise BMDS methods.

@@
 
-% math notation
+% notation
 \usepackage{amssymb}
-\newcommand{\point}{f_{\scriptscriptstyle \vert}}
-\newcommand{\threshold}{f_{\textstyle \lrcorner}}
-\newcommand{\interval}{f_{\sqcup}}
+\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
+\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
+\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
+\newcommand{\mdssd}{\varphi}						% MDS standard deviation
+\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
+\newcommand{\tree}{\tau}							% Phylogeny
+\newcommand{\vn}{n}									% Number of viruses
+\newcommand{\sn}{k}									% Number of sera
 
@@
 
-\subsection*{Bayesian multidimensional scaling}
+\subsection*{Antigenic cartography and Bayesian multidimensional scaling}
 
-We follow Smith et al. \cite{Smith04} and represent antigenic locations on a 2D antigenic map. 
-Through the hemagglutination inhibition (HI) assay, there exist measurements of the cross-reactivity of hemagglutinin (HA) from one virus strain to serum raised against another strain \cite{Hirst43}. 
-Thus, antigenic phenotype is measured through a series a pairwise comparisons $H_{ij}$, comparing virus from strain $i$ to serum from strain $j$. 
-Due to experimental constraints, the distance matrix $\mathbf{H}$ is sparse; most comparisons have not be made. 
-Our goal is to find an optimal projection of the high-dimensional distance matrix into a lower number of dimensions. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of HI measurements.
+Antigenic characteristics of viral strains are often assessed through immunological assays such as hemagglutination inhibition (HI) \cite{Hirst43}.  
+At heart, these assays compare the reactivity $H_{ij}$ of one virus strain $i$ to antibodies raised against another virus strain $j$ via challenge or vaccination.  
+In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.
+These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
+However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
+Further, measurements are usually interval and truncated, e.g.\ inhibition or neutralization may cease somewhere between the serial titers of 160 and 320, or inhibition or neutralization may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
 
-Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots, n$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots, s$.
+Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
+These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
+Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data has not yet been attained.
+
+Here, we follow previous models in representing antigenic locations as points in an $N$-dimensional antigenic map. 
+Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions. 
+We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
+
+Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
 Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
@@
 \end{equation}
-Here, $H_{ij}$ represents the $\mathrm{log}_2$ HI titer of virus $i$ against serum $j$, and HI distance is defined as
+Here, $H_{ij}$ represents the log$_2$ titer of virus $i$ against serum $j$, and immunological distance can be defined as
 \begin{equation}
-	d_{ij} =  \max{ H_j } - H_{ij}.
+	d_{ij} =  S_j - H_{ij},
 \end{equation}
+where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
 Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
@@
 is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
-Here, we instead assume a probabilistic interpretation in which a point estimate of HI distance is normally distributed 
+Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
 \begin{equation} 
-	d_{ij} \sim \mbox{Normal}( \delta_{ij}, \sigma^2 ),
+	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
 \end{equation}
-and the likelihood of observing some HI distance given the placement of antigenic locations is 
+and the likelihood of observing a particular titer given the placement of antigenic locations is 
 \begin{equation} 
-	\point(d_{ij}) = \phi \left( \frac{x-\delta_{ij}}{\sigma} \right),
+	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
 \end{equation}
-where $\sigma^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
-However, HI assays sometimes show no inhibition of agglutination at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
-In this case, the likelihood of observing the threshold measurement follows the cumulative density of the upper tail of the normal distribution
+where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
+HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
+In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
 \begin{equation} 
-	\threshold(d_{ij}) = 1 - \Phi \left( \frac{x-\delta_{ij}}{\sigma} \right),
+	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
 \end{equation}
 where $\Phi(\cdot)$ represents the standard normal CDF.
-Although it is simplest to assume that HI measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80.
-Rather taking the HI titer as 1:40, we can instead treat this as an interval measurement, assuming that the exact HI titer for this measurement would occur somewhere between 1:40 and 1:80.
-Assuming an HI distance represents an interval estimate, we calculcate its probability according to
+Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
+Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
+HI titers are usually reported as the highest titer that successfully inhibits virus binding or growth, so that in this case, we calculate the likelihood of an interval measurement as
 \begin{equation} 
-	\interval(d_{ij}) = \Phi \left( \frac{x-\delta_{ij}}{\sigma} \right) - \Phi \left( \frac{x-\delta_{ij}-1}{\sigma} \right).
+	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
 \end{equation}
-Throughout the analysis, we use interval probabilities $\interval$ rather than point probabilities $\point$ unless otherwise noted.
+These likelihoods are illustrated in Figure \ref{hij_likelihood}.
+Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
 
-We calculcate the likelihood of a set of antigenic locations by multiplying probabilities of individual measurements
+%%% hij_likelihood %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
+	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
+	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $S_j = \mathrm{log}_2 1280 = 10.32$.  
+	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
+	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
+	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
+	} 
+	\label{hij_likelihood} 
+\end{figure}
+
+We calculate the overall likelihood by multiplying probabilities of individual measurements
 \begin{equation} 
-	L(\mathbf{X}) = \prod_{(i,j) \in \cal I} f(d_{ij}),
+	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
 \end{equation}
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $\omega = 1/\sigma^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we assume a uniform prior over $\mathbf{X}$ and $\mathbf{Y}$.
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a uniform prior over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$.
 
-The preceding model represents HI distance as a drop in titer against the most reactive comparison for a particular antiserum.  
-Here, we relax this assumption and treat the expected homologous titer as a random variable.
-In this case, HI distance follows
-\begin{equation} 
-	d_{ij} =  C_j - H_{ij},
-\end{equation}
-with the set of column effects $C_j$ for $j = 1,\ldots, s$ estimated.
-We assume that $C_j$ values are hierarchically distributed according to a normal distribution.
-The mean value of this distribution is taken from the empirical mean across columns in the $\mathrm{log}_2$ HI matrix, and the variance across sera is estimated.
-This formulation assumes that particular sera are more reactive in general than other sera, we follow the same logic and by including an effective of a particular viruses being more reactive in general than other viruses.
-Here, HI distance follows
-\begin{equation} 
-	d_{ij} =  \frac{1}{2} (R_i+C_j) - H_{ij},
-\end{equation}
-and the set of row effects $R_i$ for $i = 1,\ldots, n$ is modeled in an analogous hierarchical fashion, with the variance across viruses estimated.
-We assume separate Gamma(0.001,1000) hyperpriors on the precisions of the $C$ and $R$ distributions.
-
-We integrate over uncertainty using the Markov chain Monto Carlo (MCMC) procedures implemented in the phylogenetic package BEAST.
-Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, moves to MDS precision $\omega$ and moves to $C$ and $R$ precision.
+We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST}.
+Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, scaling of the entire set of virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ and scaling $\mdssd$.
 

### 2012-11-29 flux.tex
Include section on virus/serum effects.

@@
 
+\subsection*{Incorporating virus and serum effects}
+
+The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular antiserum.  
+Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
+In this case, $H_{ij}$ still follows
+\begin{equation} 
+	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
+\end{equation}
+but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
+We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
+The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
+This formulation assumes that particular sera are more reactive in general than other sera.
+
+Additionally, we follow the same logic and assume that particular virus isolates are more reactive in general than other virus isolates.
+With virus reactivity included, $H_{ij}$ follows
+\begin{equation} 
+	H_{ij} \sim \mbox{Normal}\left(  \frac{V_i+S_j}{2} - \delta_{ij}, \mdssd^2 \right),
+\end{equation}
+and the vector of `virus effects' $V_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\mathbf{V}$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+
+With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $S_j$ and virus effects $V_i$.
+
 %%% ACKNOWLEDGMENTS %%%

### 2012-11-29 flux.tex
Include section on antigenic drift prior.

@@
 \newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
+\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
 \newcommand{\tree}{\tau}							% Phylogeny
@@
 
+\subsection*{Antigenic drift prior}
+
+As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
+However, these distances represent only a local description of antigenic space and do not provide a global picture.
+An example of this phenomena is shown in Figure \ref{schematic_map}.
+In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
+This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
+In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
+In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}, and so we impose a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
+\begin{equation} 
+	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
+\end{equation}
+for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
+Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
+The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
+We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
+We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.
+
+%%% schematic_map %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
+	\caption{\textbf{Schematic antigenic map with three viruses and two sera.} 
+	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
+	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
+	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 1 and 3.
+	Left and right panels represent cartographic models that would give equal likelihoods of a given set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
+	} 
+	\label{schematic_map} 
+\end{figure}
+
 %%% ACKNOWLEDGMENTS %%%

### 2012-11-30 flux.tex
Include section on diffusion model.

@@
 %%% ABSTRACT %%%
+\section*{Abstract}
 
@@
 The abstract should be 125 words or less.
-
-\section*{Abstract}
 
@@
 
-There are currently three major clades of influenza circulating within the human population: influenza A subtype H3N2, influenza A subtype H1N1 and influenza B. 
+Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
 Subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
@@
 
-\subsection*{Antigenic cartography and Bayesian multidimensional scaling}
+\subsection*{Bayesian multidimensional scaling}
 
@@
 
+\subsection*{Diffusion model of antigenic evolution}
+
+Most previous work on virus antigenic and genetic evolution has reconstructed antigenic and genetic patterns in isolation, and then compared the results of each reconstruction to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
+Here, we simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
+We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations
+\begin{eqnarray}
+	(X_1,\ldots,X_n) &\sim& \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd) \nonumber \\
+	Y_j &\sim& \mathrm{Uniform}( -\infty, \infty ),
+\end{eqnarray}
+where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
+Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
+The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
+The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
+Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
+Metropolis-Hastings proposals scale $\diffusionsd$.
+The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
+\begin{equation}
+	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
+	p(\mathbf{X} | \beta, \vartheta) \;
+	p(\mathbf{X} | \tree, \diffusionsd) \; 
+	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
+\end{equation}
+
 %%% ACKNOWLEDGMENTS %%%

### 2012-11-30 flux.tex
Include short (125-word) abstract.

@@
 The abstract should be 125 words or less.
+
+Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
+However, the extent to which antigenic evolution determines competitive dynamics within and between virus clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata has remained unclear.
+Here, we develope a novel cartographic approach to simultaneously characterize the genetic and antigenic evolution of viral lineages.
+Through this approach, we determine historical rates and patterns of antigenic drift in all four viral clades and we show that antigenic drift influences year-to-year variability in clade dominance.
+We investigate the selective underpinnings for differing antigenic dynamics across clades and suggest that pleiotropic effects linked to antigenic mutation determine selective outcomes.
 

### 2012-11-30 flux.tex
Include minor revisions.

@@
 Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
-Subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
+In the case of influenza A, subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
+In the case of influenza B, lineages Victoria and Yamagata diverged from one another around 1980 \cite{Rota92}.
 Currently, seasonal influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain of influenza B. 

### 2012-11-30 flux.tex
Include revisions to introduction and BMDS validation.

@@
 
-% Abstracts explain to the general reader why the research was done and why the results are important.
-They should start with some brief BACKGROUND information: a sentence giving a broad introduction to the field comprehensible to the general reader, and then a sentence of more detailed background specific to your study.
-This should be followed by the RESULTS, or if the paper is more methods/technique oriented an explanation of OBJECTIVES/METHODS and then the RESULTS.
-The final sentence should outline the main CONCLUSIONS of the study, in terms that will be comprehensible to all our readers.
-The abstract should be 125 words or less.
-
 Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
@@
 In the case of influenza A, subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, lineages Victoria and Yamagata diverged from one another around 1980 \cite{Rota92}.
-Currently, seasonal influenza is treated with a trivalent vaccine containing one strain of H3N2, one strain of H1N1 and one strain of influenza B. 
-The World Health Organization (WHO) Global Influenza Surveillance Network issues twice-yearly recommendations on which strains of influenza to use as vaccine strains in 9--12 month’s time, i.e.\ a February recommendation for the Northern Hemisphere flu season and an August recommendation for the Southern Hemisphere flu season.
-These recommendations are provided by a panel of experts after review of the available data.
+In the case of influenza B, lineages Victoria and Yamagata diverged from one prior to 1980 \cite{Rota92}.
+Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
+Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
+Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
+HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata \cite{Hay01}.
 
-Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
-Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain.
-Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains. 
 The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
@@
 However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
-Here, we implement a flexible Bayesian approach to jointly analyze the antigenic and genetic dynamics of the influenza virus population. 
-We apply this approach to investigate the dynamics of influenza lineages A/H3N2,  A/H1N1, B/Victoria and B/Yamagata.
+Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
+We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata.
 
 %%% RESULTS %%%
-\section*{Results}
+\section*{Results and Discussion}
 
-\subsection*{Bayesian multidimensional scaling}
+\subsection*{Antigenic cartography}
 
-In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS.
-In this model, viruses and sera are given $N$ dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
+In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS (see Methods).
+In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
 In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
@@
 We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
+We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata (see Methods).
 
-We began with a BMDS analog of the MDS model used by Smith et al.
-\cite{Smith04}, where viruses and sera are represented as 2D locations and HI expectation is relative to the maximum titer of a particular ferret sera.
-This model performed well, yielding an average absolute predictive error of 0.66 log$_2$ HI titers on the 6545 training measurements used to build the model, and an average error of 0.86 on an additional 723 test measurements (Table \ref{errortable}).
-We find that a model with two dimensional antigenic locations better predicts test data than models with higher (or lower) dimensionality (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
-We extend this model by estimating the strength of overall reactivity of each sera rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus in calculating an expectated HI titer.
-We refer to these estimates as column effects and row effects, respectively.
-We found that including these effects decreased training error to 0.50 and test error to 0.75 log$_2$ HI titers.
+We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
+By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare predicted titers to observed titers in the test dataset. 
+We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret sera, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
+We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
+We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
+We refer to these estimates as serum effects and virus effects, respectively.
+We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
+We arrive at similar precisions when repeating this analysis for A/H1N1, B/Victoria and B/Yamagata.
+For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Victoria shows a test error of 0.74 and B/Yamagata shows a test error of 0.80.
 
 %%% errortable %%%
-\begin{table}[tb]
+\begin{table}[h]
 	\centering
-	\caption{\textbf{Absolute prediction error of log$_2$ HI titer for training and test data across models.}}
+	\caption{\textbf{Absolute prediction error of log$_2$ HI titer against A/H3N2 test data across models.}}
 	\label{errortable}		
@@
 	\hline
-	MDS 	& 	Column effects 	&	Row effects	& 	Training error	& 	Test error	\\
+	BMDS dimension 	& 	Serum effects 	&	Virus effects	& 	Test error	\\
 	\hline	
-	1D 		&	Fixed 			&	None		& 	0.91			&	1.03		\\	
-	2D 		&	Fixed 			&	None		& 	0.66			&	0.86		\\
-	3D 		&	Fixed 			&	None		& 	0.65			&	0.88		\\
-	4D 		&	Fixed 			&	None		& 	0.71			&	0.96		\\
-	5D 		&	Fixed 			&	None		& 	0.71			&	1.06		\\	
-	2D 		&	Estimated 		&	None		& 	0.55			&	0.77		\\	
-	2D 		&	Estimated 		&	Estimated	& 	0.50			&	0.75		\\		
+	1D 				&	Fixed 			&	None			&	1.05		\\	
+	2D 				&	Fixed 			&	None			&	0.87 		\\
+	3D 				&	Fixed 			&	None			&	0.87		\\
+	4D 				&	Fixed 			&	None			&	0.91		\\
+	5D 				&	Fixed 			&	None			&	0.97		\\	
+	2D 				&	Estimated 		&	None			&	0.78		\\	
+	2D 				&	Estimated 		&	Estimated		&	0.75		\\		
 	\hline
@@
 	\caption{\textbf{Antigenic dynamics of influenza lineages.}}
-	\label{errortable}
+	\label{dynamicstable}
 	\begin{tabular}{ l c c c c } 
@@
 \section*{Methods}
-
-\subsection*{Genetic and antigenic data}
-
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
-We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
-If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
-This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
-However, the majority of isolates were present from 2002 to 2007. 
-Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
-We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
-This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.
-
-Antigenic data for influenza A (H1N1) was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
-The same procedure was followed as was followed for H3N2 to match sequence data and to censor antigenic comparisons.
-This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
-
-Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Hay01} \tbc{Include in BibTex: Ansaldi04, AusWHO06, Barr06, Daum06, Lin07, Muyanga01, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
-Here, the sequence matching and censoring procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
-
-Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Hay01} \tbc{Include in BibTex: Abed03, Ansaldi03, Ansaldi04, AusWHO06, Barr06, Daum06, Kanegae90, Lin07, Matsuzaki04, Muyanga01, Nakagawa02, Nakajima92, Nerome98, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
-For B/Yamagata, the matching and censoring procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
-
-Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
 
@@
 
+\subsection*{Genetic and antigenic data}
+
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
+We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
+If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
+This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
+However, the majority of isolates were present from 2002 to 2007. 
+Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
+We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
+This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.
+
+Antigenic data for influenza A (H1N1) was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
+The same procedure was followed as was followed for H3N2 to match sequence data and to censor antigenic comparisons.
+This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
+
+Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Hay01} \tbc{Include in BibTex: Ansaldi04, AusWHO06, Barr06, Daum06, Lin07, Muyanga01, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+Here, the sequence matching and censoring procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
+
+Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Hay01} \tbc{Include in BibTex: Abed03, Ansaldi03, Ansaldi04, AusWHO06, Barr06, Daum06, Kanegae90, Lin07, Matsuzaki04, Muyanga01, Nakagawa02, Nakajima92, Nerome98, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+For B/Yamagata, the matching and censoring procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
+
+Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
+
 %%% ACKNOWLEDGMENTS %%%

### 2012-11-30 flux.tex
Revise results and include new figures.

@@
 
-Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
-In the case of influenza A, subtypes refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, lineages Victoria and Yamagata diverged from one prior to 1980 \cite{Rota92}.
+Before 2009, there were four major clades of influenza circulating within the human population: influenza A H3N2 and H1N1, and influenza B Victoria and Yamagata. 
+In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
+In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
 Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
@@
 Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
-HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata \cite{Hay01}.
+HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.
 
@@
 Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata.
+We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Vic and B/Yam.
 
@@
 
-\subsection*{Antigenic cartography}
+\subsection*{Antigenic and evolutionary cartography}
 
@@
 We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata (see Methods).
+We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).
 
@@
 We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
-We arrive at similar precisions when repeating this analysis for A/H1N1, B/Victoria and B/Yamagata.
-For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Victoria shows a test error of 0.74 and B/Yamagata shows a test error of 0.80.
+We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
+For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.
 
@@
 
-\subsection*{Antigenic drift across influenza lineages}
+Most previous work on virus antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
+Here, we simultaneously model antigenic and genetic evolution by adapting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
+Here, the diffusion process acts as a prior, so that genetically similar viruses are expected to share similar antigenic phenotypes, and we model the relationship between genetic change and accrual of antigenic distance.
+We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
+We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.
 
-%%% dynamicstable %%%
-\begin{table}[tb]
+\subsection*{Antigenic evolution across influenza lineages}
+
+Through our modeling, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
+We observe prominent antigenic clusters in A/H3N2 and A/H1N1 with less obvious clustering in B/Vic and B/Yam.
+Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
+This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.
+
+%%% map %%%
+\begin{figure}[h]
 	\centering		
-	\caption{\textbf{Antigenic dynamics of influenza lineages.}}
-	\label{dynamicstable}
-	\begin{tabular}{ l c c c c } 
-	\hline
-	 													&	A/H3N2	& 	A/H1N1 	& 	B/Vic	& 	B/Yam	\\
-	\hline	
-	Rate of population drift 							&	1.004	&	0.554	&	0.406	&	0.205	\\
-	Population standard deviation 						&	0.787	&	0.790	&	0.917	&	0.631	\\
+	\includegraphics[width=0.85\textwidth]{figures/map}
+	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
+	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
+	Antigenic units represent two-fold dilutions of the HI assay.
+	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
+	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
+	\label{map} 
+\end{figure}
 
-	Rate of drift on side branches						&	0.067	&	0.192	&	-0.024	&	0.025	\\	
-	Rate of drift on internal branches					&	0.172	&	0.017	&	0.035	&	0.009	\\
-	Rate of drift on trunk								&	1.011	&	0.465	&	0.015	&	0.417	\\
-
-	Coefficient of diffusion on side branches 			&	0.425	&	0.387	&	0.461	&	0.251	\\
-	Coefficient of diffusion on internal branches		&	0.290	&	0.161	&	0.259	&	0.154	\\
-	Coefficient of diffusion on trunk  					&	0.511	&	0.442 	&	1.098	&	0.260	\\	
-	\hline
-	\end{tabular}
-\end{table}
-
-Through this analysis we find that the antigenic phenotype of influenza A (H3N2) underwent rapid turnover from 1968 to 2011 (Figure~\ref{flu_grid}A).  
-Here, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
+For A/H3N2, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
 Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
@@
 
-We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{flu_grid}C).
+We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{drift}).
 However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.40--1.79.  
-Large jumps in antigenic phenotype (Figure~\ref{flu_grid}C) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
-Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{flu_grid}E), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
+Large jumps in antigenic phenotype (Figure~\ref{drift}) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
+Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
+
+%%% drift %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.65\textwidth]{figures/drift}
+	\caption{\textbf{Antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
+	Antigenic drift is shown in terms of change of location in the first antigenic dimension through time.
+	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
+	Antigenic units represent two-fold dilutions of the HI assay.
+	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
+	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
+	\label{drift} 
+\end{figure}
 
@@
 
-%%% flu_grid %%%
+\subsection*{Competition between influenza lineages}
+
+We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Vic and B/Yam.
+From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
+Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
+Influenza A (H1N1) accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 10\% of incidence.
+These proportions are significantly similar to one another (randomization test $p = 0.010$).
+
+%%% incidence %%%
 \begin{figure}[tb]
 	\centering		
-	\includegraphics[width=\textwidth]{figures/flu_grid}
-	\caption{\textbf{Antigenic locations of influenza H3N2 and H1N1.} 
-	(A) and (B) Antigenic maps showing the mean posterior location of 338 strains of H3N2 influenza and 243 strains of H1N1 influenza.  
-	The model has been constructed so that the primary axis of variation lies along the $x$-axis (AG1), with the $y$-axis (AG2) orthogonal to this axis.  
-	(C) and (D) Antigenic location along the primary axis of variation (AG1) vs.\ year of virus isolation.  
-	The dashed lines show the relationship of between time and AG1 with a slope of \tbc{XXX} for H3N2 and \tbc{XXX} for H1N1.  
-	(E) and (F) Antigenic location along the secondary axis of variation (AG2) vs.\ year of virus isolation.  
-	Points represent median strain locations and translucent areas represent 50\% highest density regions.
-	Antigenic units represent two-fold dilutions of the HI assay, and strains have been colored based on year of isolation.} 
-	\label{flu_grid} 
-\end{figure}
-
-\subsection*{Competition between influenza lineages}
-
-We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata.
-From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Victoria and B/Yamagata split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
-Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
-Influenza A (H1N1) accounted for 23\% of incidence, B/Victoria accounted for 12\% of incidence and B/Yamagata accounted for 10\% of incidence.
-These proportions are significantly similar to one another (randomization test $p = 0.010$).
-
-%%% drift_incidence %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=\textwidth]{figures/drift_incidence}
+	\includegraphics[width=\textwidth]{figures/incidence}
 	\caption{\textbf{Relative antigenic drift and incidence between lineages of influenza from 1999 to 2009.} 
@@
 	} 
-	\label{drift_incidence} 
+	\label{incidence} 
 \end{figure}
-
-%%% DISCUSSION %%%
-\section*{Discussion}
 
@@
 \end{equation}
-These likelihoods are illustrated in Figure \ref{hij_likelihood}.
+These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
 Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
@@
 However, these distances represent only a local description of antigenic space and do not provide a global picture.
-An example of this phenomena is shown in Figure \ref{schematic_map}.
+An example of this phenomena is shown in Figure~\ref{schematic_map}.
 In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
@@
 
-Most previous work on virus antigenic and genetic evolution has reconstructed antigenic and genetic patterns in isolation, and then compared the results of each reconstruction to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
-Here, we simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
+We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
 We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations

### 2012-11-30 flux.tex
Revise drift results.

@@
 However, the extent to which antigenic evolution determines competitive dynamics within and between virus clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata has remained unclear.
-Here, we develope a novel cartographic approach to simultaneously characterize the genetic and antigenic evolution of viral lineages.
+Here, we develop a novel cartographic approach to simultaneously characterize the genetic and antigenic evolution of viral lineages.
 Through this approach, we determine historical rates and patterns of antigenic drift in all four viral clades and we show that antigenic drift influences year-to-year variability in clade dominance.
@@
 Through our modeling, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
+Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
 We observe prominent antigenic clusters in A/H3N2 and A/H1N1 with less obvious clustering in B/Vic and B/Yam.
@@
 
-For A/H3N2, cross-reactive measurements only exist betwen strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
+HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
 Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
+Because of this, we interpret antigenic locations locally rather than globally.
+We can determine the rate of antigenic movement of virus lineages without knowing the larger configuration that the movement occurs under.
 
-We find that influenza A (H3N2) evolved 1.16 antigenic units per year from 1968 to 2011, with 95\% highest posterior density (HPD) interval 1.13--1.20 (Figure~\ref{drift}).
-However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.40--1.79.  
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.00 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.97--1.04 (Figure~\ref{drift}).
+However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.00--1.57.  
 Large jumps in antigenic phenotype (Figure~\ref{drift}) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
-Most variation is contained within the first antigenic dimension, but dimension 2 ocassionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing 1989 and Beijing 1992 clusters.
+Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.
 
@@
 
-We find that other types and subtypes of influenza evolved in antigenic phenotype substantially slower than influenza A (H3N2).
-Influenza A (H1N1) evolved at an average rate of 0.64 antigenic units per year (HPD 0.59--0.70) since its reemergence in 1977 to 2009.  
-H1N1 influenza shows a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands 2006 cluster.  
-It is unclear whether H1N1 undergoes fewer antigenic transitions of comparable magnitude to H3N2, or whether antigenic transitions in H1N1 are of smaller scale.  
-Both the Victoria and Yamagata lineages of influenza B also evolved relatively slowely, with average rates of 0.45 (HPD 0.40--0.51) and 0.20 (HPD 0.15--0.26) antigenic units per year, respectively.  
-From 2000 to 2011, both the Victoria and Yamagata lineages show slow antigenic movement, but no major antigenic transitions.
+We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
+Influenza A/H1N1 evolved at a rate of 0.55 antigenic units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
+Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) antigenic units per year.
+Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
+Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward on antigenic dimension 1 has moved down on antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
+Influenza B/Yamagata evolves slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
+Very little punctuated evolution is apparent in the evolution of B/Yam.
 
-\subsection*{Competition between influenza lineages}
+\subsection*{Competition between influenza clades}
 

### 2012-11-30 flux.tex
Change serum prior to diffuse normal.

@@
 % notation
+\usepackage{amsmath}
 \usepackage{amssymb}
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a uniform prior over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$.
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 1000 along diagonal entries and equal to zero along off-diagonal entries.
 
@@
 In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
-In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}, and so we impose a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
+In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
+To capture this, we replace the diffuse normal prior on virus locations with a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
 \begin{equation} 
@@
 We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
-We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations
-\begin{eqnarray}
-	(X_1,\ldots,X_n) &\sim& \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd) \nonumber \\
-	Y_j &\sim& \mathrm{Uniform}( -\infty, \infty ),
-\end{eqnarray}
+To do this, we incorporate a prior on virus locations according to the evolutionary process
+\begin{equation}
+	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
+\end{equation}
 where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.

### 2012-12-01 flux.tex
Added newer BEAST citation

@@
 
-We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST}.
+We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST17}.
 Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, scaling of the entire set of virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ and scaling $\mdssd$.
@@
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
+The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
 The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.

### 2012-12-03 flux.tex
Include paragraph on surveillance data.

@@
 Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward on antigenic dimension 1 has moved down on antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
+Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
 Influenza B/Yamagata evolves slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
@@
 
-\subsection*{Competition between influenza clades}
+\subsection*{Competition between influenza lineages}
 
@@
 
-\subsection*{Genetic and antigenic data}
+\subsection*{Genetic, antigenic and surveillance data}
 
@@
 If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
+Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
 This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
@@
 
-Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
+Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
+As an example, one report states "collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
+Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses.
+Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses."
+In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
+Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
 

### 2012-12-04 flux.tex
Revise results on surveillance match.

@@
 Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
-Influenza B/Yamagata evolves slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
+Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
 Very little punctuated evolution is apparent in the evolution of B/Yam.
@@
 
-We investigated the relative rates of antigenic evolution in different influenza lineages from 1999 to 2009, finding a general correspondence between antigenic drift along antigenic dimension 1 and relative incidence between influenza A/H3N2, A/H1N1, B/Vic and B/Yam.
-From 1999 to 2009, influenza A (H3N2) accounted for the majority of antigenic evolution (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 17\%, 18\% and 14\%, respectively.
-Similarly, influenza A (H3N2) was responsible for the majority of incidence (54\%) in the USA during this time period.
-Influenza A (H1N1) accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 10\% of incidence.
-These proportions are significantly similar to one another (randomization test $p = 0.010$).
+We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
+From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (46\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 21\%, 17\% and 16\%, respectively.
+Similarly, A/H3N2 was responsible for the majority of incidence (54\%) in the USA during this time period.
+Influenza A/H1N1 accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 11\% of incidence.
+These proportions are significantly similar to one another (randomization test, $p = 0.015$).
 
@@
 	\centering		
-	\includegraphics[width=\textwidth]{figures/incidence}
-	\caption{\textbf{Relative antigenic drift and incidence between lineages of influenza from 1999 to 2009.} 
-	(A) Relative antigenic drift of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
-	Relative change in antigenic dimension 1 between year $i$ and year $i-1$ is shown for each of the 4 lineages.
-	(B) Relative incidence of influenza A H3N2 and H1N1 and influenza B Victoria and Yamagata lineages.
-	Relative isolation counts of each lineage in the USA for each winter influenza season.
+	\includegraphics[width=0.95\textwidth]{figures/incidence}
+	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008.} 
+	(A) Relative antigenic drift of influenza clades.
+	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
+	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimensions 1 of viruses collected during the previous year $i-1$.
+	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
+	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
+	Here, `99' refers to the 1999/2000 influenza season.
+	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
+	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g.
+drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
 	} 
@@
 \end{figure}
+
+Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
+Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 in viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
+For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999/2000 winter influenza season is dominanated by A/H3N2.
+Similarly, between samples from 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000/2001 season is dominated by these same two clades.
+Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
+Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.29$, $r_\mathrm{Vic} = 0.75$ and $r_\mathrm{Yam} = 0.58$.
+
+These findings clearly demonstrate the epidemiological importance of antigenic drift and show that competition between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
+Furthmore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
 
@@
 
-We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST17}.
+We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
 Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, scaling of the entire set of virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ and scaling $\mdssd$.
@@
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
+The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST,BEAST17}.
 The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.

### 2012-12-04 flux.tex
Include section on polymorphism/divergence.

@@
 
-\subsection*{Competition between influenza lineages}
+\subsection*{Epidemiological consequences of antigenic drift}
 
@@
 Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 in viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
-For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999/2000 winter influenza season is dominanated by A/H3N2.
+For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999/2000 winter influenza season is dominated by A/H3N2.
 Similarly, between samples from 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000/2001 season is dominated by these same two clades.
@@
 These findings clearly demonstrate the epidemiological importance of antigenic drift and show that competition between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
-Furthmore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
+Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
+
+\subsection*{Conversion of antigenic mutation into fixed differences}
+
+We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or influenza B.
+Previous work using general epidemiological models has suggested that a greater rate of antigenic drift may emerge from either a greater fundamental reproductive number $R_0$ or a faster rate of mutation decreasing cross-immunity \cite{Gog02,Lin03}.
+Owing to these and similar findings, subsequent work has ascribed differences in the evolution in A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+
+Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
+Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations proceding in a random direction.
+However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
+This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
+The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
+For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
+After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drift along dimension 1.
+
+%%% accumulation %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/accumulation}
+	\caption{\textbf{Rate of antigenic diffusion of influenza lineages categorized by time back from sampling.}
+	(A) Rate of diffusion in terms of antigenic units per year only along antigenic dimension 1.
+	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
+	In the diffusion model, the antigenic location of each virus sample can be calculcated back through time.
+	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
+	When the $x$-axis is close to 0, the $y$-axis shows the rate of diffusion near tips of the phylogeny, and when the $x$-axis is large, the $y$-axis shows the rate of diffusion along the trunk of the phylogeny.
+	} 
+	\label{accumulation} 
+\end{figure}
+
+If we examine the introduction of new mutations in two dimensions, we see that A/H3N2 appears to mutate in antigenic phenotype more rapidly than B/Vic, which mutates more rapidly still than A/H1N1 or B/Yam (Figure~\ref{accumulation}B).
+Thus, we have direct evidence that the rate of antigenic mutation is greater in A/H3N2 than in A/H1N1 or influenza B.
+However, we observe an interesting additional effect.
+The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
+This yields the surprising conclusion that although antigenic mutation is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.
+
+We suggest that this effect could be due to the pleiotropic effects of antigenic mutations, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
+Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
+Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantagous, while other antigenic mutations may be, on the whole, deleterious.
+In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
+It seems possibe that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
+However, more work is necessary to investigate this hypothesis.
+
+%%% CONCLUSIONS %%%
+\section*{Conclusions}
 
@@
 
+%%% AUTHOR CONTRIBUTIONS %%%
+\subsection*{Author contributions}
+
 %%% FUNDING %%%
 \subsection*{Funding}
+
+TB is supported by the Royal Society.
 

### 2012-12-11 flux.tex
Clean up flu B references.

@@
 
-Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Hay01} \tbc{Include in BibTex: Ansaldi04, AusWHO06, Barr06, Daum06, Lin07, Muyanga01, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
 Here, the sequence matching and censoring procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
 
-Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Hay01} \tbc{Include in BibTex: Abed03, Ansaldi03, Ansaldi04, AusWHO06, Barr06, Daum06, Kanegae90, Lin07, Matsuzaki04, Muyanga01, Nakagawa02, Nakajima92, Nerome98, Puzelli04, Rota90, Shaw02, Xu04} and NIMR vaccine strain selection reports for 2002--2012 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRFeb09,NIMRSep09,NIMRFeb10,NIMRSep10,NIMRFeb11,NIMRSep11,NIMRFeb12}.
+Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
 For B/Yamagata, the matching and censoring procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.

### 2012-12-17 flux.tex
Revise manuscript.

@@
 
-Before 2009, there were four major clades of influenza circulating within the human population: influenza A H3N2 and H1N1, and influenza B Victoria and Yamagata. 
+Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
@@
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
+Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over previously established strains.
 HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.
@@
 The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
-The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968. 
+The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
 Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
@@
 
-In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS (see Methods).
+In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Methods).
 In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
@@
 We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
-By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare predicted titers to observed titers in the test dataset. 
-We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret sera, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
+By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
+We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
 We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
@@
 We refer to these estimates as serum effects and virus effects, respectively.
-We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
+We find that including these effects decreases test error to 0.75 for influenza A/H3N2.
 We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
@@
 
-Most previous work on virus antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adapting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
-Here, the diffusion process acts as a prior, so that genetically similar viruses are expected to share similar antigenic phenotypes, and we model the relationship between genetic change and accrual of antigenic distance.
+Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
+Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
+The diffusion process acts as a prior, so that genetically similar viruses are expected to have antigenic locations that lie close to one another, and thus we model the relationship between genetic change and accrual of antigenic distance.
 We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
@@
 
-Through our modeling, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
+Through our analysis, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
 Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
-We observe prominent antigenic clusters in A/H3N2 and A/H1N1 with less obvious clustering in B/Vic and B/Yam.
+We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less obvious clustering in B/Vic and B/Yam.
 Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a somewhat curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
+Incorporating this prior made no difference in predictive ability of the model; test error for A/H3N2 was 0.75 log$_2$ HI titers whether the prior was included or excluded.
 Because of this, we interpret antigenic locations locally rather than globally.
@@
 We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
-Influenza A/H1N1 evolved at a rate of 0.55 antigenic units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
-Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) antigenic units per year.
+Influenza A/H1N1 evolved at a rate of 0.55 units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
+Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) units per year.
 Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
-Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
+Interestingly, a minor lineage of B/Vic has persisted, which rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
+The eventual fate of this lineage has yet to be determined.
+Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) units per year.
 Very little punctuated evolution is apparent in the evolution of B/Yam.
@@
 Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
-Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 in viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
-For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999/2000 winter influenza season is dominated by A/H3N2.
-Similarly, between samples from 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000/2001 season is dominated by these same two clades.
+Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
+For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999-2000 winter influenza season is dominated by A/H3N2.
+Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
 Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
@@
 For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
-After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drift along dimension 1.
+After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.
 
@@
 The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
-This yields the surprising conclusion that although antigenic mutation is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.
+This yields the surprising conclusion that although antigenic mutation in general is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.
 
-We suggest that this effect could be due to the pleiotropic effects of antigenic mutations, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
+We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
 Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
@@
 In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
-It seems possibe that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
+It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
 However, more work is necessary to investigate this hypothesis.
+
+In this context, it is especially interesting that adaptive fixations in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
+Thus, we conclude that additional adaptive fixations at the amino acid level translate to greater rates of adaptive antigenic change.
 
@@
 
-Antigenic characteristics of viral strains are often assessed through immunological assays such as hemagglutination inhibition (HI) \cite{Hirst43}.  
-At heart, these assays compare the reactivity $H_{ij}$ of one virus strain $i$ to antibodies raised against another virus strain $j$ via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.
+Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
+At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
+In the case of HI, the measurement of cross-reactivity takes the form of a titer $H_{ij}$ representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.
 These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
 However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
-Further, measurements are usually interval and truncated, e.g.\ inhibition or neutralization may cease somewhere between the serial titers of 160 and 320, or inhibition or neutralization may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
+Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
 
@@
 Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
-Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
+Virus and serum may be isolated from / raised against the same strain and have different cartographic locations.
 Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
-This gives set of distances between virus and serum cartographic locations 
+This gives a set of distances between virus and serum cartographic locations 
 \begin{equation}
@@
 \end{equation}
-Here, $H_{ij}$ represents the log$_2$ titer of virus $i$ against serum $j$, and immunological distance can be defined as
+With $H_{ij}$ as the log$_2$ titer of virus $i$ against serum $j$, immunological distance can be defined as
 \begin{equation}
@@
 Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
-HI titers are usually reported as the highest titer that successfully inhibits virus binding or growth, so that in this case, we calculate the likelihood of an interval measurement as
+HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
 \begin{equation} 
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 1000 along diagonal entries and equal to zero along off-diagonal entries.
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 10,000 along diagonal entries and equal to zero along off-diagonal entries.
 
@@
 
-The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular antiserum.  
+The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.  
 Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
@@
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace the diffuse normal prior on virus locations with a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
+To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a multivariate normal distribution in which the expectation of the first antigenic dimension increases with date of virus sampling
 \begin{equation} 
@@
 Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
-The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
+The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
 We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
@@
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST,BEAST17}.
+The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
 The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
@@
 \end{equation}
+Source code implementing these models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
 
@@
 
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A (H3N2) by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
 We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
@@
 However, the majority of isolates were present from 2002 to 2007. 
-Because we are interested in longer-term antigenic evolution, we censored the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
+Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
 We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
@@
 
-Antigenic data for influenza A (H1N1) was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
-The same procedure was followed as was followed for H3N2 to match sequence data and to censor antigenic comparisons.
+Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
+The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
 This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
 
-Antigenic comparisons for influenza B Victoria-lineage were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-Here, the sequence matching and censoring procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
+Antigenic comparisons for influenza B/Victoria were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
+Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
 
-Antigenic comparisons for influenza B Yamagata-lineage were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-For B/Yamagata, the matching and censoring procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
+Antigenic comparisons for influenza B/Yamagata were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
+For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
 
 Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
-As an example, one report states "collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
+As an example, one report states ``collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
 Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses.
-Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses."
+Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
 In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
 Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
+
+Antigenic, genetic and surveillance data is available as supplemental datasets \tbc{PREPARE}.
 
@@
 
+We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
+We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
+
 %%% AUTHOR CONTRIBUTIONS %%%
 \subsection*{Author contributions}
+
+AR, MAS, TB, and PL conceived the study.
+TB, GD, CR, DS and AR gathered antigenic and genetic data.
+AR, MAS, TB, and PL designed statistical phylogenetic procedures.
+TB performed the analysis.
+TB, MAS, AR wrote the paper.
 
@@
 
-\pagebreak
+%\pagebreak
 
-\setcounter{figure}{0}
-\setcounter{table}{0}
-\setcounter{page}{1}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thepage}{S\arabic{page}}
+%\setcounter{figure}{0}
+%\setcounter{table}{0}
+%\setcounter{page}{1}
+%\renewcommand{\thefigure}{S\arabic{figure}}
+%\renewcommand{\thetable}{S\arabic{table}}
+%\renewcommand{\thepage}{S\arabic{page}}
 
 %%% SUPPORTING INFORMATION %%%
-\section*{Supporting Information}
-
-\pagebreak
+%\section*{Supporting Information}
 

### 2012-12-17 flux.tex
Acknowledge GISAID laboratories.

@@
 Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
-Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations proceding in a random direction.
+Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations progressing in a random direction.
 However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
@@
 	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
-	In the diffusion model, the antigenic location of each virus sample can be calculcated back through time.
+	In the diffusion model, the antigenic location of each virus sample can be calculated back through time.
 	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
@@
 Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
-Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantagous, while other antigenic mutations may be, on the whole, deleterious.
-In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
+Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantageous, while other antigenic mutations may be, on the whole, deleterious.
+In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantageous in A/H3N2.
 It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
@@
 \section*{Conclusions}
+
+In this study we provide a foundation for evolutionary antigenic cartography in which to simultaneously assess antigenic phenotype and patterns of antigenic evolution.
+We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
+However, A/H3N2 shows substantially stronger signals of adaptive antigenic evolution, evolving faster in antigenic phenotype and showing more efficient conversion of transient antigenic polymorphism into fixed differences.
+Consequently, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
+We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
 
@@
 
-Antigenic, genetic and surveillance data is available as supplemental datasets \tbc{PREPARE}.
+Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
 
@@
 We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
+We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Hospital Clinic de Barcelona, Institut National d'Hygi\`{e}ne du Maroc, National Institute for Medical Research (UK), National Institute of Infectious Diseases (Japan), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, WHO Chinese National Influenza Center, WHO Collaborating Centre for Reference and Research on Influenza (Australia).
 

### 2012-12-18 flux.tex
Include paragraph and figure on year-to-year drift.

@@
 We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.00 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.97--1.04 (Figure~\ref{drift}).
-However, the rate of antigenic drift was not constant, with year-to-year movement of mean antigenic phenotype showing an interquartile range of 0.00--1.57.  
-Large jumps in antigenic phenotype (Figure~\ref{drift}) correspond to cluster transitions identified by Smith et al. \cite{Smith04}.  
+However, we observe occassional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
 Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.
@@
 
+Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift remains relatively constant through time (Figure~\ref{jumps}).
+For example, the largest cluster transition we observe is the emergence of the Sydney/97 cluster in 1997, evident as a large gap in virus locations along dimension 1 in Figure~\ref{drift}.
+However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
+In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
+In 1999, the transition is complete.
+Thus, we observe nearly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than an immediate jump in mean antigenic phenotype in 1997 (Figure~\ref{jumps}).
+A similar pattern is evident in the Fujian/02 transition (Figure~\ref{jumps}).
+Although A/H3N2 shows significantly greater mean rates of drift than A/H1N1 and influenza B, year-to-year variation in antigenic drift is fairly similar across influenza clades.
+We find that the standard deviation across years in year-to-year drift is 0.74 antigenic units for A/H3N2, 0.56 units for A/H1N1, 0.72 units for B/Vic and 0.38 units for B/Yam.
+
+%%% jumps %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.55\textwidth]{figures/jumps}
+	\caption{\textbf{Year-to-year antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
+	Bars represent interquartile range across MCMC samples.
+	} 
+	\label{jumps} 
+\end{figure}
+
 \subsection*{Epidemiological consequences of antigenic drift}
@@
 	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
-	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimensions 1 of viruses collected during the previous year $i-1$.
+	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
 	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
@@
 Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.29$, $r_\mathrm{Vic} = 0.75$ and $r_\mathrm{Yam} = 0.58$.
+Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.20$, $r_\mathrm{Vic} = 0.74$ and $r_\mathrm{Yam} = 0.58$.
 
@@
 
+We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
 We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
-We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
-We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Hospital Clinic de Barcelona, Institut National d'Hygi\`{e}ne du Maroc, National Institute for Medical Research (UK), National Institute of Infectious Diseases (Japan), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, WHO Chinese National Influenza Center, WHO Collaborating Centre for Reference and Research on Influenza (Australia).
+We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Chinese Center of Disease Prevention and Control, Hospital Clinic of Barcelona, National Institute of Hygiene of Morocco, National Institute of Infectious Diseases (Japan), National Institute for Medical Research (UK), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, Victorian Infectious Diseases Reference Laboratory (Australia).
 

### 2012-12-18 flux.tex
Revise incidence comparison with better numbers for year-to-year drift.
Instead of taking mean location averaged across posterior for year (i-1) and
comparing this to mean location averaged across posterior for year i.  I take
each posterior slice and compare mean location of year (i-1) to year i. I then
average this measurements of year-to-year antigenic drift.

@@
 	\centering		
-	\includegraphics[width=0.55\textwidth]{figures/jumps}
+	\includegraphics[width=0.5\textwidth]{figures/jumps}
 	\caption{\textbf{Year-to-year antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
@@
 We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
-From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (46\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 21\%, 17\% and 16\%, respectively.
+To rule out effects caused by the 2009 pandemic, incidence data were examined only until 2008.
+From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 22\%, 15\% and 13\%, respectively.
 Similarly, A/H3N2 was responsible for the majority of incidence (54\%) in the USA during this time period.
 Influenza A/H1N1 accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 11\% of incidence.
-These proportions are significantly similar to one another (randomization test, $p = 0.015$).
+These proportions are significantly similar to one another (randomization test, $p = 0.001$).
 
@@
 	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
+	Years with negative absolute drift were assigned a relative antigenic drift of 0.
 	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
@@
 drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
+	Years lacking antigenic information (Figure~\ref{jumps}) were not included in the correlation.
 	} 
@@
 Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
-Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.20$, $r_\mathrm{Vic} = 0.74$ and $r_\mathrm{Yam} = 0.58$.
+Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.74$ (Pearson correlation, $p = 2.43 \times 10^{-7}$).
+Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.61$, $r_\mathrm{H1} = 0.68$, $r_\mathrm{Vic} = 0.23$ and $r_\mathrm{Yam} = 0.56$.
 
-These findings clearly demonstrate the epidemiological importance of antigenic drift and show that competition between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
+These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence of influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
 Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
+However, these findings alone do not conclusively show interference between clades; more work needs to be done to establish whether antigenic drift within one influenza clades drives down incidence in other clades.
 
@@
 \end{equation}
-Source code implementing these models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
 
@@
 
+\subsection*{Implemention details}
+
+\tbc{PREPARE}.
+
+\subsection*{Availability}
+
+Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
 Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
+Also included as supplemental information are XML control files to load antigenic and phylogenetic data into BEAST and perform cartographic analyses \tbc{PREPARE}.
 

### 2012-12-19 flux.tex
Rerevise numbers for year-to-year antigenic drift.
I did not like the holes in the dataset that previously existed because
H1 and B had years with no virus isolations.  Here, I've sliced the tree
at each year and taken the mean location on dimension 1 of these slices.
This also allowed me to take the surveillance data back to 1998.

@@
 	\centering		
-	\includegraphics[width=0.85\textwidth]{figures/map}
+	\includegraphics[width=0.95\textwidth]{figures/map}
 	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
@@
 	\centering		
-	\includegraphics[width=0.65\textwidth]{figures/drift}
+	\includegraphics[width=0.75\textwidth]{figures/drift}
 	\caption{\textbf{Antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
@@
 
-Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift remains relatively constant through time (Figure~\ref{jumps}).
-For example, the largest cluster transition we observe is the emergence of the Sydney/97 cluster in 1997, evident as a large gap in virus locations along dimension 1 in Figure~\ref{drift}.
+Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
+We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.
+For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster in 1997, evident as a large gap in virus locations along dimension 1 in Figure~\ref{drift}.
 However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
 In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
-In 1999, the transition is complete.
+In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
 Thus, we observe nearly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than an immediate jump in mean antigenic phenotype in 1997 (Figure~\ref{jumps}).
-A similar pattern is evident in the Fujian/02 transition (Figure~\ref{jumps}).
-Although A/H3N2 shows significantly greater mean rates of drift than A/H1N1 and influenza B, year-to-year variation in antigenic drift is fairly similar across influenza clades.
-We find that the standard deviation across years in year-to-year drift is 0.74 antigenic units for A/H3N2, 0.56 units for A/H1N1, 0.72 units for B/Vic and 0.38 units for B/Yam.
+However, once the transition is completed in 1999, year-to-year antigenic drift drops, showing very low values in 2000 and 2001 before the Fujian/02 variant emerges.
+In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
+We find that the standard deviation in year-to-year drift is 0.80 antigenic units for A/H3N2, 0.38 units for A/H1N1, 0.44 units for B/Vic and 0.19 units for B/Yam.
 
@@
 	\centering		
-	\includegraphics[width=0.5\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
-	Bars represent interquartile range across MCMC samples.
+	\includegraphics[width=0.85\textwidth]{figures/jumps}
+	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	(A) Histograms of year-to-year antigenic drift across influenza clades.
+	(B) Timeseries of year-to-year antigenic drift across influenza clades from 1992 to 2011.
+	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
+	For example, 2000 signifies difference in antigenic dimension 1 between 1999 and 2000.
 	} 
@@
 
-We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
-To rule out effects caused by the 2009 pandemic, incidence data were examined only until 2008.
-From 1999 to 2008, A/H3N2 accounted for the majority of antigenic drift (51\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 22\%, 15\% and 13\%, respectively.
-Similarly, A/H3N2 was responsible for the majority of incidence (54\%) in the USA during this time period.
-Influenza A/H1N1 accounted for 23\% of incidence, B/Vic accounted for 12\% of incidence and B/Yam accounted for 11\% of incidence.
-These proportions are significantly similar to one another (randomization test, $p = 0.001$).
+We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1998 to 2010, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
+Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
+We compare these measurements of relative antigenic drift to relative isolations of viruses form each clade in the USA from 1998 to 2011 (see Methods).
+From 1998 to 2010, A/H3N2 accounted for the majority of antigenic drift (56\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 15\%, 21\% and 7\%, respectively.
+Similarly, A/H3N2 was responsible for the majority of incidence (56\%) during this time period.
+Influenza A/H1N1 accounted for 20\% of incidence, B/Vic accounted for 13\% of incidence and B/Yam accounted for 11\% of incidence.
+These proportions are significantly similar to one another (randomization test, $p = 0.011$).
 
@@
 	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
-	Drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of viruses collected in year $i$ compared to the mean of antigenic dimension 1 of viruses collected during the previous year $i-1$.
-	For example, `99' signifies difference in antigenic dimension 1 between 1998 and 1999.
+	For example, `00' signifies difference in antigenic dimension 1 between 1999 and 2000.
 	Years with negative absolute drift were assigned a relative antigenic drift of 0.
@@
 drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
-	Years lacking antigenic information (Figure~\ref{jumps}) were not included in the correlation.
 	} 
@@
 Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
-For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999-2000 winter influenza season is dominated by A/H3N2.
-Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
-Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.74$ (Pearson correlation, $p = 2.43 \times 10^{-7}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.61$, $r_\mathrm{H1} = 0.68$, $r_\mathrm{Vic} = 0.23$ and $r_\mathrm{Yam} = 0.56$.
+For example, we find that every year that A/H3N2 is responsible for the majority of antigenic drift (1998, 1998, 2001, 2002, 2003, 2004, 2005, 2009) shows A/H3N2 as dominating viral isolations.
+Additionally, the two strongest years of A/H1N1 drift (2000 and 2006) are dominated by A/H1N1 over A/H3N2.
+We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 2.14 \times 10^{-9}$).
+Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.57$, $r_\mathrm{H1} = 0.38$, $r_\mathrm{Vic} = 0.46$ and $r_\mathrm{Yam} = 0.40$.
+Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.
 

### 2012-12-19 flux.tex
Finish conclusions section.

@@
 
-In this context, it is especially interesting that adaptive fixations in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
+In this context, it is especially interesting that adaptive fixation in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
 Thus, we conclude that additional adaptive fixations at the amino acid level translate to greater rates of adaptive antigenic change.
@@
 
-In this study we provide a foundation for evolutionary antigenic cartography in which to simultaneously assess antigenic phenotype and patterns of antigenic evolution.
+In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
 We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
 However, A/H3N2 shows substantially stronger signals of adaptive antigenic evolution, evolving faster in antigenic phenotype and showing more efficient conversion of transient antigenic polymorphism into fixed differences.
-Consequently, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
+Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
 We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
+
+The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
+For example, our likelihood-based model faciliates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type is used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
+Additionally, this framework should be ideally suited to uncovering genetic determinates of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
+In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
+
+Further work characterizing antigenic evolution in the human influenza virus may eventually prove to be crucial in improving vaccine strain selection.
 

### 2012-12-19 flux.tex
Include section on implementation considerations.

@@
 We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.00 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.97--1.04 (Figure~\ref{drift}).
-However, we observe occassional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
+However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
 Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.
@@
 The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
-For example, our likelihood-based model faciliates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type is used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
-Additionally, this framework should be ideally suited to uncovering genetic determinates of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
+For example, our likelihood-based model facilities the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type is used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
+Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
 In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
@@
 	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
-	p(\mathbf{X} | \beta, \vartheta) \;
+	p(\mathbf{X} | \beta, \driftsd) \;
 	p(\mathbf{X} | \tree, \diffusionsd) \; 
-	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
+	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\driftsd,\diffusionsd)}{ p(\mathbf{H}) }.
 \end{equation}
@@
 
-\subsection*{Implemention details}
+\subsection*{Implementation considerations}
 
-\tbc{PREPARE}.
+There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
+The first issue lay in summarizing the posterior distribution of virus and serum locations.
+Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift dramatically.
+Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at $\{0,0\}$.
+However, local isometries are often still a problem.
+For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
+These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
+Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
+Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
+Then, for quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
+
+Running the fully parameterized model including the phylogenetic diffusion prior resulted a pathological outcome in which all virus and serum locations are compressed into a single point, diffusion volatility $\diffusionsd$ is tends toward zero and MDS variance $\mdssd^2$ tends towards infinity.
+In this case, $p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd)$ is reduced, but $p(\mathbf{X} | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
+To avoid this problematic outcome we conducted a two stage analysis.
+We first estimated values of $\mdssd$, $\beta$ and $\driftsd$ using HI data alone, and then held these values constant in a subsequent analysis that included the phylogenetic diffusion prior.
+
+Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
+MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
+These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis.
+
+MCMC was used to sample virus locations $\mathbf{X}$, serum locations $\mathbf{Y}$, virus effects $\mathbf{V}$, serum effects $\mathbf{S}$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
+MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
 

### 2012-12-20 flux.tex
Improve matrix notation.

@@
 \newcommand{\sn}{k}									% Number of sera
+\setlength{\arraycolsep}{2pt}
+\newcommand{\twomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
+\newcommand{\fourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}		% pretty inline matrix
 
@@
 Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
-\begin{equation} 
-	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
+\begin{equation} \label{hij}
+	H_{ij} \sim \mbox{TruncatedNormal}( S_j - \delta_{ij}, \mdssd^2 ),
 \end{equation}
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 10,000 along diagonal entries and equal to zero along off-diagonal entries.
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.
 
@@
 Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
-In this case, $H_{ij}$ still follows
-\begin{equation} 
-	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
-\end{equation}
-but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
+In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $S_j - \delta_{ij}$, but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
 We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
@@
 Additionally, we follow the same logic and assume that particular virus isolates are more reactive in general than other virus isolates.
-With virus reactivity included, $H_{ij}$ follows
-\begin{equation} 
-	H_{ij} \sim \mbox{Normal}\left(  \frac{V_i+S_j}{2} - \delta_{ij}, \mdssd^2 \right),
-\end{equation}
-and the vector of `virus effects' $V_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\mathbf{V}$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+With virus reactivity included, the expectation of $H_{ij}$ becomes $(V_i+S_j)/2 - \delta_{ij}$ and the vector of `virus effects' $V_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\mathbf{V}$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
 
@@
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a multivariate normal distribution in which the expectation of the first antigenic dimension increases with date of virus sampling
-\begin{equation} 
-	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
-\end{equation}
-for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
+To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution $\mathrm{Normal}(\mu,\Sigma)$ in which the expectation of the first antigenic dimension increases with date of virus sampling.
+Here, $\mu = \twomatrix{\beta \, y_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus.
 Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
 The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
-We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
+We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
 We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.
@@
 The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
-Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
+Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
 Metropolis-Hastings proposals scale $\diffusionsd$.

### 2012-12-20 flux.tex
Fix typo.
Clarify notation.

@@
 \usepackage{amssymb}
+\newcommand{\virus}{\mathbf{x}}						% virus coordinate
+\newcommand{\serum}{\mathbf{y}}						% serum coordinate
+\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
+\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
+\newcommand{\ve}{v}									% virus effect
+\newcommand{\se}{s}									% serum effect
+\newcommand{\ves}{\mathbf{v}}						% set of virus effects
+\newcommand{\ses}{\mathbf{s}}						% set of serum effects
 \newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
@@
 At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer $H_{ij}$ representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.
+In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding of another virus.
 These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
-However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
+Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
+Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
 Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
@@
 Here, we follow previous models in representing antigenic locations as points in an $N$-dimensional antigenic map. 
-Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions. 
+Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions $P$. 
 We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
+Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
 
-Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
+Let $\virus_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = \twomatrix{x_{i1}}{x_{i2}}$ for $P=2$. 
+Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = \twomatrix{y_{j1}}{y_{j2}}$ for $P=2$.
+This yields an $\vn \times P$ matrix $\viruses$ of virus coordinates and an $\sn \times P$ matrix $\sera$ of serum coordinates.
 Virus and serum may be isolated from / raised against the same strain and have different cartographic locations. 
-Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
 This gives a set of distances between virus and serum cartographic locations 
 \begin{equation}
-	\delta_{ij} =  || X_i - Y_j ||_2.
+	\delta_{ij} =  || \virus_i - \serum_j ||_2.
 \end{equation}
-With $H_{ij}$ as the log$_2$ titer of virus $i$ against serum $j$, immunological distance can be defined as
+
+Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
 \begin{equation}
-	d_{ij} =  S_j - H_{ij},
+	d_{ij} =  \se_j - H_{ij},
 \end{equation}
-where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
-Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
+where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
+In following multidimensional scaling (MDS), these approaches attempt to optimize over $\viruses$ and $\sera$ such that
 \begin{equation}
@@
 \end{equation}
-is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
+is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
+
 Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
 \begin{equation} \label{hij}
-	H_{ij} \sim \mbox{TruncatedNormal}( S_j - \delta_{ij}, \mdssd^2 ),
+	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ),
 \end{equation}
-and the likelihood of observing a particular titer given the placement of antigenic locations is 
+and the likelihood of observing an exact titer given the placement of antigenic locations is 
 \begin{equation} 
-	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
+	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
 \end{equation}
 where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
+The `serum effect' $\se_j$ represents the expected log$_2$ titer between virus $i$ and serum $j$ when $\delta_{ij}=0$ and is initially fixed such that $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
+
 HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
@@
 \begin{equation} 
-	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
+	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
 \end{equation}
@@
 \begin{equation} 
-	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
+	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
 \end{equation}
@@
 	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
-	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $S_j = \mathrm{log}_2 1280 = 10.32$.  
+	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
 	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
@@
 \begin{equation} 
-	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
+	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
 \end{equation}
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior on virus and serum locations
+\begin{eqnarray}
+	\virus_i &\sim& \mathrm{Normal}(\mu,\Sigma) \nonumber \\
+	\serum_j &\sim& \mathrm{Normal}(\mu,\Sigma),
+\end{eqnarray}
+with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.
 
 We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
-Metropolis-Hastings proposals include moves to individual virus and serum locations $X_i$ and $Y_j$, scaling of the entire set of virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ and scaling $\mdssd$.
+Metropolis-Hastings proposals include moves to individual virus and serum locations $\virus_i$ and $\serum_j$, scaling of the entire set of virus and serum locations $\viruses$ and $\sera$ and scaling $\mdssd$.
 
@@
 The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
-Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $S_j - \delta_{ij}$, but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
-We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
+However, this model may biased in some circumstances.
+In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
+To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
+In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses$ is estimated rather than fixed.
+We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.
 The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
@@
 
-Additionally, we follow the same logic and assume that particular virus isolates are more reactive in general than other virus isolates.
-With virus reactivity included, the expectation of $H_{ij}$ becomes $(V_i+S_j)/2 - \delta_{ij}$ and the vector of `virus effects' $V_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\mathbf{V}$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+Additionally, we follow the same logic and assume that some virus isolates are more reactive in general than other virus isolates.
+With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of `virus effects' $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
 
-With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $S_j$ and virus effects $V_i$.
+With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $\se_j$ and virus effects $\ve_i$.
 
@@
 
-As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
+As presented, virus and serum locations $\viruses$ and $\sera$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
 However, these distances represent only a local description of antigenic space and do not provide a global picture.
@@
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution $\mathrm{Normal}(\mu,\Sigma)$ in which the expectation of the first antigenic dimension increases with date of virus sampling.
-Here, $\mu = \twomatrix{\beta \, y_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus.
+To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution in which the expectation of dimension 1 increases with date of virus sampling
+\begin{equation}
+	\virus_i \sim \mathrm{Normal}(\mu, \Sigma),
+\end{equation}
+where $\mu = \twomatrix{\beta \, t_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
 Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
@@
 \begin{equation}
-	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
+	(\virus_1,\ldots,\virus_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
 \end{equation}
@@
 Metropolis-Hastings proposals scale $\diffusionsd$.
-The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+The probability of observing virus locations $p(\viruses | \tree, \diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
 Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
 \begin{equation}
-	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
-	p(\mathbf{X} | \beta, \driftsd) \;
-	p(\mathbf{X} | \tree, \diffusionsd) \; 
-	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\driftsd,\diffusionsd)}{ p(\mathbf{H}) }.
+	p(\viruses,\sera | \mathbf{H},\tree) = \frac{ p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
+	p(\viruses | \beta, \driftsd) \;
+	p(\viruses | \tree, \diffusionsd) \; 
+	p(\sera, \ses, \ves, \mdssd, \beta, \driftsd, \diffusionsd)}{ p(\mathbf{H}) }.
 \end{equation}
@@
 Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift dramatically.
-Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at $\{0,0\}$.
+Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
 However, local isometries are often still a problem.
@@
 Running the fully parameterized model including the phylogenetic diffusion prior resulted a pathological outcome in which all virus and serum locations are compressed into a single point, diffusion volatility $\diffusionsd$ is tends toward zero and MDS variance $\mdssd^2$ tends towards infinity.
-In this case, $p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd)$ is reduced, but $p(\mathbf{X} | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
+In this case, $p(\mathbf{H}|\viruses,\sera,\ses,\ves, \mdssd)$ is reduced, but $p(\viruses | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
 To avoid this problematic outcome we conducted a two stage analysis.
@@
 
-MCMC was used to sample virus locations $\mathbf{X}$, serum locations $\mathbf{Y}$, virus effects $\mathbf{V}$, serum effects $\mathbf{S}$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.

### 2012-12-21 flux.tex
Some edits to the Methods

@@
 \affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[4]{Department of Biostatistics, UCLA School of Public Health, University of California, Los Angeles CA, USA}
+\affil[4]{Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles CA, USA}
 \affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
@@
 These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data has not yet been attained.
+Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data %has not yet been attained.
+remains notably lacking.
 
-Here, we follow previous models in representing antigenic locations as points in an $N$-dimensional antigenic map. 
-Our goal is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into a lower number of dimensions $P$. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which a probabilistic model is constructed to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
+Here, we follow previous models in representing antigenic locations as points in a low $P$-dimensional antigenic map. 
+One of our initial goals is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into this lower dimensional space. 
+We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which we construct a probabilistic model to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
 Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
 
-Let $\virus_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = \twomatrix{x_{i1}}{x_{i2}}$ for $P=2$. 
-Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = \twomatrix{y_{j1}}{y_{j2}}$ for $P=2$.
-This yields an $\vn \times P$ matrix $\viruses$ of virus coordinates and an $\sn \times P$ matrix $\sera$ of serum coordinates.
+Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}
+%\twomatrix{x_{i1}}{x_{i2}}
+$ for $P=2$. 
+Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}
+%\twomatrix{y_{j1}}{y_{j2}}
+$ for $P=2$.
+For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
+%This yields an $\vn \times P$ matrix $\viruses$ of virus coordinates and an $\sn \times P$ matrix $\sera$ of serum coordinates.
 Virus and serum may be isolated from / raised against the same strain and have different cartographic locations. 
@@
 \begin{equation}
-	\delta_{ij} =  || \virus_i - \serum_j ||_2.
+	\delta_{ij} =  || \virus_i - \serum_j ||_2,
 \end{equation}
+where $|| \cdot ||_2$ is an $L_2$ norm.
 
@@
 \end{equation}
-where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
-In following multidimensional scaling (MDS), these approaches attempt to optimize over $\viruses$ and $\sera$ such that
+where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
+In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
 \begin{equation}
@@
 
-Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
+Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
 \begin{equation} \label{hij}
-	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ),
+	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ).
 \end{equation}
-and the likelihood of observing an exact titer given the placement of antigenic locations is 
+Consequentially, the likelihood of observing an exact titer given the placement of antigenic locations is 
 \begin{equation} 
@@
 \end{equation}
-where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
-The `serum effect' $\se_j$ represents the expected log$_2$ titer between virus $i$ and serum $j$ when $\delta_{ij}=0$ and is initially fixed such that $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
+where 
+%$\mdssd^2$ represents variance and 
+$\phi(\cdot)$ represents the standard normal probability density function (PDF).
+%The `serum effect' $\se_j$ represents the expected log$_2$ titer between virus $i$ and serum $j$ when $\delta_{ij}=0$.
+and is initially fixed such that $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
+%% MAS: Repeated material from above.
+Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
+In the antigenic setting, these remain random and can be negative (I still don't really believe this) since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
 
@@
 \end{equation}
-where $\Phi(\cdot)$ represents the standard normal CDF.
+where $\Phi(\cdot)$ represents the standard normal cummulative distribution fucntion (CDF).
 Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming a diffuse normal prior on virus and serum locations
+We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming independent, diffuse normal prior on virus and serum locations
 \begin{eqnarray}
-	\virus_i &\sim& \mathrm{Normal}(\mu,\Sigma) \nonumber \\
-	\serum_j &\sim& \mathrm{Normal}(\mu,\Sigma),
+	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
+	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
 \end{eqnarray}
-with $\mu = \twomatrix{0}{0}$ and $\Sigma = \fourmatrix{10,000}{0}{0}{10,000}$ for $P=2$.
+with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime} 
+%\twomatrix{0}{0}
+$ and $\boldsymbol{\Sigma}
+%\fourmatrix{10,000}{0}{0}{10,000}
+$ is a diagonal matrix with diagonal elements all equalling $10000$.
+%for $P=2$.
 
 We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
-Metropolis-Hastings proposals include moves to individual virus and serum locations $\virus_i$ and $\serum_j$, scaling of the entire set of virus and serum locations $\viruses$ and $\sera$ and scaling $\mdssd$.
+Metropolis-Hastings proposals include transition kernels for individual virus and serum locations $\virus_i$ and $\serum_j$, scaling of the entire set of virus and serum locations $\viruses$ and $\sera$ and scaling $\mdssd^2$.
 
@@
 The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
-However, this model may biased in some circumstances.
+However, this model may be biased in some circumstances.
 In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
 To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses$ is estimated rather than fixed.
+In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
 We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
+We take an Empirical Bayesian approach is specifying the 
+mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
 This formulation assumes that particular sera are more reactive in general than other sera.
@@
 
-With these effects included, Metropolis-Hastings proposals additionally incorporate moves to individual serum effects $\se_j$ and virus effects $\ve_i$.
+With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$.
+\textbf{What are $\ve_i$?
+These have not been defined nor used.}
 

### 2013-01-07 flux.tex
Accept Marc's revisions.

@@
 \usepackage{ulem}
+\definecolor{blue}{rgb}{0.324,0.609,0.708}
 \definecolor{purple}{rgb}{0.459,0.109,0.538}
@@
 \def\tbc#1{\textcolor{purple}{[#1]}}
+\def\ms#1#2{\sout{#1} \textcolor{blue}{#2}} 
+\def\msc#1{\textcolor{blue}{[#1]}}
 
@@
 \title{\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography}
+
+% Antigenic drift determines evolutionary success of seasonal influenza lineages
 
@@
 These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data %has not yet been attained.
-remains notably lacking.
+Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
 
@@
 
-Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}
-%\twomatrix{x_{i1}}{x_{i2}}
-$ for $P=2$. 
-Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}
-%\twomatrix{y_{j1}}{y_{j2}}
-$ for $P=2$.
+Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
+Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}$ for $P=2$.
 For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
-%This yields an $\vn \times P$ matrix $\viruses$ of virus coordinates and an $\sn \times P$ matrix $\sera$ of serum coordinates.
 Virus and serum may be isolated from / raised against the same strain and have different cartographic locations. 
@@
 \end{equation}
-Consequentially, the likelihood of observing an exact titer given the placement of antigenic locations is 
+Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
 \begin{equation} 
@@
 \end{equation}
-where 
-%$\mdssd^2$ represents variance and 
-$\phi(\cdot)$ represents the standard normal probability density function (PDF).
-%The `serum effect' $\se_j$ represents the expected log$_2$ titer between virus $i$ and serum $j$ when $\delta_{ij}=0$.
-and is initially fixed such that $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
-%% MAS: Repeated material from above.
+where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
 Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
-In the antigenic setting, these remain random and can be negative (I still don't really believe this) since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
+In the antigenic setting, these remain random and can be negative \msc{I still don't really believe this} since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
 
@@
 \end{equation}
-where $\Phi(\cdot)$ represents the standard normal cummulative distribution fucntion (CDF).
+where $\Phi(\cdot)$ represents the standard normal cumulative distribution fucntion (CDF).
 Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
@@
 \end{eqnarray}
-with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime} 
-%\twomatrix{0}{0}
-$ and $\boldsymbol{\Sigma}
-%\fourmatrix{10,000}{0}{0}{10,000}
-$ is a diagonal matrix with diagonal elements all equalling $10000$.
-%for $P=2$.
+with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime} $ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 
@@
 We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-We take an Empirical Bayesian approach is specifying the 
-mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
+We take an Empirical Bayesian approach is specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
 This formulation assumes that particular sera are more reactive in general than other sera.
@@
 With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$.
-\textbf{What are $\ve_i$?
+\msc{What are $\ve_i$?
 These have not been defined nor used.}

### 2013-01-07 flux.tex
Clarify virus effects.

@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography}
-
-% Antigenic drift determines evolutionary success of seasonal influenza lineages
+\title{\vspace{1.0cm} \LARGE \bf 
+Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\
+or \\
+Charting antigenic drift and evolutionary success of seasonal influenza lineages
+}
 
@@
 Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
-In the antigenic setting, these remain random and can be negative \msc{I still don't really believe this} since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
+In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
 
@@
 \end{equation}
-where $\Phi(\cdot)$ represents the standard normal cumulative distribution fucntion (CDF).
+where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
 Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
@@
 \end{eqnarray}
-with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime} $ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
+with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 
@@
 We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-We take an Empirical Bayesian approach is specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
+We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
 This formulation assumes that particular sera are more reactive in general than other sera.
 
-Additionally, we follow the same logic and assume that some virus isolates are more reactive in general than other virus isolates.
-With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of `virus effects' $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
+With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
 
 With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$.
-\msc{What are $\ve_i$?
-These have not been defined nor used.}
 
@@
 \begin{equation}
-	\virus_i \sim \mathrm{Normal}(\mu, \Sigma),
+	\virus_i \sim \mathrm{Normal}(\boldsymbol{\mu}, \boldsymbol{\Sigma}),
 \end{equation}
-where $\mu = \twomatrix{\beta \, t_i}{0}$ and $\Sigma = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
+where $\boldsymbol{\mu} = (\beta \, t_i, 0)^{\prime}$ and $\boldsymbol{\Sigma} = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
 Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.

### 2013-01-14 flux.tex
Revise methods to incorporate drifted diffusion prior.

@@
 \newcommand{\mdssd}{\varphi}						% MDS standard deviation
-\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
-\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
+\newcommand{\virussd}{\sigma}						% Virus / Diffusion standard deviation
+\newcommand{\serumsd}{\vartheta}					% Serum standard deviation
 \newcommand{\tree}{\tau}							% Phylogeny
@@
 \newcommand{\sn}{k}									% Number of sera
+\newcommand{\node}{\mathbf{z}}						% Internal node in phylogeny
+\newcommand{\normal}{\mathcal{N}}					% Normal distribution
 \setlength{\arraycolsep}{2pt}
-\newcommand{\twomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
-\newcommand{\fourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}		% pretty inline matrix
+\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
+\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
+\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
+\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix
 
@@
 
-\subsection*{Bayesian multidimensional scaling}
+\subsection*{Antigenic cartography}
 
@@
 At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding of another virus.
+In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding to red blood cells of another virus.
 These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
@@
 
+\subsection*{Likelihood functions}
+
 Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
@@
 For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
-Virus and serum may be isolated from / raised against the same strain and have different cartographic locations. 
+Virus and serum may be isolated from / raised against the same strain and have different cartographic locations, and separate serum isolates raised against the same strain may also have different cartographic locations.
 This gives a set of distances between virus and serum cartographic locations 
@@
 \begin{equation} \label{hij}
-	H_{ij} \sim \mbox{Normal}( \se_j - \delta_{ij}, \mdssd^2 ).
+	H_{ij} \sim \normal( \se_j - \delta_{ij}, \, \mdssd^2 ).
 \end{equation}
@@
 where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
-Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
-Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
+Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:160 dilution and no inhibition at 1:320 dilution.
+Rather than taking the HI titer as 160, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 160 and 320.
 HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$, and we start by assuming independent, diffuse normal prior on virus and serum locations
-\begin{eqnarray}
-	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
-	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
-\end{eqnarray}
-with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 
-We integrate over uncertainty using the Markov chain Monte Carlo (MCMC) procedures implemented in the phylogenetic package BEAST \cite{BEAST,BEAST17}.
-Metropolis-Hastings proposals include transition kernels for individual virus and serum locations $\virus_i$ and $\serum_j$, scaling of the entire set of virus and serum locations $\viruses$ and $\sera$ and scaling $\mdssd^2$.
-
-\subsection*{Incorporating virus and serum effects}
+\subsection*{Virus and serum effects}
 
@@
 Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
-With virus reactivity included, the expectation of $H_{ij}$ becomes $(\ve_i+\se_j)/2 - \delta_{ij}$ and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+With virus reactivity included, observed titers follow
+\begin{equation}
+	H_{ij} \sim \normal \left( \frac{\ve_i+\se_j}{2} - \delta_{ij}, \, \mdssd^2 \right),
+\end{equation}
+and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
 
-With these effects included, Metropolis-Hastings proposals additionally incorporate transition kernels for individual serum effects $\se_j$ and virus effects $\ve_i$.
+\subsection*{Location prior}
 
-\subsection*{Antigenic drift prior}
-
-As presented, virus and serum locations $\viruses$ and $\sera$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
-However, these distances represent only a local description of antigenic space and do not provide a global picture.
-An example of this phenomena is shown in Figure~\ref{schematic_map}.
+As presented, virus and serum locations $\viruses$ and $\sera$ give a likelihood of observing the HI data $\mathbf{H}$.
+However, distances between viruses and sera represent only a local description of antigenic space and do not provide a global picture.
+An example of this phenomenon is shown in Figure~\ref{schematic_map}.
 In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
@@
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace the diffuse normal prior on virus locations with a prior where virus locations follow a normal distribution in which the expectation of dimension 1 increases with date of virus sampling
-\begin{equation}
-	\virus_i \sim \mathrm{Normal}(\boldsymbol{\mu}, \boldsymbol{\Sigma}),
-\end{equation}
-where $\boldsymbol{\mu} = (\beta \, t_i, 0)^{\prime}$ and $\boldsymbol{\Sigma} = \fourmatrix{\driftsd^2}{0}{0}{\driftsd^2}$ for $P=2$, and $t_i$ is the difference in time between virus $i$ and the earliest sampled virus.
-Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
-The parameter $\driftsd$ determines the breadth of the virus population cloud at each point in time.
-We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
-We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.
+To capture this, we assume that the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
+\begin{eqnarray}
+	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
+	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
+\end{eqnarray}
+where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
+Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
+The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
 
@@
 
-We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
-To do this, we incorporate a prior on virus locations according to the evolutionary process
+We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
+To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
 \begin{equation}
-	(\virus_1,\ldots,\virus_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
+	\virus_i \sim \twomatrix{\beta \, t_i}{0} + \mbox{Evolutionary Brownian Process}(\virussd, \tree)
 \end{equation}
-where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
+for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion and $\tree$ is a phylogeny specifying tree topology and branch lengths.
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
-The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
-Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=0.001$.
-Metropolis-Hastings proposals scale $\diffusionsd$.
-The probability of observing virus locations $p(\viruses | \tree, \diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
+The evolutionary Brownian process assumes that the location $\node_i$ of node $i$ on the phylogeny follows from its parent node $p$, and with the addition of the drift along dimension 1, is distributed as
 \begin{equation}
-	p(\viruses,\sera | \mathbf{H},\tree) = \frac{ p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
-	p(\viruses | \beta, \driftsd) \;
-	p(\viruses | \tree, \diffusionsd) \; 
-	p(\sera, \ses, \ves, \mdssd, \beta, \driftsd, \diffusionsd)}{ p(\mathbf{H}) }.
+	\node_i \sim \twomatrix{\beta \, t_i}{0} + \normal(\node_p, d \, \boldsymbol{\Sigma})
 \end{equation}
+for $P=2$, where $\node_p$ is the location of the parent node, $d$ is the length of the branch connecting nodes $i$ and $p$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd$.
+The root node $\node_0$ is assumed to follow a normal distribution with expectation $(0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
+
+\subsection*{Posterior samples}
+
+Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
+Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
+\begin{equation}
+	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
+	p(\viruses | \beta, \virussd, \tree) \;
+	p(\sera | \beta, \serumsd) \; 
+	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
+\end{equation}
+We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST,BEAST17}.
+Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
+For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
 
@@
 The first issue lay in summarizing the posterior distribution of virus and serum locations.
-Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift dramatically.
-Our `antigenic drift prior' removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
+Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift.
+Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
 However, local isometries are often still a problem.
@@
 Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
-
-Running the fully parameterized model including the phylogenetic diffusion prior resulted a pathological outcome in which all virus and serum locations are compressed into a single point, diffusion volatility $\diffusionsd$ is tends toward zero and MDS variance $\mdssd^2$ tends towards infinity.
-In this case, $p(\mathbf{H}|\viruses,\sera,\ses,\ves, \mdssd)$ is reduced, but $p(\viruses | \tree, \diffusionsd)$ becomes very large, yielding a greater overall posterior probability.
-To avoid this problematic outcome we conducted a two stage analysis.
-We first estimated values of $\mdssd$, $\beta$ and $\driftsd$ using HI data alone, and then held these values constant in a subsequent analysis that included the phylogenetic diffusion prior.
+Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
+This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
 
@@
 MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
-These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis.
+These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
 
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, population variance $\driftsd^2$, diffusion volatility $\diffusionsd$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.

### 2013-01-14 flux.tex
Include Gregory, Hay and McCauley as coauthors.

@@
 \author[1]{Gytis Dudas}
-\author[6]{Colin Russell}
-\author[6,7]{Derek Smith}
-\author[1,8]{Andrew Rambaut}
+\author[6]{Victoria Gregory}
+\author[6]{Alan J. Hay}
+\author[6]{John W. McCauley}
+\author[7]{Colin Russell}
+\author[7,8]{Derek Smith}
+\author[1,9]{Andrew Rambaut}
 
@@
 \affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
-\affil[6]{Department of Zoology, University of Cambridge, Cambridge, UK.}
-\affil[7]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
-\affil[8]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
+\affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
+\affil[7]{Department of Zoology, University of Cambridge, Cambridge, UK.}
+\affil[8]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
+\affil[9]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
 
@@
 We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
-We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
 We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Chinese Center of Disease Prevention and Control, Hospital Clinic of Barcelona, National Institute of Hygiene of Morocco, National Institute of Infectious Diseases (Japan), National Institute for Medical Research (UK), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, Victorian Infectious Diseases Reference Laboratory (Australia).
@@
 
-AR, MAS, TB, and PL conceived the study.
-TB, GD, CR, DS and AR gathered antigenic and genetic data.
-AR, MAS, TB, and PL designed statistical phylogenetic procedures.
+AR, MAS, TB and PL conceived the study.
+TB, GD, VG, AJH, JWM, CR, DS and AR gathered antigenic and genetic data.
+AR, MAS, TB and PL designed statistical phylogenetic procedures.
 TB performed the analysis.
-TB, MAS, AR wrote the paper.
+TB, MAS and AR wrote the paper.
 
@@
 TB is supported by the Royal Society. 
+VG, AJH and JWM are funded by MRC reference no U117512723.
 

### 2013-01-14 flux.tex
Include additional title possibilities.

@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \LARGE \bf 
+\title{\vspace{1.0cm} \Large \bf 
 Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\
 or \\
-Charting antigenic drift and evolutionary success of seasonal influenza lineages
+Charting antigenic drift and evolutionary success of seasonal influenza lineages \\
+or \\
+Integrating influenza antigenic dynamics with molecular evolution \\
+or \\
+Cartographic analysis of antigenic dynamics and evolutionary success of seasonal influenza lineages \\
+or \\
+Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages 
 }

### 2013-01-15 flux.tex
Revise error table and further revise methods.

@@
 \newcommand{\mdssd}{\varphi}						% MDS standard deviation
-\newcommand{\virussd}{\sigma}						% Virus / Diffusion standard deviation
-\newcommand{\serumsd}{\vartheta}					% Serum standard deviation
+\newcommand{\virussd}{\sigma_x}						% Virus / Diffusion standard deviation
+\newcommand{\serumsd}{\sigma_y}						% Serum standard deviation
 \newcommand{\tree}{\tau}							% Phylogeny
@@
 
-We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
+We test model performance by constructing training datasets representing 90\% of the HI measurements in each of the four datasets and test datasets representing the remaining 10\% of the measurements in each dataset.
 By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
-We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
-We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
-We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
+We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.91 log$_2$ HI titers for A/H3N2 (Table \ref{errortable}).
+We find that model with two dimensional antigenic locations predict test data to the same precision or slightly better precision as 3D models, but perform substantially better than 1D models or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
+The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
+We extend this 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
 We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error to 0.75 for influenza A/H3N2.
-We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
-For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.
+We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (Table \ref{errortable}).
 
@@
 	\centering
-	\caption{\textbf{Absolute prediction error of log$_2$ HI titer against A/H3N2 test data across models.}}
+	\caption{\textbf{Absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
 	\label{errortable}		
-	\begin{tabular}{ c c c c c } 
+	\begin{tabular}{ c c c c c c c c } 
 	\hline
-	BMDS dimension 	& 	Serum effects 	&	Virus effects	& 	Test error	\\
+	\multicolumn{4}{c}{Model} 	&	\multicolumn{4}{c}{Test error}  \\
+	Dimen 	& 	Serum effects 	&	Virus effects	& Covariance	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline	
-	1D 				&	Fixed 			&	None			&	1.05		\\	
-	2D 				&	Fixed 			&	None			&	0.87 		\\
-	3D 				&	Fixed 			&	None			&	0.87		\\
-	4D 				&	Fixed 			&	None			&	0.91		\\
-	5D 				&	Fixed 			&	None			&	0.97		\\	
-	2D 				&	Estimated 		&	None			&	0.78		\\	
-	2D 				&	Estimated 		&	Estimated		&	0.75		\\		
+	1D 		&	Fixed 			&	None			& None			&	1.07	&	0.92	&	0.94	&	1.08	\\	
+	2D 		&	Fixed 			&	None			& None			&	0.91 	&	0.74 	&	0.77	&	0.83	\\
+	3D 		&	Fixed 			&	None			& None			&	0.91	&	0.74 	&	0.79	&	0.84	\\
+	4D 		&	Fixed 			&	None			& None			&	0.95	&	0.75 	&	0.81	&	0.86	\\
+	5D 		&	Fixed 			&	None			& None			&	0.99	&	0.76 	&	0.84	&	0.88	\\	
+	2D 		&	Estimated 		&	None			& None			&	0.79	&	0.74 	&	0.68	&	0.76	\\	
+	2D 		&	Estimated 		&	Estimated		& None			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
+	2D 		&	Estimated 		&	Estimated		& Diffusion		&	0.76	&	0.71 	&	0.64	&	0.72	\\		
 	\hline
@@
 Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
-The diffusion process acts as a prior, so that genetically similar viruses are expected to have antigenic locations that lie close to one another, and thus we model the relationship between genetic change and accrual of antigenic distance.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
-We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.
+Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to a Brownian motion process (see Methods).
+The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
+We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (Table \ref{errortable}).
+Additionally, modeling antigenic diffusion gives the advantage of directly correlating evolutionary outcomes with changes in antigenic phenotype.
 
@@
 \begin{equation}
-	\virus_i \sim \twomatrix{\beta \, t_i}{0} + \mbox{Evolutionary Brownian Process}(\virussd, \tree)
+	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
 \end{equation}
@@
 \begin{equation}
-	\node_i \sim \twomatrix{\beta \, t_i}{0} + \normal(\node_p, d \, \boldsymbol{\Sigma})
+	\node_i \sim (\beta \, t_i, \, 0)^{\prime} + \normal(\node_p, d \, \boldsymbol{\Sigma})
 \end{equation}
 for $P=2$, where $\node_p$ is the location of the parent node, $d$ is the length of the branch connecting nodes $i$ and $p$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd$.
-The root node $\node_0$ is assumed to follow a normal distribution with expectation $(0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+The root node $\node_0$ is assumed to follow a normal distribution with expectation $(\beta \, t_0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
@@
 
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.

### 2013-01-15 flux.tex
Move data details to supplement.

@@
 
-\subsection*{Posterior samples}
+\subsection*{Posterior samples and summaries}
 
@@
 
-\subsection*{Genetic, antigenic and surveillance data}
-
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
-We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
-If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
-Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
-This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
-However, the majority of isolates were present from 2002 to 2007. 
-Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
-We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
-This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.
-
-Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
-The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
-This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
-
-Antigenic comparisons for influenza B/Victoria were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
-
-Antigenic comparisons for influenza B/Yamagata were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
-
-Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
-As an example, one report states ``collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
-Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses.
-Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
-In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
-Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
-
-\subsection*{Implementation considerations}
-
-There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
-The first issue lay in summarizing the posterior distribution of virus and serum locations.
-Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift.
+There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
 Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
@@
 Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
-Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
+For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
 Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
 This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
+
+\subsection*{Data and implementation}
+
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications and WHO vaccine strain selection reports (see Supp.\ Text).
+Hemagglutinin sequences were compiled from the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID}.
+Antigenic and genetic data were subsampled to give similar quantities of data across years of the analysis.
+Data gathering procedures resulted in 402 virus strains A/H3N2, 115 virus strains for A/H1N1, 179 virus strains for B/Vic and 174 virus strains for B/Yam.
+Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
 

### 2013-01-15 flux_si.tex
Move data details to supplement.

--- 2013-01-14 flux.tex
@@
 
-% notation
-\usepackage{amsmath}
-\usepackage{amssymb}
-\newcommand{\virus}{\mathbf{x}}						% virus coordinate
-\newcommand{\serum}{\mathbf{y}}						% serum coordinate
-\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
-\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
-\newcommand{\ve}{v}									% virus effect
-\newcommand{\se}{s}									% serum effect
-\newcommand{\ves}{\mathbf{v}}						% set of virus effects
-\newcommand{\ses}{\mathbf{s}}						% set of serum effects
-\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
-\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
-\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
-\newcommand{\mdssd}{\varphi}						% MDS standard deviation
-\newcommand{\virussd}{\sigma}						% Virus / Diffusion standard deviation
-\newcommand{\serumsd}{\vartheta}					% Serum standard deviation
-\newcommand{\tree}{\tau}							% Phylogeny
-\newcommand{\vn}{n}									% Number of viruses
-\newcommand{\sn}{k}									% Number of sera
-\newcommand{\node}{\mathbf{z}}						% Internal node in phylogeny
-\newcommand{\normal}{\mathcal{N}}					% Normal distribution
-\setlength{\arraycolsep}{2pt}
-\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
-\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
-\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
-\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix
-
 % comments
@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \Large \bf 
-Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\
-or \\
-Charting antigenic drift and evolutionary success of seasonal influenza lineages \\
-or \\
-Integrating influenza antigenic dynamics with molecular evolution \\
-or \\
-Cartographic analysis of antigenic dynamics and evolutionary success of seasonal influenza lineages \\
-or \\
-Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages 
+\title{\vspace{1.0cm} \LARGE \bf 
+Supporting Text: \\
+Revealing the competitive dynamics of influenza viruses through evolutionary cartography
 }
@@
 
-%%% ABSTRACT %%%
-\section*{Abstract}
-
-Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
-However, the extent to which antigenic evolution determines competitive dynamics within and between virus clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata has remained unclear.
-Here, we develop a novel cartographic approach to simultaneously characterize the genetic and antigenic evolution of viral lineages.
-Through this approach, we determine historical rates and patterns of antigenic drift in all four viral clades and we show that antigenic drift influences year-to-year variability in clade dominance.
-We investigate the selective underpinnings for differing antigenic dynamics across clades and suggest that pleiotropic effects linked to antigenic mutation determine selective outcomes.
-
-%%% INTRODUCTION %%%
-\section*{Introduction}
-
-Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
-While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection.
-The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
-A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
-A thorough understanding of the process of antigenic drift is essential to our efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
-
-Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
-In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
-Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
-Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over previously established strains.
-HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.
-
-The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
-The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
-Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
-The process of antigenic drift results in the rapid turnover of the virus population. 
-Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
-
-Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
-An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
-However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
-Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Vic and B/Yam.
-
-%%% RESULTS %%%
-\section*{Results and Discussion}
-
-\subsection*{Antigenic and evolutionary cartography}
-
-In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Methods).
-In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
-In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
-Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
-We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).
-
-We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
-By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
-We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
-We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
-We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
-We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error to 0.75 for influenza A/H3N2.
-We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
-For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.
-
-%%% errortable %%%
-\begin{table}[h]
-	\centering
-	\caption{\textbf{Absolute prediction error of log$_2$ HI titer against A/H3N2 test data across models.}}
-	\label{errortable}		
-	\begin{tabular}{ c c c c c } 
-	\hline
-	BMDS dimension 	& 	Serum effects 	&	Virus effects	& 	Test error	\\
-	\hline	
-	1D 				&	Fixed 			&	None			&	1.05		\\	
-	2D 				&	Fixed 			&	None			&	0.87 		\\
-	3D 				&	Fixed 			&	None			&	0.87		\\
-	4D 				&	Fixed 			&	None			&	0.91		\\
-	5D 				&	Fixed 			&	None			&	0.97		\\	
-	2D 				&	Estimated 		&	None			&	0.78		\\	
-	2D 				&	Estimated 		&	Estimated		&	0.75		\\		
-	\hline
-	\end{tabular}
-\end{table}
-
-Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
-The diffusion process acts as a prior, so that genetically similar viruses are expected to have antigenic locations that lie close to one another, and thus we model the relationship between genetic change and accrual of antigenic distance.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
-We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.
-
-\subsection*{Antigenic evolution across influenza lineages}
-
-Through our analysis, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
-Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
-We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less obvious clustering in B/Vic and B/Yam.
-Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
-This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.
-
-%%% map %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/map}
-	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
-	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
-	Antigenic units represent two-fold dilutions of the HI assay.
-	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
-	\label{map} 
-\end{figure}
-
-HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a somewhat curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
-Incorporating this prior made no difference in predictive ability of the model; test error for A/H3N2 was 0.75 log$_2$ HI titers whether the prior was included or excluded.
-Because of this, we interpret antigenic locations locally rather than globally.
-We can determine the rate of antigenic movement of virus lineages without knowing the larger configuration that the movement occurs under.
-
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.00 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.97--1.04 (Figure~\ref{drift}).
-However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
-Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.
-
-%%% drift %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.75\textwidth]{figures/drift}
-	\caption{\textbf{Antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
-	Antigenic drift is shown in terms of change of location in the first antigenic dimension through time.
-	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
-	Antigenic units represent two-fold dilutions of the HI assay.
-	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
-	\label{drift} 
-\end{figure}
-
-We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
-Influenza A/H1N1 evolved at a rate of 0.55 units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
-Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) units per year.
-Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted, which rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
-The eventual fate of this lineage has yet to be determined.
-Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) units per year.
-Very little punctuated evolution is apparent in the evolution of B/Yam.
-
-Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
-We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.
-For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster in 1997, evident as a large gap in virus locations along dimension 1 in Figure~\ref{drift}.
-However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
-In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
-In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
-Thus, we observe nearly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than an immediate jump in mean antigenic phenotype in 1997 (Figure~\ref{jumps}).
-However, once the transition is completed in 1999, year-to-year antigenic drift drops, showing very low values in 2000 and 2001 before the Fujian/02 variant emerges.
-In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
-We find that the standard deviation in year-to-year drift is 0.80 antigenic units for A/H3N2, 0.38 units for A/H1N1, 0.44 units for B/Vic and 0.19 units for B/Yam.
-
-%%% jumps %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.85\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Histograms of year-to-year antigenic drift across influenza clades.
-	(B) Timeseries of year-to-year antigenic drift across influenza clades from 1992 to 2011.
-	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
-	For example, 2000 signifies difference in antigenic dimension 1 between 1999 and 2000.
-	} 
-	\label{jumps} 
-\end{figure}
-
-\subsection*{Epidemiological consequences of antigenic drift}
-
-We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1998 to 2010, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
-Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
-We compare these measurements of relative antigenic drift to relative isolations of viruses form each clade in the USA from 1998 to 2011 (see Methods).
-From 1998 to 2010, A/H3N2 accounted for the majority of antigenic drift (56\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 15\%, 21\% and 7\%, respectively.
-Similarly, A/H3N2 was responsible for the majority of incidence (56\%) during this time period.
-Influenza A/H1N1 accounted for 20\% of incidence, B/Vic accounted for 13\% of incidence and B/Yam accounted for 11\% of incidence.
-These proportions are significantly similar to one another (randomization test, $p = 0.011$).
-
-%%% incidence %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/incidence}
-	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008.} 
-	(A) Relative antigenic drift of influenza clades.
-	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
-	For example, `00' signifies difference in antigenic dimension 1 between 1999 and 2000.
-	Years with negative absolute drift were assigned a relative antigenic drift of 0.
-	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
-	Here, `99' refers to the 1999/2000 influenza season.
-	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
-	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g.
-drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
-	} 
-	\label{incidence} 
-\end{figure}
-
-Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
-Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
-For example, we find that every year that A/H3N2 is responsible for the majority of antigenic drift (1998, 1998, 2001, 2002, 2003, 2004, 2005, 2009) shows A/H3N2 as dominating viral isolations.
-Additionally, the two strongest years of A/H1N1 drift (2000 and 2006) are dominated by A/H1N1 over A/H3N2.
-We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 2.14 \times 10^{-9}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.57$, $r_\mathrm{H1} = 0.38$, $r_\mathrm{Vic} = 0.46$ and $r_\mathrm{Yam} = 0.40$.
-Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.
-
-These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence of influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
-Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
-However, these findings alone do not conclusively show interference between clades; more work needs to be done to establish whether antigenic drift within one influenza clades drives down incidence in other clades.
-
-\subsection*{Conversion of antigenic mutation into fixed differences}
-
-We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or influenza B.
-Previous work using general epidemiological models has suggested that a greater rate of antigenic drift may emerge from either a greater fundamental reproductive number $R_0$ or a faster rate of mutation decreasing cross-immunity \cite{Gog02,Lin03}.
-Owing to these and similar findings, subsequent work has ascribed differences in the evolution in A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
-
-Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
-Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations progressing in a random direction.
-However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
-This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
-The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
-For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
-After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.
-
-%%% accumulation %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/accumulation}
-	\caption{\textbf{Rate of antigenic diffusion of influenza lineages categorized by time back from sampling.}
-	(A) Rate of diffusion in terms of antigenic units per year only along antigenic dimension 1.
-	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
-	In the diffusion model, the antigenic location of each virus sample can be calculated back through time.
-	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
-	When the $x$-axis is close to 0, the $y$-axis shows the rate of diffusion near tips of the phylogeny, and when the $x$-axis is large, the $y$-axis shows the rate of diffusion along the trunk of the phylogeny.
-	} 
-	\label{accumulation} 
-\end{figure}
-
-If we examine the introduction of new mutations in two dimensions, we see that A/H3N2 appears to mutate in antigenic phenotype more rapidly than B/Vic, which mutates more rapidly still than A/H1N1 or B/Yam (Figure~\ref{accumulation}B).
-Thus, we have direct evidence that the rate of antigenic mutation is greater in A/H3N2 than in A/H1N1 or influenza B.
-However, we observe an interesting additional effect.
-The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
-This yields the surprising conclusion that although antigenic mutation in general is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.
-
-We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
-Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
-Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantageous, while other antigenic mutations may be, on the whole, deleterious.
-In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantageous in A/H3N2.
-It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
-However, more work is necessary to investigate this hypothesis.
-
-In this context, it is especially interesting that adaptive fixation in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
-Thus, we conclude that additional adaptive fixations at the amino acid level translate to greater rates of adaptive antigenic change.
-
-%%% CONCLUSIONS %%%
-\section*{Conclusions}
-
-In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
-We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
-However, A/H3N2 shows substantially stronger signals of adaptive antigenic evolution, evolving faster in antigenic phenotype and showing more efficient conversion of transient antigenic polymorphism into fixed differences.
-Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
-We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
-
-The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
-For example, our likelihood-based model facilities the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type is used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
-Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
-In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
-
-Further work characterizing antigenic evolution in the human influenza virus may eventually prove to be crucial in improving vaccine strain selection.
-
-%%% METHODS %%%
-\section*{Methods}
-
-\subsection*{Antigenic cartography}
-
-Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
-At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding to red blood cells of another virus.
-These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
-Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
-Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
-Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
-
-Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
-These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
-
-Here, we follow previous models in representing antigenic locations as points in a low $P$-dimensional antigenic map. 
-One of our initial goals is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into this lower dimensional space. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which we construct a probabilistic model to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
-Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
-
-\subsection*{Likelihood functions}
-
-Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
-Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}$ for $P=2$.
-For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
-Virus and serum may be isolated from / raised against the same strain and have different cartographic locations, and separate serum isolates raised against the same strain may also have different cartographic locations.
-This gives a set of distances between virus and serum cartographic locations 
-\begin{equation}
-	\delta_{ij} =  || \virus_i - \serum_j ||_2,
-\end{equation}
-where $|| \cdot ||_2$ is an $L_2$ norm.
-
-Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
-\begin{equation}
-	d_{ij} =  \se_j - H_{ij},
-\end{equation}
-where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
-In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
-\begin{equation}
-	\sum_{(i,j) \in \cal I} 
-	\left(
-		\delta_{ij} - d_{ij}
-	\right)^2
-\end{equation}
-is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
-
-Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
-\begin{equation} \label{hij}
-	H_{ij} \sim \normal( \se_j - \delta_{ij}, \, \mdssd^2 ).
-\end{equation}
-Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
-\begin{equation} 
-	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
-Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
-In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
-
-HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
-In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
-\begin{equation} 
-	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
-Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:160 dilution and no inhibition at 1:320 dilution.
-Rather than taking the HI titer as 160, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 160 and 320.
-HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
-\begin{equation} 
-	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
-\end{equation}
-These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
-Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
-
-%%% hij_likelihood %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
-	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
-	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
-	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
-	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
-	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
-	} 
-	\label{hij_likelihood} 
-\end{figure}
-
-We calculate the overall likelihood by multiplying probabilities of individual measurements
-\begin{equation} 
-	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
-\end{equation}
-using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-
-\subsection*{Virus and serum effects}
-
-The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
-However, this model may be biased in some circumstances.
-In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
-To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
-We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
-This formulation assumes that particular sera are more reactive in general than other sera.
-
-Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
-With virus reactivity included, observed titers follow
-\begin{equation}
-	H_{ij} \sim \normal \left( \frac{\ve_i+\se_j}{2} - \delta_{ij}, \, \mdssd^2 \right),
-\end{equation}
-and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
-
-\subsection*{Location prior}
-
-As presented, virus and serum locations $\viruses$ and $\sera$ give a likelihood of observing the HI data $\mathbf{H}$.
-However, distances between viruses and sera represent only a local description of antigenic space and do not provide a global picture.
-An example of this phenomenon is shown in Figure~\ref{schematic_map}.
-In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
-This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
-In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
-In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we assume that the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
-\begin{eqnarray}
-	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
-	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
-\end{eqnarray}
-where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
-Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
-The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
-
-%%% schematic_map %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
-	\caption{\textbf{Schematic antigenic map with three viruses and two sera.} 
-	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
-	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
-	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 1 and 3.
-	Left and right panels represent cartographic models that would give equal likelihoods of a given set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
-	} 
-	\label{schematic_map} 
-\end{figure}
-
-\subsection*{Diffusion model of antigenic evolution}
-
-We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
-To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
-\begin{equation}
-	\virus_i \sim \twomatrix{\beta \, t_i}{0} + \mbox{Evolutionary Brownian Process}(\virussd, \tree)
-\end{equation}
-for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion and $\tree$ is a phylogeny specifying tree topology and branch lengths.
-Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The evolutionary Brownian process assumes that the location $\node_i$ of node $i$ on the phylogeny follows from its parent node $p$, and with the addition of the drift along dimension 1, is distributed as
-\begin{equation}
-	\node_i \sim \twomatrix{\beta \, t_i}{0} + \normal(\node_p, d \, \boldsymbol{\Sigma})
-\end{equation}
-for $P=2$, where $\node_p$ is the location of the parent node, $d$ is the length of the branch connecting nodes $i$ and $p$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd$.
-The root node $\node_0$ is assumed to follow a normal distribution with expectation $(0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
-The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
-
-\subsection*{Posterior samples}
-
-Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
-Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
-\begin{equation}
-	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
-	p(\viruses | \beta, \virussd, \tree) \;
-	p(\sera | \beta, \serumsd) \; 
-	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
-\end{equation}
-We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST,BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
-For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
+%%% SUPPORTING METHODS %%%
+\section*{Supporting Methods}
 
@@
 We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
-This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.
+This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements.
 
@@
 The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
-This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
+This procedure yielded 115 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
 
@@
 
-\subsection*{Implementation considerations}
-
-There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
-The first issue lay in summarizing the posterior distribution of virus and serum locations.
-Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift.
-Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
-However, local isometries are often still a problem.
-For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
-These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
-Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
-Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
-This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
-
-Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
-MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
-These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
-
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd$ and phylogenetic tree $\tree$.
-MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
-
-\subsection*{Availability}
-
-Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
-Also included as supplemental information are XML control files to load antigenic and phylogenetic data into BEAST and perform cartographic analyses \tbc{PREPARE}.
-
-%%% ACKNOWLEDGMENTS %%%
-\subsection*{Acknowledgments}
-
-We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
-We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Chinese Center of Disease Prevention and Control, Hospital Clinic of Barcelona, National Institute of Hygiene of Morocco, National Institute of Infectious Diseases (Japan), National Institute for Medical Research (UK), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, Victorian Infectious Diseases Reference Laboratory (Australia).
-
-%%% AUTHOR CONTRIBUTIONS %%%
-\subsection*{Author contributions}
-
-AR, MAS, TB and PL conceived the study.
-TB, GD, VG, AJH, JWM, CR, DS and AR gathered antigenic and genetic data.
-AR, MAS, TB and PL designed statistical phylogenetic procedures.
-TB performed the analysis.
-TB, MAS and AR wrote the paper.
-
-%%% FUNDING %%%
-\subsection*{Funding}
-
-TB is supported by the Royal Society. 
-VG, AJH and JWM are funded by MRC reference no U117512723.
-
 %%% REFERENCES %%%
@@
 
-%\pagebreak
-
-%\setcounter{figure}{0}
-%\setcounter{table}{0}
-%\setcounter{page}{1}
-%\renewcommand{\thefigure}{S\arabic{figure}}
-%\renewcommand{\thetable}{S\arabic{table}}
-%\renewcommand{\thepage}{S\arabic{page}}
-
-%%% SUPPORTING INFORMATION %%%
-%\section*{Supporting Information}
-
 \end{document}

### 2013-01-15 flux.tex
Clarify section on model comparison.

@@
 
-We test model performance by constructing training datasets representing 90\% of the HI measurements in each of the four datasets and test datasets representing the remaining 10\% of the measurements in each dataset.
-By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
-We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.91 log$_2$ HI titers for A/H3N2 (Table \ref{errortable}).
-We find that model with two dimensional antigenic locations predict test data to the same precision or slightly better precision as 3D models, but perform substantially better than 1D models or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
-The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
-We extend this 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
-We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (Table \ref{errortable}).
+We test model performance by constructing training datasets representing 90\% of the HI measurements for each of the four influenza lineages and test datasets representing the remaining 10\% of the measurements for each lineage.
+By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers. 
+We begin with a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a flat prior.
+We find that this model performs well, yielding an average absolute predictive error of between 0.79 and 0.95 log$_2$ HI titers across influenza lineages (model 1; Table \ref{errortable}).
+We extend this model to include a prior on virus and serum locations that assumes antigenic location along dimension 1 increases in proportion to date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
 
@@
 	\centering
-	\caption{\textbf{Absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
+	\caption{\textbf{Average absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
 	\label{errortable}	
+	\makebox[\textwidth][c]{
 	\begin{tabular}{ c c c c c c c c } 
 	\hline
-	\multicolumn{4}{c}{Model} 	&	\multicolumn{4}{c}{Test error}  \\
-	Dimen 	& 	Serum effects 	&	Virus effects	& Covariance	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	\multicolumn{4}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
+	Model	&	Dimen 	& 	Serum/virus effects 	& Location Prior					& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline		
-	1D 		&	Fixed 			&	None			& None			&	1.07	&	0.92	&	0.94	&	1.08	\\	
-	2D 		&	Fixed 			&	None			& None			&	0.91 	&	0.74 	&	0.77	&	0.83	\\
-	3D 		&	Fixed 			&	None			& None			&	0.91	&	0.74 	&	0.79	&	0.84	\\
-	4D 		&	Fixed 			&	None			& None			&	0.95	&	0.75 	&	0.81	&	0.86	\\
-	5D 		&	Fixed 			&	None			& None			&	0.99	&	0.76 	&	0.84	&	0.88	\\	
-	2D 		&	Estimated 		&	None			& None			&	0.79	&	0.74 	&	0.68	&	0.76	\\	
-	2D 		&	Estimated 		&	Estimated		& None			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
-	2D 		&	Estimated 		&	Estimated		& Diffusion		&	0.76	&	0.71 	&	0.64	&	0.72	\\		
+	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.95 	&	0.79 	&	0.80	&	0.93	\\
+	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
+	3		&	1D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	1.07	&	0.92	&	0.94	&	1.08	\\		
+	4		&	3D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91	&	0.74 	&	0.79	&	0.84	\\
+	5		&	4D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.95	&	0.75 	&	0.81	&	0.86	\\
+	6		&	5D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.99	&	0.76 	&	0.84	&	0.88	\\				
+	7		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd$			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
+	8		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd,\tree$	&	0.76	&	0.71 	&	0.64	&	0.72	\\		
 	\hline
 	\end{tabular}	
+	}
 \end{table}
+
+We find that decreasing dimensionality to a single antigenic dimension results in a substantially less predictive model (model 3; Table \ref{errortable}), and that increasing dimensionality to three dimensions is equal or slightly worse in predictive power (model 4; Table \ref{errortable}).
+Increasing dimensionality further to four or five dimensions results in continued declines in test error (models 5 and 6; Table \ref{errortable}).
+Consequently, we specify a two dimensional model in all subsequent analyses.
+The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
+We extend the 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
+We refer to these estimates as serum effects and virus effects, respectively.
+We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (model 7; Table \ref{errortable}).
 
@@
 The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (Table \ref{errortable}).
-Additionally, modeling antigenic diffusion gives the advantage of directly correlating evolutionary outcomes with changes in antigenic phenotype.
+We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (model 8; Table \ref{errortable}).
+Additionally, modeling antigenic diffusion has the advantage of allowing direct correlation of evolutionary outcomes with changes in antigenic phenotype.
 
@@
 
-\subsection*{Posterior samples and summaries}
+\subsection*{Posterior samples}
 
@@
 For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
-
-There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
-Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
-However, local isometries are often still a problem.
-For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
-These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
-Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
-For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
-This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
 
@@
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
+
+There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
+Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
+However, local isometries are often still a problem.
+For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
+These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
+Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
+For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
+Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
+This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
 

### 2013-01-16 flux.tex
Prior is not flat

@@
 	\hline		
-	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.95 	&	0.79 	&	0.80	&	0.93	\\
+	1		&	2D 		&	Fixed $\ses$ 			& Uninformed								&	0.95 	&	0.79 	&	0.80	&	0.93	\\
 	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	

### 2013-01-17 flux.tex
Minor revisions.

--- 2013-01-15 flux.tex
@@
 	\hline		
-	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.95 	&	0.79 	&	0.80	&	0.93	\\
+	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.93 	&	0.79 	&	0.80	&	0.93	\\
 	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
@@
 	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
-	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
+	Circles represent a posterior sample of virus locations and have been shaded based on year of isolation.
 	Antigenic units represent two-fold dilutions of the HI assay.
@@
 Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a somewhat curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
-Incorporating this prior made no difference in predictive ability of the model; test error for A/H3N2 was 0.75 log$_2$ HI titers whether the prior was included or excluded.
+To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 2 through 8; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
 Because of this, we interpret antigenic locations locally rather than globally.
@@
 	Antigenic drift is shown in terms of change of location in the first antigenic dimension through time.
-	Circles represent a posterior sample of virus locations and have been colored based on year of isolation in 4-year intervals.
+	Circles represent a posterior sample of virus locations and have been shaded based on year of isolation.
 	Antigenic units represent two-fold dilutions of the HI assay.
@@
 Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted, which rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
-The eventual fate of this lineage has yet to be determined.
+Interestingly, a minor lineage of B/Vic has persisted, appearing antigenically distinct from 2011 viruses (Figure~\ref{drift}), though the eventual fate of this lineage has yet to be determined.
 Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) units per year.
-Very little punctuated evolution is apparent in the evolution of B/Yam.
+Little punctuated evolution is apparent in the evolution of B/Yam.
 
@@
 Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
-We compare these measurements of relative antigenic drift to relative isolations of viruses form each clade in the USA from 1998 to 2011 (see Methods).
-From 1998 to 2010, A/H3N2 accounted for the majority of antigenic drift (56\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 15\%, 21\% and 7\%, respectively.
+We compare these measurements of relative antigenic drift to relative isolations of viruses from each clade in the USA from 1998 to 2011 (see Methods).
+From 1998 to 2011, A/H3N2 accounted for the majority of antigenic drift (56\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 15\%, 21\% and 7\%, respectively.
 Similarly, A/H3N2 was responsible for the majority of incidence (56\%) during this time period.

### 2013-01-17 flux.tex

@@
 	\hline		
-	1		&	2D 		&	Fixed $\ses$ 			& Flat								&	0.93 	&	0.79 	&	0.80	&	0.93	\\
+	1		&	2D 		&	Fixed $\ses$ 			& Uninformed						&	0.93 	&	0.78 	&	0.80	&	0.93	\\
 	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	

### 2013-01-18 flux.tex
Include info on error by distance and location accuracy.

@@
 By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers. 
-We begin with a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a flat prior.
+We begin with a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
 We find that this model performs well, yielding an average absolute predictive error of between 0.79 and 0.95 log$_2$ HI titers across influenza lineages (model 1; Table \ref{errortable}).
-We extend this model to include a prior on virus and serum locations that assumes antigenic location along dimension 1 increases in proportion to date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
+We extend this model to include a time-informed prior on virus and serum locations that assumes antigenic location along dimension 1 increases in proportion to date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
 
@@
 We find that decreasing dimensionality to a single antigenic dimension results in a substantially less predictive model (model 3; Table \ref{errortable}), and that increasing dimensionality to three dimensions is equal or slightly worse in predictive power (model 4; Table \ref{errortable}).
-Increasing dimensionality further to four or five dimensions results in continued declines in test error (models 5 and 6; Table \ref{errortable}).
+Increasing dimensionality further to four or five dimensions results in continued declines in predictive power (models 5 and 6; Table \ref{errortable}).
 Consequently, we specify a two dimensional model in all subsequent analyses.
@@
 The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the diffusion (model 8; Table \ref{errortable}).
+We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the phylogenetic diffusion (model 8; Table \ref{errortable}).
 Additionally, modeling antigenic diffusion has the advantage of allowing direct correlation of evolutionary outcomes with changes in antigenic phenotype.
+
+We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Figure~\ref{error_by_distance_and_year}A), thus supporting our model assumption that drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
+Additionally, we find that absolute error in predicted titer is nearly constant with time (Figure~\ref{error_by_distance_and_year}B).
+Antigenic locations inferred by the model are well resolved; estimates of antigenic distance between pairs of viruses show relatively little variation across the posterior.
+We estimate that virus distances have, on average, a 50\% credible interval of $\pm0.45$ antigenic units for A/H3N2, $\pm0.57$ units for A/H1N1, $\pm0.76$ units for B/Vic, and $\pm0.65$ units for B/Yam.
 
@@
 
-%\pagebreak
+\pagebreak
 
-%\setcounter{figure}{0}
-%\setcounter{table}{0}
+\setcounter{figure}{0}
+\setcounter{table}{0}
 %\setcounter{page}{1}
-%\renewcommand{\thefigure}{S\arabic{figure}}
-%\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 %\renewcommand{\thepage}{S\arabic{page}}
@@
 %%% SUPPORTING INFORMATION %%%
-%\section*{Supporting Information}
+\section*{Supporting Information}
+
+%%% error_by_distance_and_year %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/error_by_distance_and_year}
+	\caption{\textbf{Absolute error in log$_2$ HI titer against antigenic distance and date of virus sampling in influenza A/H3N2.} 
+	(A) Points show the absolute deviation from the predicted log$_2$ HI titer of the antigenic model, computed from virus location $\virus_i$, serum location $\serum_j$, virus effect $\ve_i$ and serum effect $\se_j$, based upon antigenic distance between virus and serum $\delta_{ij}$.
+	The red line shows a LOESS regression curve of these points.
+	(B) Points show the absolute deviation from the predicted log$_2$ HI titers of the antigenic model, based upon the date of virus sampling.
+	The red line shows average deviations for each year.
+	Relationships for A/H1N1, B/Vic and B/Yam are similar.
+	} 
+	\label{error_by_distance_and_year} 
+\end{figure}
 

### 2013-01-21 flux.tex
Replace figures and numbers with results from new method.

@@
 Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
-We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less obvious clustering in B/Vic and B/Yam.
-Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
+We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less prominent, though still apparent, clustering in B/Vic and B/Yam.
+Antigenic clusters show high genetic similarity, so that we observe very few mutation events leading to each cluster, rather than the repeated emergence of clusters.
 This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.
@@
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/map}
+	\includegraphics[width=1.0\textwidth]{figures/map}
 	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a somewhat curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 2 through 8; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
@@
 
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.00 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.97--1.04 (Figure~\ref{drift}).
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of $\beta=1.01$ antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04 (Figure~\ref{drift}).
 However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
@@
 We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
-Influenza A/H1N1 evolved at a rate of 0.55 units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
-Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) units per year.
+Influenza A/H1N1 evolved at a rate of 0.62 units per year (HPD 0.56--0.67), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
+Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.42 (HPD 0.32--0.51) units per year.
 Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted, appearing antigenically distinct from 2011 viruses (Figure~\ref{drift}), though the eventual fate of this lineage has yet to be determined.
-Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) units per year.
-Little punctuated evolution is apparent in the evolution of B/Yam.
+Interestingly, a minor lineage of B/Vic has persisted through 2011, appearing antigenically distinct from other 2011 viruses (Figure~\ref{drift}), though the eventual fate of this lineage has yet to be determined.
+Influenza B/Yamagata evolved slower still, with an average rate of and 0.32 (HPD 0.25--0.39) units per year.
+Little punctuated evolution is obvious in the evolution of B/Yam.
 
@@
 We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.
-For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster in 1997, evident as a large gap in virus locations along dimension 1 in Figure~\ref{drift}.
+For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster, evident as the appearance in 1997 of a large gap in virus locations along antigenic dimension 1 (Figure~\ref{drift}).
 However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
@@
 In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
-Thus, we observe nearly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than an immediate jump in mean antigenic phenotype in 1997 (Figure~\ref{jumps}).
-However, once the transition is completed in 1999, year-to-year antigenic drift drops, showing very low values in 2000 and 2001 before the Fujian/02 variant emerges.
+Thus, we observe fairly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
 In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
-We find that the standard deviation in year-to-year drift is 0.80 antigenic units for A/H3N2, 0.38 units for A/H1N1, 0.44 units for B/Vic and 0.19 units for B/Yam.
+We find that the standard deviation in year-to-year drift is 0.88 antigenic units for A/H3N2, 0.51 units for A/H1N1, 0.54 units for B/Vic and 0.27 units for B/Yam.
 
@@
 	\centering		
-	\includegraphics[width=0.85\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Histograms of year-to-year antigenic drift across influenza clades.
-	(B) Timeseries of year-to-year antigenic drift across influenza clades from 1992 to 2011.
+	\includegraphics[width=0.95\textwidth]{figures/jumps}
+	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	(A) Timeseries of year-to-year antigenic drift across influenza clades from 1992 to 2011.
 	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
-	For example, 2000 signifies difference in antigenic dimension 1 between 1999 and 2000.
+	For example, 2000 represents difference in antigenic dimension 1 between 1999 and 2000.
+	Error bars represent 50\% credible intervals across posterior samples of year-to-year drift.
+	(B) Histograms of year-to-year antigenic drift across influenza clades summarizing observations from 1992 to 2011.
 	} 
@@
 
-We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 1998 to 2010, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
+We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
 Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
-We compare these measurements of relative antigenic drift to relative isolations of viruses from each clade in the USA from 1998 to 2011 (see Methods).
-From 1998 to 2011, A/H3N2 accounted for the majority of antigenic drift (56\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 15\%, 21\% and 7\%, respectively.
-Similarly, A/H3N2 was responsible for the majority of incidence (56\%) during this time period.
-Influenza A/H1N1 accounted for 20\% of incidence, B/Vic accounted for 13\% of incidence and B/Yam accounted for 11\% of incidence.
-These proportions are significantly similar to one another (randomization test, $p = 0.011$).
+We compare these measurements of relative antigenic drift to relative isolations of viruses from each clade in the USA from the 2001--2002 to the 2011--2012 influenza seasons (see Methods).
+From 2000 to 2011, A/H3N2 accounted for the majority of antigenic drift (46\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 18\%, 21\% and 15\%, respectively.
+Similarly, A/H3N2 was responsible for the majority of incidence (58\%) during this time period.
+Influenza A/H1N1 accounted for 18\% of incidence, B/Vic accounted for 16\% of incidence and B/Yam accounted for 7\% of incidence.
+These proportions are significantly similar to one another (randomization test, $p = 0.030$).
 
@@
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/incidence}
-	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 1999 to 2008.} 
+	\includegraphics[width=0.9\textwidth]{figures/incidence}
+	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011.} 
 	(A) Relative antigenic drift of influenza clades.
 	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
-	For example, `00' signifies difference in antigenic dimension 1 between 1999 and 2000.
+	For example, `01' signifies difference in antigenic dimension 1 between 2000 and 2001.
 	Years with negative absolute drift were assigned a relative antigenic drift of 0.
 	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
-	Here, `99' refers to the 1999/2000 influenza season.
+	Here, `01' refers to the 2001/2002 influenza season.
 	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
@@
 
-Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
-Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
-For example, we find that every year that A/H3N2 is responsible for the majority of antigenic drift (1998, 1998, 2001, 2002, 2003, 2004, 2005, 2009) shows A/H3N2 as dominating viral isolations.
-Additionally, the two strongest years of A/H1N1 drift (2000 and 2006) are dominated by A/H1N1 over A/H3N2.
-We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 2.14 \times 10^{-9}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.57$, $r_\mathrm{H1} = 0.38$, $r_\mathrm{Vic} = 0.46$ and $r_\mathrm{Yam} = 0.40$.
+Additionally, year-to-year comparisons show a similar pattern, in which years with high levels of antigenic drift coming into an influenza season, e.g.\ drift from 2000 to 2001, show high levels of relative incidence in that season, e.g.\ incidence in the 2001/2002 season (Figure~\ref{incidence}).
+We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 6.43 \times 10^{-8}$).
+Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.56$, $r_\mathrm{H1} = 0.46$, $r_\mathrm{Vic} = 0.49$ and $r_\mathrm{Yam} = 0.23$.
 Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.

### 2013-01-21 flux.tex
Simplify and rewrite section on polymorphism and divergence.

@@
 	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g.
-drift from 1998 to 1999 compared to incidence in the 1999/2000 influenza season.
+drift from 2000 to 2001 compared to incidence in the 2001/2002 influenza season.
 	} 
@@
 
-We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or influenza B.
-Previous work using general epidemiological models has suggested that a greater rate of antigenic drift may emerge from either a greater fundamental reproductive number $R_0$ or a faster rate of mutation decreasing cross-immunity \cite{Gog02,Lin03}.
-Owing to these and similar findings, subsequent work has ascribed differences in the evolution in A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
+Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both fundamental reproductive number $R_0$ and rate cross-immunity decreasing mutation \cite{Gog02,Lin03}.
+Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
+To facilitate comparisons, we measure the diffusion coefficient $D$ characterizing the antigenic movement of each influenza lineage \cite{Pybus12}.
+This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
+Higher diffusion coefficients mean a lineage is diffusing faster through antigenic space.
+We compare diffusion coefficients between trunk branches and side branches, where trunk branches are defined as branches ancestral to the most recent virus sample and side branches are defined as those branches not ancestral to the most recent virus sample.
 
-Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
-Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations progressing in a random direction.
-However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
-This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
-The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
-For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
-After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.
+%%% diffusiontable %%%
+\begin{table}[h]
+	\centering
+	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including 50\% highest posterior density intervals.}}
+	\label{diffusiontable}	
+	\begin{tabular}{ c c c c } 
+	\hline
+	Lineage	&	Trunk branch $D$ 	& 	Side branch $D$		& 	Ratio \\
+	\hline		
+	A/H3N2	&	0.90 (0.79--0.97)	&	0.62 (0.57--0.65)	&	1.45 (1.31--1.57) \\
+	A/H1N1	&	0.62 (0.50--0.71)	&	0.44 (0.38--0.48)	&	1.45 (1.18--1.70) \\
+	B/Vic	&	0.63 (0.51--0.71)	&	0.62 (0.54--0.68)	&	1.02 (0.87--1.14) \\
+	B/Yam	&	0.35 (0.29--0.40)	&	0.33 (0.28--0.36)	&	1.09 (0.93--1.23) \\
+	\hline
+	\end{tabular}	
+\end{table}
 
-%%% accumulation %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/accumulation}
-	\caption{\textbf{Rate of antigenic diffusion of influenza lineages categorized by time back from sampling.}
-	(A) Rate of diffusion in terms of antigenic units per year only along antigenic dimension 1.
-	(B) Rate of diffusion in terms of antigenic units per year along both antigenic dimensions 1 and 2.
-	In the diffusion model, the antigenic location of each virus sample can be calculated back through time.
-	Here, when the $x$-axis is 1, we show the rate of diffusion 1 year back from from the date of sampling averaged across all virus samples.
-	When the $x$-axis is close to 0, the $y$-axis shows the rate of diffusion near tips of the phylogeny, and when the $x$-axis is large, the $y$-axis shows the rate of diffusion along the trunk of the phylogeny.
-	} 
-	\label{accumulation} 
-\end{figure}
+Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
+We expect the diffusion coefficient on side branches to more closely resemble the raw rate of antigenic mutation.
+Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
+Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
+We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
+Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
+Thus, although drift forward along dimension 1 is promoted in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be promoted more strongly in influenza A than in influenza B, giving an advantage to influenza A in terms of overall rate of antigenic drift.
+Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them an advantage over their sister lineages A/H1N1 and B/Yam also in terms of overall rate of antigenic drift.
 
-If we examine the introduction of new mutations in two dimensions, we see that A/H3N2 appears to mutate in antigenic phenotype more rapidly than B/Vic, which mutates more rapidly still than A/H1N1 or B/Yam (Figure~\ref{accumulation}B).
-Thus, we have direct evidence that the rate of antigenic mutation is greater in A/H3N2 than in A/H1N1 or influenza B.
-However, we observe an interesting additional effect.
-The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
-This yields the surprising conclusion that although antigenic mutation in general is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.
+In this context, it is especially interesting that adaptive evolution in the HA1 protein determined through analysis of nonsynonymous and synonymous substitutions is substantially stronger in A/H3N2 than in A/H1N1 \cite{Wolf06}; H3 appears to accumulate adaptive fixations at $\sim$150\% the rate of H1 \cite{Bhatt11}.
+Here, we observe that A/H3N2 has a diffusion coefficient $\sim$141\% greater than A/H1N1 along side branches of the phylogeny.
+These results suggest that the greater rate of adaptive fixation in A/H3N2 may be simply due to a greater abundance of new antigenic mutations to fix.
 
-We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
-Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
-Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantageous, while other antigenic mutations may be, on the whole, deleterious.
-In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantageous in A/H3N2.
-It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
-However, more work is necessary to investigate this hypothesis.
-
-In this context, it is especially interesting that adaptive fixation in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
-Thus, we conclude that additional adaptive fixations at the amino acid level translate to greater rates of adaptive antigenic change.
+We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
+In the later case, we expect that many antigenic variants may be at inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
+These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
+Still, it is clear that more work is necessarily to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
 
@@
 
+\tbc{Describe calculation of diffusion coefficient $D$.}
+
 \subsection*{Availability}

### 2013-01-21 flux.tex
Revise Conclusions.

@@
 We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
-However, A/H3N2 shows substantially stronger signals of adaptive antigenic evolution, evolving faster in antigenic phenotype and showing more efficient conversion of transient antigenic polymorphism into fixed differences.
+We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
+Influenza A/H3N2 benefits from a higher rate of new antigenic mutation coupled with strong adaptive evolution and evolves in antigenic phenotype faster than A/H1N1, B/Vic and B/Yam.
 Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
@@
 The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
-For example, our likelihood-based model facilities the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type is used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
+For example, our likelihood-based model facilities the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
 Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.

### 2013-01-21 flux.tex
Minor revisions to methods.

@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
+We begin by assuming independent, diffuse normal prior on virus and serum locations
+\begin{eqnarray}
+	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
+	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
+\end{eqnarray}
+with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 
@@
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we assume that the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
+To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
 \begin{eqnarray}
@@
 To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
-\begin{equation}
+\begin{equation} \label{ebpprior}
 	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
@@
 \begin{equation} 
-	\node_i \sim (\beta \, t_i, \, 0)^{\prime} + \normal(\node_p, d \, \boldsymbol{\Sigma})
+	\node_i \sim (\beta \, d, \, 0)^{\prime} + \normal(\node_p, d \, \boldsymbol{\Sigma})
 \end{equation}
@@
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+This formulation corresponds to a Weiner process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \tbc{CITE}.
+This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
+
 Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.

### 2013-01-22 flux.tex
Rewrite abstract.

@@
 \affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
-\affil[7]{Department of Zoology, University of Cambridge, Cambridge, UK.}
-\affil[8]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands.}
-\affil[9]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA.}
+\affil[7]{Department of Zoology, University of Cambridge, Cambridge, UK}
+\affil[8]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
+\affil[9]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
 
@@
 Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
-However, the extent to which antigenic evolution determines competitive dynamics within and between virus clades A/H3N2, A/H1N1, B/Victoria and B/Yamagata has remained unclear.
-Here, we develop a novel cartographic approach to simultaneously characterize the genetic and antigenic evolution of viral lineages.
-Through this approach, we determine historical rates and patterns of antigenic drift in all four viral clades and we show that antigenic drift influences year-to-year variability in clade dominance.
-We investigate the selective underpinnings for differing antigenic dynamics across clades and suggest that pleiotropic effects linked to antigenic mutation determine selective outcomes.
+Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
+Here, we extend previous approaches to antigenic cartography, which seeks to place strains on an antigenic map, such that distances on this map best recapitulate titers observed across multiple HI assays.
+In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
+Using HI data for influenza lineages A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine average rates of antigenic drift for each clade, as well as year-to-year variability in the rate of drift.
+Through comparison with epidemiological data, we show that antigenic drift influences overall incidence across clades, and demonstrate a year-to-year correlation between drift and incidence.
+We investigate the selective underpinnings for differing antigenic dynamics across clades and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza clades.
+This work does much to elucidate the competitive dynamics of influenza lineages, but also makes possible substantial future advances in influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+
+\pagebreak
 
@@
 	\hline		
-	1		&	2D 		&	Fixed $\ses$ 			& Uninformed						&	0.93 	&	0.78 	&	0.80	&	0.93	\\
+	1		&	2D 		&	Fixed $\ses$ 			& Diffuse							&	0.93 	&	0.78 	&	0.80	&	0.93	\\
 	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	

### 2013-01-22 flux.tex
Include section on diffusion coefficient estimation in SI methods.

@@
 Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-To facilitate comparisons, we measure the diffusion coefficient $D$ characterizing the antigenic movement of each influenza lineage \cite{Pybus12}.
+To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Supp.\ Methods).
 This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
@@
 	\centering
-	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including 50\% highest posterior density intervals.}}
+	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including posterior means and 50\% highest posterior density intervals.}}
 	\label{diffusiontable}	
@@
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Weiner process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \tbc{CITE}.
+This formulation corresponds to a Weiner process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \cite{BrownianMotionHandbook}.
 This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
@@
 
-\tbc{Describe calculation of diffusion coefficient $D$.}
-
 \subsection*{Availability}
@@
 Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
-Also included as supplemental information are XML control files to load antigenic and phylogenetic data into BEAST and perform cartographic analyses \tbc{PREPARE}.
 

### 2013-01-22 flux_si.tex
Include section on diffusion coefficient estimation in SI methods.

@@
 \def\msc#1{\textcolor{blue}{[#1]}}
+
+% notation
+\usepackage{amsmath}
+\usepackage{amssymb}
+\newcommand{\normal}{\mathcal{N}}					% Normal distribution
+\setlength{\arraycolsep}{2pt}
+\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
+\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix
 
@@
 
+\subsection*{Estimation of diffusion coefficient}
+
+The distance $d$ traveled by two-dimensional diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
+Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
+\begin{equation}
+	X_t \sim \normal(0, \sigma^2 t),
+\end{equation}
+where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
+\begin{equation}
+	\mathrm{E}[X_t^2] = \sigma^2 t.
+\end{equation}
+In moving to a two-dimensional diffusion
+\begin{equation}
+	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma^2 t}{0}{0}{\sigma^2 t} \right),
+\end{equation}
+a factor of 2 is included when calculating expected squared displacement
+\begin{equation}
+	\mathrm{E}[X_t^2 + Y_t^2] = 2 \sigma^2 t.
+\end{equation}
+Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$, in this case $D=\sigma^2/2$ and 
+\begin{equation}
+	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
+\end{equation}
+
+Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
+\begin{equation} \label{pybusest}
+	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
+\end{equation}
+in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
+However, short branches along the phylogeny may cause $d_i^2/t_i$ to have high variance and lead to uncertainty in the resulting estimates of $D$.
+Consequently, we estimate $D$ by looking at the relationship between total squared displacement along branches of a phylogeny and total time along branches of a phylogeny
+\begin{eqnarray}
+	d_\mathrm{total} = \frac{1}{n} \sum^n_{i=1} d_i^2 \nonumber \\
+	t_\mathrm{total} = \frac{1}{n} \sum^n_{i=1} t_i^2 \nonumber \\
+	\hat{D} = \frac{ d_\mathrm{total} }{ 4 t_\mathrm{total} }.
+\end{eqnarray}
+This estimator showed less variance than the estimator in equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
+When estimating $D$ specific to trunk branches or side branches, we calculate $d_\mathrm{total}$ and $t_\mathrm{total}$ only from the subset of trunk branches or side branches as appropriate.
+
 %%% REFERENCES %%%

### 2013-01-22 flux.tex
BEAST 1.5 will not supported most of the analysis in the paper; do not cite.

@@
 
-Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST, BEAST17}.
+Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
 
@@
 \end{equation}
-We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST,BEAST17}.
+We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST17}.
 Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
@@
 
-Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
+Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
 MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
@@
 
-Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST, BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
+Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
 Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.

### 2013-01-22 flux_si.tex
Fix math notation problem (but this estimator still may not be correct).

@@
 \begin{eqnarray}
-	d_\mathrm{total} = \frac{1}{n} \sum^n_{i=1} d_i^2 \nonumber \\
-	t_\mathrm{total} = \frac{1}{n} \sum^n_{i=1} t_i^2 \nonumber \\
-	\hat{D} = \frac{ d_\mathrm{total} }{ 4 t_\mathrm{total} }.
+	d^2_\mathrm{total} = \sum^n_{i=1} d^2_i \nonumber \\
+	t_\mathrm{total} = \sum^n_{i=1} t_i^2 \nonumber \\
+	\hat{D} = \frac{ d^2_\mathrm{total} }{ 4 t_\mathrm{total} }.
 \end{eqnarray}
 This estimator showed less variance than the estimator in equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
-When estimating $D$ specific to trunk branches or side branches, we calculate $d_\mathrm{total}$ and $t_\mathrm{total}$ only from the subset of trunk branches or side branches as appropriate.
+When estimating $D$ specific to trunk branches or side branches, we calculate $d^2_\mathrm{total}$ and $t_\mathrm{total}$ only from the subset of trunk branches or side branches as appropriate.
 

### 2013-01-22 flux_si.tex
Fix typo.
Confirm that diffusion coefficient estimator has lower mean squared error than estimator used in Pybus et al 2012 with some estimation error on node locations and branch lengths.

@@
 \end{equation}
-Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$, in this case $D=\sigma^2/2$ and 
+Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$.
+In this case $D=\sigma^2/2$ and 
 \begin{equation}
@@
 in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
-However, short branches along the phylogeny may cause $d_i^2/t_i$ to have high variance and lead to uncertainty in the resulting estimates of $D$.
-Consequently, we estimate $D$ by looking at the relationship between total squared displacement along branches of a phylogeny and total time along branches of a phylogeny
-\begin{eqnarray}
-	d^2_\mathrm{total} = \sum^n_{i=1} d^2_i \nonumber \\
-	t_\mathrm{total} = \sum^n_{i=1} t_i^2 \nonumber \\
-	\hat{D} = \frac{ d^2_\mathrm{total} }{ 4 t_\mathrm{total} }.
+However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
+Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
+\begin{eqnarray} \label{ourest}
+	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
+	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
+	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
 \end{eqnarray}
-This estimator showed less variance than the estimator in equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
-When estimating $D$ specific to trunk branches or side branches, we calculate $d^2_\mathrm{total}$ and $t_\mathrm{total}$ only from the subset of trunk branches or side branches as appropriate.
+This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
+Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
+When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
 

### 2013-01-23 flux.tex
Added MAS grants
Include Gytis's changes.

@@
 A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
-A thorough understanding of the process of antigenic drift is essential to our efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
+A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
 
@@
 
-Through our analysis, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
+Through our analysis, we reveal the antigenic, as well as evolutionary, relationships among viruses in influenza A/H3N2, A/H1N1, B/Vic and B/Yam, quantifying both antigenic and evolutionary distances between strains (Figure~\ref{map}).
 Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
@@
 These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
-Still, it is clear that more work is necessarily to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
+Still, it is clear that more work is necessary to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
 
@@
 TB is supported by the Royal Society. 
+MAS is supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
 VG, AJH and JWM are funded by MRC reference no U117512723.

### 2013-01-24 flux.tex
Make Table 1 more approachable per Marc's suggestion.

@@
 We find that this model performs well, yielding an average absolute predictive error of between 0.79 and 0.95 log$_2$ HI titers across influenza lineages (model 1; Table \ref{errortable}).
-We extend this model to include a time-informed prior on virus and serum locations that assumes antigenic location along dimension 1 increases in proportion to date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
+We extend this model to include a time-informed prior that models the process of antigenic drift, in which virus and serum locations  along dimension 1 increase with date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
 
@@
 	\makebox[\textwidth][c]{
-	\begin{tabular}{ c c c c c c c c } 
+	\begin{tabular}{ c c c c c c c c c } 
 	\hline
-	\multicolumn{4}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Dimen 	& 	Serum/virus effects 	& Location Prior					& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	\multicolumn{5}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
+	Model	&	Dimen 	& 	Serum effects 	&	Virus effects	& Location Prior	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline		
-	1		&	2D 		&	Fixed $\ses$ 			& Diffuse							&	0.93 	&	0.78 	&	0.80	&	0.93	\\
-	2		&	2D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
-	3		&	1D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	1.07	&	0.92	&	0.94	&	1.08	\\		
-	4		&	3D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.91	&	0.74 	&	0.79	&	0.84	\\
-	5		&	4D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.95	&	0.75 	&	0.81	&	0.86	\\
-	6		&	5D 		&	Fixed $\ses$			& $\beta,\virussd,\serumsd$			&	0.99	&	0.76 	&	0.84	&	0.88	\\				
-	7		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd$			&	0.77	&	0.71 	&	0.65	&	0.73	\\	
-	8		&	2D 		&	Estimated $\ses, \ves$	& $\beta,\virussd,\serumsd,\tree$	&	0.76	&	0.71 	&	0.64	&	0.72	\\		
+	1		&	2D 		&	Fixed 			&	None			& Uninformed		&	0.93 	&	0.78 	&	0.80	&	0.93	\\
+	2		&	2D 		&	Fixed			&	None			& Drift				&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
+	3		&	1D 		&	Fixed			&	None			& Drift				&	1.07	&	0.92	&	0.94	&	1.08	\\		
+	4		&	3D 		&	Fixed			&	None			& Drift				&	0.91	&	0.74 	&	0.79	&	0.84	\\
+	5		&	4D 		&	Fixed			&	None			& Drift				&	0.95	&	0.75 	&	0.81	&	0.86	\\
+	6		&	5D 		&	Fixed			&	None			& Drift				&	0.99	&	0.76 	&	0.84	&	0.88	\\				
+	7		&	2D 		&	Estimated		&	Estimated		& Drift				&	0.77	&	0.71 	&	0.65	&	0.73	\\	
+	8		&	2D 		&	Estimated		&	Estimated		& Diffusion/Drift	&	0.76	&	0.71 	&	0.64	&	0.72	\\		
 	\hline

### 2013-01-24 flux_si.tex
Generalize diffusion coefficient derivation per Marc's suggestion.

@@
 
-The distance $d$ traveled by two-dimensional diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
+The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
 Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
@@
 \begin{equation}
-	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma^2 t}{0}{0}{\sigma^2 t} \right),
+	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
 \end{equation}
-a factor of 2 is included when calculating expected squared displacement
+where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
 \begin{equation}
-	\mathrm{E}[X_t^2 + Y_t^2] = 2 \sigma^2 t.
+	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
 \end{equation}
-Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$.
-In this case $D=\sigma^2/2$ and 
+Such diffusions are often characterized with diffusion coefficient $D$.
+In this case
+\begin{equation}
+	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
+\end{equation}
+so that expected squared displacement follows from $D$ alone
 \begin{equation}

### 2013-01-24 flux.tex
Added some of the standard grant numbers.
These are mainly copied from other papers so may need adjustment

@@
 MAS is supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
+The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
+278433-PREDEMICS and ERC Grant agreement no.
+260864.
 VG, AJH and JWM are funded by MRC reference no U117512723.
+AR acknowledges the support of the Wellcome Trust (grant no. 092807).
+Collaboration between MAS, AR and PL was supported by the National Evolutionary Synthesis Center (NESCent), NSF EF-0423641.
+DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant
+P0050/2008, Wellcome 087982AIA, and NIH Director's Pioneer Award DP1-OD000490-01.
+CAR was supported by a University Research Fellowship from the Royal Society.
 

### 2013-01-25 flux.tex
Fix author initials.
Minor text changes and a comment on the figures

@@
 Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages 
+or \\
+Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution
+
 }
@@
 \author[6]{John W. McCauley}
-\author[7]{Colin Russell}
-\author[7,8]{Derek Smith}
+\author[7]{Colin A. Russell}
+\author[7,8]{Derek J. Smith}
 \author[1,9]{Andrew Rambaut}
@@
 
-Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
+Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity to previous virus strains acquired by the host population.
 Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
@@
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
+In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single one prior to 1980 \cite{Rota92}.
 Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
@@
 The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
-Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
+Evolution of antigenic phenotype appears punctuated with episodes of more rapid innovation interspersed by periods of stasis, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
 The process of antigenic drift results in the rapid turnover of the virus population. 
@@
 Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to a Brownian motion process (see Methods).
-The diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
+The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
 We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the phylogenetic diffusion (model 8; Table \ref{errortable}).
@@
 %%% map %%%
+
+%% AR - Figure improvements: Give proper axis labels. Label clusters (Texas/77, Beijing/89, Beijing/92, Wuhan/95, Sydney/97 that are referenced in the text)
 \begin{figure}[h]
@@
 	Antigenic units represent two-fold dilutions of the HI assay.
-	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
+	Positioning of lineages, e.g.\ A/H3N2 and A/H1N1, relative to each other is arbitrary.
 	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
@@
 
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of $\beta=1.01$ antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04 (Figure~\ref{drift}).
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of $\beta=1.01$ antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04.
 However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
@@
 %%% drift %%%
+
+%% AR - Figure improvements: Give proper axis labels. Label clusters (Texas/77, Beijing/89, Beijing/92, Sydney/97 that are referenced in the text)
 \begin{figure}[h]
@@
 	Antigenic units represent two-fold dilutions of the HI assay.
-	Distances between clades, e.g.\ A/H3N2 and A/H1N1, are arbitrary.
+	Relative positioning of lineages, e.g.\ A/H3N2 and A/H1N1, in the vertical axis is arbitrary.
 	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
@@
 
-Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
+Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
 We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.

### 2013-01-25 flux.tex
Fix compile.

@@
 or \\
-Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages 
+Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages \\
 or \\
 Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution   
-
 }
@@
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single one prior to 1980 \cite{Rota92}.
+In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota92}.
 Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 

### 2013-01-27 flux.tex
Implement textual revisions suggested by John and Alan.

@@
 In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
-Using HI data for influenza lineages A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine average rates of antigenic drift for each clade, as well as year-to-year variability in the rate of drift.
-Through comparison with epidemiological data, we show that antigenic drift influences overall incidence across clades, and demonstrate a year-to-year correlation between drift and incidence.
-We investigate the selective underpinnings for differing antigenic dynamics across clades and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza clades.
-This work does much to elucidate the competitive dynamics of influenza lineages, but also makes possible substantial future advances in influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+Using HI data for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
+Through comparison with epidemiological data, we show that antigenic drift influences overall incidence across lineages, and demonstrate a year-to-year correlation between drift and incidence.
+We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza lineages.
+This work does much to elucidate the antigenic dynamics of influenza lineages, but also makes possible substantial future advances in influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
 
@@
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota92}.
-Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
+In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota90}.
+Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Wiley81, Nelson07NatRevGenet}. 
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
 Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over previously established strains.
-HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.
 
-The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
+The results of many HI assays across a multitude of virus strains of a single subtype can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
 The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
@@
 The process of antigenic drift results in the rapid turnover of the virus population. 
-Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
+Although mutation occurs rapidly, genetic diversity among contemporaneous viruses is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
 
@@
 However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
-Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of clades A/H3N2, A/H1N1, B/Vic and B/Yam.
+Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
+We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
 
@@
 We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).
+We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).
 
@@
 Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to a Brownian motion process (see Methods).
+Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Methods).
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
@@
 
-We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Figure~\ref{error_by_distance_and_year}A), thus supporting our model assumption that drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
-Additionally, we find that absolute error in predicted titer is nearly constant with time (Figure~\ref{error_by_distance_and_year}B).
+We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Figure~\ref{error_by_distance_and_year}A), thus supporting our model assumption that the drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
+Additionally, we find that the absolute error in predicted titer is nearly constant with time (Figure~\ref{error_by_distance_and_year}B).
 Antigenic locations inferred by the model are well resolved; estimates of antigenic distance between pairs of viruses show relatively little variation across the posterior.
@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 2 through 8; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
-Because of this, we interpret antigenic locations locally rather than globally.
-We can determine the rate of antigenic movement of virus lineages without knowing the larger configuration that the movement occurs under.
+Because of this, we interpret map locations locally rather than globally.
+We can determine the rate of antigenic movement of viruses withen each clade without knowing the larger configuration under which the movement occurs.
 
@@
 
-Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
+Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift in the mean antigenic phenotype of the virus population is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
 We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.
@@
 In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
-Thus, we observe fairly equal rates of year-to-year antigenic drift in mean phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
+Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
 In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
@@
 	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Timeseries of year-to-year antigenic drift across influenza clades from 1992 to 2011.
+	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011.
 	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
@@
 
-We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within each influenza clade (Figure~\ref{incidence}).
+We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within the USA for each influenza clade (Figure~\ref{incidence}).
+Here, we take incidence to be equivalent to frequency of isolations of a particular clade relative to all virus isolations. 
 Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
@@
 We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 6.43 \times 10^{-8}$).
-Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.56$, $r_\mathrm{H1} = 0.46$, $r_\mathrm{Vic} = 0.49$ and $r_\mathrm{Yam} = 0.23$.
+Correlation is present to greater or lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.56$, $r_\mathrm{H1} = 0.46$, $r_\mathrm{Vic} = 0.49$ and $r_\mathrm{Yam} = 0.23$.
 Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.
@@
 These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence of influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
-Furthermore, these findings suggest that the composition of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
+Furthermore, these findings suggest that the predominate clade of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
 However, these findings alone do not conclusively show interference between clades; more work needs to be done to establish whether antigenic drift within one influenza clades drives down incidence in other clades.
@@
 We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
-Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both fundamental reproductive number $R_0$ and rate cross-immunity decreasing mutation \cite{Gog02,Lin03}.
+Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
 Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
@@
 Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
-We expect the diffusion coefficient on side branches to more closely resemble the raw rate of antigenic mutation.
-Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
+We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
+Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
 Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
@@
 Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
-Thus, although drift forward along dimension 1 is promoted in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be promoted more strongly in influenza A than in influenza B, giving an advantage to influenza A in terms of overall rate of antigenic drift.
+Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving an advantage to influenza A in terms of overall rate of antigenic drift.
 Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them an advantage over their sister lineages A/H1N1 and B/Yam also in terms of overall rate of antigenic drift.
@@
 We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
-In the later case, we expect that many antigenic variants may be at inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
+In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
 These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
@@
 The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
-For example, our likelihood-based model facilities the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
+For example, our likelihood-based model facilitates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
 Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
@@
 At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting binding to red blood cells of another virus.
+In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting the binding of another virus to red blood cells.
 These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
@@
 	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
-	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 1 and 3.
-	Left and right panels represent cartographic models that would give equal likelihoods of a given set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
+	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
+	Left and right panels represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
 	} 
@@
 AR, MAS, TB and PL conceived the study.
-TB, GD, VG, AJH, JWM, CR, DS and AR gathered antigenic and genetic data.
+TB, GD, VG, JWM, AJH, CAR, DJS and AR gathered antigenic and genetic data.
 AR, MAS, TB and PL designed statistical phylogenetic procedures.
 TB performed the analysis.
-TB, MAS and AR wrote the paper.
+TB, MAS, AR, JWM, AJH wrote the paper.
 

### 2013-01-28 flux.tex
Improve figures.

@@
 %%% map %%%
-
-%% AR - Figure improvements: Give proper axis labels. Label clusters (Texas/77, Beijing/89, Beijing/92, Wuhan/95, Sydney/97 that are referenced in the text)
 \begin{figure}[h]
@@
 	Antigenic units represent two-fold dilutions of the HI assay.
-	Positioning of lineages, e.g.\ A/H3N2 and A/H1N1, relative to each other is arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
+	Absolute positioning of lineages, e.g.\ A/H3N2 and A/H1N1, is arbitrary.
+	Lines represent mean posterior diffusion paths when virus locations are fixed.
+	Antigenic clusters are labeled after vaccine strains present within clusters.
+	Abbreviated strains are as follows Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
 	\label{map} 
@@
 %%% drift %%%
-
-%% AR - Figure improvements: Give proper axis labels. Label clusters (Texas/77, Beijing/89, Beijing/92, Sydney/97 that are referenced in the text)
 \begin{figure}[h]

### 2013-01-28 flux.tex
Minor revisions.

@@
 \title{\vspace{1.0cm} \Large \bf 
-Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\
-or \\
 Charting antigenic drift and evolutionary success of seasonal influenza lineages \\
@@
 or \\
-Cartographic analysis of antigenic dynamics and evolutionary success of seasonal influenza lineages \\
-or \\
 Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages \\
 or \\
-Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution   
+Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution \\
+or \\
+Cartographic characterization of antigenic and evolutionary dynamics of seasonal influenza viruses
 }
@@
 In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
-Using HI data for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
+Using HI data for four lineages of influenza, encompassing influenza A subtypes H3N2 and H1N1, and influenza B lineages Victoria and Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
 Through comparison with epidemiological data, we show that antigenic drift influences overall incidence across lineages, and demonstrate a year-to-year correlation between drift and incidence.
@@
 
-Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
+Before 2009, there were four major clades or lineages of influenza circulating within the human population: the H3N2 and H1N1 subtypes of influenza A, and the Victoria and Yamagata lineages of influenza B.
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota90}.
+In the case of influenza B, Victoria (B/Vic) and Yamagata (B/Yam) refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota90}.
 Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Wiley81, Nelson07NatRevGenet}. 
@@
 Antigenic clusters show high genetic similarity, so that we observe very few mutation events leading to each cluster, rather than the repeated emergence of clusters.
-This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.
+This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Victoria/75 and Beijing/89) appearing to be evolutionary dead-ends.
 

### 2013-01-28 flux.tex
Begin description of interference analysis.

@@
 Interestingly, a minor lineage of B/Vic has persisted through 2011, appearing antigenically distinct from other 2011 viruses (Figure~\ref{drift}), though the eventual fate of this lineage has yet to be determined.
-Influenza B/Yamagata evolved slower still, with an average rate of and 0.32 (HPD 0.25--0.39) units per year.
+Influenza B/Yamagata evolved slower still, with an average rate of 0.32 (HPD 0.25--0.39) units per year.
 Little punctuated evolution is obvious in the evolution of B/Yam.
@@
 
-We investigated the relative rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and relative incidence within the USA for each influenza clade (Figure~\ref{incidence}).
-Here, we take incidence to be equivalent to frequency of isolations of a particular clade relative to all virus isolations. 
-Here, we take the measurements of year-to-year antigenic drift from the preceding section (Figure~\ref{jumps}) and calculate the relative contribution to total antigenic drift for each clade in each year.
-We compare these measurements of relative antigenic drift to relative isolations of viruses from each clade in the USA from the 2001--2002 to the 2011--2012 influenza seasons (see Methods).
-From 2000 to 2011, A/H3N2 accounted for the majority of antigenic drift (46\%), while A/H1N1, B/Vic and B/Yam split the remainder, accounting for 18\%, 21\% and 15\%, respectively.
-Similarly, A/H3N2 was responsible for the majority of incidence (58\%) during this time period.
-Influenza A/H1N1 accounted for 18\% of incidence, B/Vic accounted for 16\% of incidence and B/Yam accounted for 7\% of incidence.
-These proportions are significantly similar to one another (randomization test, $p = 0.030$).
+We investigated the rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and incidence within the USA for each influenza clade.
+We take the estimates of the rate of antigenic drift from the preceding section and compare these rates to a measurements of relative incidence for each influenza clade calculated from proportion of influenza isolations attributable to each clade (see Methods).
+We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid interactions with the 2009 H1N1 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of influenza isolations during this time period.
+We find a significant correlation between rate of antigenic drift and relative incidence across the four clades with $r=0.96$ (Pearson correlation, $p = 0.021$).
 
-%%% incidence %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.9\textwidth]{figures/incidence}
-	\caption{\textbf{Relative antigenic drift and relative incidence of A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011.} 
-	(A) Relative antigenic drift of influenza clades.
-	Relative antigenic drift for A/H3N2 is measured as the drift of A/H3N2 viruses divided by the sum of drift for A/H3N2, A/H1N1, B/Vic and B/Yam.
-	For example, `01' signifies difference in antigenic dimension 1 between 2000 and 2001.
-	Years with negative absolute drift were assigned a relative antigenic drift of 0.
-	(B) Relative incidence of influenza clades determined through relative isolation counts in the USA for winter influenza seasons.
-	Here, `01' refers to the 2001/2002 influenza season.
-	(C) Correlation of relative antigenic drift and relative incidence in each influenza clade.
-	Scatterplots represent drift from year $i-1$ to year $i$ compared to incidence in winter season beginning in year $i$, e.g.
-drift from 2000 to 2001 compared to incidence in the 2001/2002 influenza season.
-	} 
-	\label{incidence} 
-\end{figure}
-
+We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
 Additionally, year-to-year comparisons show a similar pattern, in which years with high levels of antigenic drift coming into an influenza season, e.g.\ drift from 2000 to 2001, show high levels of relative incidence in that season, e.g.\ incidence in the 2001/2002 season (Figure~\ref{incidence}).
@@
 Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.
+
+%%% interferencetable %%%
+\begin{table}[h]
+	\centering
+	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\beta_\mathrm{within} \times \mathrm{drift}_\mathrm{H3}$ + $\beta_\mathrm{sister} \times \mathrm{drift}_\mathrm{H1}$ + $\beta_\mathrm{across} \times \mathrm{drift}_\mathrm{Vic}$ + $\beta_\mathrm{across} \times \mathrm{drift}_\mathrm{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
+	\label{interferencetable}	
+	\begin{tabular}{ c c c c } 
+	\hline
+	Parameter				&	Estimate	&	Standard error 	& 	$p$-value \\
+	\hline		
+	$\beta_\mathrm{within}$	&	$1.19$		&	0.22			&	$2.5 \times 10^{-6}$ \\
+	$\beta_\mathrm{sister}$	&	$-0.67$		&	0.22			&	$0.004$ \\
+	$\beta_\mathrm{across}$	&	$-0.42$		&	0.16			&	$0.011$ \\
+	\hline
+	\end{tabular}
+\end{table}
 
@@
 
+%%% incidence %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/incidence}
+	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
+	(A) Incidence within a clade compared to antigenic drift within the same clade.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
+	(B) Incidence within a clade compared to antigenic drift in sister clades.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, and incidence of B/Vic in 1998/1999 is measured against antigenic drift of B/Yam from 1997 to 1998.
+	(B) Incidence within a clade compared to antigenic drift in non-type clades.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of B/Vic + B/Yam from 1997 to 1998, and incidence of B/Vic in 1998/1999 is measured against antigenic drift of A/H3N2  A/H1N1 from 1997 to 1998.
+	Incidence is measured as yearly ILI percentage multipled by proportion of virus isolations of a particular clade.
+	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
+	} 
+	\label{incidence} 
+\end{figure}
+
 \end{document}

### 2013-01-29 flux.tex
Clarify notation regarding internal nodes and phylogenetic diffusion.

--- 2013-01-28 flux.tex
@@
 \newcommand{\mdssd}{\varphi}						% MDS standard deviation
-\newcommand{\virussd}{\sigma_x}						% Virus / Diffusion standard deviation
-\newcommand{\serumsd}{\sigma_y}						% Serum standard deviation
-\newcommand{\tree}{\tau}							% Phylogeny
-\newcommand{\vn}{n}									% Number of viruses
-\newcommand{\sn}{k}									% Number of sera
-\newcommand{\node}{\mathbf{z}}						% Internal node in phylogeny
-\newcommand{\normal}{\mathcal{N}}					% Normal distribution
+\newcommand{\virussd}{\sigma_x}						% virus / diffusion standard deviation
+\newcommand{\serumsd}{\sigma_y}						% serum standard deviation
+\newcommand{\tree}{\tau}							% phylogeny
+\newcommand{\vn}{n}									% number of viruses
+\newcommand{\sn}{k}									% number of sera
+\newcommand{\normal}{\mathcal{N}}					% normal distribution
 \setlength{\arraycolsep}{2pt}
@@
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-The evolutionary Brownian process assumes that the location $\node_i$ of node $i$ on the phylogeny follows from its parent node $p$, and with the addition of the drift along dimension 1, is distributed as
+In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_0$.
+This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
 \begin{equation} 
-	\node_i \sim (\beta \, d, \, 0)^{\prime} + \normal(\node_p, d \, \boldsymbol{\Sigma})
+	\virus_i \sim (\beta \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
 \end{equation}
-for $P=2$, where $\node_p$ is the location of the parent node, $d$ is the length of the branch connecting nodes $i$ and $p$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd$.
-The root node $\node_0$ is assumed to follow a normal distribution with expectation $(\beta \, t_0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
+The root virus location $\virus_0$ is assumed to follow a normal distribution with expectation $(\beta \, t_0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.

### 2013-01-29 flux.tex
Annotate drift figure with cluster labels.

@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Antigenic clusters are labeled after vaccine strains present within clusters.
-	Abbreviated strains are as follows Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
+	Antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
 	\label{map} 
@@
 	Relative positioning of lineages, e.g.\ A/H3N2 and A/H1N1, in the vertical axis is arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.} 
+	Lines represent mean posterior diffusion paths when virus locations are fixed.
+	Antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
 	\label{drift} 

### 2013-01-29 flux.tex
Rewrite incidence comparison based on new regression model.

@@
 \newcommand{\normal}{\mathcal{N}}					% normal distribution
+\newcommand{\bwithin}{\beta_\mathrm{within}}		% within clade drift coefficient
+\newcommand{\bsister}{\beta_\mathrm{sister}}		% sister clade drift coefficient
+\newcommand{\bother}{\beta_\mathrm{other}}			% across clade drift coefficient
+\newcommand{\driftclade}[1]{x_\mathrm{#1}}
 \setlength{\arraycolsep}{2pt}
@@
 Using HI data for four lineages of influenza, encompassing influenza A subtypes H3N2 and H1N1, and influenza B lineages Victoria and Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
-Through comparison with epidemiological data, we show that antigenic drift influences overall incidence across lineages, and demonstrate a year-to-year correlation between drift and incidence.
+Through comparison with epidemiological data, we demonstrate a year-to-year correlation between drift and incidence and present evidence that antigenic drift mediates interference between influenza lineages.
 We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza lineages.
@@
 Because of this, we interpret map locations locally rather than globally.
-We can determine the rate of antigenic movement of viruses withen each clade without knowing the larger configuration under which the movement occurs.
+We can determine the rate of antigenic movement of viruses within each clade without knowing the larger configuration under which the movement occurs.
 
@@
 	\includegraphics[width=0.95\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2009 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
 	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011.
@@
 
-We investigated the rates of antigenic evolution in A/H3N2, A/H1N1, B/Vic and B/Yam from 2000 to 2011, finding a general correspondence between antigenic drift in dimension 1 and incidence within the USA for each influenza clade.
-We take the estimates of the rate of antigenic drift from the preceding section and compare these rates to a measurements of relative incidence for each influenza clade calculated from proportion of influenza isolations attributable to each clade (see Methods).
-We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid interactions with the 2009 H1N1 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of influenza isolations during this time period.
-We find a significant correlation between rate of antigenic drift and relative incidence across the four clades with $r=0.96$ (Pearson correlation, $p = 0.021$).
+We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
+We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza isolations attributable to each clade (see Methods).
+We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of influenza isolations in the USA during this time period.
+We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).
 
 We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
-Additionally, year-to-year comparisons show a similar pattern, in which years with high levels of antigenic drift coming into an influenza season, e.g.\ drift from 2000 to 2001, show high levels of relative incidence in that season, e.g.\ incidence in the 2001/2002 season (Figure~\ref{incidence}).
-We find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.71$ (Pearson correlation, $p = 6.43 \times 10^{-8}$).
-Correlation is present to greater or lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.56$, $r_\mathrm{H1} = 0.46$, $r_\mathrm{Vic} = 0.49$ and $r_\mathrm{Yam} = 0.23$.
-Pandemic H1N1 is present in the incidence data in 2009 and afterwards, however, our estimates of relative incidence only include isolations of seasonal A/H3N2, A/H1N1, B/Vic and B/Yam.
+We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
+We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of virus isolations attributable to a clade for each season.
+This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
+In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
+We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A).
+Here, within-clade drift explains 35\% of the variance in standardized incidence and presents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
+
+We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
+This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
+We fit a linear model that predicts standardized incidence based upon drift within clades, drift in sister clades, and drift in other clades (Table~\ref{interferencetable}).
+For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in sister clade A/H1N1 and by drift in other clades B/Vic and B/Yam.
+We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
+We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of antigenic drift in a sister clade decreases incidence by 0.61 standard deviations and 1 unit of antigenic drift in other clades decreases incidence by 0.32 standard deviations.
 
@@
 	\centering
-	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\beta_\mathrm{within} \times \mathrm{drift}_\mathrm{H3}$ + $\beta_\mathrm{sister} \times \mathrm{drift}_\mathrm{H1}$ + $\beta_\mathrm{across} \times \mathrm{drift}_\mathrm{Vic}$ + $\beta_\mathrm{across} \times \mathrm{drift}_\mathrm{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
+	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\bwithin \times \driftclade{H3} + \bsister \times \driftclade{H1} + \bother \times \driftclade{Vic} + \bother \times \driftclade{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
 	\label{interferencetable}	
@@
 	\hline		
-	$\beta_\mathrm{within}$	&	$1.19$		&	0.22			&	$2.5 \times 10^{-6}$ \\
-	$\beta_\mathrm{sister}$	&	$-0.67$		&	0.22			&	$0.004$ \\
-	$\beta_\mathrm{across}$	&	$-0.42$		&	0.16			&	$0.011$ \\
+	$\bwithin$	&	$1.26$		&	0.22			&	$1.1 \times 10^{-6}$ \\
+	$\bsister$	&	$-0.61$		&	0.22			&	$0.008$ \\
+	$\bother$	&	$-0.32$		&	0.11			&	$0.005$ \\
 	\hline
@@
 
-These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence of influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly impacted by levels of antigenic drift within each clade.
+These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly influenced by levels of antigenic drift within each clade, consistent with antigenic drift increasing the susceptible pool available to a clade.
+The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
+Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2.
+If interference is mediated by CTL epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.
+
 Furthermore, these findings suggest that the predominate clade of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
-However, these findings alone do not conclusively show interference between clades; more work needs to be done to establish whether antigenic drift within one influenza clades drives down incidence in other clades.
+However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
+Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.
 
@@
 In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
-We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each influenza clade.
+We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each clade and interference between clades.
 We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
@@
 	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
-	(A) Incidence within a clade compared to antigenic drift within the same clade.
+	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
 	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
-	(B) Incidence within a clade compared to antigenic drift in sister clades.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, and incidence of B/Vic in 1998/1999 is measured against antigenic drift of B/Yam from 1997 to 1998.
-	(B) Incidence within a clade compared to antigenic drift in non-type clades.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of B/Vic + B/Yam from 1997 to 1998, and incidence of B/Vic in 1998/1999 is measured against antigenic drift of A/H3N2  A/H1N1 from 1997 to 1998.
-	Incidence is measured as yearly ILI percentage multipled by proportion of virus isolations of a particular clade.
+	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
+	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of virus isolations of a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
 	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.

### 2013-01-29 flux.tex
Fix root node notation.

@@
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_0$.
+In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
 This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
@@
 for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
-The root virus location $\virus_0$ is assumed to follow a normal distribution with expectation $(\beta \, t_0, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\beta \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.

### 2013-01-29 flux.tex
Minor revision.

@@
 
-Further work characterizing antigenic evolution in the human influenza virus may eventually prove to be crucial in improving vaccine strain selection.
+A deeper understanding of the process of antigenic evolution in the human influenza virus may eventually prove to be crucial in improving vaccine strain selection.
 

### 2013-01-29 flux.tex
Include revisions based on Philippe's suggestions.

@@
 Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
+We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza clades.
 
@@
 These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
+Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
 Still, it is clear that more work is necessary to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
@@
 In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
+Here, methods to efficiently calculate branch- and site-specific codon substitutions may be necessary to easily correlate sequence substitutions with antigenic change \cite{Lemey12}.
 
@@
 Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
-
-\subsection*{Likelihood functions}
 
@@
 using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We begin by assuming independent, diffuse normal prior on virus and serum locations
+We begin by assuming independent, diffuse normal priors on virus and serum locations
 \begin{eqnarray}
@@
 \end{eqnarray}
-with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
+where $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 
@@
 
-\subsection*{Location prior}
+\subsection*{Drift model of antigenic evolution}
 
-As presented, virus and serum locations $\viruses$ and $\sera$ give a likelihood of observing the HI data $\mathbf{H}$.
-However, distances between viruses and sera represent only a local description of antigenic space and do not provide a global picture.
+As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
 An example of this phenomenon is shown in Figure~\ref{schematic_map}.
-In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
-This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
-In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
+In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B).
+This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from the observing the serological data alone.
+Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
 In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
@@
 	\caption{\textbf{Schematic antigenic map with three viruses and two sera.}
+	(A) Map with virus 1 and virus 3 antigenically similar.
+	(B) Map with virus 1 and virus 3 antigenically divergent.
 	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
@@
 	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
-	Left and right panels represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
+	Configurations (A) and (B) represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
 	} 
@@
 
-\subsection*{Diffusion model of antigenic evolution}
+\subsection*{Phylogenetic diffusion model of antigenic evolution}
 
@@
 
-Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
+Phylogenetic trees were estimated for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
 MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
 These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
+However, it would be possible to jointly sample from sequence data and serological data using these methods.
 
@@
 For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
-These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
+These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Victoria/75 in place.
 Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.

### 2013-01-29 flux.tex
Minor revisions.

@@
 or \\
-Integrating antigenic cartography and molecular evolution to characterize predictors of evolutionary success of seasonal influenza lineages \\
-or \\
 Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution \\
 or \\
-Cartographic characterization of antigenic and evolutionary dynamics of seasonal influenza viruses
+Cartographic characterization of the antigenic, evolutionary and competitive dynamics of seasonal influenza viruses \\
+or \\
+Integrating antigenic dynamics with molecular evolution to characterize competition within and between seasonal influenza lineages \\
 }

### 2013-01-30 flux.tex
Give Table 1 a more logical organization that mirrors Methods development.

@@
 By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers. 
-We begin with a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
-We find that this model performs well, yielding an average absolute predictive error of between 0.79 and 0.95 log$_2$ HI titers across influenza lineages (model 1; Table \ref{errortable}).
-We extend this model to include a time-informed prior that models the process of antigenic drift, in which virus and serum locations  along dimension 1 increase with date of virus or serum sampling, resulting in a small improvement in predictive error (model 2; Table \ref{errortable}).
+We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
+After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
+We arrive at a similar conclusion, finding that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
+We find that this 2D model performs well, yielding an average absolute predictive error of between XXX and XXX log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}).
+Consequently, we specify a two dimensional model in all subsequent analyses.
+The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting and fundamental empirical observation.
 
@@
 	\hline	
-	1		&	2D 		&	Fixed 			&	None			& Uninformed		&	0.93 	&	0.78 	&	0.80	&	0.93	\\
-	2		&	2D 		&	Fixed			&	None			& Drift				&	0.91 	&	0.74 	&	0.77	&	0.83	\\	
-	3		&	1D 		&	Fixed			&	None			& Drift				&	1.07	&	0.92	&	0.94	&	1.08	\\		
-	4		&	3D 		&	Fixed			&	None			& Drift				&	0.91	&	0.74 	&	0.79	&	0.84	\\
-	5		&	4D 		&	Fixed			&	None			& Drift				&	0.95	&	0.75 	&	0.81	&	0.86	\\
-	6		&	5D 		&	Fixed			&	None			& Drift				&	0.99	&	0.76 	&	0.84	&	0.88	\\				
+	1		&	1D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.94 	&	0.90	&	1.08	\\
+	2		&	2D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.78 	&	0.82	&	0.90	\\
+	3		&	3D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.80 	&	0.85	&	0.92	\\
+	4		&	4D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.84 	&	0.90	&	0.97	\\	
+	5		&	5D 		&	Fixed 			&	None			& Uninformed		&	X 		&	X 		&	0.98	&	1.04	\\	
+	6		&	2D 		&	Estimated 		&	Estimated		& Uninformed		&	X 		&	X 		&	X		&	X		\\
 	7		&	2D 		&	Estimated		&	Estimated		& Drift				&	0.77	&	0.71 	&	0.65	&	0.73	\\
@@
 
-We find that decreasing dimensionality to a single antigenic dimension results in a substantially less predictive model (model 3; Table \ref{errortable}), and that increasing dimensionality to three dimensions is equal or slightly worse in predictive power (model 4; Table \ref{errortable}).
-Increasing dimensionality further to four or five dimensions results in continued declines in predictive power (models 5 and 6; Table \ref{errortable}).
-Consequently, we specify a two dimensional model in all subsequent analyses.
-The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting empirical observation.
 We extend the 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
 We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error further, ranging from an improvement of 0.03 in A/H1N1 to an improvement of 0.14 in A/H3N2 (model 7; Table \ref{errortable}).
+We find that including these effects decreases test error further, ranging in improvement from XXX to XXX (model 6; Table \ref{errortable}).
+We further extend this model to include a time-informed prior that models the process of antigenic drift, in which virus and serum locations along dimension 1 increase with date of virus or serum sampling, resulting in a small improvement in predictive error (model 7; Table \ref{errortable}).
 
-Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}.
+Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01, Smith04, Russell08}.
 Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Methods).

### 2013-01-30 flux.tex
Make revisions based on Gytis's suggestions.

@@
 We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza lineages.
-This work does much to elucidate the antigenic dynamics of influenza lineages, but also makes possible substantial future advances in influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
 
@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
 	\label{map} 
@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
 	\label{drift} 
@@
 Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift in the mean antigenic phenotype of the virus population is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
-We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present at year $i$ and comparing this location to the average location of phylogenetic lineages present at year $i-1$.
+We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
 For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster, evident as the appearance in 1997 of a large gap in virus locations along antigenic dimension 1 (Figure~\ref{drift}).
@@
 Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
-Still, it is clear that more work is necessary to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
+Still, it is clear that more work is necessary to fully clarify the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
 

### 2013-01-31 flux.tex
Further reorganize Table 1.

@@
 We arrive at a similar conclusion, finding that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
-We find that this 2D model performs well, yielding an average absolute predictive error of between XXX and XXX log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}).
+We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}).
 Consequently, we specify a two dimensional model in all subsequent analyses.
@@
 	\multicolumn{5}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Dimen 	& 	Serum effects 	&	Virus effects	& Location Prior	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	Model	&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline	
-	1		&	1D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.94 	&	0.90	&	1.08	\\
-	2		&	2D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.78 	&	0.82	&	0.90	\\
-	3		&	3D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.80 	&	0.85	&	0.92	\\
-	4		&	4D 		&	Fixed 			&	None			& Uninformed		&	X 		&	0.84 	&	0.90	&	0.97	\\	
-	5		&	5D 		&	Fixed 			&	None			& Uninformed		&	X 		&	X 		&	0.98	&	1.04	\\	
-	6		&	2D 		&	Estimated 		&	Estimated		& Uninformed		&	X 		&	X 		&	X		&	X		\\
-	7		&	2D 		&	Estimated		&	Estimated		& Drift				&	0.77	&	0.71 	&	0.65	&	0.73	\\
-	8		&	2D 		&	Estimated		&	Estimated		& Diffusion/Drift	&	0.76	&	0.71 	&	0.64	&	0.72	\\		
+	1		&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
+	2		&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
+	3		&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
+	4		&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
+	5		&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
+	6		&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
+	7		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
+	8		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
 	\hline
@@
 
-We extend the 2D model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
-We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error further, ranging in improvement from XXX to XXX (model 6; Table \ref{errortable}).
-We further extend this model to include a time-informed prior that models the process of antigenic drift, in which virus and serum locations along dimension 1 increase with date of virus or serum sampling, resulting in a small improvement in predictive error (model 7; Table \ref{errortable}).
-
 Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01, Smith04, Russell08}.
@@
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy when including covariance induced by the phylogenetic diffusion (model 8; Table \ref{errortable}).
-Additionally, modeling antigenic diffusion has the advantage of allowing direct correlation of evolutionary outcomes with changes in antigenic phenotype.
+The antigenic diffusion process includes both systematic drift along the tree and covariance induced by phylogenetic proximity.
+We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when including the phylogenetic diffusion (model 6; Table \ref{errortable}).
+
+We further extend the model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
+We refer to these estimates as serum effects and virus effects, respectively.
+We find that including these effects decreases test error further, ranging in improvement from 0.03 to 0.13 log$_2$ HI titers (models 7 and 8; Table \ref{errortable}).
 

### 2013-02-01 flux.tex
Fix typo.

@@
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Weiner process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \cite{BrownianMotionHandbook}.
+This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \cite{BrownianMotionHandbook}.
 This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.

### 2013-02-01 flux.tex
Fix typo.

@@
 The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Weiner process \cite{BrownianMotionHandbook}.
+This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
 This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.

### 2013-02-01 flux.tex
Implement revisions suggested by Alan.

@@
 
-Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity to previous virus strains acquired by the host population.
+Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
 Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
@@
 Through comparison with epidemiological data, we demonstrate a year-to-year correlation between drift and incidence and present evidence that antigenic drift mediates interference between influenza lineages.
-We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza lineages.
+We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences.
 This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
@@
 
-Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
+Seasonal influenza infects between 10\% and 20\% of the human population every year, causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
 While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection.
@@
 We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
-We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza isolations attributable to each clade (see Methods).
-We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of influenza isolations in the USA during this time period.
+We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Methods).
+We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of viruses detected in the USA during this time period.
 We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).
@@
 We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
-We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of virus isolations attributable to a clade for each season.
+We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of viruses attributable to a clade for each season.
 This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
 In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
-We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A).
+We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A).
 Here, within-clade drift explains 35\% of the variance in standardized incidence and presents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
 
-We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
+We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
 This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
@@
 
-Furthermore, these findings suggest that the predominate clade of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
+Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
 However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
@@
 Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
-Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving an advantage to influenza A in terms of overall rate of antigenic drift.
-Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them an advantage over their sister lineages A/H1N1 and B/Yam also in terms of overall rate of antigenic drift.
+Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
+Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.
 
@@
 	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
-	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of virus isolations of a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
+	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
 	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.

### 2013-02-04 flux.tex
Implement revisions based on Colin's suggestions.

@@
 This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Victoria/75 and Beijing/89) appearing to be evolutionary dead-ends.
+Labeling of prominent antigenic clusters in figures \ref{map} and \ref{drift} is intended as a rough guide for orientation and not as exhaustive catalog of antigenic variation.
 
@@
 
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of $\beta=1.01$ antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04.
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.01 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04 (Figure~\ref{drift}).
 However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
@@
 Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
+This observation remains despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that come to take over the virus population.
 In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).

### 2013-02-04 flux.tex
Further revisions to clarify what I mean by "drift" and "clade".

@@
 \newcommand{\normal}{\mathcal{N}}					% normal distribution
-\newcommand{\bwithin}{\beta_\mathrm{within}}		% within clade drift coefficient
-\newcommand{\bsister}{\beta_\mathrm{sister}}		% sister clade drift coefficient
-\newcommand{\bother}{\beta_\mathrm{other}}			% across clade drift coefficient
+\newcommand{\bwithin}{\beta_c}		% within clade drift coefficient
+\newcommand{\bsister}{\beta_s}		% sister clade drift coefficient
+\newcommand{\bother}{\beta_t}			% across clade drift coefficient
 \newcommand{\driftclade}[1]{x_\mathrm{#1}}
@@
 \title{\vspace{1.0cm} \Large \bf 
-Charting antigenic drift and evolutionary success of seasonal influenza lineages \\
-or \\
 Integrating influenza antigenic dynamics with molecular evolution \\
 or \\
-Predicting the evolutionary success of seasonal influenza lineages through the integration of antigenic cartography and molecular evolution \\
-or \\
-Cartographic characterization of the antigenic, evolutionary and competitive dynamics of seasonal influenza viruses \\
-or \\
-Integrating antigenic dynamics with molecular evolution to characterize competition within and between seasonal influenza lineages \\
+Cartographic characterization of the antigenic, evolutionary and competitive dynamics of seasonal influenza viruses 
 }
@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
 	\label{map} 
@@
 Because of this, we interpret map locations locally rather than globally.
-We can determine the rate of antigenic movement of viruses within each clade without knowing the larger configuration under which the movement occurs.
+We can determine rates of antigenic movement without knowing the larger configuration under which the movement occurs.
 
@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
 	\label{drift} 
@@
 In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
-Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype in going from 1996 to 1997, from 1997 to 1998 and from 1998 to 1999, rather than all drift in mean antigenic phenotype attributable to a shift in 1997 (Figure~\ref{jumps}).
-This observation remains despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that come to take over the virus population.
-In addition to showing greater overall rates of antigenic drift, A/H3N2 shows greater variation in year-to-year drift than seen in A/H1N1 or influenza B (Figure~\ref{jumps}).
-We find that the standard deviation in year-to-year drift is 0.88 antigenic units for A/H3N2, 0.51 units for A/H1N1, 0.54 units for B/Vic and 0.27 units for B/Yam.
+Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype between 1996 and 1999, rather than all drift in mean antigenic phenotype coinciding with the appearence of a new antigenic variant in 1997 (Figure~\ref{jumps}).
+It is interesting that this phenomenon occurs despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that each come to take over the virus population.
+Because drifted variants require, on average, 2--3 years to take over the virus population, rates of year-to-year drift do not always spike heavily at the appearance of a new variant.
 
@@
 This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
-We fit a linear model that predicts standardized incidence based upon drift within clades, drift in sister clades, and drift in other clades (Table~\ref{interferencetable}).
-For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in sister clade A/H1N1 and by drift in other clades B/Vic and B/Yam.
+We fit a linear model that predicts standardized incidence based upon drift within a clade, drift in other clades of same type, and drift in clades of differing type (Table~\ref{interferencetable}).
+For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in A/H1N1 and by drift across types in B/Vic and B/Yam.
 We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
-We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of antigenic drift in a sister clade decreases incidence by 0.61 standard deviations and 1 unit of antigenic drift in other clades decreases incidence by 0.32 standard deviations.
+We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations.
 
@@
 	\centering
-	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\bwithin \times \driftclade{H3} + \bsister \times \driftclade{H1} + \bother \times \driftclade{Vic} + \bother \times \driftclade{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
+	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
 	\label{interferencetable}	
@@
 
-These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam is strongly influenced by levels of antigenic drift within each clade, consistent with antigenic drift increasing the susceptible pool available to a clade.
+These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence across  A/H3N2, A/H1N1, B/Vic and B/Yam is strongly influenced by levels of antigenic drift within each clade, consistent with antigenic drift increasing the susceptible pool available to a clade.
 The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.

### 2013-02-04 flux.tex
Implement wording fix suggested by @evogytis.

@@
 Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype between 1996 and 1999, rather than all drift in mean antigenic phenotype coinciding with the appearence of a new antigenic variant in 1997 (Figure~\ref{jumps}).
-It is interesting that this phenomenon occurs despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that each come to take over the virus population.
+It is interesting that this phenomenon occurs despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that sweep through the virus population.
 Because drifted variants require, on average, 2--3 years to take over the virus population, rates of year-to-year drift do not always spike heavily at the appearance of a new variant.

### 2013-02-19 flux.tex
Fix typos.

@@
 We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A).
-Here, within-clade drift explains 35\% of the variance in standardized incidence and presents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
+Here, within-clade drift explains 35\% of the variance in standardized incidence and represents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
 
@@
 This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
-We fit a linear model that predicts standardized incidence based upon drift within a clade, drift in other clades of same type, and drift in clades of differing type (Table~\ref{interferencetable}).
+We fit a linear model that predicts standardized incidence based upon drift within a clade, drift in other clades of the same virus type, and drift in clades of differing type (Table~\ref{interferencetable}).
 For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in A/H1N1 and by drift across types in B/Vic and B/Yam.

### 2013-02-19 cover_letter.tex
Include cover letter.

\documentclass[stdletter,letterpaper,addrfromright,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
\topmarginskip{-0.4in}
\bottommarginskip{-1.5in}
\leftmarginsize{1in}
\rightmarginsize{1.25in}
\sigskipbefore{0.2in}
\sigskipafter{0in}
\noLines
\nolines
\noHeadline
\noheadline
\signature{Trevor Bedford}

\namefrom{}
\addrfrom{Inst.\ of Evolutionary Biology \\ University of Edinburgh \\ Ashworth Laboratories \\ King's Buildings \\ Edinburgh, UK}
\phonefrom{(617) 285-2542}
\faxfrom{(734) 763-0544}
\emailfrom{bedfordt@umich.edu}

\greetto{Dear Prof.\ Levin,}
\closeline{Sincerely,}

% comments
\usepackage{color} 
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

\begin{document}

\begin{newlfm}

Please find, attached, our manuscript entitled ``Integrating influenza antigenic dynamics with molecular evolution,'' which we would be grateful for your consideration for publication in \textit{Proc Natl Acad Sci USA}.  
We believe that \textit{Proc Natl Acad Sci USA} is an appropriate venue for this work, as it clarifies long-standing questions about the antigenic and evolutionary dynamics of influenza viruses, while simultaneously presenting a major methodological advance in the characterization of antigenic phenotype.

Since the publication of ``Mapping the Antigenic and Genetic Evolution of Influenza Virus'' by Smith et al. in 2004, there has been significant interest in characterizing patterns of antigenic and genetic evolution in seasonal influenza.
Such cartographic approaches use multidimensional scaling (MDS) to position viruses and antisera on an antigenic map so that distances between viruses and antisera predict observed serological data.
Here, we present a substantial advance to previous approaches that uses Bayesian multidimensional scaling (BMDS) to place antigenic cartography in a fully probabilistic framework.
Doing so has allowed us to build in important covariates into the antigenic model; most importantly, phylogenetic relationships.
Thus, our method simultaneously models evolutionary and antigenic dynamics, and allows us to infer which phylogeny branches have evolved in antigenic phenotype and which branches have remained static.

We test our method with substantial new datasets for influenza A/H3N2, A/H1N1, B/Victoria and B/Yamagata, presenting, for the first time, detailed evolutionary and antigenic reconstructions for all four major circulating lineages.
We find that influenza A/H3N2 shows the fastest and most punctuated evolution of antigenic phenotype relative to the other three circulating lineages.
We connect antigenic evolution to incidence patterns, showing that antigenic drift leading into an influenza season accounts for 46\% of the variance in seasonal incidence.
This analysis shows a strong signal of interference between lineages, with antigenic drift in a sister lineage having a negative impact on within-lineage incidence.
Additionally, we've used phylogenetic information to characterize fundamental differences in antigenic dynamics between lineages, finding that A/H3N2 benefits from both a greater influx of new antigenic mutations and from a more rapid spread of these mutations through the population.

Both the link between incidence and antigenic drift and the presence of interference across influenza subtypes have remained large-standing questions in the influenza community.
We believe our results do much to address these questions.
Additionally, we believe our methodological advances will prove fundamental to future research in influenza and other antigenically variable pathogens.
Our unified genetic / antigenic approach makes it possible to look closely for causal relationships between genetic mutations and antigenic change, and to characterize, for the first time, the antigenic mutations that are ultimately successful in the global competition among virus lineages.

\end{newlfm}
\end{document}

### 2013-02-19 flux.tex
Move methods to supplement.

@@
 %%% TITLE %%%
-\title{\vspace{1.0cm} \Large \bf 
-Integrating influenza antigenic dynamics with molecular evolution \\
-or \\
-Cartographic characterization of the antigenic, evolutionary and competitive dynamics of seasonal influenza viruses 
+\title{\vspace{1.0cm} \LARGE \bf 
+Integrating influenza antigenic dynamics with molecular evolution
 }
@@
 \maketitle
+
+\pagebreak
 
@@
 
-\subsection*{Antigenic cartography}
-
-Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
-At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting the binding of another virus to red blood cells.
-These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
-Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
-Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
-Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
-
-Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
-These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
-
-Here, we follow previous models in representing antigenic locations as points in a low $P$-dimensional antigenic map. 
-One of our initial goals is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into this lower dimensional space. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which we construct a probabilistic model to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
-Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
-
-Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
-Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}$ for $P=2$.
-For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
-Virus and serum may be isolated from / raised against the same strain and have different cartographic locations, and separate serum isolates raised against the same strain may also have different cartographic locations.
-This gives a set of distances between virus and serum cartographic locations 
-\begin{equation}
-	\delta_{ij} =  || \virus_i - \serum_j ||_2,
-\end{equation}
-where $|| \cdot ||_2$ is an $L_2$ norm.
-
-Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
-\begin{equation}
-	d_{ij} =  \se_j - H_{ij},
-\end{equation}
-where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
-In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
-\begin{equation}
-	\sum_{(i,j) \in \cal I} 
-	\left(
-		\delta_{ij} - d_{ij}
-	\right)^2
-\end{equation}
-is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
-
-Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
-\begin{equation} \label{hij}
-	H_{ij} \sim \normal( \se_j - \delta_{ij}, \, \mdssd^2 ).
-\end{equation}
-Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
-\begin{equation} 
-	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
-Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
-In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
-
-HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
-In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
-\begin{equation} 
-	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
-Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:160 dilution and no inhibition at 1:320 dilution.
-Rather than taking the HI titer as 160, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 160 and 320.
-HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
-\begin{equation} 
-	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
-\end{equation}
-These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
-Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
-
-%%% hij_likelihood %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
-	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
-	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
-	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
-	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
-	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
-	} 
-	\label{hij_likelihood} 
-\end{figure}
-
-We calculate the overall likelihood by multiplying probabilities of individual measurements
-\begin{equation} 
-	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
-\end{equation}
-using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We begin by assuming independent, diffuse normal priors on virus and serum locations
-\begin{eqnarray}
-	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
-	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
-\end{eqnarray}
-where $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
-
-\subsection*{Virus and serum effects}
-
-The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
-However, this model may be biased in some circumstances.
-In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
-To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
-We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
-This formulation assumes that particular sera are more reactive in general than other sera.
-
-Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
-With virus reactivity included, observed titers follow
-\begin{equation}
-	H_{ij} \sim \normal \left( \frac{\ve_i+\se_j}{2} - \delta_{ij}, \, \mdssd^2 \right),
-\end{equation}
-and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
-
-\subsection*{Drift model of antigenic evolution}
-
-As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
-An example of this phenomenon is shown in Figure~\ref{schematic_map}.
-In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B).
-This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from the observing the serological data alone.
-Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
-In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
-\begin{eqnarray}
-	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
-	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
-\end{eqnarray}
-where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
-Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
-The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
-
-%%% schematic_map %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
-	\caption{\textbf{Schematic antigenic map with three viruses and two sera.}
-	(A) Map with virus 1 and virus 3 antigenically similar.
-	(B) Map with virus 1 and virus 3 antigenically divergent.
-	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
-	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
-	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
-	Configurations (A) and (B) represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
-	} 
-	\label{schematic_map} 
-\end{figure}
-
-\subsection*{Phylogenetic diffusion model of antigenic evolution}
-
-We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
-To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
-\begin{equation} \label{ebpprior}
-	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
-\end{equation}
-for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion and $\tree$ is a phylogeny specifying tree topology and branch lengths.
-Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
-This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
-\begin{equation} 
-	\virus_i \sim (\beta \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
-\end{equation}
-for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
-The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\beta \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
-The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
-This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
-
-Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
-
-\subsection*{Posterior samples}
-
-Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
-Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
-\begin{equation}
-	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
-	p(\viruses | \beta, \virussd, \tree) \;
-	p(\sera | \beta, \serumsd) \; 
-	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
-\end{equation}
-We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
-For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
-
-\subsection*{Data and implementation}
+\subsection*{Methods summary}
 
@@
 
-Phylogenetic trees were estimated for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
-MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
-These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
-However, it would be possible to jointly sample from sequence data and serological data using these methods.
-
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
-MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
-
-There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
-Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
-However, local isometries are often still a problem.
-For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
-These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Victoria/75 in place.
-Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
-For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
-This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
+HI titers were modeled by positioning viruses and sera in a $N$-d antigenic landscape, in which distances between viruses and sera yield an expectation for log$_2$ HI titer (see Supp.\ Text).
+Thus, given a set of virus and serum locations we calculate the likelihood of observing a set of HI titers.
+We initially begin with uniformed priors on virus and serum locations, but proceed to relax this assumption and incorporate priors informed by time and phylogenetic distance.
+Here, we assume that viruses diffuse in antigenic space following a Brownian motion process, so that evolutionary similar viruses are expected to have similar antigenic locations.
+We integrate across virus and serum locations and other parameters through the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST17}.
 
@@
 Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Antigenic, genetic and surveillance data is available as supplemental information \tbc{PREPARE}.
+Antigenic, genetic and surveillance data is available as supplemental information.
 

### 2013-02-19 flux_si.tex
Move methods to supplement.

@@
 \usepackage{amssymb}
-\newcommand{\normal}{\mathcal{N}}					% Normal distribution
+\newcommand{\virus}{\mathbf{x}}						% virus coordinate
+\newcommand{\serum}{\mathbf{y}}						% serum coordinate
+\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
+\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
+\newcommand{\ve}{v}									% virus effect
+\newcommand{\se}{s}									% serum effect
+\newcommand{\ves}{\mathbf{v}}						% set of virus effects
+\newcommand{\ses}{\mathbf{s}}						% set of serum effects
+\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
+\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
+\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
+\newcommand{\mdssd}{\varphi}						% MDS standard deviation
+\newcommand{\virussd}{\sigma_x}						% virus / diffusion standard deviation
+\newcommand{\serumsd}{\sigma_y}						% serum standard deviation
+\newcommand{\tree}{\tau}							% phylogeny
+\newcommand{\vn}{n}									% number of viruses
+\newcommand{\sn}{k}									% number of sera
+\newcommand{\normal}{\mathcal{N}}					% normal distribution
+\newcommand{\bwithin}{\beta_c}						% within clade drift coefficient
+\newcommand{\bsister}{\beta_s}						% sister clade drift coefficient
+\newcommand{\bother}{\beta_t}						% across clade drift coefficient
+\newcommand{\driftclade}[1]{x_\mathrm{#1}}
 \setlength{\arraycolsep}{2pt}
+\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
+\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
 \newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
@@
 Supporting Text: \\
-Revealing the competitive dynamics of influenza viruses through evolutionary cartography
+Integrating influenza antigenic dynamics with molecular evolution
 }
@@
 \section*{Supporting Methods}
+
+\subsection*{Antigenic cartography}
+
+Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
+At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
+In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting the binding of another virus to red blood cells.
+These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
+Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
+Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
+Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
+
+Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
+These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
+Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
+
+Here, we follow previous models in representing antigenic locations as points in a low $P$-dimensional antigenic map. 
+One of our initial goals is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into this lower dimensional space. 
+We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which we construct a probabilistic model to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
+Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
+
+Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
+Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}$ for $P=2$.
+For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
+Virus and serum may be isolated from / raised against the same strain and have different cartographic locations, and separate serum isolates raised against the same strain may also have different cartographic locations.
+This gives a set of distances between virus and serum cartographic locations 
+\begin{equation}
+	\delta_{ij} =  || \virus_i - \serum_j ||_2,
+\end{equation}
+where $|| \cdot ||_2$ is an $L_2$ norm.
+
+Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
+\begin{equation}
+	d_{ij} =  \se_j - H_{ij},
+\end{equation}
+where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
+In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
+\begin{equation}
+	\sum_{(i,j) \in \cal I} 
+	\left(
+		\delta_{ij} - d_{ij}
+	\right)^2
+\end{equation}
+is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
+
+Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
+\begin{equation} \label{hij}
+	H_{ij} \sim \normal( \se_j - \delta_{ij}, \, \mdssd^2 ).
+\end{equation}
+Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
+\begin{equation} 
+	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
+\end{equation}
+where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
+Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
+In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
+
+HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
+In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
+\begin{equation} 
+	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
+\end{equation}
+where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
+Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:160 dilution and no inhibition at 1:320 dilution.
+Rather than taking the HI titer as 160, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 160 and 320.
+HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
+\begin{equation} 
+	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
+\end{equation}
+These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
+Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
+
+%%% hij_likelihood %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
+	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
+	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
+	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
+	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
+	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
+	} 
+	\label{hij_likelihood} 
+\end{figure}
+
+We calculate the overall likelihood by multiplying probabilities of individual measurements
+\begin{equation} 
+	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
+\end{equation}
+using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
+We begin by assuming independent, diffuse normal priors on virus and serum locations
+\begin{eqnarray}
+	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
+	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
+\end{eqnarray}
+where $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
+
+\subsection*{Virus and serum effects}
+
+The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
+However, this model may be biased in some circumstances.
+In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
+To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
+In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
+We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
+We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
+This formulation assumes that particular sera are more reactive in general than other sera.
+
+Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
+With virus reactivity included, observed titers follow
+\begin{equation}
+	H_{ij} \sim \normal \left( \frac{\ve_i+\se_j}{2} - \delta_{ij}, \, \mdssd^2 \right),
+\end{equation}
+and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+
+\subsection*{Drift model of antigenic evolution}
+
+As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
+An example of this phenomenon is shown in Figure~\ref{schematic_map}.
+In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B).
+This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from the observing the serological data alone.
+Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
+In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
+To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
+\begin{eqnarray}
+	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
+	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
+\end{eqnarray}
+where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
+Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
+The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
+
+%%% schematic_map %%%
+\begin{figure}[tb]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
+	\caption{\textbf{Schematic antigenic map with three viruses and two sera.}
+	(A) Map with virus 1 and virus 3 antigenically similar.
+	(B) Map with virus 1 and virus 3 antigenically divergent.
+	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
+	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
+	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
+	Configurations (A) and (B) represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
+	} 
+	\label{schematic_map} 
+\end{figure}
+
+\subsection*{Phylogenetic diffusion model of antigenic evolution}
+
+We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
+To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
+\begin{equation} \label{ebpprior}
+	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
+\end{equation}
+for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion and $\tree$ is a phylogeny specifying tree topology and branch lengths.
+Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
+In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
+This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
+\begin{equation} 
+	\virus_i \sim (\beta \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
+\end{equation}
+for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
+The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\beta \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
+This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
+
+Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
+
+\subsection*{Posterior samples}
+
+Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
+Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
+\begin{equation}
+	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
+	p(\viruses | \beta, \virussd, \tree) \;
+	p(\sera | \beta, \serumsd) \; 
+	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
+\end{equation}
+We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST17}.
+Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
+For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
 
@@
 Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
+
+\subsection*{Implementation}
+
+Phylogenetic trees were estimated for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
+MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
+These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
+However, it would be possible to jointly sample from sequence data and serological data using these methods.
+
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
+MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
+
+There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
+Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
+However, local isometries are often still a problem.
+For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
+These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Victoria/75 in place.
+Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
+For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~1, Figure~2).
+Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
+This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
 

### 2013-02-22 flux.tex
Small cleanup.

--- 2013-02-19 flux.tex
@@
 
-% comments
-\usepackage{ulem}
-\definecolor{blue}{rgb}{0.324,0.609,0.708}
-\definecolor{purple}{rgb}{0.459,0.109,0.538}
-\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
-\def\tbc#1{\textcolor{purple}{[#1]}}
-\def\ms#1#2{\sout{#1} \textcolor{blue}{#2}} 
-\def\msc#1{\textcolor{blue}{[#1]}}
-
 %%% TITLE %%%
 \title{\vspace{1.0cm} \Large \bf 
-Integrating influenza antigenic dynamics with molecular evolution \\
-or \\
-Cartographic characterization of the antigenic, evolutionary and competitive dynamics of seasonal influenza viruses 
+Integrating influenza antigenic dynamics with molecular evolution
 }
@@
 \maketitle
+\pagebreak
 

### 2013-02-22 cover_letter.tex
Revise cover letter.

@@
 Since the publication of ``Mapping the Antigenic and Genetic Evolution of Influenza Virus'' by Smith et al.
-in 2004, there has been significant interest in characterizing patterns of antigenic and genetic evolution in seasonal influenza.
+in 2004, there has been significant interest in assessing patterns of antigenic and genetic evolution in seasonal influenza.
 Such cartographic approaches use multidimensional scaling (MDS) to position viruses and antisera on an antigenic map so that distances between viruses and antisera predict observed serological data.
 Here, we present a substantial advance to previous approaches that uses Bayesian multidimensional scaling (BMDS) to place antigenic cartography in a fully probabilistic framework.
-Doing so has allowed us to build in important covariates into the antigenic model; most importantly, phylogenetic relationships.
+Doing so has allowed us to build in important covariates into the statistical model; most importantly, phylogenetic relationships.
 Thus, our method simultaneously models evolutionary and antigenic dynamics, and allows us to infer which phylogeny branches have evolved in antigenic phenotype and which branches have remained static.
@@
 
-Both the link between incidence and antigenic drift and the presence of interference across influenza subtypes have remained large-standing questions in the influenza community.
+Both the link between incidence and antigenic drift and the presence of interference across influenza subtypes have remained long-standing questions in the influenza community.
 We believe our results do much to address these questions.
 Additionally, we believe our methodological advances will prove fundamental to future research in influenza and other antigenically variable pathogens.
-Our unified genetic / antigenic approach makes it possible to look closely for causal relationships between genetic mutations and antigenic change, and to characterize, for the first time, the antigenic mutations that are ultimately successful in the global competition among virus lineages.
+Our unified genetic / antigenic approach will make it possible to look closely for causal relationships between genetic mutations and antigenic change, and to characterize, for the first time, the antigenic mutations that are ultimately successful in the global competition among virus lineages.
 

### 2013-02-25 flux.tex
Improve title page formatting.
Revise paragraph on year-to-year drift.

@@
 \renewcommand\Affilfont{\small \normalfont}
+\makeatletter
+\renewcommand\AB@affilsepx{, \protect\Affilfont}
+\makeatother
 
@@
 \maketitle
-\pagebreak
 
 %%% ABSTRACT %%%
-\section*{Abstract}
+\begin{abstract}
 
@@
 This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+
+\end{abstract}
 
@@
 
-Interestingly, although antigenic clusters seem to emerge in a punctuated fashion in influenza A/H3N2, year-to-year antigenic drift in the mean antigenic phenotype of the virus population is relatively constant through time with most years exhibiting an intermediate level of drift (Figure~\ref{jumps}).
-We calculate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
-For A/H3N2, the largest cluster transition we observe is the emergence of the Sydney/97 cluster, evident as the appearance in 1997 of a large gap in virus locations along antigenic dimension 1 (Figure~\ref{drift}).
-However, in 1997 only three of the nine viruses sampled show the Sydney/97 phenotype, while the remaining six viruses exhibit the old Wuhan/95 phenotype.
-In 1998, three of the four viruses sampled show the Sydney/97 phenotype, with the remaining virus still showing the Wuhan/95 phenotype.
-In 1999, all sampled viruses exhibit the Sydney/97 phenotype.
-Thus, we observe fairly equal rates of year-to-year drift in mean antigenic phenotype between 1996 and 1999, rather than all drift in mean antigenic phenotype coinciding with the appearence of a new antigenic variant in 1997 (Figure~\ref{jumps}).
-It is interesting that this phenomenon occurs despite the fact that antigenic drift is accomplished, not by steady movement of the entire virus population, but by the successive emergence of drifted variants that sweep through the virus population.
-Because drifted variants require, on average, 2--3 years to take over the virus population, rates of year-to-year drift do not always spike heavily at the appearance of a new variant.
+We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}).
+We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
+There may often be large discontinuities in the virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
+We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift of 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
 

### 2013-02-25 flux.tex
Revise figure 3 to show incidence timeseries.

@@
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/jumps}
+	\includegraphics[width=1.0\textwidth]{figures/jumps}
 	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2009 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011.
-	Year-to-year antigenic drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
+	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011 and seasonal incidence from 1998/1999 to 2008/2009.
+	Colored lines represent year-to-year antigenic drift, and drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
 	For example, 2000 represents difference in antigenic dimension 1 between 1999 and 2000.
 	Error bars represent 50\% credible intervals across posterior samples of year-to-year drift.
+	Gray dotted lines represent clade-specific USA incidence taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a clade for each season.
+	Here, 2000 represents the 2000/2001 influenza season.
 	(B) Histograms of year-to-year antigenic drift across influenza clades summarizing observations from 1992 to 2011.

### 2013-02-25 flux.tex
Fix typos.

@@
 We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
-There may often be large discontinuities in the virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
-We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift of 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
+There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
+We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
 

### 2013-02-27 flux.tex
Small textual revision.
Implement revisions based on Derek's comments.

@@
 We arrive at a similar conclusion, finding that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
-We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}).
+We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}), in line with the results of Smith et al.\ \cite{Smith04}.
 Consequently, we specify a two dimensional model in all subsequent analyses.
@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
+Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 2 through 8; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
-Because of this, we interpret map locations locally rather than globally.
-We can determine rates of antigenic movement without knowing the larger configuration under which the movement occurs.
+Because of this, we interpret map locations locally rather than globally, and assess rates of antigenic movement without knowing the larger configuration under which the movement occurs.
 
@@
 Influenza B/Yamagata evolved slower still, with an average rate of 0.32 (HPD 0.25--0.39) units per year.
-Little punctuated evolution is obvious in the evolution of B/Yam.
+Although less punctuated evolution is apparent in B/Yam, we still observe three distinct clusters in the last decade, comprising the Shanghai/02, Florida/06 and Wisconsin/10 antigenic clusters (Figure~\ref{drift}).
 
@@
 We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
+This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
 

### 2013-02-27 flux_si.tex
Revert "Move methods to supplement."
This reverts commit e2bcc6cafec6b6e6fdc3e0e74602d2916fb67456.
Small cleanup.
Implement revisions based on Derek's comments.

--- 2013-01-24 flux_si.tex
@@
 This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements. 
+Each virus isolate was compared to an average of 21.9 serum isolates, and each serum isolate was compared to an average of 18.0 virus isolates.
 
@@
 This procedure yielded 115 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
+Each virus isolate was compared to an average of 10.0 serum isolates, and each serum isolate was compared to an average of 16.2 virus isolates.
 
@@
 Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
+Each virus isolate was compared to an average of 6.5 serum isolates, and each serum isolate was compared to an average of 16.7 virus isolates.
 
@@
 For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
+Each virus isolate was compared to an average of 6.9 serum isolates, and each serum isolate was compared to an average of 17.3 virus isolates.
 

### 2013-02-27 flux.tex
Add figure reference.

@@
 In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
-We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A).
+We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{jumps}, Figure~\ref{incidence}A).
 Here, within-clade drift explains 35\% of the variance in standardized incidence and represents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).

### 2013-03-09 flux.tex
Fix author contributions.

@@
 
-AR, MAS, TB and PL conceived the study.
+AR, MAS, TB conceived the study.
+AR, MAS, TB, DJS, PL designed the antigenic model.
 TB, GD, VG, JWM, AJH, CAR, DJS and AR gathered antigenic and genetic data.
-AR, MAS, TB and PL designed statistical phylogenetic procedures.
+AR, MAS, TB and PL implemented statistical phylogenetic procedures.
 TB performed the analysis.
+TB, AR, DJS, JWM, PL, CAR, AJH, GD, MAS interpreted the results.
 TB, MAS, AR, JWM, AJH wrote the paper.

### 2013-03-10 flux.tex
Fix funding acknowledgments.

@@
 
-TB is supported by the Royal Society. 
-MAS is supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
+TB was supported by a Newton International Fellowship from the Royal Society. 
+MAS was supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
 The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
@@
 260864.
-VG, AJH and JWM are funded by MRC reference no U117512723.
+VG, AJH and JWM were funded by MRC reference no U117512723.
 AR acknowledges the support of the Wellcome Trust (grant no. 092807).
@@
 DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant
-P0050/2008, Wellcome 087982AIA, and NIH Director's Pioneer Award DP1-OD000490-01.
+P0050/2008, Wellcome 087982AIA, and NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
 CAR was supported by a University Research Fellowship from the Royal Society.

### 2013-03-10 flux.tex
Fix author affiliations.

@@
 \author[6]{John W. McCauley}
-\author[7]{Colin A. Russell}
-\author[7,8]{Derek J. Smith}
-\author[1,9]{Andrew Rambaut}
+\author[7,8]{Colin A. Russell}
+\author[7,8,9]{Derek J. Smith}
+\author[1,10]{Andrew Rambaut}
 
@@
 \affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
-\affil[7]{Department of Zoology, University of Cambridge, Cambridge, UK}
-\affil[8]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
-\affil[9]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[7]{Centre for Pathogen Evolution, Department of Zoology, University of Cambridge, Cambridge, UK}
+\affil[8]{WHO Collaborating Center for Modeling, Evolution, and Control of Emerging Infectious Diseases, University of Cambridge, Cambridge, UK}
+\affil[9]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
+\affil[10]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
 
@@
 Collaboration between MAS, AR and PL was supported by the National Evolutionary Synthesis Center (NESCent), NSF EF-0423641.
-DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant
-P0050/2008, Wellcome 087982AIA, and NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
+DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant P0050/2008, Wellcome 087982AIA, NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
 CAR was supported by a University Research Fellowship from the Royal Society.

### 2013-03-10 flux.tex
Revise model introduction.

@@
 
+We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
+After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
+Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Methods).
+Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Methods) and seek to estimate model parameters through Bayesian inference and Markov chain Monte Carlo (MCMC).
+However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.
+
 We test model performance by constructing training datasets representing 90\% of the HI measurements for each of the four influenza lineages and test datasets representing the remaining 10\% of the measurements for each lineage.
 By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers.
-We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
-After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
-We arrive at a similar conclusion, finding that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
+We find that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
 We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}), in line with the results of Smith et al.\ \cite{Smith04}.
@@
 We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when including the phylogenetic diffusion (model 6; Table \ref{errortable}).
+Thus, antigenic locations are inferred using both genetic sequences and HI data and will differ from locations inferred from HI data alone.
+If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in influencing antigenic locations (as may be the case for B/Vic and B/Yam).
 
@@
 We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations.
+The term for across lineage interference $\bother$ is small in magnitude, but remains statistically significant (Table~\ref{interferencetable}).
 
@@
 In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
-\begin{equation}
+\begin{equation} \label{mds}
 	\sum_{(i,j) \in \cal I} 
@@
 We calculate the overall likelihood by multiplying probabilities of individual measurements
-\begin{equation} 
+\begin{equation}  \label{bmds}
 	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
@@
 AR, MAS, TB, DJS, PL designed the antigenic model.
-TB, GD, VG, JWM, AJH, CAR, DJS and AR gathered antigenic and genetic data.
-AR, MAS, TB and PL implemented statistical phylogenetic procedures.
+TB, GD, VG, JWM, AJH, CAR, DJS, AR gathered antigenic and genetic data.
+AR, MAS, TB, PL implemented statistical phylogenetic procedures.
 TB performed the analysis.

### 2013-03-10 flux.tex
Include H3 Brisbane/07 in map and drift figures.

@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Brisbane/07, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
 	\label{map} 
@@
 	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
+	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Brisbane/07, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
 	\label{drift} 

### 2013-03-13 flux.tex
Bayesian inference and MCMC are not the same thing

--- 2013-03-10 flux.tex
@@
 Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Methods).
-Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Methods) and seek to estimate model parameters through Bayesian inference and Markov chain Monte Carlo (MCMC).
+Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
 However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.
@@
 \end{equation}
-We sample from this posterior distribution using the Markov chain Monte Carlo (MCMC) procedures implemented in the software package BEAST \cite{BEAST17}.
+We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
 Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.

### 2013-03-15 flux.tex
Include drift-only model in training/test comparisons.

--- 2013-03-10 flux.tex
@@
 	5		&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
-	6		&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
-	7		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
-	8		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
+	6		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
+	7		&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
+	8		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
+	9		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
 	\hline
@@
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
-The antigenic diffusion process includes both systematic drift along the tree and covariance induced by phylogenetic proximity.
-We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process, and find a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when including the phylogenetic diffusion (model 6; Table \ref{errortable}).
-Thus, antigenic locations are inferred using both genetic sequences and HI data and will differ from locations inferred from HI data alone.
-If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in influencing antigenic locations (as may be the case for B/Vic and B/Yam).
+The antigenic diffusion process includes both systematic drift with time and covariance induced by phylogenetic proximity.
+We examine the effects of including only systematic drift (model 6; Table \ref{errortable}) and systematic drift plus phylogenetic diffusion (model 7; Table \ref{errortable}), finding a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when both processes are included.
+The phylogenetic diffusion requires genetic data for each virus isolate.
+Thus, in these models, antigenic locations are inferred using both genetic sequences and HI data and will differ from locations inferred from HI data alone.
+If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in determining antigenic locations (as may be the case for B/Vic and B/Yam).
 
@@
 We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error further, ranging in improvement from 0.03 to 0.13 log$_2$ HI titers (models 7 and 8; Table \ref{errortable}).
+We find that including these effects decreases test error further, ranging in improvement from 0.03 to 0.13 log$_2$ HI titers (models 8 and 9; Table \ref{errortable}).
 

### 2013-03-15 flux.tex
Include new analysis comparing Smith locations to BMDS locations.

@@
 We examine the effects of including only systematic drift (model 6; Table \ref{errortable}) and systematic drift plus phylogenetic diffusion (model 7; Table \ref{errortable}), finding a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when both processes are included.
-The phylogenetic diffusion requires genetic data for each virus isolate.
-Thus, in these models, antigenic locations are inferred using both genetic sequences and HI data and will differ from locations inferred from HI data alone.
+The systematic drift process informs virus and serum locations by dates of isolation and the phylogenetic diffusion process informs virus locations by genetic sequences.
+Thus, in these models, antigenic locations are inferred using both genetic data and HI data and will differ from locations inferred from HI data alone.
 If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in determining antigenic locations (as may be the case for B/Vic and B/Yam).
@@
 We estimate that virus distances have, on average, a 50\% credible interval of $\pm0.45$ antigenic units for A/H3N2, $\pm0.57$ units for A/H1N1, $\pm0.76$ units for B/Vic, and $\pm0.65$ units for B/Yam.
+We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (see Supp.\ Discussion).
 
@@
 
-HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 2 through 8; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
-Because of this, we interpret map locations locally rather than globally, and assess rates of antigenic movement without knowing the larger configuration under which the movement occurs.
+HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 14 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
+Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory (see Supp.\ Discussion).
+To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
+Because of this, we interpret map locations locally rather than globally, and assess rates of antigenic movement without making strong statements about the larger configuration under which the movement occurs.
 
@@
 is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
+In the case of threshold measurements, this error function is modified slightly; see \cite{Smith04} for further details.
 
@@
 
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications and WHO vaccine strain selection reports (see Supp.\ Text).
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications and WHO vaccine strain selection reports (see Supp.\ Methods).
 Hemagglutinin sequences were compiled from the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID}.
@@
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
+In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same posterior distribution from different starting points.
+However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution (see Supp.\ Discussion).
+We favor models that include $\beta$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
 

### 2013-03-15 flux_si.tex
Include new analysis comparing Smith locations to BMDS locations.

@@
 \newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix
+
+\setcounter{figure}{2}
+\setcounter{table}{0}
+\setcounter{page}{1}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thepage}{S\arabic{page}}
 
@@
 
+%%% SUPPORTING DISCUSSION %%%
+\section*{Supporting Discussion}
+
+\subsection*{Comparison with previous results}
+
+Here, we attempt to compare antigenic locations inferred by our BMDS model to antigenic locations previously inferred by the error minimization methods of Smith et al.\ \cite{Smith04}, referred to here as antigenic cartography by MDS.
+For this comparison, we use exactly the same HI data used to produce the results in \cite{Smith04}, consisting of 273 virus isolates, 79 serum isolates and a total of 4252 HI measurements taken between 1968 and 2003.
+We begin with a BMDS analog of the antigenic model used in \cite{Smith04}, where serum effects are taken as the maximum titer of a particular ferret serum and the expected log$_2$ drop in HI titer is proportional to Euclidean distance between virus and serum locations.
+To bring models into further alignment, we use a Uniform$(-100,100)$ distribution over virus locations and serum locations.
+Unsurprisingly, we find that this BMDS model produces results that are strongly congruent with MDS results (Figure~\ref{smith_comparison}).
+Antigenic cluster locations are consistent between methods (Figure~\ref{smith_comparison}A--B) and antigenic distances between pairs of viruses are consistent between temporally similar and temporally divergent viruses (Figure~\ref{smith_comparison}C--E), suggesting that the resulting maps are consistent at both local and global scales.
+Credible intervals of antegenic distances for the BMDS model remain narrow across the temporal spectrum (Figure~\ref{smith_comparison}C--E), implying a fair degree of rigidity to the map.
+
+%%% smith_comparison %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from an equivalent BMDS model.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 
+	} 
+	\label{smith_comparison} 
+\end{figure}
+
+Smith et al.\ \cite{Smith04} show that there exists two global solutions in their assignment of antigenic locations, involving the rotation of clusters HK/68, EN/72 and VI/75 (shown in Figure S2 of \cite{Smith04}).
+We observe the same metastable behavior in our analysis; some MCMC chains converge on the solution shown in Figure \ref{smith_comparison}B, while other MCMC chains converge on the alternative solution shown in Figure \ref{smith_comparison_rotation}B.
+The distribution of likelihood values appear highly similar between these two solutions.
+The rotation of the HK/68, EN/72 and VI/75 clusters creates a map that bends slightly, so that temporally distant viruses appear closer in the rotated solution than in the original solution (Figure~\ref{smith_comparison_rotation}C--E).
+In this case, it is clear that the solutions are locally consistent between viruses up to $\sim$15 years divergent, even if there some degree of global flexibility.
+
+%%% smith_comparison_rotation %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_rotation}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model, showing an alternative solution found by MCMC.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from an equivalent BMDS model that has converged on the alternative solution.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	} 
+	\label{smith_comparison_rotation} 
+\end{figure}
+
+As discussed in the main text, the presence of multiple solutions with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
+Because of this issue, and to more easily estimate rates of antigenic drift, we included a model of systematic drift in antigenic location (discussed in Methods) that favors linear movement in the antigenic map.
+We find that including this drift prior on antigenic locations removes the problem of identifiability.
+Antigenic locations produced by this model remain locally consistent with MDS results between viruses $\sim$15 years divergent, but global comparisons show that this BMDS model has partitioned more variance to the first antigenic dimension (Figure~\ref{smith_comparison_drift}).
+We additionally find that including the drift prior on antigenic locations often results in greater predictive power, with a slight improvement of test error for the A/H1N1, B/Vic and B/Yam datasets (Table~1).
+
+%%% smith_comparison_drift %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_drift}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from a BMDS model that includes date-informed priors on antigenic locations.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a date-informed prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	} 
+	\label{smith_comparison_drift} 
+\end{figure}
+
+Our final BMDS model (model 9, Table 1) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects (discussed in Methods).
+Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.
+To isolate this difference, we use a Uniform$(-100,100)$ prior on antigenic locations.
+Surprisingly, estimating virus and serum effects results in a more linear antigenic map (Figure~\ref{smith_comparison_effects}), resembling the appearance of the map incorporating the antigenic drift prior, while preserving local consistency.
+We generally observe congruence between MDS and BMDS antigenic locations for viruses less than $\sim$10 years divergent (Figure~\ref{smith_comparison_effects}E).
+However, specific viruses may be affected, for instance A/Bilthoven/16190/1968 (Figure~\ref{smith_comparison_effects}C), which appears more distant from all other viruses when serum and virus effects are included.
+
+%%% smith_comparison_effects %%%
+\begin{figure}[H]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_effects}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus and serum effects.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and virus and serum effects estimated in a hierarchical Bayesian fashion. 	
+	} 
+	\label{smith_comparison_effects} 
+\end{figure}
+
+In this dataset, viruses 15 or more years divergent always yield threshold titers, and hence, their relative locations must be indirectly inferred rather than through direct comparison.
+This may explain why we observe local consistency between models at scales less than $\sim$15 years, but some degree of global inconsistency.
+Still, these results suggest that, when making local comparisons, such as those used to calculcate year-to-year antigenic drift (Figure 3), outcomes are expected to be robust to  many model particulars.
+
 %%% REFERENCES %%%

### 2013-03-15 flux.tex
Revise text.

@@
 We estimate that virus distances have, on average, a 50\% credible interval of $\pm0.45$ antigenic units for A/H3N2, $\pm0.57$ units for A/H1N1, $\pm0.76$ units for B/Vic, and $\pm0.65$ units for B/Yam.
+
 We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (see Supp.\ Discussion).
+Differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
+If a single map is desired, then the MDS method may be preferred, while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.
 
@@
 However, local isometries are often still a problem.
-For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
-These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Victoria/75 in place.
-Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
+For example, in A/H3N2 the HK/68, EN/72 and VI/75 clusters may rotate in relation to other clusters (see Supp.\ Discussion).
+Consequently, it may be difficult to fully align MCMC samples using Procrustes analysis.
 For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
+Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate the quantity across MCMC samples to yield an expectation and a credible interval.
 This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.

### 2013-03-15 flux_si.tex
Working edits; TB should confirm that he agrees with changes to SI.

@@
 
-Smith et al.\ \cite{Smith04} show that there exists two global solutions in their assignment of antigenic locations, involving the rotation of clusters HK/68, EN/72 and VI/75 (shown in Figure S2 of \cite{Smith04}).
+Smith et al.\ \cite{Smith04} show that there exist at least two solutions in their assignment of antigenic locations, involving the rotation of clusters HK/68, EN/72 and VI/75 (shown in Figure S2 of \cite{Smith04}).
 We observe the same metastable behavior in our analysis; some MCMC chains converge on the solution shown in Figure \ref{smith_comparison}B, while other MCMC chains converge on the alternative solution shown in Figure \ref{smith_comparison_rotation}B.
-The distribution of likelihood values appear highly similar between these two solutions.
+The distribution of likelihood values appear highly similar between these two solutions, suggesting that they represent global optima.
 The rotation of the HK/68, EN/72 and VI/75 clusters creates a map that bends slightly, so that temporally distant viruses appear closer in the rotated solution than in the original solution (Figure~\ref{smith_comparison_rotation}C--E).
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_rotation}
-	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model, showing an alternative solution found by MCMC.} 
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
 	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
@@
 
-As discussed in the main text, the presence of multiple solutions with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
-Because of this issue, and to more easily estimate rates of antigenic drift, we included a model of systematic drift in antigenic location (discussed in Methods) that favors linear movement in the antigenic map.
+As discussed in the main text, the presence of multiple optima with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
+Because of this issue, and to more easily estimate rates of antigenic drift, we include a model of systematic drift in antigenic location (discussed in Methods) that favors linear movement in the antigenic map.
 We find that including this drift prior on antigenic locations removes the problem of identifiability.

### 2013-03-15 flux.tex
There is only one posterior distribution; but there can be many estimates of it

@@
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
-In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same posterior distribution from different starting points.
+In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
 However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution (see Supp.\ Discussion).

### 2013-03-17 flux.tex
Add references for flu B clusters.

@@
 Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.42 (HPD 0.32--0.51) units per year.
-Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
-Interestingly, a minor lineage of B/Vic has persisted through 2011, appearing antigenically distinct from other 2011 viruses (Figure~\ref{drift}), though the eventual fate of this lineage has yet to be determined.
 Influenza B/Yamagata evolved slower still, with an average rate of 0.32 (HPD 0.25--0.39) units per year.
-Although less punctuated evolution is apparent in B/Yam, we still observe three distinct clusters in the last decade, comprising the Shanghai/02, Florida/06 and Wisconsin/10 antigenic clusters (Figure~\ref{drift}).
+Punctuated evolution is less obvious in B/Yam and B/Vic compared to A/H3N2 and A/H1N1, but antigenic clusters are still apparent, with recent transitions to the Brisbane/08 cluster in B/Vic \cite{Barr10} and to the Wisconsin/10 cluster in B/Yam \cite{Klimov12}.
+Interestingly, a minor lineage of B/Vic, denoted B/Hubei-Songzi/51/2008 \cite{Barr10}, has persisted through 2011, while remaining antigenically distinct from B/Brisbane/60/2008 viruses (Figure~\ref{drift}).
 

### 2013-03-25 flux.tex
Include 100-word statement of significance for PNAS.

@@
 \pagebreak
+
+%%% SIGNIFICANCE %%%
+\section*{Significance}
+
+Influenza viruses continually evolve in antigenic phenotype to escape human immunity.
+In this study, we develop a new method to characterize influenza antigenic phenotype from immunological and genetic data.
+We use this method to assess patterns of antigenic evolution in four different influenza viruses and show that antigenic drift strongly influences viral success.
+Drifted viruses cause larger seasonal outbreaks and outcompete less evolved viruses.
+A more thorough understanding of the predictors of viral success may allow selection of vaccine strains better matched to the population of influenza viruses circulating when the vaccine is deployed.
 

### 2013-03-25 cover_letter.tex
Fix typo.

@@
 
-Since the publication of ``Mapping the Antigenic and Genetic Evolution of Influenza Virus'' by Smith et al.
-in 2004, there has been significant interest in assessing patterns of antigenic and genetic evolution in seasonal influenza.
+Since the publication of ``Mapping the Antigenic and Genetic Evolution of Influenza Virus'' by Smith et al.\ in 2004, there has been significant interest in assessing patterns of antigenic and genetic evolution in seasonal influenza.
 Such cartographic approaches use multidimensional scaling (MDS) to position viruses and antisera on an antigenic map so that distances between viruses and antisera predict observed serological data.

### 2013-04-01 flux.tex
Reformat for the arXiv, condensing main text and supplement.

--- 2013-03-17 flux.tex
@@
 
-We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (see Supp.\ Discussion).
+%%% error_by_distance_and_year %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/error_by_distance_and_year}
+	\caption{\textbf{Absolute error in log$_2$ HI titer against antigenic distance and date of virus sampling in influenza A/H3N2.} 
+	(A) Points show the absolute deviation from the predicted log$_2$ HI titer of the antigenic model, computed from virus location $\virus_i$, serum location $\serum_j$, virus effect $\ve_i$ and serum effect $\se_j$, based upon antigenic distance between virus and serum $\delta_{ij}$.
+	The red line shows a LOESS regression curve of these points.
+	(B) Points show the absolute deviation from the predicted log$_2$ HI titers of the antigenic model, based upon the date of virus sampling.
+	The red line shows average deviations for each year.
+	Relationships for A/H1N1, B/Vic and B/Yam are similar.
+	} 
+	\label{error_by_distance_and_year} 
+\end{figure}
+
+We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (Figures~\ref{smith_comparison}--\ref{smith_comparison_effects}).
 Differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 14 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory (see Supp.\ Discussion).
+Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory.
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
@@
 	For example, 2000 represents difference in antigenic dimension 1 between 1999 and 2000.
-	Error bars represent 50\% credible intervals across posterior samples of year-to-year drift.
+	Error bars represent 50\% Bayesian credible intervals of year-to-year drift.
 	Gray dotted lines represent clade-specific USA incidence taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a clade for each season.
@@
 Here, within-clade drift explains 35\% of the variance in standardized incidence and represents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
+
+%%% incidence %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{figures/incidence}
+	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
+	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
+	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
+	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
+	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
+	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
+	} 
+	\label{incidence} 
+\end{figure}
 
@@
 Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Supp.\ Methods).
+To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Methods).
 This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
@@
 
-\subsection*{Posterior samples}
+\subsection*{Posterior inference}
 
@@
 
-\subsection*{Data and implementation}
+\subsection*{Genetic, antigenic and surveillance data}
 
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications and WHO vaccine strain selection reports (see Supp.\ Methods).
-Hemagglutinin sequences were compiled from the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID}.
-Antigenic and genetic data were subsampled to give similar quantities of data across years of the analysis.
-Data gathering procedures resulted in 402 virus strains A/H3N2, 115 virus strains for A/H1N1, 179 virus strains for B/Vic and 174 virus strains for B/Yam.
+We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
+We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
+If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
+Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
+This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
+However, the majority of isolates were present from 2002 to 2007. 
+Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
+We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
+This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements. 
+Each virus isolate was compared to an average of 21.9 serum isolates, and each serum isolate was compared to an average of 18.0 virus isolates.
+
+Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
+The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
+This procedure yielded 115 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
+Each virus isolate was compared to an average of 10.0 serum isolates, and each serum isolate was compared to an average of 16.2 virus isolates.
+
+Antigenic comparisons for influenza B/Victoria were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
+Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
+Each virus isolate was compared to an average of 6.5 serum isolates, and each serum isolate was compared to an average of 16.7 virus isolates.
+
+Antigenic comparisons for influenza B/Yamagata were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
+For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
+Each virus isolate was compared to an average of 6.9 serum isolates, and each serum isolate was compared to an average of 17.3 virus isolates.
+
 Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
+As an example, one report states ``collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
+Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses.
+Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
+In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
+Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
+
+\subsection*{Implementation}
 
@@
 In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
-However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution (see Supp.\ Discussion).
+However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
 We favor models that include $\beta$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
@@
 However, local isometries are often still a problem.
-For example, in A/H3N2 the HK/68, EN/72 and VI/75 clusters may rotate in relation to other clusters (see Supp.\ Discussion).
+For example, in A/H3N2 the HK/68, EN/72 and VI/75 clusters may rotate in relation to other clusters.
 Consequently, it may be difficult to fully align MCMC samples using Procrustes analysis.
@@
 
+\subsection*{Estimation of diffusion coefficient}
+
+The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
+Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
+\begin{equation}
+	X_t \sim \normal(0, \sigma^2 t),
+\end{equation}
+where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
+\begin{equation}
+	\mathrm{E}[X_t^2] = \sigma^2 t.
+\end{equation}
+In moving to a two-dimensional diffusion
+\begin{equation}
+	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
+\end{equation}
+where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
+\begin{equation}
+	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
+\end{equation}
+Such diffusions are often characterized with diffusion coefficient $D$.
+In this case
+\begin{equation}
+	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
+\end{equation}
+so that expected squared displacement follows from $D$ alone
+\begin{equation}
+	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
+\end{equation}
+
+Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
+\begin{equation} \label{pybusest}
+	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
+\end{equation}
+in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
+However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
+Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
+\begin{eqnarray} \label{ourest}
+	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
+	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
+	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
+\end{eqnarray}
+This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
+Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
+When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
+
+\subsection*{Comparison with previous results}
+
+Here, we attempt to compare antigenic locations inferred by our BMDS model to antigenic locations previously inferred by the error minimization methods of Smith et al.\ \cite{Smith04}, referred to here as antigenic cartography by MDS.
+For this comparison, we use exactly the same HI data used to produce the results in \cite{Smith04}, consisting of 273 virus isolates, 79 serum isolates and a total of 4252 HI measurements taken between 1968 and 2003.
+We begin with a BMDS analog of the antigenic model used in \cite{Smith04}, where serum effects are taken as the maximum titer of a particular ferret serum and the expected log$_2$ drop in HI titer is proportional to Euclidean distance between virus and serum locations.
+To bring models into further alignment, we use a Uniform$(-100,100)$ distribution over virus locations and serum locations.
+Unsurprisingly, we find that this BMDS model produces results that are strongly congruent with MDS results (Figure~\ref{smith_comparison}).
+Antigenic cluster locations are consistent between methods (Figure~\ref{smith_comparison}A--B) and antigenic distances between pairs of viruses are consistent between temporally similar and temporally divergent viruses (Figure~\ref{smith_comparison}C--E), suggesting that the resulting maps are consistent at both local and global scales.
+Credible intervals of antegenic distances for the BMDS model remain narrow across the temporal spectrum (Figure~\ref{smith_comparison}C--E), implying a fair degree of rigidity to the map.
+
+%%% smith_comparison %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from an equivalent BMDS model.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 
+	} 
+	\label{smith_comparison} 
+\end{figure}
+
+Smith et al.\ \cite{Smith04} show that there exist at least two solutions in their assignment of antigenic locations, involving the rotation of clusters HK/68, EN/72 and VI/75 (shown in Figure S2 of \cite{Smith04}).
+We observe the same metastable behavior in our analysis; some MCMC chains converge on the solution shown in Figure \ref{smith_comparison}B, while other MCMC chains converge on the alternative solution shown in Figure \ref{smith_comparison_rotation}B.
+The distribution of likelihood values appear highly similar between these two solutions, suggesting that they represent global optima.
+The rotation of the HK/68, EN/72 and VI/75 clusters creates a map that bends slightly, so that temporally distant viruses appear closer in the rotated solution than in the original solution (Figure~\ref{smith_comparison_rotation}C--E).
+In this case, it is clear that the solutions are locally consistent between viruses up to $\sim$15 years divergent, even if there some degree of global flexibility.
+
+%%% smith_comparison_rotation %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_rotation}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from an equivalent BMDS model that has converged on the alternative solution.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	} 
+	\label{smith_comparison_rotation} 
+\end{figure}
+
+As discussed in the main text, the presence of multiple optima with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
+Because of this issue, and to more easily estimate rates of antigenic drift, we include a model of systematic drift in antigenic location (discussed in Methods) that favors linear movement in the antigenic map.
+We find that including this drift prior on antigenic locations removes the problem of identifiability.
+Antigenic locations produced by this model remain locally consistent with MDS results between viruses $\sim$15 years divergent, but global comparisons show that this BMDS model has partitioned more variance to the first antigenic dimension (Figure~\ref{smith_comparison_drift}).
+We additionally find that including the drift prior on antigenic locations often results in greater predictive power, with a slight improvement of test error for the A/H1N1, B/Vic and B/Yam datasets (Table~\ref{errortable}).
+
+%%% smith_comparison_drift %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_drift}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from a BMDS model that includes date-informed priors on antigenic locations.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a date-informed prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	} 
+	\label{smith_comparison_drift} 
+\end{figure}
+
+Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects (discussed in Methods).
+Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.
+To isolate this difference, we use a Uniform$(-100,100)$ prior on antigenic locations.
+Surprisingly, estimating virus and serum effects results in a more linear antigenic map (Figure~\ref{smith_comparison_effects}), resembling the appearance of the map incorporating the antigenic drift prior, while preserving local consistency.
+We generally observe congruence between MDS and BMDS antigenic locations for viruses less than $\sim$10 years divergent (Figure~\ref{smith_comparison_effects}E).
+However, specific viruses may be affected, for instance A/Bilthoven/16190/1968 (Figure~\ref{smith_comparison_effects}C), which appears more distant from all other viruses when serum and virus effects are included.
+
+%%% smith_comparison_effects %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_effects}
+	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
+	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus and serum effects.
+	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
+	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
+	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
+	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
+	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and virus and serum effects estimated in a hierarchical Bayesian fashion. 	
+	} 
+	\label{smith_comparison_effects} 
+\end{figure}
+
+In this dataset, viruses 15 or more years divergent always yield threshold titers, and hence, their relative locations must be indirectly inferred rather than through direct comparison.
+This may explain why we observe local consistency between models at scales less than $\sim$15 years, but some degree of global inconsistency.
+Still, these results suggest that, when making local comparisons, such as those used to calculcate year-to-year antigenic drift (Figure~\ref{drift}), outcomes are expected to be robust to  many model particulars.
+
 \subsection*{Availability}
@@
 Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Antigenic, genetic and surveillance data is available as supplemental information.
 
@@
 
-\pagebreak
-
-\setcounter{figure}{0}
-\setcounter{table}{0}
-%\setcounter{page}{1}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-%\renewcommand{\thepage}{S\arabic{page}}
-
-%%% SUPPORTING INFORMATION %%%
-\section*{Supporting Information}
-
-%%% error_by_distance_and_year %%%
-\begin{figure}[H]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/error_by_distance_and_year}
-	\caption{\textbf{Absolute error in log$_2$ HI titer against antigenic distance and date of virus sampling in influenza A/H3N2.} 
-	(A) Points show the absolute deviation from the predicted log$_2$ HI titer of the antigenic model, computed from virus location $\virus_i$, serum location $\serum_j$, virus effect $\ve_i$ and serum effect $\se_j$, based upon antigenic distance between virus and serum $\delta_{ij}$.
-	The red line shows a LOESS regression curve of these points.
-	(B) Points show the absolute deviation from the predicted log$_2$ HI titers of the antigenic model, based upon the date of virus sampling.
-	The red line shows average deviations for each year.
-	Relationships for A/H1N1, B/Vic and B/Yam are similar.
-	} 
-	\label{error_by_distance_and_year} 
-\end{figure}
-
-%%% incidence %%%
-\begin{figure}[H]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/incidence}
-	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
-	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
-	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
-	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
-	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
-	} 
-	\label{incidence} 
-\end{figure}
-
 \end{document}

### 2013-04-02 flux_si.tex
Remove comment functions.
Fix typo.

@@
 \makeatother
-
-% comments
-\usepackage{ulem}
-\definecolor{blue}{rgb}{0.324,0.609,0.708}
-\definecolor{purple}{rgb}{0.459,0.109,0.538}
-\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
-\def\tbc#1{\textcolor{purple}{[#1]}}
-\def\ms#1#2{\sout{#1} \textcolor{blue}{#2}} 
-\def\msc#1{\textcolor{blue}{[#1]}}
 
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison}
-	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
+	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
 	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_rotation}
-	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
+	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
 	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_drift}
-	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
+	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
 	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_effects}
-	\caption{\textbf{Comparison of A/H3N3 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
+	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
 	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.

### 2013-04-04 flux.tex
Change section title to "Materials and methods".

@@
 %%% RESULTS %%%
-\section*{Results and Discussion}
+\section*{Results and discussion}
 
@@
 
-In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Methods).
+In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Materials and methods).
 In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
@@
 We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).
+We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Materials and methods).
 
@@
 After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
-Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Methods).
-Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
+Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Materials and methods).
+Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Materials and methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
 However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.
@@
 Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01, Smith04, Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Methods).
+Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
@@
 Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory.
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
+To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Materials and methods).
 Because of this, we interpret map locations locally rather than globally, and assess rates of antigenic movement without making strong statements about the larger configuration under which the movement occurs.
@@
 We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
-We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Methods).
+We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Materials and methods).
 We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of viruses detected in the USA during this time period.
@@
 Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Methods).
+To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Materials and methods).
 This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
@@
 %%% METHODS %%%
-\section*{Methods}
+\section*{Materials and methods}
 
@@
 As discussed in the main text, the presence of multiple optima with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
-Because of this issue, and to more easily estimate rates of antigenic drift, we include a model of systematic drift in antigenic location (discussed in Methods) that favors linear movement in the antigenic map.
+Because of this issue, and to more easily estimate rates of antigenic drift, we include a model of systematic drift in antigenic location that favors linear movement in the antigenic map.
 We find that including this drift prior on antigenic locations removes the problem of identifiability.
@@
 
-Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects (discussed in Methods).
+Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects.
 Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.

### 2013-04-04 cover_letter.tex
Address eLife rather than PNAS.

@@
 
-\greetto{Dear Prof.\ Levin,}
+\greetto{Dear Editor,}
 \closeline{Sincerely,}
@@
 
-Please find, attached, our manuscript entitled ``Integrating influenza antigenic dynamics with molecular evolution,'' which we would be grateful for your consideration for publication in \textit{Proc Natl Acad Sci USA}.
-We believe that \textit{Proc Natl Acad Sci USA} is an appropriate venue for this work, as it clarifies long-standing questions about the antigenic and evolutionary dynamics of influenza viruses, while simultaneously presenting a major methodological advance in the characterization of antigenic phenotype.
+Please find, attached, our manuscript entitled ``Integrating influenza antigenic dynamics with molecular evolution,'' which we would be grateful for your consideration for publication in \textit{eLife}.
+We believe that \textit{eLife} is an appropriate venue for this work, as it clarifies long-standing questions about the antigenic and evolutionary dynamics of influenza viruses, while simultaneously presenting a major methodological advance in the characterization of antigenic phenotype.
 

### 2013-04-12 flux.tex
Shorten abstract for eLife requirements.

@@
 
-Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
+Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade host immunity acquired to previous virus strains.
 Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
-Here, we extend previous approaches to antigenic cartography, which seeks to place strains on an antigenic map, such that distances on this map best recapitulate titers observed across multiple HI assays.
-In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
-Using HI data for four lineages of influenza, encompassing influenza A subtypes H3N2 and H1N1, and influenza B lineages Victoria and Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
-Through comparison with epidemiological data, we demonstrate a year-to-year correlation between drift and incidence and present evidence that antigenic drift mediates interference between influenza lineages.
-We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences.
-This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+Here, we extend previous approaches to antigenic cartography, and simultaneously characterize antigenic and genetic evolution by modeling the diffusion of antigenic phenotype over a shared virus phylogeny.
+Using HI data from influenza A subtypes H3N2 and H1N1 and influenza B lineages Victoria and Yamagata, we determine patterns of antigenic drift across lineages and show that antigenic drift drives incidence rates and mediates interference between influenza lineages.
+Additionally, we describe the selective underpinnings of differences in antigenic drift across lineages.
+This work makes possible substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
 

### 2013-04-12 flux.tex
Include Impact statement for eLife's requirements.
Fix typo.

@@
 \end{abstract}
+
+%%% IMPACT %%%
+% Combined evolutionary and antigenic analysis shows that human influenza viruses differ dramatically in rates of antigenic drift and these rates significantly impact seasonal incidence patterns.
 
@@
 Collaboration between MAS, AR and PL was supported by the National Evolutionary Synthesis Center (NESCent), NSF EF-0423641.
-DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant P0050/2008, Wellcome 087982AIA, NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
+DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant P0050/2008, Wellcome Trust 087982AIA, NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
 CAR was supported by a University Research Fellowship from the Royal Society.

### 2013-05-14 response.tex
Include beginnings of response letter.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TITLE %%%
\title{\vspace{1.0cm} \Large \bf
Reviewer responses for\\
Integrating influenza antigenic dynamics with molecular evolution
}

\author[1]{Trevor Bedford}
\author[2,3,4]{Marc A. Suchard}
\author[5]{Philippe Lemey}
\author[1]{Gytis Dudas}
\author[6]{Victoria Gregory}
\author[6]{Alan J. Hay}
\author[6]{John W. McCauley}
\author[7,8]{Colin A. Russell}
\author[7,8,9]{Derek J. Smith}
\author[1,10]{Andrew Rambaut}

\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
\affil[4]{Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles CA, USA}
\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
\affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
\affil[7]{Centre for Pathogen Evolution, Department of Zoology, University of Cambridge, Cambridge, UK}
\affil[8]{WHO Collaborating Center for Modeling, Evolution, and Control of Emerging Infectious Diseases, University of Cambridge, Cambridge, UK}
\affil[9]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
\affil[10]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}

\date{}

\begin{document}

\maketitle

%%% REVIEWER 1 %%%
\section*{Reviewer 1}

\textsl{This paper is exceptional in many ways, but also contains some flaws that need to be addressed by revisions. If these flaws are addressed, I would be highly supportive of publication. This paper is complex, so I have not been able put all of my major critiques in the first 500 words. Instead, I summarize my major critiques here, and then further detail them along with the minor points.}

\textsl{This paper deals with the antigenic evolution of influenza. A foundational paper in Science in 2004 by Smith et al introduced ``antigenic cartography,'' the technique used to monitor antigenic drift and select the influenza vaccine. This paper makes a clear methodological improvement on the original antigenic cartography technique. The new approach integrates phylogenetics, allowing inclusion of viral sequences and isolation dates. This paper also shows that accounting for virus and serum effects improves cartography. As the authors briefly note, the new method is also extensible to evolutionary questions. All of these improvements are made using a freely available phylogenetic software program that is the standard in the field. These methodological improvements would immediately merit publication in a technique-specific journal.}

\textsl{In addition, this paper presents biological results that should interest virologists, public health officials, and evolutionary biologists. While these results do not really alter the current understanding of influenza antigenic evolution, they extend this understanding from H3N2 to other clades, provide a more detailed analysis than has previously been performed for any clade, and quantify factors that have previously been characterized only qualitatively.}

\textsl{I do, however, have some major critiques:}

\textsl{* It is difficult to understand the biological results without grasping technical aspects of the computational approach. In my specific comments, I suggest some changes.}

\textsl{* The authors need to explain where they got the incidence data, how they chose the date ranges, specify when they are using relative versus absolute incidence, and provide raw data as supporting material. I describe my concerns in more detail in the specific comments.}

\textsl{* The improvement from inclusion of serum effects deserves comment and explanation if possible (are the effects due to the serum source)? Table 1 should include a model with virus effects but no serum effects. I describe my concerns in more detail in the specific comments.}

\textsl{MAJOR POINTS IN MORE DETAIL:}

\textsl{* I have concerns about Figure 5, which the authors use to support their claim that ``antigenic drift drives incidence rates.'' The authors first establish the well-known fact that H3N2 incidence is highest, followed by H1N1 and then B. They also find that relative incidence is correlated with the rate of antigenic drift. However, this does not imply causality in terms of the higher average drift driving the higher average incidence. The authors do not claim any causality in this correlation between average drift and incidence in the four clades, and I am therefore fine with this part of the analysis. The problems arise in Figure 5A. The authors find a significant correlation between drift and incidence taken over all years and clades, and then use this to argue that drift drives incidence. I would agree with this conclusion if higher drift was associated with higher incidence with a correlation that is better than that obtained simply by comparing the averages of the four clades. But I don't think this is true. In Figure 5A, the authors are averaging over all clades, making the correlation in 5A a trivial consequence of the fact that H3N2 has higher average incidence and drift. Even if the rate of drift and incidence were constant year-over-year for each clade (and simply differed among clades), the authors would observe the correlation that they report. Put another way, imagine that every year the drift and incidence in each clade was equal to its average. Then Figure 5A would in effect correspond to simply recalculating the correlation between average drift and incidence in a scenario where each point is included as many times as there are years being examined. The more years that were included the higher the P-value would get because the number of data points would increase, but really the single actual trend (that H3N2 is higher in average drift and incidence) is just being counted multiple times. In order to draw a conclusion stronger that drift drives incidence, the authors need to show that incidence and drift are correlated within each clade individually, or else standardize the incidence so that the mean and variance for each individual clade is zero and one. It is not clear to me that there is any correlation between the drift and incidence within any of the individual clades. A similar problem affects Figure 5B - for example, since H3N2 has the highest drift, it will always tend to be associated with lower drift in other clades since the other clades inherently have lower average drift. But this does not imply direct interference between clades. I am open to further discussion about this point, but I do not see how Figure 5 implies that antigenic drift drives incidence, or that there is dynamical interference across clades. By the way, I do not think that a lack of correlation would be a disqualifying concern in terms of publication. Finding no evidence that antigenic drift drives incidence is just as important as finding that it does. But the claim needs to be correct.}

\textsl{* The incidence data are poorly explained. In the first paragraph of the Epidemiological consequences of antigenic drift section, the authors use relative incidence. In the next paragraph, are they switching to absolute incidence? This is not clear. If they are switching to absolute incidence, I don't think they ever explain in the Methods where that data comes from. In either case, what are the justifications for using absolute versus relative incidence? Also, why do they not analyze incidence prior to 1998/1999? Finally, because each influenza season spans two calendar years, why do they not break up the data by season rather than year?}

\textsl{* An interesting result is that estimating the serum effects improves the performance a great deal (more than including the phylogenetic data, in fact!). It suggests that there may be systematic differences between the sera. Can this be attributed to the serum itself (such as whether it is goat versus ferret), or to the particular lab or study? Could the authors correlate the estimated serum effects back to the originating source for that serum to check? As a related point, I was surprised that inclusion of the virus effects led to only minor improvement, since prior work (http://www.ncbi.nlm.nih.gov/pubmed/19900932) suggests that some viruses are generally more resistant to HI than others. However, the virus effects are only included in a model (\#9 in Table 1) where the serum effects are already included, and it is possible to imagine that including just one of these terms somehow helps correct for both effects. Can the authors include a tenth model in Table 1 that includes the virus effects but not the serum effects? This would clarify if serum effects are actually more important than virus effects, or simply whether inclusion of either one individually is more helpful than both together.}

IMPORTANT BUT NOT SCIENTIFICALLY SUBSTANTIVE POINTS:

\textsl{* The authors should be commended for making publicly available all of the HI data that they have assembled (a good scientific practice that sadly has not been adhered to in some previous publications on antigenic cartography). However, they should mention in the text the availability of these data on datadryad - I almost missed it since it is only mentioned on the cover page. They could also consider more clearly specifying where the code relevant to this paper is found in the BEAST repository.}

\textsl{* The one exception to the aforementioned praise on data availability is the influenza incidence data - I could not find any tables providing these data. They need to be provided.}

\textsl{* The authors need to define their nomenclature regarding clade / lineage / type / strain. For instance, they say: ``We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations. The term for across lineage Bt is small in magnitude.'' I had to read this many times before concluding that type refers to A or B, clade to refers to any of the four classifications. I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.}

\textsl{* The comparison in Table 1 is crucial, but the model descriptions are very hard to understand until after reading the Methods carefully. The difficulty is because the models are described in technical terms (dimensionality, location prior, serum effects, virus effects). The most confusing part is the location prior. My understanding (and if this is wrong then things really need to be clarified) is that uninformed refers to making the predictions just from HI data, drift refers to making the predictions from HI data + date of isolation, and diffusion / drift refers to making the predictions from HI data + date of isolation + phylogenetic relationship. If the models were named in terms of the data used to make the predictions rather than in terms of the prior, it would be easier to understand intuitively. For instance, a reader could then simply see, ``OK, they included the isolation date in the model and the predictions improved by this much.'' I recognize that the prior plays a role in addition to the data, but I feel that a more intuitive naming scheme should be considered with detailed descriptions in the Methods. There also might be a more intuitive way to denote serum effects and virus effects, although this is less confusing than the location prior.}

\textsl{* The first figure of the paper (Figure 1) is basically methodological rather than presenting a biological result. Although the data in this figure may be important, making it the first thing the reader encounters makes the paper less accessible to readers primarily interested in the biological results.}

\textsl{* The authors suggest that interference between clades could be due to CTL epitopes. While this is one plausible explanation, it could also be due to other forms of cross-reactive immunity such as broadly reactive anti-hemagglutinin antibodies (http://www.ncbi.nlm.nih.gov/pubmed/21878571, http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2860935/) or antibodies against internal proteins (http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3597515/).}

MINOR COMMENTS:

\textsl{* The first paragraph says: ``A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective.'' In fact, influenza vaccines are probably relatively ineffective for a number of reasons, and there is no correlation between vaccine match and efficacy (although the data is admittedly noisy), as discussed on page 28 of http://www.cidrap.umn.edu/cidrap/files/80/ccivi\%20report.pdf So it might be more accurate to say that ``antigenic drift is probably one reason why influenza vaccines are only transiently effective.''}

\textsl{* In the second paragraph of the Introduction, the authors refer to ``more evolved viruses.'' Because the concept of directionality in evolution is poorly defined, ``more recent'' would be better than ``more evolved.''}

\textsl{* In the third paragraph, the authors refer to ``largely linear movement'' of the virus. Although such a description becomes meaningful later in the paper when the authors start specifying antigenic dimensions, at this point in the introduction it is unclear what the virus is moving through in a linear fashion.}

\textsl{* The authors say, ``If a single map is desired, then the MDS method may be preferred while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.'' Is this the first part of this sentence true? The BMDS method introduced in this paper is such a clear improvement on the original method in the 2004 Smith et al Science paper that this new approach should always be preferred. I don't see any difficulties with summarizing the posterior of the BMDS as done in this paper. It might be possible to make some non-Bayesian (such as maximum-likelihood version) of the method in this paper that included genetic data / dates / virus+serum effects, and then argue that such a maximum-likelihood version could be preferred to the BMDS method here. But if the comparison is simply with the 2004 Smith et al Science approach (which only uses HI data without dates / phylogeny / virus+serum effects), then the current approach is clearly better due to the inclusion of additional data, regardless of one's philosophical preference for Bayesian versus maximum likelihood.}

\textsl{* In the text, the authors introduce the three beta coefficients without first defining what they mean in either intuitive or mathematical terms.}

\textsl{* The authors say, ``we find that interference between sister clades, e.g. B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g. B/Vic and A/H3N2.'' However, this statement is not linked to any analysis/data - where is the figure or table?}

\textsl{* Could the authors elaborate in the Methods about the meaning of the lines connecting points in Figure 2 and 3? They say these represent ``mean posterior diffusions paths,'' but I'm having trouble grasping what that means.}

\textsl{* It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map. Also, are the phylogenetic trees built on just HA1 or the whole HA protein?}

\textsl{* Although this is a stylistic rather than a substantive comment, the paper ends with a weak conclusion. The authors present a lot of complex results that could use some intuitive interpretations. Instead, their second-to-last sentence is a technical point about the phylogenetic methodology.}

\textsl{* The authors should be commended on writing a clear Methods section, particularly as it relates to the antigenic cartography.}

\textsl{* Page 16, there is an extraneous ``the'' before the word ``observing.''}

\textsl{* In equation 12, what is meant by the x subscript given to the sigma?}

%%% REVIEWER 2 %%%
\section*{Reviewer 2}

\textsl{The authors extend earlier work on ``antigenic cartography'' -- that is, low-dimensional embeddings of viral strains using dimensionality reduction techniques, in order to quantify antigenic groupings of viral isolates. What is novel here compared to prior work is the introduciton of a Baysian framework for this mapping process, the observation that antigenic drift within a clade correlates with clade-specific disease incidence each season, and anti-correlates with drift in other clades. The authors also introduce a phenomonological model of antigenic variation based on diffusion, and interpret antigenic drift in terms of fitted diffusion coeffficiants. Finally, the authors provide antigenic maps for more clades of influenza (h3n2, h1n1, b/vic, and b/yam) than in previous publications.}

\textsl{Understanding antigenic drift of influenza in quantitative terms is clearly of great practical importance, and also an interesting intellectual problem in its own right. And it is nice to see antigenic maps for so many viral clades. However, I do not feel that the technical advances in ``mapping'' strains introduced here provide us much more confidence in the antigenic locations of strains than prior techniques do. Indeed, the authors themselves note (page 3) that the average predictive error of this new methods is comparable to that of Smith et all (2004). What might be more valuable, perhaps, are the results relating inferred antigenic drift and influenza incidence each season, by clade. But these results are quite possibly driven by sampling biases (see below), and I am not really convinced of their validity -- at least not without sereious efforts to remove such sampling biases. Finally, the discussion of a ``diffusion'' model of drift was extremely confusing to me, and the fitted diffusion coefficients do not have a clear meaning or interpretation.}

\textsl{* The authors do not seem to discuss this a problematic aspect of their results relating inferred drift to influenza disease incidence -- namely that biased sampling of clades, in each season, could cause the observed correlations. If clades were sampled for HI assays unequally in each season, as I suspect, then this fact alone might cause the trend towards apparently more drift in the clade that was sampled most intensively in each season (and to which most ILI was attributed). This is an obvious possible confound for the analysis presented in Fig. 4, and it should be ruled out by simulation of constant true rates of antigenic drift, varible sampling proportions in each season, and reconstructed drift using BMDS.}

\textsl{* Likewise, in Fig. 5, the negative correlation between drift in different clades in each season could, again, be driven by variation in the number of strains per clade that were subjected to HI each season. The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.}

\textsl{* I found the section fitting diffusion coefficients (in different branches) to the antigenic maps confusing -- if only because strict diffusion (in one or two dimensions) has no advection term and therefore no tendency to move in one direction or another; whereas the data clearly show an advective tendency (moving to the ``right'' over time in antigenic dimension 1). I presume the authors attempt to deal with this by imposing a prior (eq 12) for the location of a virus based on its location in a molecular phylogeny. But this would seem to conflate antigenic drift, which the authors wish to quantify using a diffusion coefficient, with genetic mutations (many of which are driven by selection for antigenic escape). In other words, this hybrid approach of assuming a prior based on a strain's genotype, and then fitting a diffusion coefficient to summarize strength of drift, does not seem to to be a pure measure of drift. What exactly the fitted diffusion coefficient means is extremely unclear to me, in any formal sense, as the procedure is so ad-hoc.}

\textsl{* How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone? How many serious discrpencies that are outside the confidence intervals on map reconstruction? In other words, the authors should demonstrate that the HI data actually provide substantively different results for antigenic grouping than genetic data do alone. This was an important contention of Smith et al (2004), but it was based on the position of only a handful of viruses in that paper. Do the authors have more convincing evidence in this study that the antigenic data are providing substantial extra information beyond the genetic data?}

\textsl{Minor questions:}

\textsl{* What prior was used for the Baysian MDS? How does this choice of prior influence the resulting antigenic maps? How have the authors justified their choice of prior?}

\textsl{* Would not BMDS provide an entire posterior distribution for location of each strain and sera in a 2-d projection? Are authors plotting the posterior mean? What about the confidence region?}

\textsl{* Does a similar correlation analysis to the one shown in Fig. 4 find a relationship between estimated within-clade antigenic drift and excess P \& I mortality attributed to influenza?}

\textsl{* The ``uninformed'' prior used in Table 1 is not described in detail, but it actually hides a subtle issue. For an unbounded variable like the x- and y- positions of a mapped isolate, the ``uninformed'' prior cannot simply be uniform on $R^2$, but rather is uniform on some compact subset of $R^2$. In other words, the authors had to assume a prior with a maximum possible x- and y-position. They don't describe what maximum bounds went into their ``uniformed'' prior, but this is well-known to influence the results of Bayesian analyses in general. This is a relatively minor issue -- authors should simply show how the form of his uniformed prior influences the resulting maps.}

\textsl{* The authors should add a column to Table 1 showing the performance of the best possible model arising from purely genetic data -- that is, 2-D or 3-D embeddings based on aa hamming distances alone, ignoring all HI data.}

%%% REFERENCES %%%
\bibliographystyle{plos}
\bibliography{flux}

\end{document}

### 2013-05-14 flux.tex
Fixes #20 - Typo.

@@
 In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B).
-This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from the observing the serological data alone.
+This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from observing the serological data alone.
 Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.

### 2013-05-14 flux.tex
Fixes #10 - Evolution wording.

@@
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over previously established strains.
+Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over previously established strains.
 

### 2013-05-14 flux.tex
Fixes #21 - Subscripting.

@@
 \end{equation}
-for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion and $\tree$ is a phylogeny specifying tree topology and branch lengths.
+for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion over virus locations and $\tree$ is a phylogeny specifying tree topology and branch lengths.
 Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.

### 2013-05-14 response.tex
Fixes #20 - Typo.
Fixes #10 - Evolution wording.
Fixes #21 - Subscripting.
Fixes #19 - Methods clarity.
Fixes #15 - Strength of sister clade interference.

@@
 
+We agree. This has been corrected.
+
 \textsl{* In the third paragraph, the authors refer to ``largely linear movement'' of the virus.
@@
 
+This statement comes from the relative magnitudes of the $\beta_s$, representing the effect of antigenic drift on sister clades, and $\beta_t$, representing the effect of drift on more divergent clades.
+We find $\beta_s = -0.61$ and $\beta_t = -0.32$.
+We have clarified the text accordingly.
+
 \textsl{* Could the authors elaborate in the Methods about the meaning of the lines connecting points in Figure 2 and 3?
@@
 
+Thank you. We attempted to make the cartography methods as clear as possible.
+
 \textsl{* Page 16, there is an extraneous ``the'' before the word ``observing.''}
 
+Fixed.
+
 \textsl{* In equation 12, what is meant by the x subscript given to the sigma?}
+
+The $\sigma_x$ in the diffusion process is describing variance among virus locations $\mathbf{X}$.
+The $x$ subscript is meant to differentiate this from serum locations $\mathbf{Y}$.
+This has been clarified in the text.
 

### 2013-05-14 flux.tex
Fixes #15 - Strength of sister clade interference.

@@
 The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
-Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2.
+Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2 (Table~\ref{interferencetable}, $\bsister = -0.61$ vs $\bother = -0.32$).
 If interference is mediated by CTL epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.

### 2013-05-14 response.tex
Fix spelling.

@@
 \textsl{The authors extend earlier work on ``antigenic cartography'' -- that is, low-dimensional embeddings of viral strains using dimensionality reduction techniques, in order to quantify antigenic groupings of viral isolates.
-What is novel here compared to prior work is the introduciton of a Baysian framework for this mapping process, the observation that antigenic drift within a clade correlates with clade-specific disease incidence each season, and anti-correlates with drift in other clades.
-The authors also introduce a phenomonological model of antigenic variation based on diffusion, and interpret antigenic drift in terms of fitted diffusion coeffficiants.
+What is novel here compared to prior work is the introduction of a Bayesian framework for this mapping process, the observation that antigenic drift within a clade correlates with clade-specific disease incidence each season, and anti-correlates with drift in other clades.
+The authors also introduce a phenomenological model of antigenic variation based on diffusion, and interpret antigenic drift in terms of fitted diffusion coefficients.
 Finally, the authors provide antigenic maps for more clades of influenza (h3n2, h1n1, b/vic, and b/yam) than in previous publications.}
@@
 What might be more valuable, perhaps, are the results relating inferred antigenic drift and influenza incidence each season, by clade.
-But these results are quite possibly driven by sampling biases (see below), and I am not really convinced of their validity -- at least not without sereious efforts to remove such sampling biases.
+But these results are quite possibly driven by sampling biases (see below), and I am not really convinced of their validity -- at least not without serious efforts to remove such sampling biases.
 Finally, the discussion of a ``diffusion'' model of drift was extremely confusing to me, and the fitted diffusion coefficients do not have a clear meaning or interpretation.}
@@
 This is an obvious possible confound for the analysis presented in Fig.
-4, and it should be ruled out by simulation of constant true rates of antigenic drift, varible sampling proportions in each season, and reconstructed drift using BMDS.}
+4, and it should be ruled out by simulation of constant true rates of antigenic drift, variable sampling proportions in each season, and reconstructed drift using BMDS.}
 
@@
 \textsl{* How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone?
-How many serious discrpencies that are outside the confidence intervals on map reconstruction?
+How many serious discrepancies that are outside the confidence intervals on map reconstruction?
 In other words, the authors should demonstrate that the HI data actually provide substantively different results for antigenic grouping than genetic data do alone.
@@
 
-\textsl{* What prior was used for the Baysian MDS? How does this choice of prior influence the resulting antigenic maps? How have the authors justified their choice of prior?}
+\textsl{* What prior was used for the Bayesian MDS? How does this choice of prior influence the resulting antigenic maps? How have the authors justified their choice of prior?}
 

### 2013-05-14 flux.tex
Fixes #14 - Beta coefficients.

@@
 \newcommand{\normal}{\mathcal{N}}					% normal distribution
-\newcommand{\bwithin}{\beta_c}		% within clade drift coefficient
+\newcommand{\bwithin}{\beta_w}		% within clade drift coefficient
 \newcommand{\bsister}{\beta_s}		% sister clade drift coefficient
 \newcommand{\bother}{\beta_t}			% across clade drift coefficient
+\newcommand{\incclade}[1]{y_\mathrm{#1}}
 \newcommand{\driftclade}[1]{x_\mathrm{#1}}
@@
 This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
-We fit a linear model that predicts standardized incidence based upon drift within a clade, drift in other clades of the same virus type, and drift in clades of differing type (Table~\ref{interferencetable}).
-For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in A/H1N1 and by drift across types in B/Vic and B/Yam.
+We fit a linear model that predicts standardized incidence based on drift within each of the four lineages, according to the following equations
+\begin{eqnarray}
+	\incclade{H3} &=& \bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
+	\incclade{H1} &=& \bsister \, \driftclade{H3} + \bwithin \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
+	\incclade{Vic} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bwithin \, \driftclade{Vic} + \bsister \, \driftclade{Yam} \nonumber \\	
+	\incclade{Yam} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bsister \, \driftclade{Vic} + \bwithin \, \driftclade{Yam},
+\end{eqnarray}
+where $y$ represents standardized incidence, $x$ represents drift in antigenic units, $\bwithin$ scales within-clade drift, $\bsister$ scales drift across sister viruses of the same type, and $\bother$ scales drift across viruses of differing type (Table~\ref{interferencetable}).
+For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2 scaled by $\bwithin$, drift in A/H1N1 scaled by $\bsister$ and by drift in B/Vic and B/Yam scaled by $\bother$.
 We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.

### 2013-05-14 response.tex
Fixes #14 - Beta coefficients.
Fixes #11 - Linear movement wording.
Fixes #16 - Diffusion paths.

@@
 
+We concur and have revised this sentence.
+
 \textsl{* The authors say, ``If a single map is desired, then the MDS method may be preferred while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.'' Is this the first part of this sentence true?
@@
 \textsl{* In the text, the authors introduce the three beta coefficients without first defining what they mean in either intuitive or mathematical terms.}
+
+We have revised the text to include a more thorough introduction to the linear model that better describes the meaning of the beta coefficients.
 
@@
 They say these represent ``mean posterior diffusions paths,'' but I'm having trouble grasping what that means.}
+
+We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.
 

### 2013-05-14 flux.tex
Fixes #11 - Linear movement wording.
Fixes #16 - Diffusion paths.

@@
 The results of many HI assays across a multitude of virus strains of a single subtype can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
-The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
+The antigenic map of influenza A/H3N2 has shown substantial evolution of the influenza virus population since its emergence in 1968. 
 Evolution of antigenic phenotype appears punctuated with episodes of more rapid innovation interspersed by periods of stasis, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
@@
 
+We summarize diffusion paths of viral lineages (Figure~\ref{map}, Figure~\ref{drift}) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
+In the course of the MCMC, we sample tip locations and internal node locations.
+At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
+Then, these $x$ and $y$ values are analyzed across MCMC samples at each timepoint to arrive at a mean posterior location.
+We draw lines between these locations to show mean posterior diffusion paths.
+As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.
+
 \subsection*{Estimation of diffusion coefficient}

### 2013-05-14 flux.tex
Fixes #8 - Interference references.

@@
 The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
+The immunological basis of between-clade interference are not entirely clear, however cytotoxic T cells (CTLs) \cite{Voeten00}, antibodies that broadly neutralize HA \cite{Corti10} and antibodies to internal proteins \cite{Laidlaw13} have been hypothesized as possible mechansisms.
 Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2 (Table~\ref{interferencetable}, $\bsister = -0.61$ vs $\bother = -0.32$).
-If interference is mediated by CTL epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.
+If interference is mediated by CTLs or by antibodies to evolutionary conserved epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.
 

### 2013-05-14 flux.tex
Fix spelling.

@@
 The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
-The immunological basis of between-clade interference are not entirely clear, however cytotoxic T cells (CTLs) \cite{Voeten00}, antibodies that broadly neutralize HA \cite{Corti10} and antibodies to internal proteins \cite{Laidlaw13} have been hypothesized as possible mechansisms.
+The immunological basis of between-clade interference are not entirely clear, however cytotoxic T cells (CTLs) \cite{Voeten00}, antibodies that broadly neutralize HA \cite{Corti10} and antibodies to internal proteins \cite{Laidlaw13} have been hypothesized as possible mechanisms.
 Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2 (Table~\ref{interferencetable}, $\bsister = -0.61$ vs $\bother = -0.32$).
@@
 At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, these $x$ and $y$ values are analyzed across MCMC samples at each timepoint to arrive at a mean posterior location.
+Then, these $x$ and $y$ values are analyzed across MCMC samples at each time point to arrive at a mean posterior location.
 We draw lines between these locations to show mean posterior diffusion paths.
@@
 Antigenic cluster locations are consistent between methods (Figure~\ref{smith_comparison}A--B) and antigenic distances between pairs of viruses are consistent between temporally similar and temporally divergent viruses (Figure~\ref{smith_comparison}C--E), suggesting that the resulting maps are consistent at both local and global scales.
-Credible intervals of antegenic distances for the BMDS model remain narrow across the temporal spectrum (Figure~\ref{smith_comparison}C--E), implying a fair degree of rigidity to the map.
+Credible intervals of antigenic distances for the BMDS model remain narrow across the temporal spectrum (Figure~\ref{smith_comparison}C--E), implying a fair degree of rigidity to the map.
 
@@
 	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
-	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
 	(B) A posterior sample of antigenic locations from an equivalent BMDS model.
@@
 	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
-	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
 	(B) A posterior sample of antigenic locations from an equivalent BMDS model that has converged on the alternative solution.
@@
 	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
-	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
 	(B) A posterior sample of antigenic locations from a BMDS model that includes date-informed priors on antigenic locations.
@@
 	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
-	(A) MDS antigenic locations, reorientated so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
+	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
 	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus and serum effects.
@@
 This may explain why we observe local consistency between models at scales less than $\sim$15 years, but some degree of global inconsistency.
-Still, these results suggest that, when making local comparisons, such as those used to calculcate year-to-year antigenic drift (Figure~\ref{drift}), outcomes are expected to be robust to  many model particulars.
+Still, these results suggest that, when making local comparisons, such as those used to calculate year-to-year antigenic drift (Figure~\ref{drift}), outcomes are expected to be robust to  many model particulars.
 

### 2013-05-14 flux.tex
Fixes #18 - Overly technical conclusion.

@@
 
+Understanding antigenic evolution in seasonal influenza is crucial to our efforts of surveillance and control.
+Cartographic methods allow complex HI datasets to be compressed to more approachable location-based summaries that quantify antigenic relationships between strains, including relationships not directly assessed via HI.
 In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
@@
 In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
-Here, methods to efficiently calculate branch- and site-specific codon substitutions may be necessary to easily correlate sequence substitutions with antigenic change \cite{Lemey12}.
 
-A deeper understanding of the process of antigenic evolution in the human influenza virus may eventually prove to be crucial in improving vaccine strain selection.
+Identifying viruses that will come to predominate in the global virus population while they are still at low frequency remains an enormous challenge.
+However, combining evolutionary and antigenic information may eventually prove useful in identifying low-frequency, but expanding, lineages of antigenically novel viruses that represent ideal targets for vaccine strain selection.
 

### 2013-05-14 response.tex
Fixes #8 - Interference references.
Fixes #18 - Overly technical conclusion.
Fixes #12 - Incidence data availability.

@@
 
+We apologize for this oversight.  The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.
+
 \textsl{* The authors need to define their nomenclature regarding clade / lineage / type / strain.
@@
 While this is one plausible explanation, it could also be due to other forms of cross-reactive immunity such as broadly reactive anti-hemagglutinin antibodies (http://www.ncbi.nlm.nih.gov/pubmed/21878571, http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2860935/) or antibodies against internal proteins (http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3597515/).}
+
+We have expanded this paragraph to encompasss broadly neutralizing anti-HA antibodies and antibodies to internal proteins and included the above references.
 
@@
 Instead, their second-to-last sentence is a technical point about the phylogenetic methodology.}
+
+We agree and have revised the conclusion accordingly.
 

### 2013-05-14 response.tex
Fixes #13 - MDS vs BMDS.

@@
 
+We agree that we may have been too hesitant in promoting the advances included in the BMDS method.
+To clarify this point, we've separated the comparison of ML vs Bayes from the comparison of the 2004 Smith et al.
+cartographic model and the new model that includes virus effects, serum effects and evolutionary priors.
+MDS vs BMDS is in some ways a matter of preference; MDS will be faster and have the convenience of giving a single map to work with, while BMDS will be slower to produce a full posterior sample and this sample will be less convenient to work with than a single set of points.
+We did find, however, the model extensions to be quite useful in the analysis of this data.
+
 \textsl{* In the text, the authors introduce the three beta coefficients without first defining what they mean in either intuitive or mathematical terms.}

### 2013-05-14 response.tex
Reformat responses.

--- 2013-05-14 response.tex
@@
 
-\textsl{This paper is exceptional in many ways, but also contains some flaws that need to be addressed by revisions.
+This paper is exceptional in many ways, but also contains some flaws that need to be addressed by revisions.
 If these flaws are addressed, I would be highly supportive of publication.
 This paper is complex, so I have not been able put all of my major critiques in the first 500 words.
-Instead, I summarize my major critiques here, and then further detail them along with the minor points.}
+Instead, I summarize my major critiques here, and then further detail them along with the minor points.
 
-\textsl{This paper deals with the antigenic evolution of influenza.
+This paper deals with the antigenic evolution of influenza.
 A foundational paper in Science in 2004 by Smith et al introduced ``antigenic cartography,'' the technique used to monitor antigenic drift and select the influenza vaccine.
@@
 All of these improvements are made using a freely available phylogenetic software program that is the standard in the field.
-These methodological improvements would immediately merit publication in a technique-specific journal.}
+These methodological improvements would immediately merit publication in a technique-specific journal.
 
-\textsl{In addition, this paper presents biological results that should interest virologists, public health officials, and evolutionary biologists.
-While these results do not really alter the current understanding of influenza antigenic evolution, they extend this understanding from H3N2 to other clades, provide a more detailed analysis than has previously been performed for any clade, and quantify factors that have previously been characterized only qualitatively.}
+In addition, this paper presents biological results that should interest virologists, public health officials, and evolutionary biologists.
+While these results do not really alter the current understanding of influenza antigenic evolution, they extend this understanding from H3N2 to other clades, provide a more detailed analysis than has previously been performed for any clade, and quantify factors that have previously been characterized only qualitatively.
 
-\textsl{I do, however, have some major critiques:}
+I do, however, have some major critiques:
 
-\textsl{* It is difficult to understand the biological results without grasping technical aspects of the computational approach. In my specific comments, I suggest some changes.}
+It is difficult to understand the biological results without grasping technical aspects of the computational approach. In my specific comments, I suggest some changes.
 
-\textsl{* The authors need to explain where they got the incidence data, how they chose the date ranges, specify when they are using relative versus absolute incidence, and provide raw data as supporting material.
-I describe my concerns in more detail in the specific comments.}
+The authors need to explain where they got the incidence data, how they chose the date ranges, specify when they are using relative versus absolute incidence, and provide raw data as supporting material.
+I describe my concerns in more detail in the specific comments.
 
-\textsl{* The improvement from inclusion of serum effects deserves comment and explanation if possible (are the effects due to the serum source)?
+The improvement from inclusion of serum effects deserves comment and explanation if possible (are the effects due to the serum source)?
 Table 1 should include a model with virus effects but no serum effects.
-I describe my concerns in more detail in the specific comments.}
+I describe my concerns in more detail in the specific comments.
 
-\textsl{MAJOR POINTS IN MORE DETAIL:}
+MAJOR POINTS IN MORE DETAIL:
 
-\textsl{* I have concerns about Figure 5, which the authors use to support their claim that ``antigenic drift drives incidence rates.'' The authors first establish the well-known fact that H3N2 incidence is highest, followed by H1N1 and then B.
+I have concerns about Figure 5, which the authors use to support their claim that ``antigenic drift drives incidence rates.'' The authors first establish the well-known fact that H3N2 incidence is highest, followed by H1N1 and then B.
 They also find that relative incidence is correlated with the rate of antigenic drift.
@@
 Finding no evidence that antigenic drift drives incidence is just as important as finding that it does.
-But the claim needs to be correct.}
+But the claim needs to be correct.
 
-\textsl{* The incidence data are poorly explained.
+The incidence data are poorly explained.
 In the first paragraph of the Epidemiological consequences of antigenic drift section, the authors use relative incidence.
@@
 Also, why do they not analyze incidence prior to 1998/1999?
-Finally, because each influenza season spans two calendar years, why do they not break up the data by season rather than year?}
+Finally, because each influenza season spans two calendar years, why do they not break up the data by season rather than year?
 
-\textsl{* An interesting result is that estimating the serum effects improves the performance a great deal (more than including the phylogenetic data, in fact!).
+An interesting result is that estimating the serum effects improves the performance a great deal (more than including the phylogenetic data, in fact!).
 It suggests that there may be systematic differences between the sera.
@@
 Can the authors include a tenth model in Table 1 that includes the virus effects but not the serum effects?
-This would clarify if serum effects are actually more important than virus effects, or simply whether inclusion of either one individually is more helpful than both together.}
+This would clarify if serum effects are actually more important than virus effects, or simply whether inclusion of either one individually is more helpful than both together.
 
@@
 
-\textsl{* The authors should be commended for making publicly available all of the HI data that they have assembled (a good scientific practice that sadly has not been adhered to in some previous publications on antigenic cartography).
+The authors should be commended for making publicly available all of the HI data that they have assembled (a good scientific practice that sadly has not been adhered to in some previous publications on antigenic cartography).
 However, they should mention in the text the availability of these data on datadryad - I almost missed it since it is only mentioned on the cover page.
-They could also consider more clearly specifying where the code relevant to this paper is found in the BEAST repository.}
+They could also consider more clearly specifying where the code relevant to this paper is found in the BEAST repository.
 
-\textsl{* The one exception to the aforementioned praise on data availability is the influenza incidence data - I could not find any tables providing these data. They need to be provided.}
+The one exception to the aforementioned praise on data availability is the influenza incidence data - I could not find any tables providing these data. They need to be provided.
 
-\textsl{* The authors need to define their nomenclature regarding clade / lineage / type / strain.
+\textbf{We apologize for this oversight.  The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.}
+
+The authors need to define their nomenclature regarding clade / lineage / type / strain.
 For instance, they say: ``We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations.
 The term for across lineage Bt is small in magnitude.'' I had to read this many times before concluding that type refers to A or B, clade to refers to any of the four classifications.
-I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.}
+I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.
 
-\textsl{* The comparison in Table 1 is crucial, but the model descriptions are very hard to understand until after reading the Methods carefully.
+The comparison in Table 1 is crucial, but the model descriptions are very hard to understand until after reading the Methods carefully.
 The difficulty is because the models are described in technical terms (dimensionality, location prior, serum effects, virus effects).
@@
 For instance, a reader could then simply see, ``OK, they included the isolation date in the model and the predictions improved by this much.'' I recognize that the prior plays a role in addition to the data, but I feel that a more intuitive naming scheme should be considered with detailed descriptions in the Methods.
-There also might be a more intuitive way to denote serum effects and virus effects, although this is less confusing than the location prior.}
+There also might be a more intuitive way to denote serum effects and virus effects, although this is less confusing than the location prior.
 
-\textsl{* The first figure of the paper (Figure 1) is basically methodological rather than presenting a biological result.
-Although the data in this figure may be important, making it the first thing the reader encounters makes the paper less accessible to readers primarily interested in the biological results.}
+The first figure of the paper (Figure 1) is basically methodological rather than presenting a biological result.
+Although the data in this figure may be important, making it the first thing the reader encounters makes the paper less accessible to readers primarily interested in the biological results.
 
-\textsl{* The authors suggest that interference between clades could be due to CTL epitopes.
-While this is one plausible explanation, it could also be due to other forms of cross-reactive immunity such as broadly reactive anti-hemagglutinin antibodies (http://www.ncbi.nlm.nih.gov/pubmed/21878571, http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2860935/) or antibodies against internal proteins (http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3597515/).}
+The authors suggest that interference between clades could be due to CTL epitopes.
+While this is one plausible explanation, it could also be due to other forms of cross-reactive immunity such as broadly reactive anti-hemagglutinin antibodies (http://www.ncbi.nlm.nih.gov/pubmed/21878571, http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2860935/) or antibodies against internal proteins (http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3597515/).
+
+\textbf{We have expanded this paragraph to encompasss broadly neutralizing anti-HA antibodies and antibodies to internal proteins and included the above references.}
 
@@
 
-\textsl{* The first paragraph says: ``A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective.'' In fact, influenza vaccines are probably relatively ineffective for a number of reasons, and there is no correlation between vaccine match and efficacy (although the data is admittedly noisy), as discussed on page 28 of http://www.cidrap.umn.edu/cidrap/files/80/ccivi\%20report.pdf So it might be more accurate to say that ``antigenic drift is probably one reason why influenza vaccines are only transiently effective.''}
+The first paragraph says: ``A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective.'' In fact, influenza vaccines are probably relatively ineffective for a number of reasons, and there is no correlation between vaccine match and efficacy (although the data is admittedly noisy), as discussed on page 28 of http://www.cidrap.umn.edu/cidrap/files/80/ccivi\%20report.pdf So it might be more accurate to say that ``antigenic drift is probably one reason why influenza vaccines are only transiently effective.''
 
-\textsl{* In the second paragraph of the Introduction, the authors refer to ``more evolved viruses.'' Because the concept of directionality in evolution is poorly defined, ``more recent'' would be better than ``more evolved.''}
+In the second paragraph of the Introduction, the authors refer to ``more evolved viruses.'' Because the concept of directionality in evolution is poorly defined, ``more recent'' would be better than ``more evolved.''
 
-We agree. This has been corrected.
+\textbf{We agree. This has been corrected.}
 
-\textsl{* In the third paragraph, the authors refer to ``largely linear movement'' of the virus.
-Although such a description becomes meaningful later in the paper when the authors start specifying antigenic dimensions, at this point in the introduction it is unclear what the virus is moving through in a linear fashion.}
+In the third paragraph, the authors refer to ``largely linear movement'' of the virus.
+Although such a description becomes meaningful later in the paper when the authors start specifying antigenic dimensions, at this point in the introduction it is unclear what the virus is moving through in a linear fashion.
 
-We concur and have revised this sentence.
+\textbf{We concur and have revised this sentence.}
 
-\textsl{* The authors say, ``If a single map is desired, then the MDS method may be preferred while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.'' Is this the first part of this sentence true?
+The authors say, ``If a single map is desired, then the MDS method may be preferred while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.'' Is this the first part of this sentence true?
 The BMDS method introduced in this paper is such a clear improvement on the original method in the 2004 Smith et al Science paper that this new approach should always be preferred.
@@
 It might be possible to make some non-Bayesian (such as maximum-likelihood version) of the method in this paper that included genetic data / dates / virus+serum effects, and then argue that such a maximum-likelihood version could be preferred to the BMDS method here.
-But if the comparison is simply with the 2004 Smith et al Science approach (which only uses HI data without dates / phylogeny / virus+serum effects), then the current approach is clearly better due to the inclusion of additional data, regardless of one's philosophical preference for Bayesian versus maximum likelihood.}
+But if the comparison is simply with the 2004 Smith et al Science approach (which only uses HI data without dates / phylogeny / virus+serum effects), then the current approach is clearly better due to the inclusion of additional data, regardless of one's philosophical preference for Bayesian versus maximum likelihood.
 
-\textsl{* In the text, the authors introduce the three beta coefficients without first defining what they mean in either intuitive or mathematical terms.}
+\textbf{We agree that we may have been too hesitant in promoting the advances included in the BMDS method.
+To clarify this point, we've separated the comparison of ML vs Bayes from the comparison of the 2004 Smith et al.
+cartographic model and the new model that includes virus effects, serum effects and evolutionary priors.
+MDS vs BMDS is in some ways a matter of preference; MDS will be faster and have the convenience of giving a single map to work with, while BMDS will be slower to produce a full posterior sample and this sample will be less convenient to work with than a single set of points.
+We did find, however, the model extensions to be quite useful in the analysis of this data.}
 
-We have revised the text to include a more thorough introduction to the linear model that better describes the meaning of the beta coefficients.
+In the text, the authors introduce the three beta coefficients without first defining what they mean in either intuitive or mathematical terms.
 
-\textsl{* The authors say, ``we find that interference between sister clades, e.g.
+\textbf{We have revised the text to include a more thorough introduction to the linear model that better describes the meaning of the beta coefficients.}
+
+The authors say, ``we find that interference between sister clades, e.g.
 B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.
-B/Vic and A/H3N2.'' However, this statement is not linked to any analysis/data - where is the figure or table?}
+B/Vic and A/H3N2.'' However, this statement is not linked to any analysis/data - where is the figure or table?
 
-This statement comes from the relative magnitudes of the $\beta_s$, representing the effect of antigenic drift on sister clades, and $\beta_t$, representing the effect of drift on more divergent clades.
+\textbf{This statement comes from the relative magnitudes of the $\beta_s$, representing the effect of antigenic drift on sister clades, and $\beta_t$, representing the effect of drift on more divergent clades.
 We find $\beta_s = -0.61$ and $\beta_t = -0.32$.
-We have clarified the text accordingly.
+We have clarified the text accordingly.}
 
-\textsl{* Could the authors elaborate in the Methods about the meaning of the lines connecting points in Figure 2 and 3?
-They say these represent ``mean posterior diffusions paths,'' but I'm having trouble grasping what that means.}
+Could the authors elaborate in the Methods about the meaning of the lines connecting points in Figure 2 and 3?
+They say these represent ``mean posterior diffusions paths,'' but I'm having trouble grasping what that means.
 
-We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.
+\textbf{We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.}
 
-\textsl{* It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map.
-Also, are the phylogenetic trees built on just HA1 or the whole HA protein?}
+It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map.
+Also, are the phylogenetic trees built on just HA1 or the whole HA protein?
 
-\textsl{* Although this is a stylistic rather than a substantive comment, the paper ends with a weak conclusion.
+Although this is a stylistic rather than a substantive comment, the paper ends with a weak conclusion.
 The authors present a lot of complex results that could use some intuitive interpretations.
-Instead, their second-to-last sentence is a technical point about the phylogenetic methodology.}
+Instead, their second-to-last sentence is a technical point about the phylogenetic methodology.
 
-\textsl{* The authors should be commended on writing a clear Methods section, particularly as it relates to the antigenic cartography.}
+\textbf{We agree and have revised the conclusion accordingly.}
 
-Thank you. We attempted to make the cartography methods as clear as possible.
+The authors should be commended on writing a clear Methods section, particularly as it relates to the antigenic cartography.
 
-\textsl{* Page 16, there is an extraneous ``the'' before the word ``observing.''}
+\textbf{Thank you. We attempted to make the cartography methods as clear as possible.}
 
-Fixed.
+Page 16, there is an extraneous ``the'' before the word ``observing.''
 
-\textsl{* In equation 12, what is meant by the x subscript given to the sigma?}
+\textbf{Fixed.}
 
-The $\sigma_x$ in the diffusion process is describing variance among virus locations $\mathbf{X}$.
+In equation 12, what is meant by the x subscript given to the sigma?
+
+\textbf{The $\sigma_x$ in the diffusion process is describing variance among virus locations $\mathbf{X}$.
 The $x$ subscript is meant to differentiate this from serum locations $\mathbf{Y}$.
-This has been clarified in the text.
+This has been clarified in the text.}
 
@@
 
-\textsl{The authors extend earlier work on ``antigenic cartography'' -- that is, low-dimensional embeddings of viral strains using dimensionality reduction techniques, in order to quantify antigenic groupings of viral isolates.
+The authors extend earlier work on ``antigenic cartography'' -- that is, low-dimensional embeddings of viral strains using dimensionality reduction techniques, in order to quantify antigenic groupings of viral isolates.
 What is novel here compared to prior work is the introduction of a Bayesian framework for this mapping process, the observation that antigenic drift within a clade correlates with clade-specific disease incidence each season, and anti-correlates with drift in other clades.
 The authors also introduce a phenomenological model of antigenic variation based on diffusion, and interpret antigenic drift in terms of fitted diffusion coefficients.
-Finally, the authors provide antigenic maps for more clades of influenza (h3n2, h1n1, b/vic, and b/yam) than in previous publications.}
+Finally, the authors provide antigenic maps for more clades of influenza (h3n2, h1n1, b/vic, and b/yam) than in previous publications.
 
-\textsl{Understanding antigenic drift of influenza in quantitative terms is clearly of great practical importance, and also an interesting intellectual problem in its own right.
+Understanding antigenic drift of influenza in quantitative terms is clearly of great practical importance, and also an interesting intellectual problem in its own right.
 And it is nice to see antigenic maps for so many viral clades.
@@
 But these results are quite possibly driven by sampling biases (see below), and I am not really convinced of their validity -- at least not without serious efforts to remove such sampling biases.
-Finally, the discussion of a ``diffusion'' model of drift was extremely confusing to me, and the fitted diffusion coefficients do not have a clear meaning or interpretation.}
+Finally, the discussion of a ``diffusion'' model of drift was extremely confusing to me, and the fitted diffusion coefficients do not have a clear meaning or interpretation.
 
-\textsl{* The authors do not seem to discuss this a problematic aspect of their results relating inferred drift to influenza disease incidence -- namely that biased sampling of clades, in each season, could cause the observed correlations.
+The authors do not seem to discuss this a problematic aspect of their results relating inferred drift to influenza disease incidence -- namely that biased sampling of clades, in each season, could cause the observed correlations.
 If clades were sampled for HI assays unequally in each season, as I suspect, then this fact alone might cause the trend towards apparently more drift in the clade that was sampled most intensively in each season (and to which most ILI was attributed).
 This is an obvious possible confound for the analysis presented in Fig.
-4, and it should be ruled out by simulation of constant true rates of antigenic drift, variable sampling proportions in each season, and reconstructed drift using BMDS.}
+4, and it should be ruled out by simulation of constant true rates of antigenic drift, variable sampling proportions in each season, and reconstructed drift using BMDS.
 
-\textsl{* Likewise, in Fig.
+Likewise, in Fig.
 5, the negative correlation between drift in different clades in each season could, again, be driven by variation in the number of strains per clade that were subjected to HI each season.
-The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.}
+The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.
 
-\textsl{* I found the section fitting diffusion coefficients (in different branches) to the antigenic maps confusing -- if only because strict diffusion (in one or two dimensions) has no advection term and therefore no tendency to move in one direction or another; whereas the data clearly show an advective tendency (moving to the ``right'' over time in antigenic dimension 1).
+I found the section fitting diffusion coefficients (in different branches) to the antigenic maps confusing -- if only because strict diffusion (in one or two dimensions) has no advection term and therefore no tendency to move in one direction or another; whereas the data clearly show an advective tendency (moving to the ``right'' over time in antigenic dimension 1).
 I presume the authors attempt to deal with this by imposing a prior (eq 12) for the location of a virus based on its location in a molecular phylogeny.
@@
 In other words, this hybrid approach of assuming a prior based on a strain's genotype, and then fitting a diffusion coefficient to summarize strength of drift, does not seem to to be a pure measure of drift.
-What exactly the fitted diffusion coefficient means is extremely unclear to me, in any formal sense, as the procedure is so ad-hoc.}
+What exactly the fitted diffusion coefficient means is extremely unclear to me, in any formal sense, as the procedure is so ad-hoc.
 
-\textsl{* How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone?
+How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone?
 How many serious discrepancies that are outside the confidence intervals on map reconstruction?
@@
 This was an important contention of Smith et al (2004), but it was based on the position of only a handful of viruses in that paper.
-Do the authors have more convincing evidence in this study that the antigenic data are providing substantial extra information beyond the genetic data?}
+Do the authors have more convincing evidence in this study that the antigenic data are providing substantial extra information beyond the genetic data?
 
-\textsl{Minor questions:}
+Minor questions:
 
-\textsl{* What prior was used for the Bayesian MDS? How does this choice of prior influence the resulting antigenic maps? How have the authors justified their choice of prior?}
+What prior was used for the Bayesian MDS? How does this choice of prior influence the resulting antigenic maps? How have the authors justified their choice of prior?
 
-\textsl{* Would not BMDS provide an entire posterior distribution for location of each strain and sera in a 2-d projection? Are authors plotting the posterior mean? What about the confidence region?}
+Would not BMDS provide an entire posterior distribution for location of each strain and sera in a 2-d projection? Are authors plotting the posterior mean? What about the confidence region?
 
-\textsl{* Does a similar correlation analysis to the one shown in Fig. 4 find a relationship between estimated within-clade antigenic drift and excess P \& I mortality attributed to influenza?}
+Does a similar correlation analysis to the one shown in Fig. 4 find a relationship between estimated within-clade antigenic drift and excess P \& I mortality attributed to influenza?
 
-\textsl{* The ``uninformed'' prior used in Table 1 is not described in detail, but it actually hides a subtle issue.
+The ``uninformed'' prior used in Table 1 is not described in detail, but it actually hides a subtle issue.
 For an unbounded variable like the x- and y- positions of a mapped isolate, the ``uninformed'' prior cannot simply be uniform on $R^2$, but rather is uniform on some compact subset of $R^2$.
@@
 They don't describe what maximum bounds went into their ``uniformed'' prior, but this is well-known to influence the results of Bayesian analyses in general.
-This is a relatively minor issue -- authors should simply show how the form of his uniformed prior influences the resulting maps.}
+This is a relatively minor issue -- authors should simply show how the form of his uniformed prior influences the resulting maps.
 
-\textsl{* The authors should add a column to Table 1 showing the performance of the best possible model arising from purely genetic data -- that is, 2-D or 3-D embeddings based on aa hamming distances alone, ignoring all HI data.}
+The authors should add a column to Table 1 showing the performance of the best possible model arising from purely genetic data -- that is, 2-D or 3-D embeddings based on aa hamming distances alone, ignoring all HI data.
 

### 2013-05-15 response.tex
Fixes #28 - Drift vs P&I mortality.

@@
 
-\textbf{We have expanded this paragraph to encompasss broadly neutralizing anti-HA antibodies and antibodies to internal proteins and included the above references.}
+\textbf{We have expanded this paragraph to encompass broadly neutralizing anti-HA antibodies and antibodies to internal proteins and included the above references.}
 
@@
 
+\textbf{We have begun to extend the correlation analysis between within-clade antigenic drift and multiple measures of influenza incidence, including ILI, virus isolation, excess P\&I mortality and influenza-associated hospitalization rates.
+This analysis is also attempting to look at a finer temporal scale, rather than the course-grained 1-year bins in the current study and attempting to better match geographic bounds in antigenic and incidence data.
+This said, we believe that further extending the incidence correlation requires a separate publication to  describe adequately, and is beyond the scope of the current analysis.}
+
 The ``uninformed'' prior used in Table 1 is not described in detail, but it actually hides a subtle issue.

### 2013-05-16 response.tex
Updates response to issue #16.

@@
 
-\textbf{We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.}
+\textbf{We apologize for the lack of clarity here.
+We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.
+This paragraph reads as follows:}
+
+\textbf{``We summarize diffusion paths of viral lineages (Figure~2, Figure~3) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
+In the course of the MCMC, we sample tip locations and internal node locations.
+At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
+Then, these $x$ and $y$ values are extracted across MCMC samples at each time point to arrive at a mean posterior location.
+We draw lines between these locations to show mean posterior diffusion paths.
+As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.''}
 

### 2013-05-16 flux.tex
Fixes #12 - Incidence data availability.
Updates response to issue #16.

@@
 We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (Figures~\ref{smith_comparison}--\ref{smith_comparison_effects}).
-Differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
-If a single map is desired, then the MDS method may be preferred, while the BMDS method provides flexibility, allowing other sources of information, such as genetic data, to be more easily incorporated.
+When implementing the same underlying model, differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
+Additionally, the BMDS method improves flexibility, allowing extensions to the basic cartographic model, such as the incorporation of serum effects and evolutionary priors, that improve fit and add biological interpretability.
 
@@
 At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, these $x$ and $y$ values are analyzed across MCMC samples at each time point to arrive at a mean posterior location.
+Then, these $x$ and $y$ values are extracted across MCMC samples at each time point to arrive at a mean posterior location.
 We draw lines between these locations to show mean posterior diffusion paths.
@@
 Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
+Incidence data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
 

### 2013-05-24 response.tex
Small wording change.
Poor description of our Bayesian inference.

@@
 \textbf{``We summarize diffusion paths of viral lineages (Figure~2, Figure~3) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
-In the course of the MCMC, we sample tip locations and internal node locations.
-At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, these $x$ and $y$ values are extracted across MCMC samples at each time point to arrive at a mean posterior location.
-We draw lines between these locations to show mean posterior diffusion paths.
+In the course of the MCMC, we sample tip locations and internal node locations. [THIS IS NOT TRUE]
+At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus. [THIS IS NOT TRUE]
+Then, these $x$ and $y$ values are analyzed across MCMC samples at each time point to arrive at a mean posterior location. [analyzed -> summarized, location -> location estimate]
+We draw lines between these locations to show mean posterior diffusion paths. [show -> approximate?]
 As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.''}

### 2013-07-16 response.tex
Attempt to fix description of diffusion path calculcation.

--- 2013-05-16 response.tex
@@
 \textbf{``We summarize diffusion paths of viral lineages (Figure~2, Figure~3) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
-In the course of the MCMC, we sample tip locations and internal node locations.
-At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, these $x$ and $y$ values are extracted across MCMC samples at each time point to arrive at a mean posterior location.
-We draw lines between these locations to show mean posterior diffusion paths.
+We use MCMC to sample tip locations, but when outputting trees sample internal node locations using a peeling algorithm as described in [22].
+Thus, after the MCMC is finished we have a posterior sample of 2000 trees each tagged with estimated tip locations and internal node locations.
+We post-processed each posterior tree by conducting a linear interpolation between parent-child node locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
+Then, for each interval, $x$ and $y$ values are averaged across the sample of posterior trees.
+We draw lines between these locations to approximate mean posterior diffusion paths.
 As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.''}

### 2013-07-16 flux.tex
Small wording change.
Attempt to fix description of diffusion path calculcation.

@@
 We summarize diffusion paths of viral lineages (Figure~\ref{map}, Figure~\ref{drift}) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
-In the course of the MCMC, we sample tip locations and internal node locations.
-At each MCMC step, we conduct a linear interpolation between these sampled locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, these $x$ and $y$ values are extracted across MCMC samples at each time point to arrive at a mean posterior location.
-We draw lines between these locations to show mean posterior diffusion paths.
+We use MCMC to sample tip locations, but when outputting trees sample internal node locations using a peeling algorithm as described in \cite{Pybus12}.
+Thus, after the MCMC is finished we have a posterior sample of 2000 trees each tagged with estimated tip locations and internal node locations.
+We post-processed each posterior tree by conducting a linear interpolation between parent-child node locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
+Then, for each interval, $x$ and $y$ values are averaged across the sample of posterior trees.
+We draw lines between these locations to approximate mean posterior diffusion paths.
 As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.

### 2013-07-16 response.tex
Fix #30 - performance of purely genetic model.

@@
 
+\textbf{We agree that it would be ideal to provide a test error associated with a purely genetic model.
+However, test errors in Table 1 are based on the prediction of a set of HI titers based on other HI titers or other HI titers plus genetic data.
+Without including some form of HI information, it's difficult to see how to predict HI titer.
+One could have a `distance' in terms of AA rather than antigenic map units, but in this case, one would still need to include a scaling for `serum effects', which could not be made without reference to HI data.
+Regardless, we've attempted to address this issue in our discussion of correlation between antigenic and genetic distances, as described above.}
+
 %%% REFERENCES %%%

### 2013-07-23 flux.tex
Small notation fix (change \mu to m in diffuse normal mean).

@@
 \begin{eqnarray}
-	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
-	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
+	\virus_i &\sim& \normal(\mathbf{m},\boldsymbol{\Sigma}) \nonumber \\
+	\serum_j &\sim& \normal(\mathbf{m},\boldsymbol{\Sigma}),
 \end{eqnarray}
-where $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
+where $\mathbf{m} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
 

### 2013-07-23 flux.tex
Notation change: drift term from \beta to \mu to agree with standard diffusion notation and also to avoid confusion with incidence regression terms.

@@
 \newcommand{\serumsd}{\sigma_y}						% serum standard deviation
+\newcommand{\drift}{\mu}							% drift / advection
 \newcommand{\tree}{\tau}							% phylogeny
@@
 \begin{eqnarray}
-	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
-	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
+	x_{i1} &\sim& \drift \, t_i + \normal(0, \virussd^2) \nonumber \\
+	y_{j1} &\sim& \drift \, t_j + \normal(0, \serumsd^2),
 \end{eqnarray}
 where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
-Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
+Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\drift$.
 The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
@@
 \begin{equation} \label{ebpprior}
-	\viruses \sim \left( \begin{matrix} \beta \, t_1 & 0 \\ \vdots & \vdots \\ \beta \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
+	\viruses \sim \left( \begin{matrix} \drift \, t_1 & 0 \\ \vdots & \vdots \\ \drift \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
 \end{equation}
@@
 \begin{equation} 
-	\virus_i \sim (\beta \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
+	\virus_i \sim (\drift \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
 \end{equation}
 for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
-The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\beta \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
-The probability of virus locations $p(\viruses | \beta, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Wiener process with drift, in which the drift term $\beta$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
+The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\drift \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
+The probability of virus locations $p(\viruses | \drift, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
+This formulation corresponds to a Wiener process with drift, in which the drift term $\drift$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
 This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
@@
 
-Top-level priors for $1/\mdssd^2$, $\beta$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
+Top-level priors for $1/\mdssd^2$, $\drift$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
 Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
@@
 	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
-	p(\viruses | \beta, \virussd, \tree) \;
-	p(\sera | \beta, \serumsd) \; 
-	p(\ses, \ves, \mdssd, \beta, \virussd, \serumsd, \tree).
+	p(\viruses | \drift, \virussd, \tree) \;
+	p(\sera | \drift, \serumsd) \; 
+	p(\ses, \ves, \mdssd, \drift, \virussd, \serumsd, \tree).
 \end{equation}
 We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\beta$, $\virussd$ and $\serumsd$.
+Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
 For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
@@
 
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\beta$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
-In all cases when drift parameter $\beta$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
-However, without drift parameter $\beta$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
-We favor models that include $\beta$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
+In all cases when drift parameter $\drift$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
+However, without drift parameter $\drift$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
+We favor models that include $\drift$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
 

### 2013-07-23 flux.tex
Fix #7 - figure ordering.

@@
 
-We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Figure~\ref{error_by_distance_and_year}A), thus supporting our model assumption that the drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
-Additionally, we find that the absolute error in predicted titer is nearly constant with time (Figure~\ref{error_by_distance_and_year}B).
+We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Pearson correlation, $r = 0.098$), thus supporting our model assumption that the drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
+Additionally, we find that the absolute error in predicted titer is nearly constant with time (Pearson correlation, $r = -0.007$).
 Antigenic locations inferred by the model are well resolved; estimates of antigenic distance between pairs of viruses show relatively little variation across the posterior.
 We estimate that virus distances have, on average, a 50\% credible interval of $\pm0.45$ antigenic units for A/H3N2, $\pm0.57$ units for A/H1N1, $\pm0.76$ units for B/Vic, and $\pm0.65$ units for B/Yam.
-
-%%% error_by_distance_and_year %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/error_by_distance_and_year}
-	\caption{\textbf{Absolute error in log$_2$ HI titer against antigenic distance and date of virus sampling in influenza A/H3N2.} 
-	(A) Points show the absolute deviation from the predicted log$_2$ HI titer of the antigenic model, computed from virus location $\virus_i$, serum location $\serum_j$, virus effect $\ve_i$ and serum effect $\se_j$, based upon antigenic distance between virus and serum $\delta_{ij}$.
-	The red line shows a LOESS regression curve of these points.
-	(B) Points show the absolute deviation from the predicted log$_2$ HI titers of the antigenic model, based upon the date of virus sampling.
-	The red line shows average deviations for each year.
-	Relationships for A/H1N1, B/Vic and B/Yam are similar.
-	} 
-	\label{error_by_distance_and_year} 
-\end{figure}
 

### 2013-07-23 response.tex
Include response in fix to #7.
Fix #25 - antigenic vs genetic distances. Include new figure 1.

@@
 
+\textbf{We agree and have revised the manuscript to remove this figure.
+Instead of the figure we report Pearson's correlations between antigenic distance and error ($r = 0.098$) and between year and error ($r = -0.007$).}
+
 The authors suggest that interference between clades could be due to CTL epitopes.
@@
 
+\textbf{We have revised the manuscript to include a thorough analysis of the relationship between genetic distance and antigenic distance.
+This is new figure 1.
+Rather than constructing BMDS maps with pairwise AA distances, we take a more direct approach and compare pairwise antigenic distance from a BMDS model using only HI data (model 2 in Table 1) and pairwise genetic distances.
+We examine both pairwise AA distance and pairwise phylogenetic distance, as phylogenetic distance was the basis for our evolutionary diffusion model.
+We find fairly modest correlations in most cases case (between 0.10 for B/Yam and 0.68 for A/H3N2).
+Genetic data provides some predictive power for antigenic distance, but is a rather weak predictor alone.
+This was a big part of our rationale for attempting a joint genetic / antigenic model.}
+
 Minor questions:

### 2013-07-23 flux.tex
Fix #25 - antigenic vs genetic distances. Include new figure 1.

@@
 
-Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01, Smith04, Russell08}.
-Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
+Previous work on influenza antigenic and genetic evolution has shown that antigenic distance accumulates with increasing genetic distance \cite{Hay01, Smith04, Russell08}.
+Here, we examine pairwise relationships between viruses and observe a correlation between amino acid mutations and antigenic distance (Figure~\ref{seq_grid}) and a similar correlation between phylogenetic distance and antigenic distance (Figure~\ref{seq_grid}).
+Thus, genetic relationships between viruses provide some predictive power to estimate antigenic distances in the absence of HI data.
+However, the magnitudes of the coefficients of determination $R^2$ are low (Figure~\ref{seq_grid}), suggesting that genetic relationships alone will not completely resolve antigenic distances.
+
+%%% map %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{figures/seq_grid}
+	\caption{\textbf{Pairwise correlations between genetic distance, measured as amino acid mutations or as phylogenetic distance, and antigenic distance for influenza A/H3N2, A/H1N1, B/Vic and B/Yam.} 
+	The top row shows correlations between number of amino acid mutations and average antigenic distance between 10,000 random pairs of viruses.
+	The bottom row shows correlations between average phylogenetic distance, measured in terms of years, and average antigenic distance between 10,000 random pairs of viruses.
+	Dashed lines show linear model fits, with $R^2$ and slope noted, while solid lines show LOESS fits.
+	Antigenic distances derive from model 2 of Table \ref{errortable}.
+	} 
+	\label{seq_grid} 
+\end{figure}
+
+Consequently, we seek to flexibly incorporate genetic data by simultaneously modeling antigenic and genetic evolution through an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.

### 2013-07-23 response.tex
Fix #29 - use of "uninformed" prior.

@@
 
+\textbf{We apologize for the lack of clarity.
+The `Uninformed location prior' in Table 1 was described in Materials and methods as a diffuse normal prior (equation 10) with each coordinate having a variance of 10000.
+We've made this clearer in the test.
+We chose this variance based on prior knowledge of the scale of antigenic maps, i.e.
+around 50 units in breadth for A/H3N2 from Smith et al.\ 2004.
+In general, our use of diffuse priors limits their influence on the resulting antigenic maps.
+In this case, choosing a variance of 1000 gives very similar results.}
+
 The authors should add a column to Table 1 showing the performance of the best possible model arising from purely genetic data -- that is, 2-D or 3-D embeddings based on aa hamming distances alone, ignoring all HI data.

### 2013-07-23 flux.tex
Fix #29 - use of "uninformed" prior.

@@
 
-We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations, expected titer is relative to the maximum titer of a particular ferret serum and virus and serum locations follow a diffuse prior.
+We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations.
+In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera, and virus and serum locations follow an uninformative diffuse normal prior as described in the `Antigenic cartography' section of Materials and methods.
 After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.

### 2013-07-23 response.tex
Fix typo.
Fix #31 - virus effects.

@@
 Can this be attributed to the serum itself (such as whether it is goat versus ferret), or to the particular lab or study?
+
 Could the authors correlate the estimated serum effects back to the originating source for that serum to check?
@@
 This would clarify if serum effects are actually more important than virus effects, or simply whether inclusion of either one individually is more helpful than both together.
+
+\textbf{The inclusion of `virus effects' was meant to capture exactly this obvservation, that some viruses may be generally more resistant to HI.
+To further assess their importance, we followed this suggestion and included test error for a model that estimates virus effects, but leaves serum effects fixed at maximum titers.
+In this case, we observe similar, but not quite as substantial, improvements to test error as the model that estimates serum effects, but does not include virus effects.
+From this observation, we conclude that estimating either virus or serum effects both improve the model.}
 
@@
 The `Uninformed location prior' in Table 1 was described in Materials and methods as a diffuse normal prior (equation 10) with each coordinate having a variance of 10000.
-We've made this clearer in the test.
+We've made this clearer in the text.
 We chose this variance based on prior knowledge of the scale of antigenic maps, i.e.

### 2013-07-23 flux.tex
Fix #31 - virus effects.

@@
 	8		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
-	9		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
+	9		&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
+	10		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
 	\hline
@@
 We refer to these estimates as serum effects and virus effects, respectively.
-We find that including these effects decreases test error further, ranging in improvement from 0.03 to 0.13 log$_2$ HI titers (models 8 and 9; Table \ref{errortable}).
+We find that including these effects decreases test error further, resulting in comparable improvements from estimating virus and serum effects individually and giving an improvement of between 0.03 to 0.13 log$_2$ HI titers when estimating both virus and serum effects together (models 8--10; Table \ref{errortable}).
 

### 2013-07-24 response.tex
Fix typo.
Fix #27 - posterior location summaries.

@@
 
-\textbf{The inclusion of `virus effects' was meant to capture exactly this obvservation, that some viruses may be generally more resistant to HI.
+\textbf{The inclusion of `virus effects' was meant to capture exactly this observation, that some viruses may be generally more resistant to HI.
 To further assess their importance, we followed this suggestion and included test error for a model that estimates virus effects, but leaves serum effects fixed at maximum titers.
@@
 
+\textbf{The BMDS does give a posterior sample of virus locations.  However, summarizing this posterior distribution proved difficult.  We discussed these difficulties in Materials and Methods:}
+
+\textbf{``There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
+Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
+However, local isometries are often still a problem.
+For example, in A/H3N2 the HK/68, EN/72 and VI/75 clusters may rotate in relation to other clusters.
+Consequently, it may be difficult to fully align MCMC samples using Procrustes analysis.
+For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~2, Figure~3).
+Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate the quantity across MCMC samples to yield an expectation and a credible interval.
+This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.''}
+
 Does a similar correlation analysis to the one shown in Fig. 4 find a relationship between estimated within-clade antigenic drift and excess P \& I mortality attributed to influenza?

### 2013-07-24 response.tex
Fix #26 - BMDS priors.

@@
 \makeatother
+
+% notation
+\usepackage{amsmath}
+\usepackage{amssymb}
+\newcommand{\virus}{\mathbf{x}}						% virus coordinate
+\newcommand{\serum}{\mathbf{y}}						% serum coordinate
+\newcommand{\viruses}{\mathbf{X}}					% set of virus coordinates
+\newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
+\newcommand{\ve}{v}									% virus effect
+\newcommand{\se}{s}									% serum effect
+\newcommand{\ves}{\mathbf{v}}						% set of virus effects
+\newcommand{\ses}{\mathbf{s}}						% set of serum effects
+\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
+\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
+\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
+\newcommand{\mdssd}{\varphi}						% MDS standard deviation
+\newcommand{\virussd}{\sigma_x}						% virus / diffusion standard deviation
+\newcommand{\serumsd}{\sigma_y}						% serum standard deviation
+\newcommand{\drift}{\mu}							% drift / advection
+\newcommand{\tree}{\tau}							% phylogeny
+\newcommand{\vn}{n}									% number of viruses
+\newcommand{\sn}{k}									% number of sera
+\newcommand{\normal}{\mathcal{N}}					% normal distribution
+\newcommand{\bwithin}{\beta_w}		% within clade drift coefficient
+\newcommand{\bsister}{\beta_s}		% sister clade drift coefficient
+\newcommand{\bother}{\beta_t}			% across clade drift coefficient
+\newcommand{\incclade}[1]{y_\mathrm{#1}}
+\newcommand{\driftclade}[1]{x_\mathrm{#1}}
+\setlength{\arraycolsep}{2pt}
+\newcommand{\smalltwomatrix}[2]{\scriptsize \Big( \begin{matrix} #1 \\ #2 \end{matrix} \Big)}				% pretty inline matrix 
+\newcommand{\smallfourmatrix}[4]{\scriptsize \Big( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \Big)}	% pretty inline matrix 
+\newcommand{\twomatrix}[2]{\left( \begin{matrix} #1 \\ #2 \end{matrix} \right)}								% pretty inline matrix 
+\newcommand{\fourmatrix}[4]{\left( \begin{matrix} #1 & #2 \\ #3 & #4 \end{matrix} \right)}					% pretty inline matrix
 
@@
 Can this be attributed to the serum itself (such as whether it is goat versus ferret), or to the particular lab or study?
+Could the authors correlate the estimated serum effects back to the originating source for that serum to check?
 
-Could the authors correlate the estimated serum effects back to the originating source for that serum to check?
 As a related point, I was surprised that inclusion of the virus effects led to only minor improvement, since prior work (http://www.ncbi.nlm.nih.gov/pubmed/19900932) suggests that some viruses are generally more resistant to HI than others.
@@
 
+\textbf{We assume priors on several BMDS parameters: MDS variance $\mdssd^2$, drift rate $\drift$, virus variance $\virussd^2$ and serum variance $\serumsd^2$.  We state in Materials and methods:}
+
+\textbf{``Top-level priors for $1/\mdssd^2$, $\drift$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.''}
+
+\textbf{These are extremely diffuse priors and should have very little impact on the resulting antigenic maps.
+Running the analysis with improper uniform priors has almost no impact on the results.
+We've made note of this in the text.}
+
 Would not BMDS provide an entire posterior distribution for location of each strain and sera in a 2-d projection? Are authors plotting the posterior mean? What about the confidence region?

### 2013-07-24 flux.tex
Fix #26 - BMDS priors.
Fix #9 - Antigenic drift wording.

@@
 The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
-A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
+A large proportion of the disease burden of influenza stems from antigenic drift, which allows individuals to be infected multiple times throughout their lives.
+Although influenza vaccines may lack efficacy for a variety of reasons \cite{Osterholm12}, antigenic drift causes efficacy against a fixed vaccine formulation to decline over time.
 A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
@@
 Top-level priors for $1/\mdssd^2$, $\drift$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
+These diffuse priors were chosen to be non-informative and provide little-to-no weight on the resulting posterior distributions.
 Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored

### 2013-07-24 flux.tex
Minor revision.

@@
 	\caption{\textbf{Pairwise correlations between genetic distance, measured as amino acid mutations or as phylogenetic distance, and antigenic distance for influenza A/H3N2, A/H1N1, B/Vic and B/Yam.} 
-	The top row shows correlations between number of amino acid mutations and average antigenic distance between 10,000 random pairs of viruses.
+	The top row shows correlations between number of amino acid mutations in HA1 and average antigenic distance between 10,000 random pairs of viruses.
 	The bottom row shows correlations between average phylogenetic distance, measured in terms of years, and average antigenic distance between 10,000 random pairs of viruses.

### 2013-07-24 response.tex
Fix #9 - Antigenic drift wording.
Revise to note HA vs HA1 sequences.

@@
 
+\textbf{We have revised this paragraph to make it clear that antigenic drift causes efficacy to decline over time, but that there are other factors that generally limit influenza vaccine efficacy.}
+
 In the second paragraph of the Introduction, the authors refer to ``more evolved viruses.'' Because the concept of directionality in evolution is poorly defined, ``more recent'' would be better than ``more evolved.''
@@
 It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map.
+
 Also, are the phylogenetic trees built on just HA1 or the whole HA protein?
+
+\textbf{The trees were inferred from the whole HA gene when available, but for many viruses only HA1 was available.
+The phylogenetic methods available in BEAST are able to properly deal with this sort of missing data.
+We've made note of this in the text.}
 

### 2013-07-24 response.tex
Fix #17 - phylogenetic trees.

@@
 
+\textbf{The virus phylogenies can be seen to some extent in Figures 2 and 3, i.e.
+phylogenetically distinct clades are apparent.
+However, we agree that it would be helpful to include a cleaner picture of the virus phylogenies.
+We've included a new figure (current figure 4) that shows more standard time-resolved phylogenies.
+We've also included discussion of the negative correlation between antigenic drift and phylogenetic diversity.}
+
 Also, are the phylogenetic trees built on just HA1 or the whole HA protein?

### 2013-07-24 flux.tex
Revise to note HA vs HA1 sequences.
Small reorganization.
Fix #17 - phylogenetic trees.

@@
 
+These patterns of antigenic drift are reflected in the virus phylogenies (Figure~\ref{trees}).
+Influenza A/H3N2 has characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
+The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
+Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift with A/H3N2 showing fast drift and reduced effective population size (Figure~\ref{trees}), as expected from basic epidemiological models \cite{Bedford12}.
+Antigenic drift results in the replacement of antigenically primitive lineages by antigenically advanced lineages, thereby reducing genealogical diversity.
+
+%%% trees %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.85\textwidth]{figures/trees}
+	\caption{\textbf{Time-resolved phylogenetic trees of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	The maximium-clade credibility (MCC) tree is shown for each virus.
+	These trees show genealogical relationships, so that branches are measured in terms of years rather than substitutions.
+	The mean posterior scaled effective population size $N_e \tau$, measured in years, and its 95\% credible interval is shown for each virus.} 
+	\label{trees} 
+\end{figure}
+
+\subsection*{Punctuated evolution and its epidemiological consequences}
+
 We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}).
@@
 \end{figure}
-
-\subsection*{Epidemiological consequences of antigenic drift}
 
@@
 If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD.
+Many strains had full length HA sequences, while other strains only possessed HA1 sequences.
 Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.

### 2013-07-24 response.tex
Fix #6 - model description.

@@
 
+\textbf{This is a very helpful suggestion.
+We've included a new column in Table 1 labeled `Data' that specifies the combination of data sources that go into the model (HI vs HI/year vs HI/year/seq).
+We've also clarified that serum and virus effects act to give a baseline expectation for titer when virus and serum have identical antigenic locations.}
+
 The first figure of the paper (Figure 1) is basically methodological rather than presenting a biological result.

### 2013-07-24 flux.tex
Fix #6 - model description.

@@
 
-We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations.
-In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera, and virus and serum locations follow an uninformative diffuse normal prior as described in the `Antigenic cartography' section of Materials and methods.
+We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations as described in the `Antigenic cartography' section of Materials and methods.
+In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera and give the baseline expectation for titer when virus and serum are antigenically identical.
+Virus and serum locations follow an uninformative diffuse normal prior. 
 After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
@@
 	\makebox[\textwidth][c]{
-	\begin{tabular}{ c c c c c c c c c } 
+	\small
+	\begin{tabular}{ c c c c c c c c c c } 
 	\hline
-	\multicolumn{5}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	\multicolumn{6}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
+	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline	
-	1		&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
-	2		&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
-	3		&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
-	4		&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
-	5		&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
-	6		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
-	7		&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
-	8		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
-	9		&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
-	10		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
+	1		&	HI			&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
+	2		&	HI			&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
+	3		&	HI			&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
+	4		&	HI			&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
+	5		&	HI			&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
+	6		&	HI/year		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
+	7		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
+	8		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
+	9		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
+	10		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
 	\hline
@@
 We refer to these estimates as serum effects and virus effects, respectively.
+These effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
 We find that including these effects decreases test error further, resulting in comparable improvements from estimating virus and serum effects individually and giving an improvement of between 0.03 to 0.13 log$_2$ HI titers when estimating both virus and serum effects together (models 8--10; Table \ref{errortable}).

### 2013-07-24 response.tex
Fix #3 - serum effects.

@@
 Could the authors correlate the estimated serum effects back to the originating source for that serum to check?
+
+\textbf{We would suggest that a serum with a larger effect contains a more concentrated and active set of antibodies than a serum with a smaller effect.
+The increased number of antibodies could be due to experimental variation in serum extraction and processing or due to variation in immune response and timing between ferrets, i.e.
+some ferrets may mount an immediate and strong immune response, while others may mount a weaker response.
+This causes variation in overall strength of the serum.
+We try to control for the overall strength when looking at patterns of cross-reactivity through the use of `serum effects'.
+Some differences in serum effects could be tracked down to particular studies, but this work is beyond the scope of the current study.
+Here, we've essentially treated serum effect as a nuisance parameter.
+We've revised the text to discuss this as well as describe why estimating serum effects improves model performance over fixing serum effects.}
 

### 2013-07-24 flux.tex
Fix #3 - serum effects.

@@
 
-We further extend the model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
-We refer to these estimates as serum effects and virus effects, respectively.
-These effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
-We find that including these effects decreases test error further, resulting in comparable improvements from estimating virus and serum effects individually and giving an improvement of between 0.03 to 0.13 log$_2$ HI titers when estimating both virus and serum effects together (models 8--10; Table \ref{errortable}).
+We further extend the model by estimating `serum effects' to quantify the overall strength reactivity of each serum, rather than fixing serum effects at maximum titers.
+Serum effects likely derive from experimental variation in serum extraction and processing or from variation in immune responses between ferrets.
+Serum effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
+However, if serum effect is fixed to the serum's maximum titer, this will often not be the case, as the virus giving the maximum titer is likely to be antigenically distinct from the serum.
+Thus, fixing serum effects will tend to under-estimate effect size; we observe a mean effect of 10.42 log$_2$ HI titers for A/H3N2 when fixing serum effects and a mean of 10.94 when estimating serum effects.
+We find that estimating serum effects improves test error further (model 8, Table \ref{errortable}), with improvements of between 0.01 and 0.12 log$_2$ HI titers.
+
+Additionally, we include and estimate `virus effects' in an analogous fashion, which are intended to represent differences in overall reactivity between viruses.
+With virus effects included, baseline titer derives from both the virus and the serum used in the HI reaction.
+We find that including virus effects further improves test error, either alone with fixed serum effects (model 9, Table \ref{errortable}) or combined with estimating serum effects (model 10, Table \ref{errortable}).
+With fixed serum effects, the inclusion of virus effects results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum effects, the inclusion of virus effects results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
 

### 2013-07-25 flux.tex
Work on issue #24, getting rid of branch-specific D.

@@
 Using HI data from influenza A subtypes H3N2 and H1N1 and influenza B lineages Victoria and Yamagata, we determine patterns of antigenic drift across lineages and show that antigenic drift drives incidence rates and mediates interference between influenza lineages.
-Additionally, we describe the selective underpinnings of differences in antigenic drift across lineages.
+%Additionally, we describe the selective underpinnings of differences in antigenic drift across lineages.
 This work makes possible substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
@@
 
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at a rate of 1.01 antigenic units per year, with a 95\% highest posterior density (HPD) interval of 0.98--1.04 (Figure~\ref{drift}).
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at an estimated rate of 1.01 antigenic units per year (Figure~\ref{drift}, Table~\ref{drifttable}).
 However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
@@
 
-We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
-Influenza A/H1N1 evolved at a rate of 0.62 units per year (HPD 0.56--0.67), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
-Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.42 (HPD 0.32--0.51) units per year.
-Influenza B/Yamagata evolved slower still, with an average rate of 0.32 (HPD 0.25--0.39) units per year.
+We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}, Table~\ref{paramtable}).
+Influenza A/H1N1 evolved at a rate of 0.62 units per year, but showed a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
+Influenza B/Victoria and B/Yamagata evolved slower still, with mean estimated rates 0.42 units per year and 0.32 units per year, respectively.
 Punctuated evolution is less obvious in B/Yam and B/Vic compared to A/H3N2 and A/H1N1, but antigenic clusters are still apparent, with recent transitions to the Brisbane/08 cluster in B/Vic \cite{Barr10} and to the Wisconsin/10 cluster in B/Yam \cite{Klimov12}.
 Interestingly, a minor lineage of B/Vic, denoted B/Hubei-Songzi/51/2008 \cite{Barr10}, has persisted through 2011, while remaining antigenically distinct from B/Brisbane/60/2008 viruses (Figure~\ref{drift}).
+Interestingly, although we observe significantly different drift rates between lineages, we observe less variation in diffusion volatility (Table~\ref{paramtable}).
+This is reflected in Figure \ref{drift}, where all four lineages exhibit similar levels of standing antigenic variation, despite A/H3N2 drifting more quickly in antigenic phenotype.
+
+%%% paramtable %%%
+\begin{table}[h]
+	\centering
+	\caption{\textbf{Estimates of drift rate $\drift$ (in units per year), diffusion volatility $\virussd^2$ (in units$^2$ per year) and scaled effective population size $N_e\tau$ (in years) for influenza A/H3N2, A/H1N1, B/Vic and B/Yam including posterior means and 95\% highest posterior density intervals.}}
+	\label{paramtable}	
+	\begin{tabular}{ c c c c} 
+	\hline
+	Lineage	&	Drift $\drift$  	& 	Volatility $\virussd^2$	&	Effective pop size $N_e\tau$	\\
+	\hline		
+	A/H3N2	&	1.01 (0.98--1.04)	&	1.25 (0.98--2.35) 		&	5.03 (4.42--5.73)	\\
+	A/H1N1	&	0.62 (0.56--0.67)	&	0.92 (0.65--1.56) 		&	6.38 (4.99--8.12)	\\
+	B/Vic	&	0.42 (0.32--0.51)	&	1.22 (0.85--2.25) 		&	10.40 (8.42--12.80)	\\
+	B/Yam	&	0.32 (0.25--0.39)	&	0.71 (0.46--1.36) 		&	9.48 (7.76--11.50)	\\
+	\hline
+	\end{tabular}	
+\end{table}
 
@@
 The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
-Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift with A/H3N2 showing fast drift and reduced effective population size (Figure~\ref{trees}), as expected from basic epidemiological models \cite{Bedford12}.
+Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift (Table~\ref{paramtable}), with A/H3N2 showing fast drift and reduced effective population size as expected from basic epidemiological models \cite{Bedford12}.
 Antigenic drift results in the replacement of antigenically primitive lineages by antigenically advanced lineages, thereby reducing genealogical diversity.
@@
 
-\subsection*{Conversion of antigenic mutation into fixed differences}
+%\subsection*{Conversion of antigenic mutation into fixed differences}
 
-We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
-Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
-Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Materials and methods).
-This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
-Higher diffusion coefficients mean a lineage is diffusing faster through antigenic space.
-We compare diffusion coefficients between trunk branches and side branches, where trunk branches are defined as branches ancestral to the most recent virus sample and side branches are defined as those branches not ancestral to the most recent virus sample.
+%We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
+%Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
+%Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+%Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
+%To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Materials and methods).
+%This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
+%Higher diffusion coefficients mean a lineage is diffusing faster through antigenic space.
+%We compare diffusion coefficients between trunk branches and side branches, where trunk branches are defined as branches ancestral to the most recent virus sample and side branches are defined as those branches not ancestral to the most recent virus sample.
 
 %%% diffusiontable %%%
-\begin{table}[h]
-	\centering
-	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including posterior means and 50\% highest posterior density intervals.}}
-	\label{diffusiontable}	
-	\begin{tabular}{ c c c c } 
-	\hline
-	Lineage	&	Trunk branch $D$ 	& 	Side branch $D$		& 	Ratio \\
-	\hline		
-	A/H3N2	&	0.90 (0.79--0.97)	&	0.62 (0.57--0.65)	&	1.45 (1.31--1.57) \\
-	A/H1N1	&	0.62 (0.50--0.71)	&	0.44 (0.38--0.48)	&	1.45 (1.18--1.70) \\
-	B/Vic	&	0.63 (0.51--0.71)	&	0.62 (0.54--0.68)	&	1.02 (0.87--1.14) \\
-	B/Yam	&	0.35 (0.29--0.40)	&	0.33 (0.28--0.36)	&	1.09 (0.93--1.23) \\
-	\hline
-	\end{tabular}	
-\end{table}
+%\begin{table}[h]
+%	\centering
+%	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including posterior means and 50\% highest posterior density intervals.}}
+%	\label{diffusiontable}	
+%	\begin{tabular}{ c c c c } 
+%	\hline
+%	Lineage	&	Trunk branch $D$ 	& 	Side branch $D$		& 	Ratio \\
+%	\hline		
+%	A/H3N2	&	0.90 (0.79--0.97)	&	0.62 (0.57--0.65)	&	1.45 (1.31--1.57) \\
+%	A/H1N1	&	0.62 (0.50--0.71)	&	0.44 (0.38--0.48)	&	1.45 (1.18--1.70) \\
+%	B/Vic	&	0.63 (0.51--0.71)	&	0.62 (0.54--0.68)	&	1.02 (0.87--1.14) \\
+%	B/Yam	&	0.35 (0.29--0.40)	&	0.33 (0.28--0.36)	&	1.09 (0.93--1.23) \\
+%	\hline
+%	\end{tabular}	
+%\end{table}
 
-Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
-We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
-Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
-Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
-We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
-Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
-Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
-Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.
+%Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
+%We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
+%Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
+%Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
+%We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
+%Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
+%Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
+%Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.
 
-In this context, it is especially interesting that adaptive evolution in the HA1 protein determined through analysis of nonsynonymous and synonymous substitutions is substantially stronger in A/H3N2 than in A/H1N1 \cite{Wolf06}; H3 appears to accumulate adaptive fixations at $\sim$150\% the rate of H1 \cite{Bhatt11}.
-Here, we observe that A/H3N2 has a diffusion coefficient $\sim$141\% greater than A/H1N1 along side branches of the phylogeny.
-These results suggest that the greater rate of adaptive fixation in A/H3N2 may be simply due to a greater abundance of new antigenic mutations to fix.
+%In this context, it is especially interesting that adaptive evolution in the HA1 protein determined through analysis of nonsynonymous and synonymous substitutions is substantially stronger in A/H3N2 than in A/H1N1 \cite{Wolf06}; H3 appears to accumulate adaptive fixations at $\sim$150\% the rate of H1 \cite{Bhatt11}.
+%Here, we observe that A/H3N2 has a diffusion coefficient $\sim$141\% greater than A/H1N1 along side branches of the phylogeny.
+%These results suggest that the greater rate of adaptive fixation in A/H3N2 may be simply due to a greater abundance of new antigenic mutations to fix.
 
-We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
-In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
-These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
-Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
-Still, it is clear that more work is necessary to fully clarify the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
+%We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
+%In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
+%These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
+%Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
+%Still, it is clear that more work is necessary to fully clarify the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
 
@@
 
-\subsection*{Estimation of diffusion coefficient}
+%\subsection*{Estimation of diffusion coefficient}
 
-The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
-Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
-\begin{equation}
-	X_t \sim \normal(0, \sigma^2 t),
-\end{equation}
-where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
-\begin{equation}
-	\mathrm{E}[X_t^2] = \sigma^2 t.
-\end{equation}
-In moving to a two-dimensional diffusion
-\begin{equation}
-	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
-\end{equation}
-where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
-\begin{equation}
-	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
-\end{equation}
-Such diffusions are often characterized with diffusion coefficient $D$.
-In this case
-\begin{equation}
-	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
-\end{equation}
-so that expected squared displacement follows from $D$ alone
-\begin{equation}
-	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
-\end{equation}
+%The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
+%Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
+%\begin{equation}
+%	X_t \sim \normal(0, \sigma^2 t),
+%\end{equation}
+%where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
+%\begin{equation}
+%	\mathrm{E}[X_t^2] = \sigma^2 t.
+%\end{equation}
+%In moving to a two-dimensional diffusion
+%\begin{equation}
+%	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
+%\end{equation}
+%where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
+%\begin{equation}
+%	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
+%\end{equation}
+%Such diffusions are often characterized with diffusion coefficient $D$.
+%In this case
+%\begin{equation}
+%	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
+%\end{equation}
+%so that expected squared displacement follows from $D$ alone
+%\begin{equation}
+%	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
+%\end{equation}
 
-Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
-\begin{equation} \label{pybusest}
-	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
-\end{equation}
-in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
-However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
-Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
-\begin{eqnarray} \label{ourest}
-	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
-	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
-	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
-\end{eqnarray}
-This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
-Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
-When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
+%Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
+%\begin{equation} \label{pybusest}
+%	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
+%\end{equation}
+%in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
+%However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
+%Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
+%\begin{eqnarray} \label{ourest}
+%	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
+%	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
+%	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
+%\end{eqnarray}
+%This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
+%Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
+%When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
 

### 2013-11-13 response.tex
Work on issue #24, getting rid of branch-specific D.
More work on issue #24.

@@
 
+\textbf{We agree that applying the standard measure of diffusion coefficient D based on displacement vs time, without accounting for advection was problematic.
+We have substantial revised the text to focus on direct parameter inference of drift parameter $\drift$ and diffusion volatility parameter $\virussd$, properly separating advection and volatility in the diffusion process.
+The section ``Antigenic evolution across influenza lineages'' most reflects these changes, including the addition of a new table 2.}
+
 How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone?

### 2013-11-13 flux.tex

@@
 HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 14 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory.
+Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory (see Materials and Methods).
 To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Materials and methods).
@@
 
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at an estimated rate of 1.01 antigenic units per year (Figure~\ref{drift}, Table~\ref{drifttable}).
+We find that influenza A/H3N2 evolved along antigenic dimension 1 at an estimated rate of 1.01 antigenic units per year (Figure~\ref{drift}, Table~\ref{paramtable}).
 However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
@@
 Interestingly, a minor lineage of B/Vic, denoted B/Hubei-Songzi/51/2008 \cite{Barr10}, has persisted through 2011, while remaining antigenically distinct from B/Brisbane/60/2008 viruses (Figure~\ref{drift}).
-Interestingly, although we observe significantly different drift rates between lineages, we observe less variation in diffusion volatility (Table~\ref{paramtable}).
+Although we observe significantly different drift rates between lineages, we observe less variation in diffusion volatility (Table~\ref{paramtable}).
 This is reflected in Figure \ref{drift}, where all four lineages exhibit similar levels of standing antigenic variation, despite A/H3N2 drifting more quickly in antigenic phenotype.
@@
 These patterns of antigenic drift are reflected in the virus phylogenies (Figure~\ref{trees}).
-Influenza A/H3N2 has characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
+Influenza A/H3N2 has a characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
 The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.

### 2013-11-13 response.tex
Work on issue #4.
Text and response is finished, however need to include MDS tutorial
on the BEAST wiki.

@@
 
-\textbf{We agree that applying the standard measure of diffusion coefficient D based on displacement vs time, without accounting for advection was problematic.
+\textbf{We agree that applying the standard measure of diffusion coefficient $D$ based on displacement vs time, without accounting for advection was problematic.
 We have substantial revised the text to focus on direct parameter inference of drift parameter $\drift$ and diffusion volatility parameter $\virussd$, properly separating advection and volatility in the diffusion process.

### 2013-11-13 response.tex
More on issue #4.

@@
 
-\textbf{We apologize for this oversight.  The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.}
+\textbf{We apologize for this oversight.
+The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.
+We've also included a reference to the BEAST tutorials on implementing these cartographic models.}
 

### 2013-11-14 flux.tex
Work on issue #4.
Text and response is finished, however need to include MDS tutorial
on the BEAST wiki.
Significant revision of drift and incidence analysis per #1 and #2.

@@
 
-We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}).
+We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
 We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
 There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
-We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
+We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}B), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
 This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
@@
 	\includegraphics[width=1.0\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2009 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Time series of year-to-year antigenic drift across influenza clades from 1992 to 2011 and seasonal incidence from 1998/1999 to 2008/2009.
-	Colored lines represent year-to-year antigenic drift, and drift for a virus clade for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
-	For example, 2000 represents difference in antigenic dimension 1 between 1999 and 2000.
+	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
+	(A) Timeseries of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
+	Colored lines represent year-to-year antigenic drift, where drift for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
+	For example, 2000 represents difference in antigenic dimension 1 between viruses from 1999 and 2000.
 	Error bars represent 50\% Bayesian credible intervals of year-to-year drift.
-	Gray dotted lines represent clade-specific USA incidence taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a clade for each season.
+	Gray dotted lines represent lineage-specific seasonal incidence in the USA taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a lineage for each season.
 	Here, 2000 represents the 2000/2001 influenza season.
-	(B) Histograms of year-to-year antigenic drift across influenza clades summarizing observations from 1992 to 2011.
+	(B) Distribution of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
 	} 
@@
 
-We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
-We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Materials and methods).
-We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of viruses detected in the USA during this time period.
-We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).
+We investigate the relationship between rates of antigenic drift and seasonal incidence in the USA in A/H3N2, A/H1N1, B/Vic and B/Yam.
+We measure seasonal incidence from USA CDC data for each virus lineage (A/H3N2, A/H1N1, B/Vic, B/Yam) by taking the average influenza-like (ILI) percentage in a season and multiplying this by the relative proportion of virus isolations attributable to a particular influenza lineage in a season (see Materials and methods).
+This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
+We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic.
+We begin by comparing overall rates of antigenic drift (Figure~\ref{drift}, Table~\ref{paramtable}) to overall levels of seasonal incidence across influenza lineages, finding a significant correlation between rate of antigenic drift and relative incidence across the four lineages (Pearson correlation, $r=0.97$, $p = 0.041$).
 
-We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
-We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
-We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of viruses attributable to a clade for each season.
-This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
-In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
-We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{jumps}, Figure~\ref{incidence}A).
-Here, within-clade drift explains 35\% of the variance in standardized incidence and represents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).
+We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and lineage-specific incidence, comparing incidence in a seasonal to antigenic drift of viruses coming into this season (Figure~\ref{jumps}A).
+For example, we compare antigenic drift of viruses from 2000 to 2001 to incidence in the 2001/2002 season.
+Within each virus lineage, we find that years with pronounced antigenic drift tend to show increased incidence (Figure~\ref{corr}), finding Pearson correlation coefficients of 0.51, 0.29, 0.44 and 0.14 for A/H3N2, A/H1N1, B/Vic and B/Yam respectively.
+We calculated significance using bootstrap permutation tests finding $p$-values of 0.056, 0.201, 0.097 and 0.341 respectively.
+We applied a similar bootstrap permutation test to calculate the significance of finding the observed levels of correlation across all four lineages, arriving at a $p$-value of 0.018.
 
-%%% incidence %%%
+%%% corr %%%
 \begin{figure}[h]
 	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/incidence}
-	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
-	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
-	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
-	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
-	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
-	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
+	\includegraphics[width=0.7\textwidth]{figures/corr}
+	\caption{\textbf{Relationship between antigenic drift and seasonal incidence for years 1998 to 2009 in influenza A/H3N2, A/H1N1, B/Vic and B/Yam.}
+	Antigenic drift from year $i-1$ to year $i$ is compared to incidence in the season $i$ / $i+1$.
+	For example, year-to-year antigenic drift from 2000 to 2001 is measured against incidence in the 2001/2002 season. 
 	} 
-	\label{incidence} 
+	\label{corr} 
 \end{figure}
 
-We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
-This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
-We fit a linear model that predicts standardized incidence based on drift within each of the four lineages, according to the following equations
-\begin{eqnarray}
-	\incclade{H3} &=& \bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
-	\incclade{H1} &=& \bsister \, \driftclade{H3} + \bwithin \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam} \nonumber \\
-	\incclade{Vic} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bwithin \, \driftclade{Vic} + \bsister \, \driftclade{Yam} \nonumber \\	
-	\incclade{Yam} &=& \bother \, \driftclade{H3} + \bother \, \driftclade{H1} + \bsister \, \driftclade{Vic} + \bwithin \, \driftclade{Yam},
-\end{eqnarray}
-where $y$ represents standardized incidence, $x$ represents drift in antigenic units, $\bwithin$ scales within-clade drift, $\bsister$ scales drift across sister viruses of the same type, and $\bother$ scales drift across viruses of differing type (Table~\ref{interferencetable}).
-For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2 scaled by $\bwithin$, drift in A/H1N1 scaled by $\bsister$ and by drift in B/Vic and B/Yam scaled by $\bother$.
-We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
-We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations.
-The term for across lineage interference $\bother$ is small in magnitude, but remains statistically significant (Table~\ref{interferencetable}).
+Combining observations of year-to-year drift vs seasonal incidence across all four lineages, yields a simple linear model where one unit of antigenic drift results in an increase of 0.8\% in lineage-specific ILI.
 
-%%% interferencetable %%%
-\begin{table}[h]
-	\centering
-	\caption{\textbf{Estimates of regression coefficients for the relationship of clade-specific USA incidence to antigenic drift within and across clades, where standardized incidence in A/H3N2 is predicted by $\bwithin \, \driftclade{H3} + \bsister \, \driftclade{H1} + \bother \, \driftclade{Vic} + \bother \, \driftclade{Yam}$ and standardized incidence in other clades follows an analogous relationship.}}
-	\label{interferencetable}	
-	\begin{tabular}{ c c c c } 
-	\hline
-	Parameter	&	Estimate	&	Standard error 	& 	$p$-value \\
-	\hline		
-	$\bwithin$	&	$1.26$		&	0.22			&	$1.1 \times 10^{-6}$ \\
-	$\bsister$	&	$-0.61$		&	0.22			&	$0.008$ \\
-	$\bother$	&	$-0.32$		&	0.11			&	$0.005$ \\
-	\hline
-	\end{tabular}
-\end{table}
-
-These findings clearly demonstrate the epidemiological importance of antigenic drift and show that incidence across  A/H3N2, A/H1N1, B/Vic and B/Yam is strongly influenced by levels of antigenic drift within each clade, consistent with antigenic drift increasing the susceptible pool available to a clade.
-The finding of a negative correlation between incidence within one clade and antigenic drift in other clades supports the hypothesis that epidemiological interference exists between clades \cite{Ferguson03, Goldstein11}.
-The immunological basis of between-clade interference are not entirely clear, however cytotoxic T cells (CTLs) \cite{Voeten00}, antibodies that broadly neutralize HA \cite{Corti10} and antibodies to internal proteins \cite{Laidlaw13} have been hypothesized as possible mechanisms.
-Interestingly, we find that interference between sister clades, e.g.\ B/Vic and B/Yam, appears stronger than interference between more divergent clades, e.g.\ B/Vic and A/H3N2 (Table~\ref{interferencetable}, $\bsister = -0.61$ vs $\bother = -0.32$).
-If interference is mediated by CTLs or by antibodies to evolutionary conserved epitopes, then we would expect interference to decrease with evolutionary distance \cite{Voeten00}.
-
-Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
-However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
-Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.
+%Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
+%However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
+%Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.
 
@@
 Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Incidence data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
+Tutorials for implementing these models can be found at http://beast.bio.ed.ac.uk/Tutorials.
+Incidence data and HI data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
 

### 2013-11-14 response.tex
Significant revision of drift and incidence analysis per #1 and #2.

@@
 
+\textbf{This is a very astute criticism.
+We see how differing overall incidence between clades could have given an artifactual signal in our previous year-to-year drift vs incidence comparisons.
+In this revision, we have reanalyzed the data in the way suggested; we show that year-to-year antigenic drift and incidence are correlated within each clade individually, arriving at correlation coefficients of 0.51, 0.29, 0.44 and 0.14 for A/H3N2, A/H1N1, B/Vic and B/Yam respectively.
+None of these correlations are significant on their own, however observing four correlation coefficients of this magnitude is highly unlikely under a null model derived from bootstrap permutations ($p = 0.018$).
+Because the increase in incidence tends to follow periods of pronounced antigenic drift, we conclude that there appears to a be causal relationship between antigenic change and increased incidence.}
+
+\textbf{However, in redoing this analysis on a lineage-by-lineage basis, we lost much of the signal for interference between lineages.
+We think there may still be something there, but the nuanced analysis that this issue deserves seems beyond the scope of the paper.
+We have decided to instead drop the discussion of interference between lineages.}
+
 The incidence data are poorly explained.
@@
 In either case, what are the justifications for using absolute versus relative incidence?
+
+\textbf{In revising this section in the manuscript, we have made the source of the incidence data and its construction much more clear.
+In addition, we have removed the relative vs absolute incidence confusion, sticking instead with one measure: ILI $\times$ proportion of viral isolates attributable to a lineage.}
+
 Also, why do they not analyze incidence prior to 1998/1999?
+
+\textbf{The 1998/1999 season is the earliest we have that properly distinguishes B/Vic and B/Yam in the CDC data.}
+
 Finally, because each influenza season spans two calendar years, why do they not break up the data by season rather than year?
+
+\textbf{This would be much preferable.
+However, many virus isolates used in this analysis lack temporal resolution beyond the year of sampling.
+Because of this, we decided to stick with just year of isolation for the present analysis.}
 

### 2013-11-14 flux.tex

@@
 Here, we extend previous approaches to antigenic cartography, and simultaneously characterize antigenic and genetic evolution by modeling the diffusion of antigenic phenotype over a shared virus phylogeny.
-Using HI data from influenza A subtypes H3N2 and H1N1 and influenza B lineages Victoria and Yamagata, we determine patterns of antigenic drift across lineages and show that antigenic drift drives incidence rates and mediates interference between influenza lineages.
-%Additionally, we describe the selective underpinnings of differences in antigenic drift across lineages.
+Using HI data from influenza A subtypes H3N2 and H1N1 and influenza B lineages Victoria and Yamagata, we determine patterns of antigenic drift across viral lineages, showing that A/H3N2 evolves faster and in a more punctuated fashion than other influenza lineages.
+We also show that year-to-year antigenic drift appears to drive incidence patterns within each influenza lineage.
 This work makes possible substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
@@
 
+Thus, we observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
+Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
+Models of influenza evolution acribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+Without more detailed epidemiological modeling, the present study cannot conclusively distinguish between these causal possibilities.
+
 %%% trees %%%
@@
 We investigate the relationship between rates of antigenic drift and seasonal incidence in the USA in A/H3N2, A/H1N1, B/Vic and B/Yam.
-We measure seasonal incidence from USA CDC data for each virus lineage (A/H3N2, A/H1N1, B/Vic, B/Yam) by taking the average influenza-like (ILI) percentage in a season and multiplying this by the relative proportion of virus isolations attributable to a particular influenza lineage in a season (see Materials and methods).
+We measure seasonal incidence from USA CDC influenza surveillance reports for each virus lineage (A/H3N2, A/H1N1, B/Vic, B/Yam) by taking the average influenza-like (ILI) percentage in a season and multiplying this by the relative proportion of virus isolations attributable to a particular influenza lineage in a season (see Materials and methods).
 This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
@@
 
-Combining observations of year-to-year drift vs seasonal incidence across all four lineages, yields a simple linear model where one unit of antigenic drift results in an increase of 0.8\% in lineage-specific ILI.
+Although the general correlation between rate of antigenic drift and incidence across the four influenza lineages may not represent a causal relationship, the within-lineage year-to-year correlations suggest a causal relationship as increases in incidence tend to occur subsequent to periods of pronounced antigenic drift.
 
@@
 In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
-We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each clade and interference between clades.
-We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
-Influenza A/H3N2 benefits from a higher rate of new antigenic mutation coupled with strong adaptive evolution and evolves in antigenic phenotype faster than A/H1N1, B/Vic and B/Yam.
-Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
-We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
+We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each clade and influences incidence patterns.
+
+%We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
+%Influenza A/H3N2 benefits from a higher rate of new antigenic mutation coupled with strong adaptive evolution and evolves in antigenic phenotype faster than A/H1N1, B/Vic and B/Yam.
+%Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
+%We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
 
@@
 In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
-
 Identifying viruses that will come to predominate in the global virus population while they are still at low frequency remains an enormous challenge.

### 2013-11-14 flux.tex
Fix #5.  Cleanup language to remove references to "clades".

@@
 Here, we extend previous approaches to antigenic cartography, and simultaneously characterize antigenic and genetic evolution by modeling the diffusion of antigenic phenotype over a shared virus phylogeny.
-Using HI data from influenza A subtypes H3N2 and H1N1 and influenza B lineages Victoria and Yamagata, we determine patterns of antigenic drift across viral lineages, showing that A/H3N2 evolves faster and in a more punctuated fashion than other influenza lineages.
+Using HI data from influenza lineages A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine patterns of antigenic drift across viral lineages, showing that A/H3N2 evolves faster and in a more punctuated fashion than other influenza lineages.
 We also show that year-to-year antigenic drift appears to drive incidence patterns within each influenza lineage.
@@
 
-Before 2009, there were four major clades or lineages of influenza circulating within the human population: the H3N2 and H1N1 subtypes of influenza A, and the Victoria and Yamagata lineages of influenza B.
+Before 2009, there were four major lineages of influenza circulating within the human population: the H3N2 and H1N1 subtypes of influenza A, and the Victoria and Yamagata lineages of influenza B. 
 In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
@@
 Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza clades.
+We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza lineages.
 
@@
 
-We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}, Table~\ref{paramtable}).
+We find that other lineages of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}, Table~\ref{paramtable}).
 Influenza A/H1N1 evolved at a rate of 0.62 units per year, but showed a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
@@
 In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
-We use this approach to characterize competitive dynamics across influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution drives strain replacement in each clade and influences incidence patterns.
+We use this approach to characterize competitive dynamics across influenza lineages A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution within each lineage drives strain replacement and contributes to seasonal incidence patterns.
 
@@
 Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
-In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
+In this case, we estimate the relative proportion of A/H3N2 of the four lineages as $0.66 \times 0.13 = 0.09$.
 Similar calculations were performed for A/H1N1, B/Vic and B/Yam.

### 2013-11-14 response.tex
Fix #5.  Cleanup language to remove references to "clades".
Fix #22 and fix #23.
No correlation between sample counts and estimates of year-to-year drift.

@@
 \usepackage{float}
+\usepackage{hyperref}
 
@@
 \maketitle
+
+Original reviewer criticisms are in plain text.
+Our responses follow in \textbf{bold}.
+A detailed list of every criticism and exactly what was revised in the manuscript in response to this criticism is available at \url{https://github.com/trvrb/flux/issues?state=closed}.
 
@@
 I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.
+
+\textbf{We've cleaned up our language regarding clades and lineages in the text.
+We consider A/H3N2, A/H1N1, B/Vic and B/Yam to each represent separate \textit{lineages} of influenza.
+We no longer refer to separate \textit{clades}.}
 
@@
 
+\textbf{We tested to see if there might be an bias in the estimates of year-to-year antigenic drift based on number of virus isolates collected by correlating correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
+We found no significant correlation (Pearson's $r = -0.01$, $p = 0.411$), suggesting our results to be unbiased with regard to sample count.
+We have included this analysis in the manuscript.}
+
 Likewise, in Fig.
@@
 The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.
+
+\textbf{The above analysis applies to this correlation as well.}
 

### 2013-11-14 flux.tex
Fix #22 and fix #23.
No correlation between sample counts and estimates of year-to-year drift.

@@
 This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
+We tested to see if there could be an artifactual effect on the measurement of year-to-year drift based on the number of viral isolates collected by correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
+We found no significant correlation (Pearson's $r = -0.01$, $p = 0.411$), suggesting our results to be unbiased with regard to sample count.
 

### 2013-12-06 flux.tex
Include description of example XMLs.
Fixed MAS grant numbers

@@
 Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Tutorials for implementing these models can be found at http://beast.bio.ed.ac.uk/Tutorials.
+More details on implementing these models can be found at https://github.com/trvrb/flux/tree/master/example-xmls.
 Incidence data and HI data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
@@
 TB was supported by a Newton International Fellowship from the Royal Society. 
-MAS was supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
+MAS was supported by National Institutes of Health grants R01 HG006139 and R01 AI107034 and National Science Foundation grant DMS126153 and IIS1251151.
 The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.

### 2013-12-06 reresponse.tex
Include stem of rereview response.

--- 2013-11-14 flux.tex
@@
 \usepackage{geometry}
-\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textwidth=15.25cm} % 15.25cm for single-space, 16.25cm for double-space
 \geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space
@@
 
-%%% TITLE %%%
-\title{\vspace{1.0cm} \Large \bf 
-Integrating influenza antigenic dynamics with molecular evolution
-}
-
-\author[1]{Trevor Bedford}
-\author[2,3,4]{Marc A. Suchard}
-\author[5]{Philippe Lemey}
-\author[1]{Gytis Dudas}
-\author[6]{Victoria Gregory}
-\author[6]{Alan J. Hay}
-\author[6]{John W. McCauley}
-\author[7,8]{Colin A. Russell}
-\author[7,8,9]{Derek J. Smith}
-\author[1,10]{Andrew Rambaut}
-
-\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[4]{Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles CA, USA}
-\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
-\affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
-\affil[7]{Centre for Pathogen Evolution, Department of Zoology, University of Cambridge, Cambridge, UK}
-\affil[8]{WHO Collaborating Center for Modeling, Evolution, and Control of Emerging Infectious Diseases, University of Cambridge, Cambridge, UK}
-\affil[9]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
-\affil[10]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
-
-\date{}
-
 \begin{document}
 
-\maketitle
+\newgeometry{top=4cm}
 
-%%% ABSTRACT %%%
-\begin{abstract}
+Dear eLife editorial board,
 
-Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade host immunity acquired to previous virus strains.
-Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
-Here, we extend previous approaches to antigenic cartography, and simultaneously characterize antigenic and genetic evolution by modeling the diffusion of antigenic phenotype over a shared virus phylogeny.
-Using HI data from influenza lineages A/H3N2, A/H1N1, B/Victoria and B/Yamagata, we determine patterns of antigenic drift across viral lineages, showing that A/H3N2 evolves faster and in a more punctuated fashion than other influenza lineages.
-We also show that year-to-year antigenic drift appears to drive incidence patterns within each influenza lineage.
-This work makes possible substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.
+Thank you for conducting a thorough review of our revised manuscript entitled ``Integrating influenza antigenic dynamics with molecular evolution''.
+We have responded to the remaining reviewer criticisms and believe the manuscript is now suitable for publication in eLife.
 
-\end{abstract}
+Point-by-point responses follow, as well as a PDF showing revisions that have been made since the last submission.
 
-%%% IMPACT %%%
-% Combined evolutionary and antigenic analysis shows that human influenza viruses differ dramatically in rates of antigenic drift and these rates significantly impact seasonal incidence patterns.
+Sincerely,\\
+Trevor Bedford
 
-\pagebreak
+\restoregeometry
 
-%%% INTRODUCTION %%%
-\section*{Introduction}
+\newpage
 
-Seasonal influenza infects between 10\% and 20\% of the human population every year, causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
-While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection.
-The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
-A large proportion of the disease burden of influenza stems from antigenic drift, which allows individuals to be infected multiple times throughout their lives.
-Although influenza vaccines may lack efficacy for a variety of reasons \cite{Osterholm12}, antigenic drift causes efficacy against a fixed vaccine formulation to decline over time.
-A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.
+\section*{Reviewer responses}
 
-Before 2009, there were four major lineages of influenza circulating within the human population: the H3N2 and H1N1 subtypes of influenza A, and the Victoria and Yamagata lineages of influenza B. 
-In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
-In the case of influenza B, Victoria (B/Vic) and Yamagata (B/Yam) refer to antigenically distinct lineages which diverged from a single lineage prior to 1980 \cite{Rota90}.
-Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Wiley81, Nelson07NatRevGenet}. 
-Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over previously established strains.
+Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.
 
-The results of many HI assays across a multitude of virus strains of a single subtype can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
-The antigenic map of influenza A/H3N2 has shown substantial evolution of the influenza virus population since its emergence in 1968. 
-Evolution of antigenic phenotype appears punctuated with episodes of more rapid innovation interspersed by periods of stasis, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
-The process of antigenic drift results in the rapid turnover of the virus population. 
-Although mutation occurs rapidly, genetic diversity among contemporaneous viruses is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
+%%% REREVIEW %%%
+\section*{Rereview}
 
-Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
-An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
-However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
-Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
-We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza lineages.
+\subsection*{Main comments}
 
-%%% RESULTS %%%
-\section*{Results and discussion}
+I like how the authors have included ``virus effects'' alone in the new Table 1.
+At least one ``virus effect'' could be overall receptor avidity, and Plotkin and Hensley have a recent paper (Journal of Virology, 87:9904) showing that avidity can influence antigenic clustering.
+It might be worthwhile to include a sentence on the possibility that ``virus effects'' could be a manifestation of avidity?
 
-\subsection*{Antigenic and evolutionary cartography}
+In Table 2 and the related discussion, the authors discuss the scaled effective population size Ne * tau.
+They never define what tau represents, and I don't think it is safe to assume that the reader will know this -- I certainly don't.
+More interpretation here would be helpful.
 
-In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Materials and methods).
-In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
-In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
-Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
-We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Materials and methods).
+In the ``Punctuated evolution and its epidemiological consequences'' section, the part about subsampling to look at whether drift associates with number of isolates should be expanded.
+I would suggest starting a new paragraph at the current ``We test to see...'' sentence that briefly explains the rationale for why this test is being done (rationale articulated by other reviewer in original critiques).
+I also think that it would be nice to have a table or figure somehow representing the actual results of this analysis.
 
-We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations as described in the `Antigenic cartography' section of Materials and methods.
-In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera and give the baseline expectation for titer when virus and serum are antigenically identical.
-Virus and serum locations follow an uninformative diffuse normal prior. 
-After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
-Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Materials and methods).
-Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Materials and methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
-However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.
+Figure 4 legend refers to ``The mean posterior scaled effective populations... is shown for each virus.'' These are not actually shown in Figure 4, at least not in a way that is obvious to me.
 
-We test model performance by constructing training datasets representing 90\% of the HI measurements for each of the four influenza lineages and test datasets representing the remaining 10\% of the measurements for each lineage.
-By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers.
-We find that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
-We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}), in line with the results of Smith et al.\ \cite{Smith04}.
-Consequently, we specify a two dimensional model in all subsequent analyses.
-The finding of a low-dimensional map across influenza lineages extends previous studies in A/H3N2 \cite{Smith04} and remains an interesting and fundamental empirical observation.
+My comments where the original reason that the authors removed the argument for the across lineage correlation in incidence.
+However, now page 11 has this orphan paragraph beginning ``Although the general correlation between rate of antigenic drift...'' This paragraph doesn't seem to make sense in the context of the presented data any more, as the paper no longer has any information about across lineage correlations, so it can't even be seen what they are saying may not be causal.
+I think this paragraph either needs to be dramatically expended or probably better eliminated.
+Maybe the previous paragraph could then just get a wrap-up sentence interpreting the results to suggest a strong relationship between drift and incidence within each lineage.
 
-%%% errortable %%%
-\begin{table}[h]
-	\centering
-	\caption{\textbf{Average absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
-	\label{errortable}	
-	\makebox[\textwidth][c]{
-	\small
-	\begin{tabular}{ c c c c c c c c c c } 
-	\hline
-	\multicolumn{6}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
-	\hline	
-	1		&	HI			&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
-	2		&	HI			&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
-	3		&	HI			&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
-	4		&	HI			&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
-	5		&	HI			&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
-	6		&	HI/year		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
-	7		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
-	8		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
-	9		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
-	10		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
-	\hline
-	\end{tabular}	
-	}
-\end{table}
+Although the Conclusion is fine, I feel that it might benefit from a paragraph summarizing the main biological results as regards different rates of drift in lineages and the correlation between drift and incidence within lineages.
+These biological results are not really mentioned in the current Conclusion.
 
-Previous work on influenza antigenic and genetic evolution has shown that antigenic distance accumulates with increasing genetic distance \cite{Hay01, Smith04, Russell08}.
-Here, we examine pairwise relationships between viruses and observe a correlation between amino acid mutations and antigenic distance (Figure~\ref{seq_grid}) and a similar correlation between phylogenetic distance and antigenic distance (Figure~\ref{seq_grid}).
-Thus, genetic relationships between viruses provide some predictive power to estimate antigenic distances in the absence of HI data.
-However, the magnitudes of the coefficients of determination $R^2$ are low (Figure~\ref{seq_grid}), suggesting that genetic relationships alone will not completely resolve antigenic distances.
+\subsection*{Minor comments}
 
-%%% map %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/seq_grid}
-	\caption{\textbf{Pairwise correlations between genetic distance, measured as amino acid mutations or as phylogenetic distance, and antigenic distance for influenza A/H3N2, A/H1N1, B/Vic and B/Yam.} 
-	The top row shows correlations between number of amino acid mutations in HA1 and average antigenic distance between 10,000 random pairs of viruses.
-	The bottom row shows correlations between average phylogenetic distance, measured in terms of years, and average antigenic distance between 10,000 random pairs of viruses.
-	Dashed lines show linear model fits, with $R^2$ and slope noted, while solid lines show LOESS fits.
-	Antigenic distances derive from model 2 of Table \ref{errortable}.
-	} 
-	\label{seq_grid} 
-\end{figure}
+The first paragraph of the Introduction refers to ``efficacy against a fixed vaccine formulation to decline over time.'' Although I understand the point, the wording seems off here.
+What declines is the efficacy of the fixed vaccine formulation against circulating viruses, not the efficacy against the vaccine formulation.
 
-Consequently, we seek to flexibly incorporate genetic data by simultaneously modeling antigenic and genetic evolution through an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
-The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
-The antigenic diffusion process includes both systematic drift with time and covariance induced by phylogenetic proximity.
-We examine the effects of including only systematic drift (model 6; Table \ref{errortable}) and systematic drift plus phylogenetic diffusion (model 7; Table \ref{errortable}), finding a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when both processes are included.
-The systematic drift process informs virus and serum locations by dates of isolation and the phylogenetic diffusion process informs virus locations by genetic sequences.
-Thus, in these models, antigenic locations are inferred using both genetic data and HI data and will differ from locations inferred from HI data alone.
-If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in determining antigenic locations (as may be the case for B/Vic and B/Yam).
+Last sentence of second paragraph of abstract: might be better just to say ``an advantage'' rather than ``a transmission advantage.'' Although drift leads to an advantage that presumably improves R and so in a sense does increase transmission, this advantage probably is acting at the step of viral replication rather than the act of transmission per se.
 
-We further extend the model by estimating `serum effects' to quantify the overall strength reactivity of each serum, rather than fixing serum effects at maximum titers.
-Serum effects likely derive from experimental variation in serum extraction and processing or from variation in immune responses between ferrets.
-Serum effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
-However, if serum effect is fixed to the serum's maximum titer, this will often not be the case, as the virus giving the maximum titer is likely to be antigenically distinct from the serum.
-Thus, fixing serum effects will tend to under-estimate effect size; we observe a mean effect of 10.42 log$_2$ HI titers for A/H3N2 when fixing serum effects and a mean of 10.94 when estimating serum effects.
-We find that estimating serum effects improves test error further (model 8, Table \ref{errortable}), with improvements of between 0.01 and 0.12 log$_2$ HI titers.
+Bottom of page 3 says ``of lower of higher dimension.'' The second ``of'' should be an ``or.''
 
-Additionally, we include and estimate `virus effects' in an analogous fashion, which are intended to represent differences in overall reactivity between viruses.
-With virus effects included, baseline titer derives from both the virus and the serum used in the HI reaction.
-We find that including virus effects further improves test error, either alone with fixed serum effects (model 9, Table \ref{errortable}) or combined with estimating serum effects (model 10, Table \ref{errortable}).
-With fixed serum effects, the inclusion of virus effects results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum effects, the inclusion of virus effects results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
+On page 9, ``acribe differences'' should be ``ascribe differences.''
 
-We find that the average absolute error in predicted log$_2$ HI titer is nearly constant with antigenic distance (Pearson correlation, $r = 0.098$), thus supporting our model assumption that the drop in log$_2$ titer is proportional to the Euclidean distance separating viruses and sera on the antigenic map.
-Additionally, we find that the absolute error in predicted titer is nearly constant with time (Pearson correlation, $r = -0.007$).
-Antigenic locations inferred by the model are well resolved; estimates of antigenic distance between pairs of viruses show relatively little variation across the posterior.
-We estimate that virus distances have, on average, a 50\% credible interval of $\pm0.45$ antigenic units for A/H3N2, $\pm0.57$ units for A/H1N1, $\pm0.76$ units for B/Vic, and $\pm0.65$ units for B/Yam.
-
-We find strong correspondence between our results and previous results by Smith et al.\ \cite{Smith04}, with equivalent models producing globally consistent antigenic maps and other models producing locally consistent maps with a small degree of global inconsistency (Figures~\ref{smith_comparison}--\ref{smith_comparison_effects}).
-When implementing the same underlying model, differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
-Additionally, the BMDS method improves flexibility, allowing extensions to the basic cartographic model, such as the incorporation of serum effects and evolutionary priors, that improve fit and add biological interpretability.
-
-\subsection*{Antigenic evolution across influenza lineages}
-
-Through our analysis, we reveal the antigenic, as well as evolutionary, relationships among viruses in influenza A/H3N2, A/H1N1, B/Vic and B/Yam, quantifying both antigenic and evolutionary distances between strains (Figure~\ref{map}).
-Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
-We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less prominent, though still apparent, clustering in B/Vic and B/Yam.
-Antigenic clusters show high genetic similarity, so that we observe very few mutation events leading to each cluster, rather than the repeated emergence of clusters.
-This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Victoria/75 and Beijing/89) appearing to be evolutionary dead-ends.
-Labeling of prominent antigenic clusters in figures \ref{map} and \ref{drift} is intended as a rough guide for orientation and not as exhaustive catalog of antigenic variation.
-
-%%% map %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/map}
-	\caption{\textbf{Antigenic locations of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
-	Circles represent a posterior sample of virus locations and have been shaded based on year of isolation.
-	Antigenic units represent two-fold dilutions of the HI assay.
-	Absolute positioning of lineages, e.g.\ A/H3N2 and A/H1N1, is arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Brisbane/07, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
-	\label{map} 
-\end{figure}
-
-HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 14 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.
-Because of the threshold of sensitivity of the HI assay, it's difficult to distinguish a linear trajectory in 2D antigenic space from a slightly curved trajectory (see Materials and Methods).
-To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space (present in models 6 through 9; Table \ref{errortable}), with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Materials and methods).
-Because of this, we interpret map locations locally rather than globally, and assess rates of antigenic movement without making strong statements about the larger configuration under which the movement occurs.
-
-We find that influenza A/H3N2 evolved along antigenic dimension 1 at an estimated rate of 1.01 antigenic units per year (Figure~\ref{drift}, Table~\ref{paramtable}).
-However, we observe occasional large jumps in antigenic phenotype (Figure~\ref{drift}), corresponding to cluster transitions identified by Smith et al. \cite{Smith04}.  
-Most variation is contained within the first antigenic dimension, but dimension 2 occasionally shows variation when two antigenically distinct lineages emerge and transiently coexist (Figure~\ref{map}), as is the case with the previously identified Beijing/89 and Beijing/92 clusters.
-
-%%% drift %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.75\textwidth]{figures/drift}
-	\caption{\textbf{Antigenic drift of A/H3N2, A/H1N1, B/Vic and B/Yam viruses showing evolutionary relationships between virus samples.} 
-	Antigenic drift is shown in terms of change of location in the first antigenic dimension through time.
-	Circles represent a posterior sample of virus locations and have been shaded based on year of isolation.
-	Antigenic units represent two-fold dilutions of the HI assay.
-	Relative positioning of lineages, e.g.\ A/H3N2 and A/H1N1, in the vertical axis is arbitrary.
-	Lines represent mean posterior diffusion paths when virus locations are fixed.
-	Prominent antigenic clusters are labeled after vaccine strains present within clusters, and are abbreviated from Hong Kong/68, England/72, Victoria/75, Bangkok/79, Sichuan/87, Beijing/89, Beijing/92, Wuhan/95, Sydney/97, Fujian/02, California/04, Wisconsin/05, Brisbane/07, Perth/09 (A/H3N2), USSR/77, Singapore/86, Beijing/95, New Caledonia/99, Solomon Islands/06 (H1N1), Victoria/87, Hong Kong/01, Malaysia/04, Brisbane/08 (Vic), Yamagata/88, Shanghai/02, Florida/06, Wisconsin/10 (Yam).}
-	\label{drift} 
-\end{figure}
-
-We find that other lineages of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}, Table~\ref{paramtable}).
-Influenza A/H1N1 evolved at a rate of 0.62 units per year, but showed a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.
-Influenza B/Victoria and B/Yamagata evolved slower still, with mean estimated rates 0.42 units per year and 0.32 units per year, respectively.
-Punctuated evolution is less obvious in B/Yam and B/Vic compared to A/H3N2 and A/H1N1, but antigenic clusters are still apparent, with recent transitions to the Brisbane/08 cluster in B/Vic \cite{Barr10} and to the Wisconsin/10 cluster in B/Yam \cite{Klimov12}.
-Interestingly, a minor lineage of B/Vic, denoted B/Hubei-Songzi/51/2008 \cite{Barr10}, has persisted through 2011, while remaining antigenically distinct from B/Brisbane/60/2008 viruses (Figure~\ref{drift}).
-Although we observe significantly different drift rates between lineages, we observe less variation in diffusion volatility (Table~\ref{paramtable}).
-This is reflected in Figure \ref{drift}, where all four lineages exhibit similar levels of standing antigenic variation, despite A/H3N2 drifting more quickly in antigenic phenotype.
-
-%%% paramtable %%%
-\begin{table}[h]
-	\centering
-	\caption{\textbf{Estimates of drift rate $\drift$ (in units per year), diffusion volatility $\virussd^2$ (in units$^2$ per year) and scaled effective population size $N_e\tau$ (in years) for influenza A/H3N2, A/H1N1, B/Vic and B/Yam including posterior means and 95\% highest posterior density intervals.}}
-	\label{paramtable}	
-	\begin{tabular}{ c c c c} 
-	\hline
-	Lineage	&	Drift $\drift$  	& 	Volatility $\virussd^2$	&	Effective pop size $N_e\tau$	\\
-	\hline		
-	A/H3N2	&	1.01 (0.98--1.04)	&	1.25 (0.98--2.35) 		&	5.03 (4.42--5.73)	\\
-	A/H1N1	&	0.62 (0.56--0.67)	&	0.92 (0.65--1.56) 		&	6.38 (4.99--8.12)	\\
-	B/Vic	&	0.42 (0.32--0.51)	&	1.22 (0.85--2.25) 		&	10.40 (8.42--12.80)	\\
-	B/Yam	&	0.32 (0.25--0.39)	&	0.71 (0.46--1.36) 		&	9.48 (7.76--11.50)	\\
-	\hline
-	\end{tabular}	
-\end{table}
-
-These patterns of antigenic drift are reflected in the virus phylogenies (Figure~\ref{trees}).
-Influenza A/H3N2 has a characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
-The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
-Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift (Table~\ref{paramtable}), with A/H3N2 showing fast drift and reduced effective population size as expected from basic epidemiological models \cite{Bedford12}.
-Antigenic drift results in the replacement of antigenically primitive lineages by antigenically advanced lineages, thereby reducing genealogical diversity.
-
-Thus, we observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
-Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-Models of influenza evolution acribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
-Without more detailed epidemiological modeling, the present study cannot conclusively distinguish between these causal possibilities.
-
-%%% trees %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.85\textwidth]{figures/trees}
-	\caption{\textbf{Time-resolved phylogenetic trees of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	The maximium-clade credibility (MCC) tree is shown for each virus.
-	These trees show genealogical relationships, so that branches are measured in terms of years rather than substitutions.
-	The mean posterior scaled effective population size $N_e \tau$, measured in years, and its 95\% credible interval is shown for each virus.} 
-	\label{trees} 
-\end{figure}
-
-\subsection*{Punctuated evolution and its epidemiological consequences}
-
-We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
-We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
-There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
-We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}B), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
-This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
-
-%%% jumps %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/jumps}
-	\caption{\textbf{Year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
-	(A) Timeseries of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
-	Colored lines represent year-to-year antigenic drift, where drift for year $i$ is measured as the mean of antigenic dimension 1 of phylogenetic lineages in year $i$ compared to the mean of antigenic dimension 1 of phylogenetic lineages from the previous year $i-1$.
-	For example, 2000 represents difference in antigenic dimension 1 between viruses from 1999 and 2000.
-	Error bars represent 50\% Bayesian credible intervals of year-to-year drift.
-	Gray dotted lines represent lineage-specific seasonal incidence in the USA taken as average influenza-like illness (ILI) multiplied by proportion of viruses attributable to a lineage for each season.
-	Here, 2000 represents the 2000/2001 influenza season.
-	(B) Distribution of year-to-year antigenic drift between 1992 and 2011 in A/H3N2, A/H1N1, B/Vic and B/Yam viruses.
-	} 
-	\label{jumps} 
-\end{figure}
-
-We investigate the relationship between rates of antigenic drift and seasonal incidence in the USA in A/H3N2, A/H1N1, B/Vic and B/Yam.
-We measure seasonal incidence from USA CDC influenza surveillance reports for each virus lineage (A/H3N2, A/H1N1, B/Vic, B/Yam) by taking the average influenza-like (ILI) percentage in a season and multiplying this by the relative proportion of virus isolations attributable to a particular influenza lineage in a season (see Materials and methods).
-This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
-We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic.
-We begin by comparing overall rates of antigenic drift (Figure~\ref{drift}, Table~\ref{paramtable}) to overall levels of seasonal incidence across influenza lineages, finding a significant correlation between rate of antigenic drift and relative incidence across the four lineages (Pearson correlation, $r=0.97$, $p = 0.041$).
-
-We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and lineage-specific incidence, comparing incidence in a seasonal to antigenic drift of viruses coming into this season (Figure~\ref{jumps}A).
-For example, we compare antigenic drift of viruses from 2000 to 2001 to incidence in the 2001/2002 season.
-Within each virus lineage, we find that years with pronounced antigenic drift tend to show increased incidence (Figure~\ref{corr}), finding Pearson correlation coefficients of 0.51, 0.29, 0.44 and 0.14 for A/H3N2, A/H1N1, B/Vic and B/Yam respectively.
-We calculated significance using bootstrap permutation tests finding $p$-values of 0.056, 0.201, 0.097 and 0.341 respectively.
-We applied a similar bootstrap permutation test to calculate the significance of finding the observed levels of correlation across all four lineages, arriving at a $p$-value of 0.018.
-
-%%% corr %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=0.7\textwidth]{figures/corr}
-	\caption{\textbf{Relationship between antigenic drift and seasonal incidence for years 1998 to 2009 in influenza A/H3N2, A/H1N1, B/Vic and B/Yam.}
-	Antigenic drift from year $i-1$ to year $i$ is compared to incidence in the season $i$ / $i+1$.
-	For example, year-to-year antigenic drift from 2000 to 2001 is measured against incidence in the 2001/2002 season. 
-	} 
-	\label{corr} 
-\end{figure}
-
-Although the general correlation between rate of antigenic drift and incidence across the four influenza lineages may not represent a causal relationship, the within-lineage year-to-year correlations suggest a causal relationship as increases in incidence tend to occur subsequent to periods of pronounced antigenic drift.
-
-%Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
-%However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
-%Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.
-
-%\subsection*{Conversion of antigenic mutation into fixed differences}
-
-%We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
-%Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-%Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
-%Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-%To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Materials and methods).
-%This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
-%Higher diffusion coefficients mean a lineage is diffusing faster through antigenic space.
-%We compare diffusion coefficients between trunk branches and side branches, where trunk branches are defined as branches ancestral to the most recent virus sample and side branches are defined as those branches not ancestral to the most recent virus sample.
-
-%%% diffusiontable %%%
-%\begin{table}[h]
-%	\centering
-%	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including posterior means and 50\% highest posterior density intervals.}}
-%	\label{diffusiontable}	
-%	\begin{tabular}{ c c c c } 
-%	\hline
-%	Lineage	&	Trunk branch $D$ 	& 	Side branch $D$		& 	Ratio \\
-%	\hline		
-%	A/H3N2	&	0.90 (0.79--0.97)	&	0.62 (0.57--0.65)	&	1.45 (1.31--1.57) \\
-%	A/H1N1	&	0.62 (0.50--0.71)	&	0.44 (0.38--0.48)	&	1.45 (1.18--1.70) \\
-%	B/Vic	&	0.63 (0.51--0.71)	&	0.62 (0.54--0.68)	&	1.02 (0.87--1.14) \\
-%	B/Yam	&	0.35 (0.29--0.40)	&	0.33 (0.28--0.36)	&	1.09 (0.93--1.23) \\
-%	\hline
-%	\end{tabular}	
-%\end{table}
-
-%Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
-%We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
-%Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
-%Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
-%We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
-%Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
-%Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
-%Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.
-
-%In this context, it is especially interesting that adaptive evolution in the HA1 protein determined through analysis of nonsynonymous and synonymous substitutions is substantially stronger in A/H3N2 than in A/H1N1 \cite{Wolf06}; H3 appears to accumulate adaptive fixations at $\sim$150\% the rate of H1 \cite{Bhatt11}.
-%Here, we observe that A/H3N2 has a diffusion coefficient $\sim$141\% greater than A/H1N1 along side branches of the phylogeny.
-%These results suggest that the greater rate of adaptive fixation in A/H3N2 may be simply due to a greater abundance of new antigenic mutations to fix.
-
-%We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
-%In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
-%These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
-%Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
-%Still, it is clear that more work is necessary to fully clarify the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
-
-%%% CONCLUSIONS %%%
-\section*{Conclusions}
-
-Understanding antigenic evolution in seasonal influenza is crucial to our efforts of surveillance and control.
-Cartographic methods allow complex HI datasets to be compressed to more approachable location-based summaries that quantify antigenic relationships between strains, including relationships not directly assessed via HI.
-In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
-We use this approach to characterize competitive dynamics across influenza lineages A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution within each lineage drives strain replacement and contributes to seasonal incidence patterns.
-
-%We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
-%Influenza A/H3N2 benefits from a higher rate of new antigenic mutation coupled with strong adaptive evolution and evolves in antigenic phenotype faster than A/H1N1, B/Vic and B/Yam.
-%Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
-%We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
-
-The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
-For example, our likelihood-based model facilitates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
-Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
-In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
-Identifying viruses that will come to predominate in the global virus population while they are still at low frequency remains an enormous challenge.
-However, combining evolutionary and antigenic information may eventually prove useful in identifying low-frequency, but expanding, lineages of antigenically novel viruses that represent ideal targets for vaccine strain selection.
-
-%%% METHODS %%%
-\section*{Materials and methods}
-
-\subsection*{Antigenic cartography}
-
-Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
-At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
-In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against a particular virus ceases to be effective at inhibiting the binding of another virus to red blood cells.
-These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
-Because experimental HI titers typically differ by factors of two, we find it convenient to work in log$_2$ space and represent the titer of virus $i$ against serum $j$ as $H_{ij} = \mathrm{log}_2 \mbox{(HI titer)}$, i.e.\ a titer of 160 has $H_{ij} = 7.32$.
-Due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
-Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.
-
-Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
-These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
-
-Here, we follow previous models in representing antigenic locations as points in a low $P$-dimensional antigenic map. 
-One of our initial goals is to find an optimal projection of the high-dimensional distance matrix $\mathbf{H}$ into this lower dimensional space. 
-We conduct this projection using Bayesian multidimensional scaling (BMDS) \cite{Oh01} in which we construct a probabilistic model to quantify the fit of a particular configuration of cartographic locations to the observed matrix of serological measurements.
-Typically, $P = 2$, but higher or lower dimensions may better reflect the data.
-
-Let $\virus_i \in \mathbb{R}^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, so that $\virus_i = (x_{i1}, x_{i2})^{\prime}$ for $P=2$. 
-Similarly, let $\serum_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$, so that $\serum_j = (y_{j1},y_{j2})^{\prime}$ for $P=2$.
-For notational compactness, we collect together all virus coordinates into an $\vn \times P$ matrix  $\viruses = (\virus_1, \ldots, \virus_{\vn})^{\prime}$ and all serum coordinates into an $\sn \times P$ matrix $\sera = (\serum_{1},\ldots,\serum_{\sn})^{\prime}$.
-Virus and serum may be isolated from / raised against the same strain and have different cartographic locations, and separate serum isolates raised against the same strain may also have different cartographic locations.
-This gives a set of distances between virus and serum cartographic locations 
-\begin{equation}
-	\delta_{ij} =  || \virus_i - \serum_j ||_2,
-\end{equation}
-where $|| \cdot ||_2$ is an $L_2$ norm.
-
-Traditional approaches to antigenic cartography \cite{Smith04} begin by defining immunological distance as
-\begin{equation}
-	d_{ij} =  \se_j - H_{ij},
-\end{equation}
-where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
-In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
-\begin{equation} \label{mds}
-	\sum_{(i,j) \in \cal I} 
-	\left(
-		\delta_{ij} - d_{ij}
-	\right)^2
-\end{equation}
-is minimized, where $\mathcal{I} = \{ (i,j) : H_{ij} \mbox{ is measured} \}$.
-In the case of threshold measurements, this error function is modified slightly; see \cite{Smith04} for further details.
-
-Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation with variance $\mdssd^2$,
-\begin{equation} \label{hij}
-	H_{ij} \sim \normal( \se_j - \delta_{ij}, \, \mdssd^2 ).
-\end{equation}
-Consequently, the likelihood of observing an exact titer given the placement of antigenic locations is 
-\begin{equation} 
-	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\phi(\cdot)$ represents the standard normal probability density function (PDF).
-Previous BMDS has employed a sampling density truncated to strictly positive quantities since $d_{ij}$ are directly observed, non-negative quantities.  
-In the antigenic setting, these remain random and can be negative since neither $\se_j$ is known nor is $H_{ij}$ observed with much precision.
-
-HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
-In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
-\begin{equation} 
-	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{ \mdssd } \right),
-\end{equation}
-where $\Phi(\cdot)$ represents the standard normal cumulative distribution function (CDF).
-Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:160 dilution and no inhibition at 1:320 dilution.
-Rather than taking the HI titer as 160, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 160 and 320.
-HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
-\begin{equation} 
-	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - \se_j }{\mdssd} \right).
-\end{equation}
-These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
-Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.
-
-%%% hij_likelihood %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.6\textwidth]{figures/hij_likelihood}
-	\caption{\textbf{Likelihood of HI titers in the BMDS model.} 
-	Here we show the likelihoods of observing three different outcomes given $\delta_{ij} = 4$, $\mdssd = 0.95$ and $\se_j = \mathrm{log}_2 (1280) = 10.32$.  
-	The likelihood of observing a threshold titer of `$<$40' is equal to the lower tail of the probability density function $\threshold(5.32) = 0.146$.
-	The likelihood of observing a point measurement with an exact inhibiting titer of `90.5' is equal to the density function $\point(6.5) = 0.413$.
-	The likelihood of observing an interval measurement with an inhibiting titer somewhere between `160' and `320' is equal to $\interval(7.32) = 0.129.$
-	} 
-	\label{hij_likelihood} 
-\end{figure}
-
-We calculate the overall likelihood by multiplying probabilities of individual measurements
-\begin{equation}  \label{bmds}
-	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
-\end{equation}
-using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
-We begin by assuming independent, diffuse normal priors on virus and serum locations
-\begin{eqnarray}
-	\virus_i &\sim& \normal(\mathbf{m},\boldsymbol{\Sigma}) \nonumber \\
-	\serum_j &\sim& \normal(\mathbf{m},\boldsymbol{\Sigma}),
-\end{eqnarray}
-where $\mathbf{m} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.
-
-\subsection*{Virus and serum effects}
-
-The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.
-However, this model may be biased in some circumstances.
-In one example, if a particular serum $j$ is only measured against distant viruses, its maximum titer will be artificially low and the likelihoods concerning this serum will appear poor. 
-To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
-We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
-We take an Empirical Bayesian approach in specifying the mean and variance of this distribution, set to the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
-This formulation assumes that particular sera are more reactive in general than other sera.
-
-Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
-With virus reactivity included, observed titers follow
-\begin{equation}
-	H_{ij} \sim \normal \left( \frac{\ve_i+\se_j}{2} - \delta_{ij}, \, \mdssd^2 \right),
-\end{equation}
-and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
-
-\subsection*{Drift model of antigenic evolution}
-
-As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
-An example of this phenomenon is shown in Figure~\ref{schematic_map}.
-In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B).
-This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from observing the serological data alone.
-Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
-In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
-To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
-\begin{eqnarray}
-	x_{i1} &\sim& \drift \, t_i + \normal(0, \virussd^2) \nonumber \\
-	y_{j1} &\sim& \drift \, t_j + \normal(0, \serumsd^2),
-\end{eqnarray}
-where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
-Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\drift$.
-The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.
-
-%%% schematic_map %%%
-\begin{figure}[tb]
-	\centering		
-	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
-	\caption{\textbf{Schematic antigenic map with three viruses and two sera.}
-	(A) Map with virus 1 and virus 3 antigenically similar.
-	(B) Map with virus 1 and virus 3 antigenically divergent.
-	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
-	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
-	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
-	Configurations (A) and (B) represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
-	} 
-	\label{schematic_map} 
-\end{figure}
-
-\subsection*{Phylogenetic diffusion model of antigenic evolution}
-
-We simultaneously model antigenic locations and genetic relatedness by assuming that virus locations are influenced by evolution following a Brownian motion process \cite{Lemey10}.
-To do this, we replace the previous prior specifying independent virus locations with a prior that incorporates covariance based on shared evolutionary history
-\begin{equation} \label{ebpprior}
-	\viruses \sim \left( \begin{matrix} \drift \, t_1 & 0 \\ \vdots & \vdots \\ \drift \, t_n & 0 \end{matrix} \right) + \, \mbox{Evolutionary Brownian Process}(\virussd, \tree)
-\end{equation}
-for $P=2$, where $\virussd$ is the volatility parameter of the Brownian motion over virus locations and $\tree$ is a phylogeny specifying tree topology and branch lengths.
-Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
-In the evolutionary Brownian process, the tips of the phylogeny $\tree$ correspond to the set of virus locations $(\virus_1, \ldots, \virus_\vn)$, and the probability of observing tip locations depends on the locations of internal nodes $(\virus_{n+1}, \ldots, \virus_{2\vn-2})$ and on the location of the root node $\virus_{2\vn-1}$.
-This process assumes that a virus location $\virus_i$ follows from the location of its parent virus $\virus_{f(i)}$, and with the addition of drift along dimension 1, is distributed as
-\begin{equation} 
-	\virus_i \sim (\drift \, d_i, \, 0)^{\prime} + \normal(\virus_{f(i)}, d_i \, \boldsymbol{\Sigma})
-\end{equation}
-for $P=2$, where $f(i)$ is a function that maps nodes to parental nodes, $d_i$ is the length of the branch connecting virus $i$ to parent virus $f(i)$, and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $\virussd^2$.
-The root virus location $\virus_{2\vn-1}$ is assumed to follow a normal distribution with expectation $(\drift \, t_{2\vn-1}, 0)^{\prime}$ for $P=2$ and variance determined by the diffusion volatility $\virussd$ \cite{Lemey10}.
-The probability of virus locations $p(\viruses | \drift, \virussd, \tree)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
-This formulation corresponds to a Wiener process with drift, in which the drift term $\drift$ only influences the expected states of nodes along the phylogeny, but does not influence the covariance structure among these nodes, which remains the same as it does in a standard Wiener process \cite{BrownianMotionHandbook}.
-This allows the separation in equation \ref{ebpprior} between drift terms affecting only expectations and the evolutionary Brownian process that includes covariance among virus locations $\virus_1,\ldots,\virus_n$.
-
-Here, the phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST17}.
-
-\subsection*{Posterior inference}
-
-Top-level priors for $1/\mdssd^2$, $\drift$, $1/\virussd^2$, and $1/\serumsd^2$ are assumed to follow diffuse $\mbox{Gamma}(a, b)$ distributions  with $a=0.001$ and $b=0.001$.
-These diffuse priors were chosen to be non-informative and provide little-to-no weight on the resulting posterior distributions.
-Under the full model, the posterior probability of observing virus and serum locations given immunological data is factored
-\begin{equation}
-	p(\viruses,\sera | \mathbf{H}) \propto p(\mathbf{H} | \viruses, \sera, \ses, \ves, \mdssd) \; 
-	p(\viruses | \drift, \virussd, \tree) \;
-	p(\sera | \drift, \serumsd) \; 
-	p(\ses, \ves, \mdssd, \drift, \virussd, \serumsd, \tree).
-\end{equation}
-We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
-For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
-
-\subsection*{Genetic, antigenic and surveillance data}
-
-We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
-We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
-If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD.
-Many strains had full length HA sequences, while other strains only possessed HA1 sequences.
-Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
-This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
-However, the majority of isolates were present from 2002 to 2007. 
-Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
-We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
-This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements. 
-Each virus isolate was compared to an average of 21.9 serum isolates, and each serum isolate was compared to an average of 18.0 virus isolates.
-
-Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
-The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
-This procedure yielded 115 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
-Each virus isolate was compared to an average of 10.0 serum isolates, and each serum isolate was compared to an average of 16.2 virus isolates.
-
-Antigenic comparisons for influenza B/Victoria were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.
-Each virus isolate was compared to an average of 6.5 serum isolates, and each serum isolate was compared to an average of 16.7 virus isolates.
-
-Antigenic comparisons for influenza B/Yamagata were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
-For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.
-Each virus isolate was compared to an average of 6.9 serum isolates, and each serum isolate was compared to an average of 17.3 virus isolates.
-
-Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
-As an example, one report states ``collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive.
-Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses.
-Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
-In this case, we estimate the relative proportion of A/H3N2 of the four lineages as $0.66 \times 0.13 = 0.09$.
-Similar calculations were performed for A/H1N1, B/Vic and B/Yam.
-
-\subsection*{Implementation}
-
-Phylogenetic trees were estimated for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
-MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
-These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
-However, it would be possible to jointly sample from sequence data and serological data using these methods.
-
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
-MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
-In all cases when drift parameter $\drift$ was included the MCMC chain mixed well and arrived at the same estimated posterior distribution from different starting points.
-However, without drift parameter $\drift$, maps for A/H3N2 showed some degree of metastability, where some chains would converge on one solution and other chains would converge on a different solution.
-We favor models that include $\drift$, because its inclusion, in addition to correcting most identifiability issues, yields much improved mixing of antigenic locations.
-
-There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
-Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
-However, local isometries are often still a problem.
-For example, in A/H3N2 the HK/68, EN/72 and VI/75 clusters may rotate in relation to other clusters.
-Consequently, it may be difficult to fully align MCMC samples using Procrustes analysis.
-For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
-Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate the quantity across MCMC samples to yield an expectation and a credible interval.
-This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
-
-We summarize diffusion paths of viral lineages (Figure~\ref{map}, Figure~\ref{drift}) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
-We use MCMC to sample tip locations, but when outputting trees sample internal node locations using a peeling algorithm as described in \cite{Pybus12}.
-Thus, after the MCMC is finished we have a posterior sample of 2000 trees each tagged with estimated tip locations and internal node locations.
-We post-processed each posterior tree by conducting a linear interpolation between parent-child node locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
-Then, for each interval, $x$ and $y$ values are averaged across the sample of posterior trees.
-We draw lines between these locations to approximate mean posterior diffusion paths.
-As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.
-
-%\subsection*{Estimation of diffusion coefficient}
-
-%The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
-%Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
-%\begin{equation}
-%	X_t \sim \normal(0, \sigma^2 t),
-%\end{equation}
-%where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
-%\begin{equation}
-%	\mathrm{E}[X_t^2] = \sigma^2 t.
-%\end{equation}
-%In moving to a two-dimensional diffusion
-%\begin{equation}
-%	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
-%\end{equation}
-%where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
-%\begin{equation}
-%	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
-%\end{equation}
-%Such diffusions are often characterized with diffusion coefficient $D$.
-%In this case
-%\begin{equation}
-%	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
-%\end{equation}
-%so that expected squared displacement follows from $D$ alone
-%\begin{equation}
-%	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
-%\end{equation}
-
-%Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
-%\begin{equation} \label{pybusest}
-%	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
-%\end{equation}
-%in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
-%However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
-%Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
-%\begin{eqnarray} \label{ourest}
-%	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
-%	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
-%	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
-%\end{eqnarray}
-%This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
-%Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
-%When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
-
-\subsection*{Comparison with previous results}
-
-Here, we attempt to compare antigenic locations inferred by our BMDS model to antigenic locations previously inferred by the error minimization methods of Smith et al.\ \cite{Smith04}, referred to here as antigenic cartography by MDS.
-For this comparison, we use exactly the same HI data used to produce the results in \cite{Smith04}, consisting of 273 virus isolates, 79 serum isolates and a total of 4252 HI measurements taken between 1968 and 2003.
-We begin with a BMDS analog of the antigenic model used in \cite{Smith04}, where serum effects are taken as the maximum titer of a particular ferret serum and the expected log$_2$ drop in HI titer is proportional to Euclidean distance between virus and serum locations.
-To bring models into further alignment, we use a Uniform$(-100,100)$ distribution over virus locations and serum locations.
-Unsurprisingly, we find that this BMDS model produces results that are strongly congruent with MDS results (Figure~\ref{smith_comparison}).
-Antigenic cluster locations are consistent between methods (Figure~\ref{smith_comparison}A--B) and antigenic distances between pairs of viruses are consistent between temporally similar and temporally divergent viruses (Figure~\ref{smith_comparison}C--E), suggesting that the resulting maps are consistent at both local and global scales.
-Credible intervals of antigenic distances for the BMDS model remain narrow across the temporal spectrum (Figure~\ref{smith_comparison}C--E), implying a fair degree of rigidity to the map.
-
-%%% smith_comparison %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/smith_comparison}
-	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model.} 
-	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
-	(B) A posterior sample of antigenic locations from an equivalent BMDS model.
-	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
-	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
-	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
-	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
-	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 
-	} 
-	\label{smith_comparison} 
-\end{figure}
-
-Smith et al.\ \cite{Smith04} show that there exist at least two solutions in their assignment of antigenic locations, involving the rotation of clusters HK/68, EN/72 and VI/75 (shown in Figure S2 of \cite{Smith04}).
-We observe the same metastable behavior in our analysis; some MCMC chains converge on the solution shown in Figure \ref{smith_comparison}B, while other MCMC chains converge on the alternative solution shown in Figure \ref{smith_comparison_rotation}B.
-The distribution of likelihood values appear highly similar between these two solutions, suggesting that they represent global optima.
-The rotation of the HK/68, EN/72 and VI/75 clusters creates a map that bends slightly, so that temporally distant viruses appear closer in the rotated solution than in the original solution (Figure~\ref{smith_comparison_rotation}C--E).
-In this case, it is clear that the solutions are locally consistent between viruses up to $\sim$15 years divergent, even if there some degree of global flexibility.
-
-%%% smith_comparison_rotation %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_rotation}
-	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an equivalent BMDS model under an alternative solution.} 
-	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
-	(B) A posterior sample of antigenic locations from an equivalent BMDS model that has converged on the alternative solution.
-	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
-	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
-	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
-	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
-	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 	
-	} 
-	\label{smith_comparison_rotation} 
-\end{figure}
-
-As discussed in the main text, the presence of multiple optima with different degrees of 2D curvature implies an issue of identifiability; the HI likelihood model alone cannot distinguish between these possibilities.
-Because of this issue, and to more easily estimate rates of antigenic drift, we include a model of systematic drift in antigenic location that favors linear movement in the antigenic map.
-We find that including this drift prior on antigenic locations removes the problem of identifiability.
-Antigenic locations produced by this model remain locally consistent with MDS results between viruses $\sim$15 years divergent, but global comparisons show that this BMDS model has partitioned more variance to the first antigenic dimension (Figure~\ref{smith_comparison_drift}).
-We additionally find that including the drift prior on antigenic locations often results in greater predictive power, with a slight improvement of test error for the A/H1N1, B/Vic and B/Yam datasets (Table~\ref{errortable}).
-
-%%% smith_comparison_drift %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_drift}
-	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that includes date-informed priors on antigenic locations.} 
-	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
-	(B) A posterior sample of antigenic locations from a BMDS model that includes date-informed priors on antigenic locations.
-	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
-	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
-	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
-	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
-	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a date-informed prior on antigenic locations and serum effects fixed at maximum titer values. 	
-	} 
-	\label{smith_comparison_drift} 
-\end{figure}
-
-Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects.
-Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.
-To isolate this difference, we use a Uniform$(-100,100)$ prior on antigenic locations.
-Surprisingly, estimating virus and serum effects results in a more linear antigenic map (Figure~\ref{smith_comparison_effects}), resembling the appearance of the map incorporating the antigenic drift prior, while preserving local consistency.
-We generally observe congruence between MDS and BMDS antigenic locations for viruses less than $\sim$10 years divergent (Figure~\ref{smith_comparison_effects}E).
-However, specific viruses may be affected, for instance A/Bilthoven/16190/1968 (Figure~\ref{smith_comparison_effects}C), which appears more distant from all other viruses when serum and virus effects are included.
-
-%%% smith_comparison_effects %%%
-\begin{figure}[h]
-	\centering		
-	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_effects}
-	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
-	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
-	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus and serum effects.
-	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
-	(C) Antigenic distances between A/Bilthoven/16190/1968 and all other viruses determined for both methods.
-	(D) Antigenic distances between A/Fujian/411/2002 and all other viruses determined for both methods.
-	(E) Antigenic distances between 750 random pairs of viruses determined for both methods.	
-	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and virus and serum effects estimated in a hierarchical Bayesian fashion. 	
-	} 
-	\label{smith_comparison_effects} 
-\end{figure}
-
-In this dataset, viruses 15 or more years divergent always yield threshold titers, and hence, their relative locations must be indirectly inferred rather than through direct comparison.
-This may explain why we observe local consistency between models at scales less than $\sim$15 years, but some degree of global inconsistency.
-Still, these results suggest that, when making local comparisons, such as those used to calculate year-to-year antigenic drift (Figure~\ref{drift}), outcomes are expected to be robust to  many model particulars.
-
-\subsection*{Availability}
-
-Source code implementing the cartographic models has been made fully available as part of the software package BEAST \cite{BEAST17}, and can be downloaded from its Google code repository (http://code.google.com/p/beast-mcmc/).
-Tutorials for implementing these models can be found at http://beast.bio.ed.ac.uk/Tutorials.
-Incidence data and HI data used in this analysis is archived with Dryad (doi:10.5061/dryad.rc515).
-
-%%% ACKNOWLEDGMENTS %%%
-\subsection*{Acknowledgments}
-
-We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.
-We acknowledge the laboratories that provided sequences to EpiFlu database: Centers for Disease Control and Prevention (USA), Chinese Center of Disease Prevention and Control, Hospital Clinic of Barcelona, National Institute of Hygiene of Morocco, National Institute of Infectious Diseases (Japan), National Institute for Medical Research (UK), Norwegian Institute of Public Health, Swedish Institute for Infectious Disease Control, Victorian Infectious Diseases Reference Laboratory (Australia).
-
-%%% AUTHOR CONTRIBUTIONS %%%
-\subsection*{Author contributions}
-
-AR, MAS, TB conceived the study.
-AR, MAS, TB, DJS, PL designed the antigenic model.
-TB, GD, VG, JWM, AJH, CAR, DJS, AR gathered antigenic and genetic data.
-AR, MAS, TB, PL implemented statistical phylogenetic procedures.
-TB performed the analysis.
-TB, AR, DJS, JWM, PL, CAR, AJH, GD, MAS interpreted the results.
-TB, MAS, AR, JWM, AJH wrote the paper.
-
-%%% FUNDING %%%
-\subsection*{Funding}
-
-TB was supported by a Newton International Fellowship from the Royal Society. 
-MAS was supported by National Institutes of Health grants R01 GM086887 and R01 HG006139 and National Science Foundation grant DMS0856099.
-The research leading to these results has received funding from the European Research Council under the European Community's Seventh Framework Programme (FP7/2007-2013) under Grant Agreement no.
-278433-PREDEMICS and ERC Grant agreement no.
-260864.
-VG, AJH and JWM were funded by MRC reference no U117512723.
-AR acknowledges the support of the Wellcome Trust (grant no. 092807).
-Collaboration between MAS, AR and PL was supported by the National Evolutionary Synthesis Center (NESCent), NSF EF-0423641.
-DJS and CAR acknowledge EU FP7 programs EMPERIE (223498) and ANTIGONE (278976), Human Frontier Science Program (HFSP) program grant P0050/2008, Wellcome Trust 087982AIA, NIH Director's Pioneer Award DP1-OD000490-01, and National Institute of Allergy and Infectious Diseases NIH CEIRS contract HHSN266200700010C.
-CAR was supported by a University Research Fellowship from the Royal Society.
-
-%%% REFERENCES %%%
-\bibliographystyle{plos}
-\bibliography{flux}
+On page 9, the two consecutive sentences beginning ``Previous work using...'' and ``Models of influenza evolution...'' seem slightly redundant.
+If they are indeed redundant that could be fixed -- if they in fact mean different things, that could be clarified.
 

### 2013-12-06 correspondence/response.tex
Consolidate correspondence and update rakefile to account for directories.

--- 2013-11-14 response.tex
@@
 
-%%% TITLE %%%
-\title{\vspace{1.0cm} \Large \bf
-Reviewer responses for\\
-Integrating influenza antigenic dynamics with molecular evolution
-}
-
-\author[1]{Trevor Bedford}
-\author[2,3,4]{Marc A. Suchard}
-\author[5]{Philippe Lemey}
-\author[1]{Gytis Dudas}
-\author[6]{Victoria Gregory}
-\author[6]{Alan J. Hay}
-\author[6]{John W. McCauley}
-\author[7,8]{Colin A. Russell}
-\author[7,8,9]{Derek J. Smith}
-\author[1,10]{Andrew Rambaut}
-
-\affil[1]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
-\affil[2]{Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[3]{Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles CA, USA}
-\affil[4]{Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles CA, USA}
-\affil[5]{Department of Microbiology and Immunology, Katholieke Universiteit Leuven, Leuven, Belgium}
-\affil[6]{Division of Virology, MRC National Institute for Medical Research, Mill Hill, London, UK}
-\affil[7]{Centre for Pathogen Evolution, Department of Zoology, University of Cambridge, Cambridge, UK}
-\affil[8]{WHO Collaborating Center for Modeling, Evolution, and Control of Emerging Infectious Diseases, University of Cambridge, Cambridge, UK}
-\affil[9]{Department of Virology, Erasmus Medical Centre, Rotterdam, Netherlands}
-\affil[10]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
-
-\date{}
-
 \begin{document}
 
-\maketitle
+\newgeometry{top=4cm}
 
-Original reviewer criticisms are in plain text.
-Our responses follow in \textbf{bold}.
-A detailed list of every criticism and exactly what was revised in the manuscript in response to this criticism is available at \url{https://github.com/trvrb/flux/issues?state=closed}.
+Dear eLife editorial board,
+
+Please find attached our extensively revised manuscript entitled ``Integrating influenza antigenic dynamics with molecular evolution''.
+This is a resubmission of manuscript number 04-04-2013-RA-eLife-00803.
+The editorial response stated that:
+
+\begin{quote}
+Given, however, the extreme importance of this work to understanding influenza evolution and vaccine selection and the potential to provide a quantitative assessment of the association between drift (as measured by cartography) and incidence, the reviewers and editor would be willing to consider a fresh submission with major revisions that convinced us that the correlation is true or that there is perhaps no correlation between drift and incidence (or conceivably that antigenic cartography is not a good way to measure drift).
+\end{quote}
+
+We have paid particular attention to the reviewer's criticisms regarding our observed correlation between year-to-year antigenic drift and influenza incidence.
+We have revised our statistics to show correlation within A/H3N2, A/H1N1, B/Vic and B/Yam individually and as well as to show that measurements of antigenic drift are robust to sampling counts.
+Additionally, we have included substantial new results regarding relationships between amino acid substitutions and antigenic drift as suggested by reviewer 2 and provided a new analysis of phylogenetic trees as suggested by reviewer 1.
+
+Point-by-point review responses are attached, as well as a diffed PDF detailing all the changes that have been made since the initial submission.
+Additionally, a detailed list of every criticism and exactly what was revised in the manuscript in response to this criticism is available at \url{https://github.com/trvrb/flux/issues?state=closed}.
+
+Sincerely,\\
+Trevor Bedford
+
+\restoregeometry
+
+\newpage
+
+\section*{Reviewer responses}
+
+Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.
 
@@
 
-%%% REFERENCES %%%
-\bibliographystyle{plos}
-\bibliography{flux}
-
 \end{document}

### 2013-12-06 flux.tex
Fix #42 - differences typo.

@@
 Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-Models of influenza evolution acribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+Models of influenza evolution ascribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
 Without more detailed epidemiological modeling, the present study cannot conclusively distinguish between these causal possibilities.

### 2013-12-06 flux.tex
Fix #41 - dimension typo.

@@
 By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers.
-We find that a two dimensional model has better predictive power than models of lower of higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
+We find that a two dimensional model has better predictive power than models of lower or higher dimension in all four influenza lineages (models 1--5; Table \ref{errortable}).
 We find that this 2D model performs well, yielding an average absolute predictive error of between 0.78 and 0.91 log$_2$ HI titers across influenza lineages (model 2; Table \ref{errortable}), in line with the results of Smith et al.\ \cite{Smith04}.

### 2013-12-06 flux.tex
Fix #40 - transmission advantage.

@@
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more recent viruses resulting in new strains having a transmission advantage over previously established strains.
+Sera from older strains react poorly with more recent viruses resulting in new strains having a selective advantage over previously established strains.
 

### 2013-12-06 correspondence/reresponse.tex
Fix #42 - differences typo.
Fix #41 - dimension typo.
Fix #40 - transmission advantage.
Fix #43 - wording redundancy.

--- 2013-12-06 reresponse.tex
@@
 
+\textbf{We agree that ``transmission advantage'' was inexact.  This has been revised to the broader ``selective advantage.''}
+
 Bottom of page 3 says ``of lower of higher dimension.'' The second ``of'' should be an ``or.''
 
+\textbf{Fixed.}
+
 On page 9, ``acribe differences'' should be ``ascribe differences.''
+
+\textbf{Fixed.}
 
@@
 
+\textbf{We've clarified that the first sentence is referring to general not-necessarily-influenza epidemiological models and the second sentence is referring to influenza-specific models.}
+
 \end{document}

### 2013-12-06 flux.tex
Fix #43 - wording redundancy.

@@
 Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-Models of influenza evolution ascribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
+Correspondingly, models specific to influenza evolution ascribe differences in the rate of antigenic drift of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
 Without more detailed epidemiological modeling, the present study cannot conclusively distinguish between these causal possibilities.

### 2013-12-06 flux.tex
Fix #39 - vaccine efficacy.

@@
 A large proportion of the disease burden of influenza stems from antigenic drift, which allows individuals to be infected multiple times throughout their lives.
-Although influenza vaccines may lack efficacy for a variety of reasons \cite{Osterholm12}, antigenic drift causes efficacy against a fixed vaccine formulation to decline over time.
+Although influenza vaccines may lack efficacy for a variety of reasons \cite{Osterholm12}, antigenic drift causes efficacy of a fixed vaccine formulation against circulating viruses to decline over time.
 A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.

### 2013-12-07 correspondence/reresponse.tex
Fix #39 - vaccine efficacy.
Fix #33 - virus avidity.

@@
 
+\textbf{We appreciate this suggestion.
+Including the biological explanation for why we observe virus effects in the form of decreased or increased overall HI reactivity is very important.
+We've revised this section to put virus avidity at the forefront and relabeled `virus effects' to `virus avidities'.
+This has the additional benefit of being more transparent of a term than the opaque `effect.'}
+
 In Table 2 and the related discussion, the authors discuss the scaled effective population size Ne * tau.
@@
 
+\textbf{We've revised the clause to ``antigenic drift causes efficacy of a fixed vaccine formulation against circulating viruses to decline over time.''}
+
 Last sentence of second paragraph of abstract: might be better just to say ``an advantage'' rather than ``a transmission advantage.'' Although drift leads to an advantage that presumably improves R and so in a sense does increase transmission, this advantage probably is acting at the step of viral replication rather than the act of transmission per se.

### 2013-12-07 flux.tex
Fix #33 - virus avidity.

@@
 \newcommand{\sera}{\mathbf{Y}}						% set of serum coordinates
-\newcommand{\ve}{v}									% virus effect
+\newcommand{\ve}{v}									% virus avidity
 \newcommand{\se}{s}									% serum effect
-\newcommand{\ves}{\mathbf{v}}						% set of virus effects
+\newcommand{\ves}{\mathbf{v}}						% set of virus avidities
 \newcommand{\ses}{\mathbf{s}}						% set of serum effects
@@
 	\multicolumn{6}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus avidities	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline	
@@
 
-Additionally, we include and estimate `virus effects' in an analogous fashion, which are intended to represent differences in overall reactivity between viruses.
-With virus effects included, baseline titer derives from both the virus and the serum used in the HI reaction.
-We find that including virus effects further improves test error, either alone with fixed serum effects (model 9, Table \ref{errortable}) or combined with estimating serum effects (model 10, Table \ref{errortable}).
-With fixed serum effects, the inclusion of virus effects results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum effects, the inclusion of virus effects results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
+Additionally, we include and estimate `virus avidities' in an analogous fashion, which are intended to represent differences in overall HI reactivity between viruses.
+Experimental work has demonstrated that influenza variants exist that differ in HA binding activity for cell surface glycan receptors, and that these high-avidity variants may arise in the presence of antibody pressure \cite{Hensley09}.
+The presence of differential virus avidity has been previously shown to distort antigenic maps constructed from a model that disregards effects from virus avidity \cite{Li13}.
+Here, with virus avidities estimated, baseline titer derives from both the virus and the serum used in the HI reaction.
+We find that including virus avidities further improves test error, either alone with fixed serum effects (model 9, Table \ref{errortable}) or combined with estimating serum effects (model 10, Table \ref{errortable}).
+With fixed serum effects, the inclusion of virus avidities results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum effects, the inclusion of virus avidities results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
 
@@
 
-Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a `virus effect' $\ve_i$ representing the general level of reactivity across HI assays.
-With virus reactivity included, observed titers follow
+Additionally, we follow the same logic and assume that some virus isolates are more reactive than other virus isolates and include a parameter for `virus avidity' $\ve_i$ representing the general level of reactivity across HI assays.
+With virus avidity included, observed titers follow
 \begin{equation}
@@
 \end{equation}
-and the vector of virus effects $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
+and the vector of virus avidities $\ve_i$ for $i = 1,\ldots, \vn$ is estimated in an analogous hierarchical fashion, with $\ves$ normally distributed with mean and variance equal to the empirical mean and variance of the set of maximum titers across viruses $\{ \max ( H_{i1},\ldots,H_{i \sn} ) : i = 1,\ldots,\vn \}$.
 
@@
 We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus effects $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
+Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus avidities $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
 For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
@@
 
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus effects $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus avidities $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
@@
 
-Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus effects.
+Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus avidities.
 Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.
@@
 We generally observe congruence between MDS and BMDS antigenic locations for viruses less than $\sim$10 years divergent (Figure~\ref{smith_comparison_effects}E).
-However, specific viruses may be affected, for instance A/Bilthoven/16190/1968 (Figure~\ref{smith_comparison_effects}C), which appears more distant from all other viruses when serum and virus effects are included.
+However, specific viruses may be affected, for instance A/Bilthoven/16190/1968 (Figure~\ref{smith_comparison_effects}C), which appears more distant from all other viruses when serum and virus avidities are included.
 
@@
 	\includegraphics[width=1.0\textwidth]{figures/smith_comparison_effects}
-	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus effects.} 
+	\caption{\textbf{Comparison of A/H3N2 antigenic locations estimated by Smith et al.\ \cite{Smith04} using MDS and an extended BMDS model that estimates serum and virus avidities.} 
 	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.

### 2013-12-07 correspondence/reresponse.tex
Move "serum effect" to "serum potency" per issue #33.

@@
 
+\textbf{With the change from `virus effect' to `virus avidity' made, we chose to make a similar biological realignment of `serum effect' to `serum potency', i.e.\ some sera have higher potency than other other sera, allowing them to inhibit hemagglutination at lower concentrations than other sera.
+The use of potency here is meant to align with the neutralizing antibody literature which distinguishes neutralization potency from neutralization breadth.
+Antigenic cartography has not traditionally measured breadth of hemagglutination inhibition.}
+
 In Table 2 and the related discussion, the authors discuss the scaled effective population size Ne * tau.

### 2013-12-07 flux.tex
Move "serum effect" to "serum potency" per issue #33.

@@
 \newcommand{\ve}{v}									% virus avidity
-\newcommand{\se}{s}									% serum effect
+\newcommand{\se}{s}									% serum potency
 \newcommand{\ves}{\mathbf{v}}						% set of virus avidities
-\newcommand{\ses}{\mathbf{s}}						% set of serum effects
+\newcommand{\ses}{\mathbf{s}}						% set of serum potencies
 \newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
@@
 We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations as described in the `Antigenic cartography' section of Materials and methods.
-In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera and give the baseline expectation for titer when virus and serum are antigenically identical.
-Virus and serum locations follow an uninformative diffuse normal prior. 
+In this case, `serum potencies' are fixed to the maximum titers exhibited by particular ferret sera and give the baseline expectation for titer when virus and serum are antigenically identical.
+Potency differs between serum isolates due to experimental noise (e.g.\ variation in serum concentration), but also due to differential ferret immune responses, causing some serum isolates to inhibit hemagglutination at generally higher titers than other isolates.
+In this model, virus and serum locations follow an uninformative diffuse normal prior. 
 After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
@@
 	\multicolumn{6}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
-	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus avidities	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
+	Model	&	Data		&	Dimen 	& Location prior	& 	Serum potency 	&	Virus avidity	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
 	\hline	
@@
 
-We further extend the model by estimating `serum effects' to quantify the overall strength reactivity of each serum, rather than fixing serum effects at maximum titers.
-Serum effects likely derive from experimental variation in serum extraction and processing or from variation in immune responses between ferrets.
-Serum effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
-However, if serum effect is fixed to the serum's maximum titer, this will often not be the case, as the virus giving the maximum titer is likely to be antigenically distinct from the serum.
-Thus, fixing serum effects will tend to under-estimate effect size; we observe a mean effect of 10.42 log$_2$ HI titers for A/H3N2 when fixing serum effects and a mean of 10.94 when estimating serum effects.
-We find that estimating serum effects improves test error further (model 8, Table \ref{errortable}), with improvements of between 0.01 and 0.12 log$_2$ HI titers.
+We further extend the model by estimating serum potencies that quantify the overall strength reactivity of each serum, rather than fixing serum potencies at maximum titers.
+Different serum potencies likely derive from experimental variation in serum extraction and processing or from variation in immune responses between ferrets.
+Serum potency determines the baseline expectation of titer when virus and serum have identical antigenic locations.
+However, if serum potency is fixed to the serum's maximum titer, this will often not be the case, as the virus giving the maximum titer may be antigenically distinct from the serum.
+Thus, fixing serum potencies will tend to under-estimate effect size; we observe a mean effect of 10.42 log$_2$ HI titers for A/H3N2 when fixing serum potencies and a mean of 10.94 when estimating serum potencies.
+We find that estimating serum potencies improves test error further (model 8, Table \ref{errortable}), with improvements of between 0.01 and 0.12 log$_2$ HI titers.
 
-Additionally, we include and estimate `virus avidities' in an analogous fashion, which are intended to represent differences in overall HI reactivity between viruses.
+Additionally, we include and estimate `virus avidity' in an analogous fashion, which are intended to represent differences in overall HI reactivity between viruses.
 Experimental work has demonstrated that influenza variants exist that differ in HA binding activity for cell surface glycan receptors, and that these high-avidity variants may arise in the presence of antibody pressure \cite{Hensley09}.
@@
 Here, with virus avidities estimated, baseline titer derives from both the virus and the serum used in the HI reaction.
-We find that including virus avidities further improves test error, either alone with fixed serum effects (model 9, Table \ref{errortable}) or combined with estimating serum effects (model 10, Table \ref{errortable}).
-With fixed serum effects, the inclusion of virus avidities results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum effects, the inclusion of virus avidities results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
+We find that including virus avidities further improves test error, either along with fixed serum potencies (model 9, Table \ref{errortable}) or combined with estimating serum potencies (model 10, Table \ref{errortable}).
+With fixed serum potencies, the inclusion of virus avidities results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum potencies, the inclusion of virus avidities results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
 
@@
 When implementing the same underlying model, differences in the MDS and BMDS approaches reflect greater philosophical differences between maximum-likelihood and Bayesian statistical approaches, with the former seeking the single most likely explanation for the data, and the latter seeking to fully characterize model uncertainty.
-Additionally, the BMDS method improves flexibility, allowing extensions to the basic cartographic model, such as the incorporation of serum effects and evolutionary priors, that improve fit and add biological interpretability.
+Additionally, the BMDS method improves flexibility, allowing extensions to the basic cartographic model, such as the incorporation of virus avidities and evolutionary priors, that improve fit and add biological interpretability.
 
@@
 \end{equation}
-where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum effect $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
+where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum potency $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
 In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
@@
 
-\subsection*{Virus and serum effects}
+\subsection*{Virus avidity and serum potency}
 
@@
 To address this issue, we relax the assumption of fixed $\se_j$ values and treat the expected log$_2$ titer when $\delta_{ij}=0$ as a random variable.
-In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum effects' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
+In this case, $H_{ij}$ still follows equation \ref{hij} with expectation $\se_j - \delta_{ij}$, but the vector of `serum potencies' $\ses = (\se_1,\ldots,\se_{\sn})$ is random and estimated rather than fixed.
 We assume that $\se_j$ values are hierarchically distributed according to a normal distribution.   
@@
 We sample from this posterior distribution using the MCMC procedures implemented in the software package BEAST \cite{BEAST17}.
-Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus avidities $\ve_i$ and serum effects $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
+Metropolis-Hastings proposals include transition kernels that translate individual virus and serum locations $\virus_i$ and $\serum_j$ and individual virus avidities $\ve_i$ and serum potencies $\se_j$, and other transition kernels that scale the entire set of virus and serum locations $\viruses$ and $\sera$ and that scale parameters $\mdssd$, $\drift$, $\virussd$ and $\serumsd$.
 For the present analysis, a two-step approach was taken to sample phylogenies, where a posterior sample of phylogenies was gathered using sequence data and then, in the cartographic analysis, trees from this set were randomly proposed and accepted following the Metropolis-Hastings algorithm \cite{Pagel04}.
@@
 
-MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus avidities $\ves$, serum effects $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
+MCMC was used to sample virus locations $\viruses$, serum locations $\sera$, virus avidities $\ves$, serum potencies $\ses$, MDS precision $1/\mdssd^2$, antigenic drift rate $\drift$, virus location precision $1/\virussd^2$, serum location precision $1/\serumsd^2$ and phylogenetic tree $\tree$.
 MCMC chains were run for 500 million steps and parameter values sampled every 200,000 steps after a burn-in of 100 million steps, yielding a total of 2000 MCMC samples.
@@
 For this comparison, we use exactly the same HI data used to produce the results in \cite{Smith04}, consisting of 273 virus isolates, 79 serum isolates and a total of 4252 HI measurements taken between 1968 and 2003.
-We begin with a BMDS analog of the antigenic model used in \cite{Smith04}, where serum effects are taken as the maximum titer of a particular ferret serum and the expected log$_2$ drop in HI titer is proportional to Euclidean distance between virus and serum locations.
+We begin with a BMDS analog of the antigenic model used in \cite{Smith04}, where serum potencies are taken as the maximum titer of a particular ferret serum and the expected log$_2$ drop in HI titer is proportional to Euclidean distance between virus and serum locations.
 To bring models into further alignment, we use a Uniform$(-100,100)$ distribution over virus locations and serum locations.
@@
 	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum potencies fixed at maximum titer values. 
 	} 
@@
 	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and serum potencies fixed at maximum titer values. 	
 	} 
@@
 	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a date-informed prior on antigenic locations and serum effects fixed at maximum titer values. 	
+	The BMDS model has a date-informed prior on antigenic locations and serum potencies fixed at maximum titer values. 	
 	} 
@@
 Our final BMDS model (model 9, Table~\ref{errortable}) differs from antigenic model used by Smith et al.\ \cite{Smith04} in including temporally- and phylogenetically-informed priors on antigenic locations and also in estimating serum and virus avidities.
-Here, we investigate the impact on antigenic locations of estimating virus and serum effects in the BMDS model.
+Here, we investigate the impact on antigenic locations of estimating virus avidity and serum potency in the BMDS model.
 To isolate this difference, we use a Uniform$(-100,100)$ prior on antigenic locations.
-Surprisingly, estimating virus and serum effects results in a more linear antigenic map (Figure~\ref{smith_comparison_effects}), resembling the appearance of the map incorporating the antigenic drift prior, while preserving local consistency.
+Surprisingly, estimating virus avidity and serum potency results in a more linear antigenic map (Figure~\ref{smith_comparison_effects}), resembling the appearance of the map incorporating the antigenic drift prior, while preserving local consistency.
 We generally observe congruence between MDS and BMDS antigenic locations for viruses less than $\sim$10 years divergent (Figure~\ref{smith_comparison_effects}E).
@@
 	(A) MDS antigenic locations, reoriented so that the primary dimension lies on the $x$-axis rather than on the $y$-axis as in Figure 1 of \cite{Smith04}.
-	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus and serum effects.
+	(B) A posterior sample of antigenic locations from a BMDS model that estimates virus avidity and serum potency.
 	In (A) and (B), viruses are shown as colored circles, with color denoting antigenic cluster inferred by \cite{Smith04}, and sera are shown as gray points.
@@
 	In (C), (D) and (E) red points show distances for the MDS model and gray bars show the 95\% credible interval of distances for the BMDS model, while the red dashed line shows a LOESS regression to MDS distances, and the black dashed line shows a LOESS regression to the BMDS distances.
-	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and virus and serum effects estimated in a hierarchical Bayesian fashion. 	
+	The BMDS model has a Uniform$(-100,100)$ prior on antigenic locations and virus avidities and serum potencies estimated in a hierarchical Bayesian fashion. 	
 	} 

### 2013-12-07 flux.tex
Fix #36 - Ne*tau estimates.

@@
 	The maximium-clade credibility (MCC) tree is shown for each virus.
-	These trees show genealogical relationships, so that branches are measured in terms of years rather than substitutions.
-	The mean posterior scaled effective population size $N_e \tau$, measured in years, and its 95\% credible interval is shown for each virus.} 
+	These trees show genealogical relationships, so that branches are measured in terms of years rather than substitutions.} 
 	\label{trees} 

### 2013-12-09 correspondence/reresponse.tex
Fix #36 - Ne*tau estimates.
Small update to issue #33.

@@
 Including the biological explanation for why we observe virus effects in the form of decreased or increased overall HI reactivity is very important.
-We've revised this section to put virus avidity at the forefront and relabeled `virus effects' to `virus avidities'.
+We've revised this section to include references to virus avidity and relabeled `virus effects' to `virus avidities'.
 This has the additional benefit of being more transparent of a term than the opaque `effect.'}
@@
 Figure 4 legend refers to ``The mean posterior scaled effective populations... is shown for each virus.'' These are not actually shown in Figure 4, at least not in a way that is obvious to me.
+
+\textbf{Thank you for catching this.
+Estimates of $N_e\tau$ had been included in figure 4, but were moved to table 2 and the figure 4 legend not updated accordingly.
+This has been fixed to leave just the estimates in table 2.}
 

### 2013-12-09 flux.tex
Fix #34 - Ne*tau description.

@@
 Influenza A/H3N2 has a characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
-The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
-Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift (Table~\ref{paramtable}), with A/H3N2 showing fast drift and reduced effective population size as expected from basic epidemiological models \cite{Bedford12}.
+The scaled effective population size $N_e \tau$ measures the timescale of coalescence of a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
+In this case, $N_e$ represents the number of concurrent infections in a panmictic population with generation interval $\tau$ (time separating infections up the genealogical tree), so that the resulting $N_e \tau$ is measured in terms of years and gives the expected waiting time for two randomly chosen lineages to coalesce in the genealogical tree.
+We see that $N_e \tau$ broadly correlates with the rate of antigenic drift (Table~\ref{paramtable}), with A/H3N2 showing fast drift and reduced effective population size as expected from basic epidemiological models \cite{Bedford12}.
 Antigenic drift results in the replacement of antigenically primitive lineages by antigenically advanced lineages, thereby reducing genealogical diversity.

### 2013-12-10 correspondence/reresponse.tex
Fix #34 - Ne*tau description.
Fix #35 - drift vs sampling count correlation.

@@
 
+\textbf{We've revised the manuscript to clarify the definition and interpretation of both $N_e$ and $\tau$.}
+
 In the ``Punctuated evolution and its epidemiological consequences'' section, the part about subsampling to look at whether drift associates with number of isolates should be expanded.
@@
 I also think that it would be nice to have a table or figure somehow representing the actual results of this analysis.
+
+\textbf{We've expanded out this sentence to a paragraph following the paragraph on the year-to-year drift vs incidence correlation.
+We give rationale for the test, explain the bootstrap procedure, give $p$-values for both combined and separate analyses across lineages and provide a figure showing a scatterplot of the data.}
 

### 2013-12-10 flux.tex
Fix #35 - drift vs sampling count correlation.

@@
 This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
-We tested to see if there could be an artifactual effect on the measurement of year-to-year drift based on the number of viral isolates collected by correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
-We found no significant correlation (Pearson's $r = -0.01$, $p = 0.411$), suggesting our results to be unbiased with regard to sample count.
 
@@
 	\label{corr} 
+\end{figure}
+
+However, it is possible that if sampling count influences cartographic estimates then a spurious correlation could arise in which years with greater incidence have higher sample counts and artifactually high estimates of drift.
+We controlled for this possibility by testing to see if viral isolate count influences estimates of year-to-year drift by correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
+We found no significant correlation when combining data across lineages (Figure~\ref{counts_corr}, Pearson's $r = -0.01$).
+We testing significance following the same bootstrap procedure we used to assess the correlation between drift and incidence, finding a $p$-value of 0.514.
+Separating lineages gave $p$-values of 0.717, 0.246, 0.337, 0.504 for A/H3N2, A/H1N1, B/Vic and B/Yam, respectively.
+These findings suggest our results to be unbiased with regard to sample count.
+
+%%% counts_corr %%%
+\begin{figure}[h]
+	\centering		
+	\includegraphics[width=0.4\textwidth]{figures/counts_corr}
+	\caption{\textbf{Relationship between antigenic drift and sample counts for years 1998 to 2011 in influenza A/H3N2, A/H1N1, B/Vic and B/Yam.}
+	Antigenic drift from year $i-1$ to year $i$ is compared to the ratio of sample counts in year $i$ to counts in year $i-1$.
+	Only comparisons which had one or more samples in years $i-1$ and $i$ were retained, leaving 11 A/H3N2, 7 A/H1N1, 9 B/Vic and 10 B/Yam comparisons.
+	Points are colored according to influenza lineage based on the color scheme in Figure \ref{corr}.
+	} 
+	\label{counts_corr} 
 \end{figure}

### 2013-12-10 flux.tex
Fix #38 - conclusion.

@@
 Understanding antigenic evolution in seasonal influenza is crucial to our efforts of surveillance and control.
-Cartographic methods allow complex HI datasets to be compressed to more approachable location-based summaries that quantify antigenic relationships between strains, including relationships not directly assessed via HI.
+Cartographic methods allow complex HI datasets to be compressed to more approachable location-based summaries that quantify antigenic relationships between strains, including relationships not directly measured via HI.
 In this study we provide a foundation for evolutionary antigenic cartography, which seeks to simultaneously assess antigenic phenotype and antigenic evolution.
 We use this approach to characterize competitive dynamics across influenza lineages A/H3N2, A/H1N1, B/Vic and B/Yam and show that antigenic evolution within each lineage drives strain replacement and contributes to seasonal incidence patterns.
-
-%We find that varying levels of mutational input and varying efficiencies of the conversion of antigenic polymorphism into fixed differences results in different rates of antigenic drift across influenza clades.
-%Influenza A/H3N2 benefits from a higher rate of new antigenic mutation coupled with strong adaptive evolution and evolves in antigenic phenotype faster than A/H1N1, B/Vic and B/Yam.
-%Correspondingly, we observe substantially greater levels of incidence in A/H3N2 than in other influenza clades.
-%We suggest that antigenic evolution strongly influences competitive dynamics both within and between influenza clades.
+We find that influenza A/H3N2 evolves faster in antigenic phenotype than A/H1N1, which in turn evolves faster than B/Vic or B/Yam.
+Consequently, the influenza A/H3N2 virus population turns over more quickly than A/H1N1 or influenza B, exhibiting a smaller effective population size and a `spindlier' phylogenetic tree.
+Furthermore, we observe a correlation between antigenic drift and viral incidence both across and within influenza lineages.
+The finding of that antigenically advancement correlates with subsequent increased incidence within a lineage suggests a causal role for antigenic drift driving influenza incidence patterns.
+The correlation between incidence and drift further suggests the possibility of aiding prediction at the start of an influenza of which lineage will predominate.
 

### 2013-12-10 flux.tex
Fix #37 - interference.

@@
 We applied a similar bootstrap permutation test to calculate the significance of finding the observed levels of correlation across all four lineages, arriving at a $p$-value of 0.018.
+The fact that we observe periods of pronounced antigenic drift preceding increased incidence in each of the four influenza lineages suggests a causal relationship, in which antigenic evolution drives increased incidence.
 
@@
 \end{figure}
-
-Although the general correlation between rate of antigenic drift and incidence across the four influenza lineages may not represent a causal relationship, the within-lineage year-to-year correlations suggest a causal relationship as increases in incidence tend to occur subsequent to periods of pronounced antigenic drift.
 

### 2013-12-11 flux.tex
added affiliation
Cleanup tex.
Text cleanup.

@@
 \author[7,8,9]{Derek J. Smith}
-\author[1,10]{Andrew Rambaut}
+\author[1,10,11]{Andrew Rambaut}
 
@@
 \affil[10]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
+\affil[11]{Centre for Immunity, Infection and Evolution, University of Edinburgh, Edinburgh, UK}
 
@@
 Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination.
-Sera from older strains react poorly with more recent viruses resulting in new strains having a selective advantage over previously established strains.
+Sera raised against older strains react poorly to more recent viruses resulting in new strains having a selective advantage over previously established strains.
 
-The results of many HI assays across a multitude of virus strains of a single subtype can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
+The results of many HI assays across a multitude of viruses of a single subtype can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
 The antigenic map of influenza A/H3N2 has shown substantial evolution of the influenza virus population since its emergence in 1968. 
-Evolution of antigenic phenotype appears punctuated with episodes of more rapid innovation interspersed by periods of stasis, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype.
-The process of antigenic drift results in the rapid turnover of the virus population. 
-Although mutation occurs rapidly, genetic diversity among contemporaneous viruses is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
+Evolution of antigenic phenotype appears punctuated with episodes of more rapid innovation interspersed by periods of relative stasis, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype \cite{Koel13}.
+The process of antigenic drift results in the rapid turnover of the virus population, so that despite mutation, genetic diversity among contemporaneous viruses remains low.
+Such population turnover is supported by phylogenetic analysis that shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.
 
@@
 However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
-Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
+Here, we implement a flexible Bayesian approach to jointly characterize the antigenic and genetic evolution of the influenza virus population. 
 We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza lineages.
@@
 We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
-We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Materials and methods).
+We apply this model to HI measurements of virus isolates against post-infection ferret antisera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam.
 
@@
 Previous work on influenza antigenic and genetic evolution has shown that antigenic distance accumulates with increasing genetic distance \cite{Hay01, Smith04, Russell08}.
-Here, we examine pairwise relationships between viruses and observe a correlation between amino acid mutations and antigenic distance (Figure~\ref{seq_grid}) and a similar correlation between phylogenetic distance and antigenic distance (Figure~\ref{seq_grid}).
+Here, we examine pairwise relationships between viruses and observe a correlation between amino acid mutations and antigenic distance (Figure~\ref{seq_grid}) and a similar correlation between phylogenetic distance, measured in years, and antigenic distance (Figure~\ref{seq_grid}).
 Thus, genetic relationships between viruses provide some predictive power to estimate antigenic distances in the absence of HI data.
@@
 
-Consequently, we seek to flexibly incorporate genetic data by simultaneously modeling antigenic and genetic evolution through an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
+Consequently, we seek to flexibly incorporate genetic data by modeling antigenic phenotype as an evolutionary diffusion \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
 The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
@@
 
-We further extend the model by estimating serum potencies that quantify the overall strength reactivity of each serum, rather than fixing serum potencies at maximum titers.
-Different serum potencies likely derive from experimental variation in serum extraction and processing or from variation in immune responses between ferrets.
+We further extend the model by estimating serum potencies, rather than fixing serum potencies at maximum titers.
+Serum potencies differ across isolates due to experimental variation in serum extraction and processing or due to variation in ferret immune response.
 Serum potency determines the baseline expectation of titer when virus and serum have identical antigenic locations.
@@
 
-Additionally, we include and estimate `virus avidity' in an analogous fashion, which are intended to represent differences in overall HI reactivity between viruses.
+Additionally, we include and estimate `virus avidity' in an analogous fashion, which is intended to represent differences in overall HI reactivity between viruses.
 Experimental work has demonstrated that influenza variants exist that differ in HA binding activity for cell surface glycan receptors, and that these high-avidity variants may arise in the presence of antibody pressure \cite{Hensley09}.
-The presence of differential virus avidity has been previously shown to distort antigenic maps constructed from a model that disregards effects from virus avidity \cite{Li13}.
+The presence of differential virus avidity has been previously shown to distort antigenic maps constructed from a model that disregards avidity effects \cite{Li13}.
 Here, with virus avidities estimated, baseline titer derives from both the virus and the serum used in the HI reaction.
-We find that including virus avidities further improves test error, either along with fixed serum potencies (model 9, Table \ref{errortable}) or combined with estimating serum potencies (model 10, Table \ref{errortable}).
+We find that including virus avidities further improves test error, either with fixed serum potencies (model 9, Table \ref{errortable}) or with estimated serum potencies (model 10, Table \ref{errortable}).
 With fixed serum potencies, the inclusion of virus avidities results in improvements of between 0.02 and 0.09 log$_2$ HI titers and with estimated serum potencies, the inclusion of virus avidities results in improvements of between 0.01 and 0.05 log$_2$ HI titers.
@@
 
-These patterns of antigenic drift are reflected in the virus phylogenies (Figure~\ref{trees}).
+These patterns of antigenic drift influence the corresponding virus phylogenies (Figure~\ref{trees}).
 Influenza A/H3N2 has a characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
 The scaled effective population size $N_e \tau$ measures the timescale of coalescence of a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
-In this case, $N_e$ represents the number of concurrent infections in a panmictic population with generation interval $\tau$ (time separating infections up the genealogical tree), so that the resulting $N_e \tau$ is measured in terms of years and gives the expected waiting time for two randomly chosen lineages to coalesce in the genealogical tree.
+In this case, $N_e$ represents the number of concurrent infections in a panmictic population with generation interval $\tau$ (time separating infections up the genealogical tree), so that $N_e \tau$ is measured in terms of years and gives the expected waiting time for two randomly chosen lineages to coalesce in the genealogical tree.
 We see that $N_e \tau$ broadly correlates with the rate of antigenic drift (Table~\ref{paramtable}), with A/H3N2 showing fast drift and reduced effective population size as expected from basic epidemiological models \cite{Bedford12}.
@@
 
-Thus, we observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
+Thus, we observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or either lineage of influenza B.
 Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
@@
 
-We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
+We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean virus location between consecutive years (Figure~\ref{jumps}A).
 We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
@@
 
-We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and lineage-specific incidence, comparing incidence in a seasonal to antigenic drift of viruses coming into this season (Figure~\ref{jumps}A).
+We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and lineage-specific incidence, comparing incidence in a season to antigenic drift of viruses coming into this season (Figure~\ref{jumps}A).
 For example, we compare antigenic drift of viruses from 2000 to 2001 to incidence in the 2001/2002 season.
@@
 We calculated significance using bootstrap permutation tests finding $p$-values of 0.056, 0.201, 0.097 and 0.341 respectively.
-We applied a similar bootstrap permutation test to calculate the significance of finding the observed levels of correlation across all four lineages, arriving at a $p$-value of 0.018.
+We applied a similar bootstrap permutation test to calculate the significance of finding the observed degree of correlation across all four lineages, arriving at a $p$-value of 0.018.
 The fact that we observe periods of pronounced antigenic drift preceding increased incidence in each of the four influenza lineages suggests a causal relationship, in which antigenic evolution drives increased incidence.
@@
 We controlled for this possibility by testing to see if viral isolate count influences estimates of year-to-year drift by correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
-We found no significant correlation when combining data across lineages (Figure~\ref{counts_corr}, Pearson's $r = -0.01$).
-We testing significance following the same bootstrap procedure we used to assess the correlation between drift and incidence, finding a $p$-value of 0.514.
+We found little correlation when combining data across lineages (Figure~\ref{counts_corr}, Pearson's $r = -0.01$).
+We tested significance following the same bootstrap procedure we used to assess the correlation between drift and incidence, finding a $p$-value of 0.514.
 Separating lineages gave $p$-values of 0.717, 0.246, 0.337, 0.504 for A/H3N2, A/H1N1, B/Vic and B/Yam, respectively.
@@
 
-%Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
-%However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
-%Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.
-
-%\subsection*{Conversion of antigenic mutation into fixed differences}
-
-%We observe a faster rate of antigenic drift in influenza A/H3N2 than in A/H1N1 or in either lineage of influenza B.
-%Previous work using general epidemiological models has suggested that rates of antigenic drift may be influenced by both the fundamental reproductive number $R_0$ and the rate at which mutation decreases cross-immunity \cite{Gog02,Lin03}.
-%Owing to these and similar findings, subsequent work ascribed differences in the evolution of A/H3N2 relative to A/H1N1 and influenza B to either greater $R_0$ or greater mutation rate \cite{Ferguson03,Bedford12}.
-%Here, we examine the conversion of recently acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches in A/H3N2, A/H1N1, B/Vic and B/Yam (Table~\ref{diffusiontable}).
-%To facilitate comparisons, we measure the diffusion coefficient $D$ \cite{Pybus12} characterizing the antigenic movement of each influenza lineage (see Materials and methods).
-%This coefficient is measured in terms of units$^2$/year and can be used to determine the diffusion length scale, that is the expected relationship between time and distance traveled, which is $2 \sqrt{t D}$ for a two-dimensional diffusion.
-%Higher diffusion coefficients mean a lineage is diffusing faster through antigenic space.
-%We compare diffusion coefficients between trunk branches and side branches, where trunk branches are defined as branches ancestral to the most recent virus sample and side branches are defined as those branches not ancestral to the most recent virus sample.
-
-%%% diffusiontable %%%
-%\begin{table}[h]
-%	\centering
-%	\caption{\textbf{Estimates of fundamental diffusion coefficient $D$ across influenza lineages and phylogenetic branches, including posterior means and 50\% highest posterior density intervals.}}
-%	\label{diffusiontable}	
-%	\begin{tabular}{ c c c c } 
-%	\hline
-%	Lineage	&	Trunk branch $D$ 	& 	Side branch $D$		& 	Ratio \\
-%	\hline		
-%	A/H3N2	&	0.90 (0.79--0.97)	&	0.62 (0.57--0.65)	&	1.45 (1.31--1.57) \\
-%	A/H1N1	&	0.62 (0.50--0.71)	&	0.44 (0.38--0.48)	&	1.45 (1.18--1.70) \\
-%	B/Vic	&	0.63 (0.51--0.71)	&	0.62 (0.54--0.68)	&	1.02 (0.87--1.14) \\
-%	B/Yam	&	0.35 (0.29--0.40)	&	0.33 (0.28--0.36)	&	1.09 (0.93--1.23) \\
-%	\hline
-%	\end{tabular}	
-%\end{table}
-
-%Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
-%We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
-%Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
-%Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
-%We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
-%Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
-%Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
-%Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.
-
-%In this context, it is especially interesting that adaptive evolution in the HA1 protein determined through analysis of nonsynonymous and synonymous substitutions is substantially stronger in A/H3N2 than in A/H1N1 \cite{Wolf06}; H3 appears to accumulate adaptive fixations at $\sim$150\% the rate of H1 \cite{Bhatt11}.
-%Here, we observe that A/H3N2 has a diffusion coefficient $\sim$141\% greater than A/H1N1 along side branches of the phylogeny.
-%These results suggest that the greater rate of adaptive fixation in A/H3N2 may be simply due to a greater abundance of new antigenic mutations to fix.
-
-%We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
-%In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
-%These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
-%Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
-%Still, it is clear that more work is necessary to fully clarify the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.
-
 %%% CONCLUSIONS %%%
@@
 Furthermore, we observe a correlation between antigenic drift and viral incidence both across and within influenza lineages.
-The finding of that antigenically advancement correlates with subsequent increased incidence within a lineage suggests a causal role for antigenic drift driving influenza incidence patterns.
-The correlation between incidence and drift further suggests the possibility of aiding prediction at the start of an influenza of which lineage will predominate.
+The finding that antigenic evolution correlates with subsequent increased incidence within a lineage suggests a causal role for antigenic drift driving influenza incidence patterns.
+The correlation between incidence and drift further suggests the possibility of using HI data at the start of an influenza season to predict which lineage will subsequently predominate.
 
@@
 
-Previous work \cite{Smith04, Cai10} has used multidimensional scaling to place viruses and sera on an `antigenic map'.  
+Previous work \cite{Smith04, Cai10} has used multidimensional scaling (MDS) to place viruses and sera on an `antigenic map'.  
 These methods heuristically optimize locations of viruses and sera by seeking to minimize the sum of squared errors between titers predicted by map locations and observed titers.  
-Antigenic maps produced by these methods have proved useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
+Antigenic maps produced by these methods have proven useful in categorizing virus phenotypes \cite{Smith04}, but the extension of these methods to integrate genetic data remains notably lacking.
 
@@
 where $H_{ij}$ is the log$_2$ titer of virus $i$ against serum $j$ and serum potency $\se_j = \max ( H_{1j},\ldots,H_{\vn j} )$ is fixed.
-In following multidimensional scaling (MDS), these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
+In following multidimensional scaling, these approaches attempt to optimize over unknown $\viruses$ and $\sera$ such that
 \begin{equation} \label{mds}
@@
 Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
-The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
+The same procedure that was followed for A/H3N2 was repeated to match sequence data and to subsample antigenic comparisons.
 This procedure yielded 115 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.
@@
 As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.
-
-%\subsection*{Estimation of diffusion coefficient}
-
-%The distance $d$ traveled by a diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
-%Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
-%\begin{equation}
-%	X_t \sim \normal(0, \sigma^2 t),
-%\end{equation}
-%where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
-%\begin{equation}
-%	\mathrm{E}[X_t^2] = \sigma^2 t.
-%\end{equation}
-%In moving to a two-dimensional diffusion
-%\begin{equation}
-%	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma_x^2 t}{\rho \sigma_x \sigma_y t}{\rho \sigma_x \sigma_y t}{\sigma_y^2 t} \right),
-%\end{equation}
-%where $\sigma_x$ represents volatility in dimension 1, $\sigma_y$ represents volatility in dimension 2 and $\rho$ represents correlation,   expected squared displacement is equal to the sum of squared displacements in each dimension
-%\begin{equation}
-%	\mathrm{E}[X_t^2 + Y_t^2] = \sigma_x^2 t + \sigma_y^2 t.
-%\end{equation}
-%Such diffusions are often characterized with diffusion coefficient $D$.
-%In this case
-%\begin{equation}
-%	D = \frac{\sigma_x^2 + \sigma_y^2}{4},
-%\end{equation}
-%so that expected squared displacement follows from $D$ alone
-%\begin{equation}
-%	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
-%\end{equation}
-
-%Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
-%\begin{equation} \label{pybusest}
-%	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
-%\end{equation}
-%in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
-%However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
-%Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
-%\begin{eqnarray} \label{ourest}
-%	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
-%	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
-%	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
-%\end{eqnarray}
-%This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
-%Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
-%When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
 
@@
 AR, MAS, TB, PL implemented statistical phylogenetic procedures.
-TB performed the analysis.
+TB, AR performed the analysis.
 TB, AR, DJS, JWM, PL, CAR, AJH, GD, MAS interpreted the results.
